### 2012-08-08 paper/Paper.tex
initial commit

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  \graphicspath{{./images/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.jpg}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  \graphicspath{{./images/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

\usepackage{tikz}
\usetikzlibrary{arrows}

\usepackage{amsmath}

%\usepackage{natbib}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Graphical formalization and automated computing of safety constraints in robotics}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Ludwig N\"{a}gele}
\IEEEauthorblockA{Department of Software Engineering and\\Programming Languages\\ 
University of Augsburg\\
86135 Augsburg, Germany\\
Email: mail@ludwig-naegele.de}
\and
\IEEEauthorblockN{Bruce MacDonald}
\IEEEauthorblockA{Department of Electrical Engineering\\ 
The University of Auckland\\
Auckland 1142, New Zealand\\
Email: b.macdonald@auckland.ac.nz}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}
}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
%\boldmath
Nowadays robotic applications are no longer confined to only industrial use but increasingly assigned to jobs within human workspaces. Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behavior in order to match the user's individual tasks and requirements.
However both the human robot collaboration as well as the behavior definition by non experts may be a risk to humans within the robots' environment.
Thus special thinking is needed about how to cope with safety requirements for robot behavior specification by consumers.
%about how safety could be achieved in this scope.

%to make human robot interaction safe and espacially how to make user defined behavior safe.
%Special highlevel tools are developed in order to allow behavior definition by non experts.
%Besides the fact that human robot collaboration is becoming highly important, it has also to be thought about convenient safety mechanisms.

In this paper we present a visual language for the defining of safety constraints for state machine definition of robot behaviour. Our approach addresses mainly non-safety-experts and our abstraction from a mathematical temporal logic expression to a more intuitive visual representation is intended to enable a wider range of software developers to do model checking.
Furthermore we propose a new concept of development support %and qualitätssicherung
by providing automated constraint generation functionality. It aims to help developers in defining reasonable constraints and finally to increase the probability of finding bugs.

A graphical editor has been developed which demonstrates both visual constraint creation functionality as well as automated constraint generation.

%Our purpose is to target non experts and to make safety checking available for a bigger community. Hence we had a special focus on a simple syntax and easy understandability as well as active support  in our approach.

%Among others robots are alredy used in elderly care centers for healthcare purpose.

\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

%----------------------------------------------------------------------
% SECTION I: Introduction
%----------------------------------------------------------------------
\section{Introduction}

%\IEEEPARstart{S}{afety}
Safety critical applications require extensive testing or formal verification in order to achieve adequate safe and predictable behavior.
Even if a more highlevel program has no physical movements and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions executed by interacting actors.
%Even if a more highlevel program has no physical actors and thus cannot cause any damage, it might still have dangerous effects on the environment.
Imagine a medication reminder application which guides a person through the process of medication intake. All interaction is done by showing textual instructions on a display. An error in the program logic could cause an already taken medication to be reminded again and may injure an Alzheimer's diseased person by medication overdosage.
This example is taken from a healthcare robotics project at the University of Auckland, New Zealand. Scientists are working on applications such as the medication reminder~\cite{medicationreminder} which run on mobile robots serving in nursing homes for the elderly. Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{5649910}\cite{healtbot_new}.
The applications are meant to be developed by healthcare professionals using Robostudio~\cite{robostudio}, a visual programming environment for rapid authoring and customization of complex services on a personal service robot.
%state machine based processes and a graphical tool for the designing of human robot interaction.
Generally these people don't have profound knowledge of the complex mathematical syntax used for formal verification methods such as computation tree logic (CTL) or linear temporal logic (LTL). Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthbots.
In case of the medication reminder mentioned above it mitght be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.

So we need a mechanism for the specification of such safety constraints which doesn't require special or expert knowledge and thus is easy to use for all kind of programmers. A visual language intended to fill this gap is presented in section~\ref{sec:visualformalism}.

Unfortunately the supply of an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer. Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
To support the user in finding significant constraints it might be useful to automatically compute constraint suggestions which make true testimonies about the current designed program. First of all the generated constraints provide an opportunity for the developer to identify reasonable constraints as well as contradictions between constraints and specification. The latter would mean there are errors in the program which lead to undesired constraints. Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.

In section~\ref{sec:automatedconstraintgeneration} we introduce an heuristic for finding reasonable safety constraints based on a state machine in the healthcare robotics domain.

%Once such constraints are defined, they determine the grade of safety coverage. Unfortunately, the program may still contain errors if there is at least one significant constraint missed out. To support the user in finding such ones the tool could furthermore automatically compute suggestions for possible constraints based on the designed statemachine. These are displayed to the programmer and may help him identifying reasonable constraints as well as contradictions between program and specification.

%Our approach presented in this paper has a special focus on the healtcare domain.

%Dieses beispiel ist eine Anwendung eines umfangreichen Healthcare-Projekts, an dem im robotics lab der Universität von Auckland gearbeitet wird. Die zustansbasierten Anwendungen laufen auf Cafero, einem mit blutdruckmessgerät und anderen tools austestatteten mobilen Roboter mit Touchdisplay.
%Die einzelnen Anwendungen werden von Healthcare-Professionals mit dem Tool Robostudio [cite] entwickelt, das eine graphische oberfläsche zur modellierung von zustandsmaschinen externen services und  bietet.
%Allerdings haben diese not to have profound knowledge of the complex mathematical syntax used for formal verification methods such as computation tree logic (CTL) or linear temporal logic (LTL)).
%Dennoch möchte man gegenüber den Patienten, die von solchen Robotern gepflegt werden, ein gewisses Level an Sicherheit gewährleisten können. Beispielsweise soll in obigem Beispiel sichergestellt werden können, dass ein Medikament nach der Einnahme nicht erneut in einer Erinnerung auftaucht, oder dass bei Nichteinnahme ein Mitarbeiter benachrichtigt wird.
%Dazu bedarf es allerdings Mechanismen zur spezifizierung solcher Sicherheitseigenschaften, die für den Programmentwickler zugänglich und einfach (am besten ohne spezialwissen) zu benutzen sind.

%Constraints can be easily defined graphically and provide the user with an instant result about validity.
%, which can make propositions about the behavior of state machines.
%This language is an abstraction layer above temporal logic and allows a rather intuitive than textual model checking.
%The abstraction from textual and mathematical temporal logic expression allows a rather intuitive way of model checking and thus makes this language accessible to non-experts.

%CTL and LTL are important concepts for model checking and verification of statemachines. But their powerfulness is reflected in the complexity of their mathematical expressions, what makes it difficult to use for programmers who haven't got deeper expertise in this subject.

%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions. The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually. A common LTL model checker will be added to it as well which fulfils the verification of the translated constraints on the designed statemachine. In case of verification failures the concerning constraints and a visual counter example can be presented.

%Once such constraints are defined, they determine the grade of safety coverage. Unfortunately, the program may still contain errors if there is at least one significant constraint missed out. To support the user in finding such ones the tool could furthermore automatically compute suggestions for possible constraints based on the designed statemachine. These are displayed to the programmer and may help him identifying reasonable constraints as well as contradictions between program and specification.

%----------------------------------------------------------------------
% SECTION I: Introduction
%----------------------------------------------------------------------
\section{Related work}

%Was gibt es schon alles?

% 1993 A Visual-Programming Environment for a Temporal Logic Language
% 1995 Visual Specification of Branching Time Temporal Logic
% 2007 HomeTL: A visual formalism, based on temporal logic, for the design of home based care
% 2007 A Visual Editor to Support the Use of Temporal Logic for ADL Monitoring

A first step towards a graphical tool for the designing of safety constraints is taken by Sisiruca and Ionescu~\cite{332301}. They developed an object-oriented graphical environment for visually creating temporal logic sentences and rules. The tool presents temporal operators as logical modules with boolean inputs and outputs which can be connected to one big expression graph.

Also Del Bimbo et al. did work on a visual tool for temporal logic, but with a special focus on hierarchical representations of formulas~\cite{520786}. Here each node of a tree is described by an abstract operator whose subformulas are branches to their replacement operators. Thus the root node represents the entire formula. They also explored the formula trees' 3D representation within a virtual space.

Both approaches are different fundamental ideas of representing temporal formulas. Whereas having a graph and inputs and outputs seems not to be applicable for our approach, the idea of hirarchical operator representation of Del Bimbo et al. will be a good base for our try to find an easy to use concept for constraint creation and understanding.

%aim for simplifying the development of constraints, but they still demand knowledge of the underlying mathematical formalism of temporal logic.

The step towards a more abstract level of development is taken by HomeTL~\cite{4341725}, a visual formalism for the design of home based care. It allows healthcare professionals to specify rules and sequences of user actions in a quite nontechnical manner. Based on these conditions a monitoring system in a patient's home environment can detect and report abnormal situations within the daily routine.

Even though this approach touches the subject of healthcare it barely matches our problem. HomeTL focuses rather on monitoring temporal boundaries of a patient's behavior than on ensuring functional safety of an implemented program. But it depicts how nontechnical constraint design can be realised and its graphical design gives a good example how to handle time dependencies.

%There are already works about constraint generation such as one approach of NASA [4], for example. They used test oracles in order to achieve consistency between program code and specification during development. 
% Das wollen wir auch erzielen mit unseren automatisch generierten constraints...

%But these kind of oracles are for automated test purpose and are rather complex than intuitive. However, our goal is to find nonmathematical and just intuitive ones. An intelligent heuristic must determine key states and possible conjunctions of them in order to build reasonable constraints. Afterwards they get filtered by an algorithm which computes importances for each constraint based on state weights. Finally the developer obtains a set of visual constraints which he can then check for sanity.

%----------------------------------------------------------------------
% SECTION II: Requirements
%----------------------------------------------------------------------
\section{Requirements}
\label{sec:requirements}

% soll hier noch gesagt werden, dass wir davon ausgehen, dass es keine endzustände gibt?

We want to introduce a visual language for the defining of safety constraints for state machine definition of robot behaviour which is suitable for non-experts in the formal methods domain. Ideally its appliance is easy to use and intuitive to read in order to facilitate maintainability.
Hence, the graphical formalism should abstract from the usually complex and mathematical concept of conventional formal methods but still stay unreduced in expressiveness compared to LTL as far as possible.

Furthermore a heuristic should be developed of how to identify reasonable constraints on an underlying state machine program. 
But since the particular domain distinguishes the significance of constraints respectively it might be difficult to show a universal heuristic working for all use cases. Hence, in this paper the concept of development support by constraint generation is examined within the domain of healthcare robotics. As a special requirement to the heuristic at least the following constraints are to be found for the medication reminder application:

\begin{itemize}
	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
% The next item is left out so that we can say, our heuristic found one we didn't think about yet :)
%	\item During the medication intake process the program can not switch to other services such as entertainment mode ore blood pressure measurement, for example.
\end{itemize}

Some of the statements don't make sense of course if patients lie or pretend to take their medication. As our approach tries to validate state machine behaviour rather than human attitudes we assume correct collaboration or the use of tools which can recognize fairly medication intake. Furthermore also the correctness of external services used by the medication reminder such as the database module holding all reminders is not taken into account.

All the functionality mentioned above shall be combined in one editor and for evaluation purpose later be integrated into a programming environment for state machine editing.

%----------------------------------------------------------------------
% SECTION II: Design
%----------------------------------------------------------------------
\section{Design}

\subsection{Visual formalism}
\label{sec:visualformalism}

% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv.

The presented visual language is made for expressing constraints for state machines. Similar to Del Bimbo's formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchical manner. However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
Each block represents either an operator or a proposition and has its specified semantics. Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.

The visual language has support for eight operator types and one proposition as shown in figure~\ref{fig:operators}.

\begin{description}
	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided. Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state. Other types such as numeric or string equations may be added but are less important for our healthcare subject.
	\item[(f)] \emph{FUTURE} operator: $F \varphi$.\\At least one time -- now or in a later state -- $\varphi$ must be true.
	\item[(g)] \emph{ALWAYS} operator: $G \varphi$.\\$\varphi$ must be true now and in all following states.
	\item[(h)] \emph{NEXT} operator´: $X \varphi$.\\$\varphi$ has to be true in the next state.
	\item[(i)] \emph{UNTIL} operator: $[\varphi U \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true. In addition eventually there has to be a future state with $\psi$ being true.
\end{description}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.5]{table}
  \caption{The visual language consists of eight unary or binary operator types and one state proposition, each of them having its own color flavor. Logical operators are arranged in a vertical row wereas time relevant statements are aligned horizontally.}
  \label{fig:operators}
\end{figure}

Regarding to the requirements the visual language should be very easy to use, especially for non-experts. Furthermore it should provide an intuitive appliance which is ideally learnable within little time. So there are some design decisions an editor for this visual language has to fulfill:

\begin{itemize}
	\item Each operator has a different significant color flavor. This kind of classification enables a faster subconscious identification by the user.
	\item All editing is simply done by only drag and drop. Operators and propositions can be moved and dropped onto other operators. A tool bar provides all kinds of operator and proposition types to be instantiated and ideally a trash can for deleting existing operators.
	\item The layout of constraints (nested operators) has two dimensions with a certain meaning. Figure~\ref{fig:directions} demonstrates the ``logical'' vertical read direction as well as the time releveant horizontal line.
	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer. All higher operators become bleached out so that the hovered operator and its children appear highlighted. It's intended to be similar to line coloring when reading a digital text.
	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appeared to make the look unnecessarily complex. These operators can be merged as shown in figure~\ref{fig:nested_and}.
	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances of them. Having no additional buttons reduces complexity and keeps it simple.
\end{itemize}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.5]{directions}
  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
  \label{fig:directions}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.5]{and_simplify} %width=0.7\linewidth
  \caption{In order to simplify the appearance of constraints certain operators might be merged. Merged nested \emph{AND} operators (center) are less complex as default ones (left). However nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
  \label{fig:nested_and}
\end{figure}

\subsection{Automated constraint generation}
\label{sec:automatedconstraintgeneration}

In order to support the developer in specifying safety constraints, automated generation and suggestion of constraints might be helpful. But it's helpful only then, if the suggested constraints are relevant and reasonable. This however depends on the domain and use case.
Regarding to the postulated constraints in the requirements in section~\ref{sec:requirements} it turned out that all therefor relevant states have got one of the following characteristics:
%following types of states are of a special interest:

\begin{enumerate}
	\item States having two or more outgoing transitions ($\alpha$-states).
	\item States in which all paths starting from one $\alpha$-state come together again ($\beta$-states).
	\item All States immediately before $\beta$ states ($\gamma$-states).
\end{enumerate}

In order to compute the required constraints and possibly even more of them automatically we manage with so called subgraphs defined as follows:
%Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

\begin{itemize}
	\item It has a start state ($\alpha$-state) with at least two outgoing transitions.
	\item There is an end state ($\beta$-state) in which all possible paths starting from start state come together again.
	\item All states between start and end state have no other incoming transitions than the ones originally coming from start state. That means there is no path to visit these states exept through start state.
	\item start state may have incoming transitions from states not contained in the subgraph.
	\item end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
\end{itemize}

\begin{figure}[htbp]
  \centering

  \tikzstyle{state} = [circle,fill=white,draw]
  \tikzstyle{statein} = [circle,fill=blue!20,draw]
  \tikzstyle{every path} = [font=\sffamily\small]

	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=15mm]

	  \node[statein] (1) {$\alpha$};
	  \node[state] (2) [above left of=1] {};
	  \node[state] (3) [above right of=1] {};
	  \node[statein] (4) [below left of=1] {};
	  \node[statein] (6) [below of=4] {$\gamma$};
	  \node[statein] (7) [below right of=4] {$\gamma$};
	  \node[statein] (5) [below right of=1] {};
	  \node[statein] (8) [below of=5] {$\gamma$};
	  \node[statein] (9) [below of=7] {$\beta$};
	  \node[state] (10) [right of=9] {};
	  \node[state] (11) [below left of=9] {};
	  \node[state] (12) [below right of=9] {};

	  \path[dashed]
	    (2) edge node [left] {} (1)
	    (3) edge node [left] {} (1)
	    (10) edge node [left] {} (9)
	    (9) edge node [left] {} (11)
	    (9) edge node [left] {} (12);
		\path[]
	  	(1) edge node [left] {} (4)
	        edge node [left] {} (5)
	    (4) edge node [left] {} (6)
	        edge node [left] {} (7)
	    (5) edge node [left] {} (8)
	    (6) edge node [left] {} (9)
	    (7) edge node [left] {} (9)
	    (8) edge node [left] {} (9)
	    ;

	\end{tikzpicture}
  \caption{Example for a subgraph.}
  \label{fig:subgraph}
\end{figure}

Using this specification all possible subgraphs can be detected within a state machine. If a subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarily for $\beta$ and all $\gamma$'s):

\begin{equation} \label{eq:first}
  \models G (\alpha \Rightarrow F \beta)
\end{equation}

\begin{equation} \label{eq:second}
  \models G (\alpha \Rightarrow F (\bigvee_{\gamma} \gamma))
\end{equation}

\begin{equation} \label{eq:third}
  \models G (\alpha \Rightarrow [\neg \beta U \bigvee_{\gamma} \gamma])
\end{equation}

Additionaly for all states not contained in the subgraph (let's call them $\delta$-states, and the proposition $\delta$ stands for ``$\delta$ is current state'') the following formula is true:

\begin{equation} \label{eq:fourth}
  \models G (\alpha \Rightarrow [(\bigwedge_{\delta} \neg \delta) U \beta])
\end{equation}

In case of $\alpha = \beta$ -- i.e. when all paths starting from $\alpha$-state eventually return again (cycle) -- even one more constraint can be assumed to be true:

\begin{equation} \label{eq:fifth}
  \models G F \alpha
\end{equation}

These five constraints apply for each subgraph matching the conditions mentioned above. Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state. Formula~\ref{eq:second} ensures that there will be always at least one $\gamma$-state be visited after each visit of an $\alpha$-state. The quite similar but stricter formula~\ref{eq:third} demands additionally a visit of at least one $\gamma$-state before the $\beta$-state can be reached. Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states. That the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

All formulas can be translated to the visual formalism presented in section~\ref{sec:visualformalism} and shown to the user. The accordant LTL formulas for the postulated constraints in section~\ref{sec:requirements} belong to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:

\begin{equation}
  \models G F \textnormal{'Polling'}
\end{equation}

\begin{equation}
	\models G (\textnormal{'Not taken yet'} \Rightarrow F (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
\end{equation}
%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.

%----------------------------------------------------------------------
% SECTION III: Prototype and evaluation
%----------------------------------------------------------------------
\section{Prototype and evaluation}

To experience and evaluate the visual formalism as well as the usefulness of constraint generation we developed an editor which provides functionalities all proposed functionalities while taking the requirements into account.
It's implemented as a java swing component, making it easy to integrate the visual language into other java applications. An API allows to set the underlying state machine what also influences the state propositions available for constraint editing. Furthermore a particular model checker can be specified to be used for constraint validation. As default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046}~\cite{NuSMV2} is used for constraint validation.

Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right which contains drag providers for the creating of all operator and proposition types as well as a trash for deleting. Constraints can be composed in the dashboard in the middle of the editor by drag and drop. The tab functionality on the left allows to manage several constraints. Each tab schows a small thumbnail of the constraint and a symbol indicating its validity what is a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress''. The magic wand button within the tab pane triggers the constraint generation.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{editor} 
  \caption{Snapshot of the visual editor.}
  \label{fig:editor}
\end{figure}

In order to evaluate the proposed safety functionalities on various programs the tool was integrated into Robostudio~\cite{robostudio}. It allows the specification of state machine based programs whose behaviour can then be examined by visual constraints. As an example of how to use them we will have a closer look at the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}. Starting from ``Menu'' the user can choose from several services such as blood pressure measurement ore entertainment mode. In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part. In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered. The patient can state if he's already taken his medication, or decide whether he wants to take it or not. In the latter he may give a reason for it and a caregiver gets notified about this incident. Otherwise it will result in ``Well done!'' after completing intake.

%we can now vergleichen die oben genannten bedingungen.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{stm} %width=0.7\linewidth
  \caption{Simplified screen-flow of the medication reminder application.}
  \label{fig:medicationreminder}
\end{figure}

%Vorstellung des medication reminders

%in particular

% TODO: hier vlt ein komponentendiagramm?

%Der Editor wurde in robostudio eingebunden, ein werkzeug zur erstellung von healthcare applicationen (elderly) basierend auf zustandsmaschinen. (bereits in der einleitung erwähnt). Über die darin definierbaren zustansmaschinen können wir nun aussagen mit der vl erstellen.

%das tool wurde den healthcare professionals vorgelegt und die ergebnisse werden gespalten in den folgenden zwei unterpunkten beschrieben (vl und constraint generierung).

%Robostudio was used as a platform to integrate the visual language described in this paper .
%basierend auf diesen healthcare applicationen können nun sicherheits constraints über unser tool definiert werden.

%rs was used as a platform to integrate the safety critical visual programming language described in this paper. fs's modular design allows the usage of the state machine model and add plugins to be integrated into the VPE, making it usabble as a plat

%personal service robot

%visual programming environment
%We embedded the editor into Robostudio~\cite{robostudio}, a tool for the creating of healthcare applications based on state machines.

%an editor for state machine based processes and a graphical tool for the designing of human robot interaction

\subsection{Visual constraints}

Before we can start creating constraints it might be useful to think about a reasonable term we want to express. Let's take the example ``Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake'' given in the requirements in section~\ref{sec:requirements}. The LTL formula for this constraint is:

\begin{equation} \label{eq:sampleconstraint}
  \models G (\textnormal{'Not taken yet'} \Rightarrow F (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'}))
\end{equation}

This expression can now be translated to the visual form as we just have to break down the formula from the outside to the inside. A surrounding \emph{GLOBALLY} encapsulates an \emph{IF} block since it is the substitution for \emph{IMPLIES}. Its first operator is a proposition \emph{'Not taken yet'} whereas the second parameter has to be a \emph{FOLLOWS} block containing a disjunction of two propositions \emph{'Well done!'} and \emph{'Staff notified'}. The resulting visual constraint is shown in figure~\ref{fig:sampleconstraint}.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.5]{sampleconstraint}
  \caption{Visual constraint creation for LTL formula ~\ref{eq:sampleconstraint}.}
  \label{fig:sampleconstraint}
\end{figure}

After each change within the dashboard the constraint is automatically revalidated by using the underlying model checker. As long as the constraint is not completed an ``incomplete'' sign is displayed in the regarding tab. Otherwise an animated ring indicates validation in progress what finally results in either a ``valid'' or ``invalid'' sign.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.5]{results}
  \caption{Status indicators displayed during or after validation: incomplete, validation in progress, invalid and valid.}
  \label{fig:results}
\end{figure}

Owing to the elaborated performance of NuSMV even the validation of constraints on huge and complex programs is fast and allows quick feedback. Besides, the validation itself runs in background without locking the dashboard. Therefore we got rid of waiting times and disruptions during constraint development what would be very well the case with conventional model checking where constraint development and constraint validation are alternating occupations. Thus it's considered as an improvement of user experience.

\subsection{Constraint generation}
\label{sec:prototypeandevaluation_constraintgeneration}

The automated constraint generation can be triggered by a click on the magic wand button in the tab area. After activation a dashboard gets opened in a new tab for each found constraint, and validation is initiated immediately. In case of the medication reminder application six constraints are found, whereof \emph{A)} and \emph{B)} shown in figure~\ref{fig:generatedconstraints} are actually matching the postulations in the requirements in section~\ref{sec:requirements}.
The constraint used for demonstration in the previous chapter gets also generated, however \emph{B)} forms an intensified restriction of it.

Constraint \emph{A)} ensures the ``Polling'' state to be always eventually visited again. If a pending medication is not already taken, constraint \emph{B)} guarantees intake or staff notification before next reminders can be read from database.

We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{C)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment mode ore blood pressure measurement.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{generatedconstraints} %width=0.7\linewidth
  \caption{Automatically generated constraints which match the postulations in the requirements.}
  \label{fig:generatedconstraints}
\end{figure}

%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde kann bei großen programmen (ab ca. 1000 Zuständen und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern. Um den warteprozess erträglicher zu machen wird während der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht

%----------------------------------------------------------------------
% SECTION III: Prototype and evaluation
%----------------------------------------------------------------------
\section{Future work}

The editor presented in this paper proofes validity of all constraints and presents the result \emph{valid} or \emph{invalid} to the user. But in contrast to conventional model checking no additional explanation for the outcome such as a counter example is given. It can be considered to extend the visual language with support for bug identification by a visualization of counter examples, for example.
%\item Substitution of parts of constraints would be nice in order to make it easier to understand. (A simple operator which abstracts from a more complex operator). It's comparable to composite states of hirachical state machines.

%\item A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.

%	\item constraints bewerten (nach gewichtung?) und nur wichtige vorschlagen? Oder zumindest sortieren

Since we focus on the healthcare domain in this paper the only proposition type for constraint creation is the state proposition. But to suit other domains additional proposition types such as number or string equations might be added. Also the portability of the constraint generation heuristic to other domains should be examined and different methods than subgraph based constraint finding might be elaborated.

%\begin{itemize}
%	\item As already mentioned
%	\item Yet, there are only states and no variable expressions etc.
%	\item Substitution of parts of constraints would be nice in order to make it easier to understand. (A simple operator which abstracts from a more complex operator). It's comparable to composite states of hirachical state machines.
%	\item A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.
%	\item Problem mit der Zustandsbenennung! Die Zustände müssen passende Namen haben, was aber nicht immer möglich ist. Daher müsste man eigentlich auch Transitionsbedingungen in den Constraints erlauben
%	\item constraints bewerten (nach gewichtung?) und nur wichtige vorschlagen? Oder zumindest sortieren
% \item es wird nur gutes gefunden (also was passieren muss). es werden keine constraints gefunden, die schlechte abläufe darstellenund verbieten
%	\item übertragbarkeit auf andere domänen untersuchen?
%	\item A bigger generation algorithm. As already metioned our approach only targets the healthcare domain. Especially the constraint generation can be improved to match also other domains. vielleicht geht beim triggern der constraint generierung ein fenster auf, wo man die heuristic angeben kann (eventuell benannt und gegliedert nach domains).
%\end{itemize}

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

\section{Conclusion}

In this paper we represented a visual formalism for the defining of safety constraints and a concept of automated constraint generation. Both features are implemented and assembled in an editor for standalone or integration use. The visual language is pretty simple what makes it accessible to a wider range of software developers. Nevertheless its expressiveness is fairly powerful and allows serious model checking.

The concept of constraint generation worked very well for our use case the medication reminder application and all postulated constraints were found. Since we even discovered a new reasonable constraint we are convinced this concept as well as the visual language for constraint defining can be a beneficial supporting thing for plenty of developers.

% conference papers do not normally have an appendix

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,Paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%\bibitem{healthcare}
%B.~MacDonald, \emph{Healthcare}.
%\bibitem{robostudio}
%C.~Datta and B.~MacDonald, \emph{Robostudio}.

%\bibitem{visualtemporallogic1993}
%\emph{1993 A Visual-Programming Environment for a Temporal Logic Language}.
%\bibitem{visualtemporallogic1995}
%\emph{1995 Visual Specification of Branching Time Temporal Logic}.
%\bibitem{hometl}
%\emph{2007 HomeTL: A visual formalism, based on temporal logic, for the design of home based care}.
%\bibitem{adl}
%\emph{2007 A Visual Editor to Support the Use of Temporal Logic for ADL Monitoring}.

%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

%\end{thebibliography}

% that's all folks
\end{document}

### 2012-08-08 src/editor/Editor.java
Commented

@@ -23,6 +23,14 @@ import editor.toolpane.ToolPane;
 import editor.validator.ConstraintValidator;
 import editor.validator.impl.NuSMVModelCheckerWrapper;
 
+/**
+ * This class allows an easy integration of the LTLCreator tool into other swing
+ * based java programs. <a>Editor</a> is a component containing a tab bar for
+ * several dashboards and a tool bar for constraint creation.
+ * 
+ * @author ludwig
+ * 
+ */
 public class Editor extends JComponent {
 	private static final long serialVersionUID = 1L;
 	
@@ -30,6 +38,9 @@ public class Editor extends JComponent {
 	private GraphTabPane graphTabPane = new GraphTabPane();
 	private Map<GraphTab, OperatorAndModelChangeListener> graphTabs = new HashMap<GraphTab, OperatorAndModelChangeListener>();
 	
+	/**
+	 * Creates a new LTLCreator component. It can be added to any swing container.
+	 */
 	public Editor() {
 		this.setLayout(new BorderLayout());
 		
@@ -43,18 +54,44 @@ public class Editor extends JComponent {
 		this.add(new ScrollColumn(new ToolPane()), BorderLayout.EAST);
 	}
 	
+	/**
+	 * Sets the new model. It causes all constraints to be revalidated, and the
+	 * list from where all possible states can be choosen will be updated.
+	 * 
+	 * @param model
+	 *            the new model.
+	 */
 	public void setModel(Fsm model) {
 		this.validator.setModel(model);
 	}
 	
+	/**
+	 * Adds a new tab with an empty dashboard to the editor.
+	 * 
+	 * @param select
+	 *            determines whether the tab should become selected.
+	 */
 	public void addNewEmptyDashboard(boolean select) {
 		addNewDashboard(null, select);
 	}
 	
+	/**
+	 * Adds a new tab with a prefilled dashboard to the editor.
+	 * 
+	 * @param operator
+	 *            the constraint contained in the dashboard.
+	 * @param select
+	 *            determines whether the tab should become selected.
+	 */
 	public void addNewDashboard(AbstractOperator operator, boolean select) {
 		this.graphTabPane.addGraphTab(new OperatorTab(operator), select);
 	}
 	
+	/**
+	 * Returns a list of all operators, ordered in the same order as the tabs.
+	 * 
+	 * @return all operators of the editor.
+	 */
 	public List<AbstractOperator> getOperators() {
[…]

### 2012-08-08 src/editor/Test.java

@@ -25,36 +25,8 @@ public class Test {
 				final Editor editor = new Editor();
 				editor.addNewEmptyDashboard(false);
 				
-
+				/* medication reminder example */
 				Fsm model = new Fsm();
-
-//				State s1 = model.addState("Menu");
-//				State s2 = model.addState("Polling");
-//				State s3 = model.addState("Are you..?");
-//				State s4 = model.addState("Person valid");
-//				State s5 = model.addState("Not taken yet");
-//				State s6 = model.addState("No help");
-//				State s7 = model.addState("Reason");
-//				State s8 = model.addState("Staff notified");
-//				State s9 = model.addState("Guided intake");
-//				State s10 = model.addState("Well done!");
-//				
-//				s1.addTransition(s2);
-//				s2.addTransition(s1);
-//				s2.addTransition(s3);
-//				s3.addTransition(s1);
-//				s3.addTransition(s4);
-//				s4.addTransition(s2);
-//				s4.addTransition(s5);
-//				s5.addTransition(s6);
-//				s5.addTransition(s9);
-//				s6.addTransition(s7);
-//				s6.addTransition(s8);
-//				s7.addTransition(s8);
-//				s8.addTransition(s2);
-//				s9.addTransition(s10);
-//				s10.addTransition(s2);
-//				model.setInitialState(s1);
 				
 				State menu = model.addState("Menu");
 				State additionalservices = model.addState("Additional services");
@@ -88,7 +60,6 @@ public class Test {
 				nohelp.addTransition(welldone);
 				
 				model.setInitialState(menu);
-				
 				editor.setModel(model);
 				
 				final JFrame frame = new JFrame("LTL Creator by Ludwig Nägele");
@@ -98,32 +69,6 @@ public class Test {
 				frame.setMinimumSize(new Dimension(400,300));
 				frame.setLocationByPlatform(true);
 				frame.setVisible(true);
-				
-//				new Thread(new Runnable() {
-//					
-//					@Override
-//					public void run() {
-//						try {
-//							Thread.sleep(6000);
-//						} catch (InterruptedException e) {
-//							// TODO Auto-generated catch block
-//							e.printStackTrace();
-//						}
-//						
-//						Fsm model = new Fsm();
-//
-//						State s1 = model.addState("Menu");
-//						State s2 = model.addState("Polling");
-//						State s3 = model.addState("Are you..?");
-//						State s4 = model.addState("Person valid");
[…]

### 2012-08-08 src/model/AbstractOperator.java

@@ -22,7 +22,13 @@ import util.Utils;
 import util.dnd.Draggable;
 import util.dnd.DropTarget;
 
-
+/**
+ * This class represents an abstract operator. Each operator can contain several
+ * child operators, but this depends on the specific subclass.
+ * 
+ * @author ludwig
+ * 
+ */
 public abstract class AbstractOperator extends JComponent implements Draggable {
 	private static final long serialVersionUID = 1L;
 
@@ -48,9 +54,30 @@ public abstract class AbstractOperator extends JComponent implements Draggable {
 		};
 	}
 	
+	/**
+	 * Returns the corresponding LTL formula for this operator and all
+	 * suboperators.
+	 * 
+	 * @return the corresponding LTL formula for this operator.
+	 * @throws NotCompleteException
+	 *             if at least one (sub-)operator is not complete.
+	 */
 	public abstract String getLTL() throws NotCompleteException;
+	
+	/**
+	 * Returns the specific color of the implemented operator type.
+	 * 
+	 * @return the specific color of the implemented operator type.
+	 */
 	public abstract Color getColor();
+	
+	/**
+	 * Creates a new instance of the implemented operator type.
+	 * 
+	 * @return a new instance of the implemented operator type.
+	 */
 	public abstract AbstractOperator createNewInstance();
+	
 	protected abstract void paintOperator(Graphics2D g, boolean considerHighlighting);
 	protected abstract void layoutBuckets();
 	protected abstract int getCenterLineX();
@@ -123,6 +150,18 @@ public abstract class AbstractOperator extends JComponent implements Draggable {
 		paintOperator(g, true, false, false);
 	}
 	
+	/**
+	 * This function paints this operator on a graphics object.
+	 * 
+	 * @param g
+	 *            the graphics object.
+	 * @param considerHighlighting
+	 *            determines, whether mouse hover effects shall be painted.
+	 * @param recursive
+	 *            if true, all suboperators will be painted recursively.
+	 * @param doLayout
+	 *            causes the operator to lay out before painting.
+	 */
 	public void paintOperator(Graphics g, boolean considerHighlighting, boolean recursive, boolean doLayout) {
 		if (doLayout) this.doLayout();
 		
@@ -187,6 +226,13 @@ public abstract class AbstractOperator extends JComponent implements Draggable {
 		return false;
 	}
 	
+	/**
[…]

### 2012-08-08 src/editor/Test.java
Comments

@@ -9,7 +9,11 @@ import javax.swing.UIManager;
 import util.fsmmodel.Fsm;
 import util.fsmmodel.State;
 
-
+/**
+ * Testclass for a standalone version of the LTLCreator tool.
+ * @author ludwig
+ *
+ */
 public class Test {
 	
 	public static void main(String[] args) {
@@ -22,55 +26,69 @@ public class Test {
 					e.printStackTrace();
 				}
 				
-				final Editor editor = new Editor();
-				editor.addNewEmptyDashboard(false);
-				
-				/* medication reminder example */
-				Fsm model = new Fsm();
-				
-				State menu = model.addState("Menu");
-				State additionalservices = model.addState("Additional services");
-				State polling = model.addState("Polling");
-				State pendingreminder = model.addState("Pending reminder");
-				State nottakenalready = model.addState("Not taken already");
-				State refuseintake = model.addState("Refuse intake");
-				State reason = model.addState("Reason");
-				State staffnotified = model.addState("Staff notified");
-				State guidedintake = model.addState("Guided intake");
-				State nohelp = model.addState("No help");
-				State welldone = model.addState("Well done!");
-				
-				menu.addTransition(additionalservices);
-				menu.addTransition(polling);
-				additionalservices.addTransition(polling);
-				polling.addTransition(menu);
-				polling.addTransition(pendingreminder);
-				pendingreminder.addTransition(polling);
-				pendingreminder.addTransition(nottakenalready);
-				nottakenalready.addTransition(refuseintake);
-				refuseintake.addTransition(reason);
-				refuseintake.addTransition(staffnotified);
-				reason.addTransition(staffnotified);
-				staffnotified.addTransition(polling);
-				nottakenalready.addTransition(guidedintake);
-				guidedintake.addTransition(welldone);
-				welldone.addTransition(polling);
-				nottakenalready.addTransition(nohelp);
-				nohelp.addTransition(guidedintake);
-				nohelp.addTransition(welldone);
-				
-				model.setInitialState(menu);
-				editor.setModel(model);
-				
-				final JFrame frame = new JFrame("LTL Creator by Ludwig Nägele");
-				frame.setDefaultCloseOperation(JFrame.EXIT_ON_CLOSE);
-				frame.add(editor);
-				frame.pack();
-				frame.setMinimumSize(new Dimension(400,300));
-				frame.setLocationByPlatform(true);
[…]

### 2012-08-08 src/editor/validator/ConstraintValidator.java

@@ -3,11 +3,10 @@ package editor.validator;
 import java.util.ArrayList;
 import java.util.List;
 
-import util.fsmmodel.Fsm;
-import util.fsmmodel.State;
-
 import model.AbstractOperator;
 import model.NotCompleteException;
+import util.fsmmodel.Fsm;
+import util.fsmmodel.State;
 
 public abstract class ConstraintValidator {
 
@@ -18,7 +17,14 @@ public abstract class ConstraintValidator {
 	public abstract boolean validate(String ltl) throws ValidationCanceledException;
 	public abstract String getLTLForVariable(String variable);
 	
-	
+	/**
+	 * Validates an operator.
+	 * 
+	 * @param operator
+	 * @return validation result.
+	 * @throws NotCompleteException if the operator is not complete.
+	 * @throws ValidationCanceledException
+	 */
 	public boolean validate(AbstractOperator operator) throws NotCompleteException, ValidationCanceledException {
 		String ltl;
 		synchronized (this.monitor) {

### 2012-08-08 src/editor/validator/impl/NuSMVModelCheckerWrapper.java

@@ -18,7 +18,6 @@ import javax.swing.filechooser.FileFilter;
 
 import util.fsmmodel.Fsm;
 import util.fsmmodel.State;
-
 import editor.validator.ConstraintValidator;
 import editor.validator.ValidationCanceledException;
 
@@ -72,15 +71,19 @@ public class NuSMVModelCheckerWrapper extends ConstraintValidator {
 				
 				return result;
 			} catch (Exception e) {
-				e.printStackTrace();
 				throw new ValidationCanceledException();
-//				System.err.println("Error during validation. Try again...");
 			} finally {
 				if (file!=null) file.delete();
 			}
 		}
 	}
 	
+	/**
+	 * Returns the path to the NuSMV.exe file. If not already set, the user may
+	 * be asked whether he wants to give the path of the model checker.
+	 * 
+	 * @return
+	 */
 	private String getNuSmvFile() {
 		if (this.nuSmvFile!=null && new File(this.nuSmvFile).isFile()) {
 			return this.nuSmvFile;
@@ -130,6 +133,14 @@ public class NuSMVModelCheckerWrapper extends ConstraintValidator {
 		return result;
 	}
 	
+	/**
+	 * Converts a state machine and a ltl formula to a NuSMV diagramm.
+	 * 
+	 * @param ltlFormula the formula to be checked.
+	 * @param model the state machine behavior.
+	 * @param replacementNames the replacements for state names.
+	 * @return
+	 */
 	private static String createNuSMVDiagram(String ltlFormula, Fsm model, Map<String, String> replacementNames) {
 		StringBuffer sb = new StringBuffer();
 		sb.append("MODULE main\n");
@@ -175,6 +186,13 @@ public class NuSMVModelCheckerWrapper extends ConstraintValidator {
 		return sb.toString();
 	}
 	
+	/**
+	 * Writes a string to file.
+	 * 
+	 * @param content
+	 * @return
+	 * @throws IOException
+	 */
 	private static File writeToFile(String content) throws IOException {
 		File temp = File.createTempFile("rsmc",".smv");
 		temp.deleteOnExit();
@@ -186,6 +204,15 @@ public class NuSMVModelCheckerWrapper extends ConstraintValidator {
 		return temp;
 	}
 	
+	/**
+	 * Executes the model checker in a shell. This function will block until the
+	 * checking process is finished.
+	 * 
+	 * @param file the NuSMV diagram.
[…]

### 2012-08-11 paper/Paper.tex
add Bruce's initial
Updates from Bruce

@@
 \and
-\IEEEauthorblockN{Bruce MacDonald}
+\IEEEauthorblockN{Bruce A. MacDonald}
 \IEEEauthorblockA{Department of Electrical Engineering\\ 
@@
 %Even if a more highlevel program has no physical actors and thus cannot cause any damage, it might still have dangerous effects on the environment.
-Imagine a medication reminder application which guides a person through the process of medication intake.
-All interaction is done by showing textual instructions on a display.
-An error in the program logic could cause an already taken medication to be reminded again and may injure an Alzheimer's diseased person by medication overdosage.
-This example is taken from a healthcare robotics project at the University of Auckland, New Zealand.
-Scientists are working on applications such as the medication reminder~\cite{medicationreminder} which run on mobile robots serving in nursing homes for the elderly.
-Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{5649910}\cite{healtbot_new}.
-The applications are meant to be developed by healthcare professionals using Robostudio~\cite{robostudio}, a visual programming environment for rapid authoring and customization of complex services on a personal service robot.
+Imagine a medication reminder application which guides a person through the process of medication intake, such as the healthcare robotics project at the University of Auckland, New Zealand \cite{jayawardena12:_desig}.
+All interaction is by showing textual instructions on a display, and by speech generation.
+An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
+Scientists are working on applications such as the medication reminder~\cite{p11:_feasib} which run on mobile robots serving in nursing homes for the elderly.
+Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{jayawardena12:_desig,5649910}.
+The applications are intended to be developed by healthcare professionals using Robostudio~\cite{robostudio}, a visual programming environment for rapid authoring and customization of complex services on a personal service robot.
 %state machine based processes and a graphical tool for the designing of human robot interaction.
 Generally these people don't have profound knowledge of the complex mathematical syntax used for formal verification methods such as computation tree logic (CTL) or linear temporal logic (LTL).
-Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthbots.
-In case of the medication reminder mentioned above it mitght be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
-
-So we need a mechanism for the specification of such safety constraints which doesn't require special or expert knowledge and thus is easy to use for all kind of programmers.
+Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthcare robots.
+In case of the medication reminder mentioned above it might be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
+
+So a mechanism is needed for the specification of such safety constraints; one that does not require special or expert knowledge and thus is easy to use for all kind of programmers.
 A visual language intended to fill this gap is presented in section~\ref{sec:visualformalism}.
@@
 
-Also Del Bimbo et al.
-did work on a visual tool for temporal logic, but with a special focus on hierarchical representations of formulas~\cite{520786}.
+Also Del Bimbo {\em et al} worked on a visual tool for temporal logic, but with a special focus on hierarchical representations of formulas~\cite{520786}.
 Here each node of a tree is described by an abstract operator whose subformulas are branches to their replacement operators.
@@
 Both approaches are different fundamental ideas of representing temporal formulas.
-Whereas having a graph and inputs and outputs seems not to be applicable for our approach, the idea of hirarchical operator representation of Del Bimbo et al.
-will be a good base for our try to find an easy to use concept for constraint creation and understanding.
+Whereas having a graph and inputs and outputs seems not to be applicable for our approach, the idea of hirarchical operator representation of Del Bimbo {\em et al} will be a good basis for our concept of constraint creation and understanding.
 
@@
 Even though this approach touches the subject of healthcare it barely matches our problem.
-HomeTL focuses rather on monitoring temporal boundaries of a patient's behavior than on ensuring functional safety of an implemented program.
-But it depicts how nontechnical constraint design can be realised and its graphical design gives a good example how to handle time dependencies.
+HomeTL focuses more on monitoring temporal boundaries of a patient's behavior than on ensuring functional safety of an implemented program.
+It does depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.
 
@@
 
-We want to introduce a visual language for the defining of safety constraints for state machine definition of robot behaviour which is suitable for non-experts in the formal methods domain.
-Ideally its appliance is easy to use and intuitive to read in order to facilitate maintainability.
-Hence, the graphical formalism should abstract from the usually complex and mathematical concept of conventional formal methods but still stay unreduced in expressiveness compared to LTL as far as possible.
-
-Furthermore a heuristic should be developed of how to identify reasonable constraints on an underlying state machine program. 
-But since the particular domain distinguishes the significance of constraints respectively it might be difficult to show a universal heuristic working for all use cases.
+We want to introduce a visual language for defining safety constraints in state machine definitions of robot behaviour, that is suitable for people who are not experts in formal methods.
+The visual language should be easy to use and intuitive to read, in order to facilitate maintainability.
+Hence, the graphical formalism should abstract from the usually complex and mathematical concept of conventional formal methods but still be unreduced in expressiveness compared to LTL, so far as possible.
+
+Furthermore a heuristic should be developed, which identifies reasonable constraints on an underlying state machine program. 
+Since the particular domain semantics distinguish the significance of constraints, it may be difficult to show a universal heuristic.
 Hence, in this paper the concept of development support by constraint generation is examined within the domain of healthcare robotics.
-As a special requirement to the heuristic at least the following constraints are to be found for the medication reminder application:
+As a special requirement for the heuristic at least the following constraints are to be found for the medication reminder application:
 
 \begin{itemize}
-	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
-	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
+	\item The program will always eventually check the database for new reminder jobs. Thus we ensure that pending reminders are processed.
+	\item Either a reminded medication is taken properly, or caregivers are notified that the patient refused medication intake.
 % The next item is left out so that we can say, our heuristic found one we didn't think about yet :)
@@
 
-Some of the statements don't make sense of course if patients lie or pretend to take their medication.
-As our approach tries to validate state machine behaviour rather than human attitudes we assume correct collaboration or the use of tools which can recognize fairly medication intake.
-Furthermore also the correctness of external services used by the medication reminder such as the database module holding all reminders is not taken into account.
-
-All the functionality mentioned above shall be combined in one editor and for evaluation purpose later be integrated into a programming environment for state machine editing.
+Some of the statements do not make sense if patients lie or pretend to take their medication but don't take it.
+As our approach validates state machine behaviour rather than human attitudes we assume there is either cooperation from the patient or a separate tool that can verify medication intake.
+Furthermore we assume the correctness of external services used by the medication reminder such as the database module holding all reminders.
+
+% Bruce: I'm not sure we need this sentence. I think you have integrated it now anyway.
+% All the functionality mentioned above shall be combined in one editor and for evaluation purpose later be integrated into a programming environment for state machine editing.
 
@@
 
-The presented visual language is made for expressing constraints for state machines.
-Similar to Del Bimbo's formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchical manner.
+** COMMENT: This is a comment between the double stars; can delete it when you read it.
+I think we need a statement about the design here; I have tried to draft it; you may likely be able to improve!
+It should link the requirements to the design; I think that is pretty straightforward so just need a simple statement I think.
+**
+The presented visual language is designed to express constraints for state machines.
+To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below.
+In addition the visual language should be capable of expressing constraints about future steps, both (f) any future state and (h) the next state, that (g) events should always happen, and that (h) a property must be true until some future event.
+
+ Like Del Bimbo {\em et al's} formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy.
 However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
@@
 
-Regarding to the requirements the visual language should be very easy to use, especially for non-experts.
-Furthermore it should provide an intuitive appliance which is ideally learnable within little time.
-So there are some design decisions an editor for this visual language has to fulfill:
+To meet the requirements the visual language should be very easy to use, especially for non-experts.
+Furthermore it should provide an intuitive application which is ideally learnable within little time.
+So there are some design decisions an editor for this visual language must fulfill:
 
 \begin{itemize}
-	\item Each operator has a different significant color flavor. This kind of classification enables a faster subconscious identification by the user.
-	\item All editing is simply done by only drag and drop.
+	\item Each operator has a different significant visual appearance; in this case we have used color.
+This kind of classification enables a faster subconscious identification by the user.
+** COMMENT: can you refer to a GUI design text book regarding this?
+And how about colour blindness?
+Don't get me wrong; I am very happy with it; just could do with some citation about it.
+**
+	\item All editing is performed only by simple drag and drop.
 Operators and propositions can be moved and dropped onto other operators.
-A tool bar provides all kinds of operator and proposition types to be instantiated and ideally a trash can for deleting existing operators.
+A tool bar provides all operator and proposition types to be instantiated and there is a trash can for deleting existing operators.
 	\item The layout of constraints (nested operators) has two dimensions with a certain meaning.
@@
 It's intended to be similar to line coloring when reading a digital text.
-	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appeared to make the look unnecessarily complex.
+	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear unnecessarily complex.
 These operators can be merged as shown in figure~\ref{fig:nested_and}.
 	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
-	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances of them.
+	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances.
 Having no additional buttons reduces complexity and keeps it simple.
@@
 In order to support the developer in specifying safety constraints, automated generation and suggestion of constraints might be helpful.
-But it's helpful only then, if the suggested constraints are relevant and reasonable.
+But it's helpful only if the suggested constraints are relevant and reasonable.
 This however depends on the domain and use case.
-Regarding to the postulated constraints in the requirements in section~\ref{sec:requirements} it turned out that all therefor relevant states have got one of the following characteristics:
+Regarding the postulated constraints in the requirements in section~\ref{sec:requirements} all the relevant states have one of the following characteristics:
 %following types of states are of a special interest:
@@
 
-In order to compute the required constraints and possibly even more of them automatically we manage with so called subgraphs defined as follows:
+In order to compute the required constraints automatically, and possibly additional ones, we manage a set of subgraphs defined as follows:
 %Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
@@
 	\item It has a start state ($\alpha$-state) with at least two outgoing transitions.
-	\item There is an end state ($\beta$-state) in which all possible paths starting from start state come together again.
-	\item All states between start and end state have no other incoming transitions than the ones originally coming from start state.
-That means there is no path to visit these states exept through start state.
-	\item start state may have incoming transitions from states not contained in the subgraph.
-	\item end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
+	\item There is one ** COMMENT: one not an? ie can only be one? ** end state ($\beta$-state) in which all possible paths starting from start state come together again.
+	\item All states between the start and end states have no other incoming transitions than the ones originally coming from the start state.
+That means there is no path to visit these states except through the start state.
+	\item The start state may have incoming transitions from states not contained in the subgraph.
+	\item The end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
 \end{itemize}
@@
 
-Using this specification all possible subgraphs can be detected within a state machine.
-If a subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarily for $\beta$ and all $\gamma$'s):
+Using this specification all possible such subgraphs can be detected within a state machine.
+If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarily for $\beta$ and all $\gamma$'s):
 
@@
 All formulas can be translated to the visual formalism presented in section~\ref{sec:visualformalism} and shown to the user.
-The accordant LTL formulas for the postulated constraints in section~\ref{sec:requirements} belong to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:
+The appropriate LTL formulas for the postulated constraints in section~\ref{sec:requirements} belong to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:
 
@@
 
-To experience and evaluate the visual formalism as well as the usefulness of constraint generation we developed an editor which provides functionalities all proposed functionalities while taking the requirements into account.
-It's implemented as a java swing component, making it easy to integrate the visual language into other java applications.
-An API allows to set the underlying state machine what also influences the state propositions available for constraint editing.
-Furthermore a particular model checker can be specified to be used for constraint validation.
-As default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046}~\cite{NuSMV2} is used for constraint validation.
-
-Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right which contains drag providers for the creating of all operator and proposition types as well as a trash for deleting.
+To explore and evaluate the visual formalism as well as the usefulness of constraint generation we developed an editor which provides all proposed functionality while taking the requirements into account.
+The implementation is a java swing component, making it easy to integrate the visual language into other java applications.
+An API enables an underlying state machine which also influences the state propositions available for constraint editing.
+Furthermore a particular model checker can be specified for constraint validation.
+As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
+
+Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right which contains drag providers for the creating of all operator and proposition types as well as a trashcan for deleting.
 Constraints can be composed in the dashboard in the middle of the editor by drag and drop.
-The tab functionality on the left allows to manage several constraints.
-Each tab schows a small thumbnail of the constraint and a symbol indicating its validity what is a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress''.
-The magic wand button within the tab pane triggers the constraint generation.
+The tab functionality on the left allows several constraints to be managed.
+Each tab shows a small thumbnail of the constraint and a symbol indicating its validity, which is a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'' The magic wand button within the tab pane triggers the constraint generation.
 
@@
 
-In order to evaluate the proposed safety functionalities on various programs the tool was integrated into Robostudio~\cite{robostudio}.
+In order to evaluate the proposed safety functionality the tool was integrated into Robostudio~\cite{robostudio}.
 It allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
-As an example of how to use them we will have a closer look at the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}.
-Starting from ``Menu'' the user can choose from several services such as blood pressure measurement ore entertainment mode.
+As an example we consider the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}.
+Starting from ``Menu'' the user can choose from several services such as blood pressure measurement or entertainment.
 In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part.
 In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered.
-The patient can state if he's already taken his medication, or decide whether he wants to take it or not.
-In the latter he may give a reason for it and a caregiver gets notified about this incident.
+The patient can state if he or she has already taken the medication, or decide whether to take it or not.
+In the latter he or she may give a reason for it and a caregiver is notified about this incident.
 Otherwise it will result in ``Well done!'' after completing intake.
@@
   \includegraphics[width=\linewidth]{stm} %width=0.7\linewidth
-  \caption{Simplified screen-flow of the medication reminder application.}
+  \caption{Simplified screen-flow of the medication reminder application. ** COMMENT: THIS ONE IS TOO SMALL! SHOULD MAKE IT BIGGER TO MAKE TEXT READABLE **}
   \label{fig:medicationreminder}
@@
 
-Owing to the elaborated performance of NuSMV even the validation of constraints on huge and complex programs is fast and allows quick feedback.
-Besides, the validation itself runs in background without locking the dashboard.
-Therefore we got rid of waiting times and disruptions during constraint development what would be very well the case with conventional model checking where constraint development and constraint validation are alternating occupations.
-Thus it's considered as an improvement of user experience.
+NuSMV performs well so even the validation of constraints on huge and complex programs is fast and allows rapid feedback.
+In addition, the validation itself runs in the background without locking the dashboard.
+Therefore we have removed waiting times and disruptions during constraint development, which would be likely with conventional model checking where constraint development and constraint validation are alternating processes.
+Thus we consider our tools an improvement for the users' experience.
 
@@
 The automated constraint generation can be triggered by a click on the magic wand button in the tab area.
-After activation a dashboard gets opened in a new tab for each found constraint, and validation is initiated immediately.
-In case of the medication reminder application six constraints are found, whereof \emph{A)} and \emph{B)} shown in figure~\ref{fig:generatedconstraints} are actually matching the postulations in the requirements in section~\ref{sec:requirements}.
-The constraint used for demonstration in the previous chapter gets also generated, however \emph{B)} forms an intensified restriction of it.
-
-Constraint \emph{A)} ensures the ``Polling'' state to be always eventually visited again.
-If a pending medication is not already taken, constraint \emph{B)} guarantees intake or staff notification before next reminders can be read from database.
-
-We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{C)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment mode ore blood pressure measurement.
+After activation a dashboard is opened in a new tab for each found constraint, and validation is initiated immediately.
+For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in figure~\ref{fig:generatedconstraints} which match the postulations in the requirements in section~\ref{sec:requirements}.
+The constraint used for demonstration in the previous section is also generated, however \emph{b)} forms an intensified restriction of it.
+
+Constraint \emph{a)} ensures the ``Polling'' state is always eventually visited again.
+If a pending medication is not already taken, constraint \emph{b)} guarantees intake or staff notification before the next reminders can be read from the database.
+
+We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{c)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment or blood pressure measurement.
+
+** COMMENT: Possible comment here about the scalability and practicality about constraint generation; do you think it will explode in larger cases? Either way a comment might be useful. **
 
@@
 
-The editor presented in this paper proofes validity of all constraints and presents the result \emph{valid} or \emph{invalid} to the user.
+The editor presented in this paper proves validity of all constraints and presents the result \emph{valid} or \emph{invalid} to the user.
 But in contrast to conventional model checking no additional explanation for the outcome such as a counter example is given.
@@
 Also the portability of the constraint generation heuristic to other domains should be examined and different methods than subgraph based constraint finding might be elaborated.
+
+A more systematic evaluation of the tool is required, however this is beyond the scope of the current project.
+Now that a visual language and tool have been proposed, the next step is to evaluate the tool on a number of applications, make any required modications, then evaluate the tool in a user study for healthcare robotic applications.
 
@@
 In this paper we represented a visual formalism for the defining of safety constraints and a concept of automated constraint generation.
-Both features are implemented and assembled in an editor for standalone or integration use.
-The visual language is pretty simple what makes it accessible to a wider range of software developers.
-Nevertheless its expressiveness is fairly powerful and allows serious model checking.
-
-The concept of constraint generation worked very well for our use case the medication reminder application and all postulated constraints were found.
-Since we even discovered a new reasonable constraint we are convinced this concept as well as the visual language for constraint defining can be a beneficial supporting thing for plenty of developers.
+Both features are implemented and assembled in an editor for standalone or integrated use.
+The visual language is reasonably simple which makes it accessible to a wider range of software developers.
+Nevertheless it is reasonably expressive and allows serious model checking.
+
+The concept of constraint generation worked very well for our use case, the medication reminder application, and all postulated constraints were found.
+Since we even discovered a new reasonable constraint we are convinced this concept as well as the visual language for defining constraints has the potential to support development of safety critical healthcare robot applications.
 

### 2012-08-12 paper/Paper.tex
"Not taken already" changed to "Not taken yet" in state machine

--- 2012-08-08 paper/Paper.tex
@@
 Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthbots.
-In case of the medication reminder mentioned above it mitght be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
+In case of the medication reminder mentioned above it might be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
 
@@
 The latter would mean there are errors in the program which lead to undesired constraints.
-Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
-
-In section~\ref{sec:automatedconstraintgeneration} we introduce an heuristic for finding reasonable safety constraints based on a state machine in the healthcare robotics domain.
+Furthermore once constraints are created and checked for plausibility they can be validated after every program change and thus ensure integrity during the development process.
+
+In section~\ref{sec:automatedconstraintgeneration} we introduce an heuristic method for finding reasonable safety constraints based on a state machine in the healthcare robotics domain.
 

### 2012-10-11 thesis/MA-Thesis.tex
Initial commit of thesis

\documentclass[10pt,a4paper,headsepline,footsepline,DIV13,BCOR12mm]{scrbook}
\usepackage[english,ngerman]{babel}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{scrpage2}
\pagestyle{scrheadings}

\usepackage{graphicx}

\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{amsmath}

% LTL Symbols
\usepackage{latexsym}
\usepackage{MnSymbol}

\usepackage{wrapfig}

\usepackage{listings}
\usepackage{color}
\definecolor{gray}{rgb}{0.4,0.4,0.4}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
\definecolor{cyan}{rgb}{0.0,0.6,0.6}

\begin{document}

%PREFACE
\selectlanguage{ngerman}
\input{titlepage} 
\newpage
%select english as language!
\selectlanguage{english}
%\pagenumbering{roman}
\pagenumbering{arabic}
\tableofcontents
\newpage
\listoffigures
\newpage
\listoftables
%\pagenumbering{arabic}

\graphicspath{{./images/}}
\DeclareGraphicsExtensions{.jpg}

\input{tex/1_abstract}
\input{tex/2_introduction}
\input{tex/3_background}
\input{tex/4_goals}
\input{tex/5_design}
\input{tex/6_implementation}
\input{tex/7_prototypeandevaluation}
\input{tex/8_conclusionandfuturework}

\chapter*{Acknowledgment}

%This work evolved from a master thesis project provided by the Elite Graduate Program Software Engineering. 
%For the received support I would like to thank the people of this program and the three patron universities: University of Augsburg, Technical University Munich and Ludwig-Maximilians-University Munich.

\bibliographystyle{unsrt}
\bibliography{Bib}

\end{document}

### 2012-10-11 thesis/tex/2_introduction.tex
Initial commit of thesis

\chapter{Introduction}

Safety critical robot applications require extensive testing or formal verification in order to achieve adequate safe and predictable behavior.
Even if a program does not control actuators itself and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions involving several actors.
Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University of Auckland, New Zealand \cite{jayawardena12:_desig}. All interaction is guided by showing textual instructions on a
\begin{wrapfigure}{r}{0.58\textwidth}
  \centering
  \includegraphics[width=0.56\textwidth]{cafero_being_used}
  \caption{Service robot being used by older person~\cite{robostudio}.}
  \label{fig:cafero_being_used}
\end{wrapfigure}
display, and by speech generation. An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
Scientists are working on applications such as the medication reminder~\cite{p11:_feasib} which run on mobile robots serving in nursing homes for the elderly. Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{jayawardena12:_desig,5649910}.

With service robots coexisting with people and acting within their work\-spa\-ces, coping with reliability and safety issues is essential for trust and acceptance of these robots. In some cases safety certifications based on international standards for functional safety, such as ISO 13482 and IEC 61508, might be even required before robots get released for use or obtain concession for sale.
However, considering functional safety while developing service robots is both complex and difficult; maybe one reason for the absense of safety checks and warranties in the current healthbot versions at the University of Auckland.
But facing the fact that safety requirements for service robots are expected to increase in future, there is a plan of slowly integrating safety functionality into the healthcare robotics project as a next step starting with the concepts introduced in this paper.

The applications for the healthcare robots mentioned above are intended to be developed by healthcare professionals using Robostudio~\cite{robostudio}, a visual programming environment for rapid authoring and customization of complex robot services.
%state machine based processes and a graphical tool for the designing of human robot interaction.
Generally these people don't have profound knowledge of the complex mathematical syntax used for formal temporal logic such as computation tree logic (CTL) or linear temporal logic (LTL). Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthcare robots.
In case of the medication reminder mentioned above it might be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
So a mechanism is needed for the specification of such safety constraints; one that does not require special or expert knowledge and thus is easy to use for all kind of programmers. A visual language intended to fill this gap is presented in section~\ref{sec:operatorconstraints}.

Unfortunately the supply of an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer. Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
To support the user in finding significant constraints it might be useful to automatically compute constraint suggestions which make true testimonies about the current designed program. First of all the generated constraints provide an opportunity for the developer to identify reasonable constraints as well as contradictions between constraints and specification. The latter would mean there are errors in the program which lead to undesired constraints. Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process. In section~\ref{sec:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine descriptions that describe robots' behaviors.

Both ideas are to be integrated into Robostudio in order to simplify respectively facilitate the use of safety mechanisms for application development for service robots. Thereby a desired safety certification for the healthcare service robots can move closer one step.

This work starts with background information about the healthcare domain, short explanations of common safety mechanisms and insights into related work in chapter~\ref{chap:background}. After the goals are mentioned in chapter~\ref{chap:goals}, design decisions are introduced in chapter~\ref{chap:design}.
A test scenario is described in chapter~\ref{chap:testscenarioandevaluation}, where the presented tools get evaluated.
Chapter~\ref{chap:conclusionandfuturework} gives a short conclusion and an outlook how the work presented in this paper could be continued.

%Once such constraints are defined, they determine the grade of safety coverage. Unfortunately, the program may still contain errors if there is at least one significant constraint missed out. To support the user in finding such ones the tool could furthermore automatically compute suggestions for possible constraints based on the designed statemachine. These are displayed to the programmer and may help him identifying reasonable constraints as well as contradictions between program and specification.

%Our approach presented in this paper has a special focus on the healtcare domain.

%Dieses beispiel ist eine Anwendung eines umfangreichen Healthcare-Projekts, an dem im robotics lab der Universitï¿½t von Auckland gearbeitet wird. Die zustansbasierten Anwendungen laufen auf Cafero, einem mit blutdruckmessgerï¿½t und anderen tools austestatteten mobilen Roboter mit Touchdisplay.
%Die einzelnen Anwendungen werden von Healthcare-Professionals mit dem Tool Robostudio [cite] entwickelt, das eine graphische oberflï¿½sche zur modellierung von zustandsmaschinen externen services und  bietet.
%Allerdings haben diese not to have profound knowledge of the complex mathematical syntax used for formal verification methods such as computation tree logic (CTL) or linear temporal logic (LTL)).
%Dennoch mï¿½chte man gegenï¿½ber den Patienten, die von solchen Robotern gepflegt werden, ein gewisses Level an Sicherheit gewï¿½hrleisten kï¿½nnen. Beispielsweise soll in obigem Beispiel sichergestellt werden kï¿½nnen, dass ein Medikament nach der Einnahme nicht erneut in einer Erinnerung auftaucht, oder dass bei Nichteinnahme ein Mitarbeiter benachrichtigt wird.
%Dazu bedarf es allerdings Mechanismen zur spezifizierung solcher Sicherheitseigenschaften, die fï¿½r den Programmentwickler zugï¿½nglich und einfach (am besten ohne spezialwissen) zu benutzen sind.

%Among others robots are alredy used in elderly care centers for healthcare purpose.

% -->

%Robostudio (oder healthcare applikation) soll irgendwann Sicherheitszertifizierung bekommen (erklaeren warum). Daher ist es wichtig, sicherheitsmechanismen zur verfuegung zu stellen. Einen kleinen Schritt in diese Richtung möchten wir mit dieser Arbeit machen...
%Anwender kommen aus verschiedenen Domaenen => need of easy-to-use safety functionality

%Aiming for safe and reliable robots,
%IEC 61508 certification. It contributes to reducing costs and development time when creating highly-dependable and safe robots.

%Glossary IEC 61508
%IEC 61508 is an international standard, produced in 2000 by the IEC, for the realisation of functional safety in devices utilising computer technology. Functional safety differs from inherent safety in that its goal is to reduce risk to permissible levels using active techniques, rather than passive techniques. IEC 61508 specifies guidelines for developing safety-related hardware and software, and management techniques for development processes to ensure accurate designs and implementations.
% <--

### 2012-10-11 thesis/tex/3_background.tex
Initial commit of thesis

\chapter{Background}
\label{chap:background}

Das Thema dieser Arbeit hat sich aus der Situation heraus entwickelt, dass die healthbots der UoA mit sicherheitsmechanismen ausgestattet werden sollen. Die Motivation hierfür, die Roboter mitsamt ihrer Anwendungen und deren technischer Hintergrund werden in Section 1 beschrieben. Auf das VPE, dass die Programmierung dieser Roboter ermöglicht, wird in section 2 eingegangen. Zum Verständnis der sicherstellungskonzepte für sicherheit, auf die diese Arbeit aufbaut, gibt Section 3 eine kurze Einführung in die Welt der funktionalen Sicherheit. In Section 4 werden andere Arbeiten und Programme untersucht, die sich ebenfalls mit Sicherheitüberprüfung beschäftigen.

\section{Healhbot application}
\label{sec:healthbotapplication}

In einigen Ländern, darunter Neuseeland, Japan und Deutschland, ist das Durchschnittsalter der Bevölkerung stets ansteigend. Dank hervorragenden Errungenschaften in der Medizin sowie höheren Lebensstandards lebt der Mensch in Durchschnitt immer länger. Das hat zur Folge, dass auch die Nachfrage an Plätzen in Altersheimen oder nach Dienstleistungen sozialer Einrichtungen wächst.
Doch ist in diesen Ländern ein enormer Mangel an professionellen Hilfskräften zu verzeichnen. Das mag mitunter daran liegen, dass in der Regel für soziale Arbeit in diesem Umfeld ein niedrigerer Lohn bezahlt wird als vergleichsweise für technische Berufe. Doch eine Anhebung der Gehälter ist nicht ohne weiteres möglich, denn auch zu den derzeitigen Preisen können sich viele ältere Leute keine persönliche Pflegekraft leisten.

Aiming for fillig the constantly increasing gap in professional personnel for healtcare of the elderly and bypassing die financial burden, service robots are supposed to more and more undertake the task of care.
There are lots of possible tasks a robot could do such as taking the blood pressure, guiding medication intake, detecting falls or just entertaining with music, video or internet.
This challenge as a goal, the healtcare robotics team at the University of Auckland researches this topic and builds service robots for supporting the elderlies in their daily life.

%\begin{wrapfigure}{r}{0.58\textwidth}
%  \centering
%  \includegraphics[width=0.56\textwidth]{irobiq_cafero}
%  \caption{IrobiQ and Cafero, two robots of the healtcare project.}
%  \label{fig:irobiq_cafero}
%\end{wrapfigure}
The two newest robots IrobiQ and Cafero are shown in figure~\ref{fig:robots}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\linewidth]{irobiq_cafero} 
  \caption{IrobiQ and Cafero, two robots of the healtcare project.}
  \label{fig:robots}
\end{figure}
These mobile robots have got a touch display for presenting screens and receiving user input. Furthermore they are equipped with additional external tools such as the blood pressure measurement device.
Applications running on the robot use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedbak. User inputs or response from devices cause dialogs to change. Figure~\ref{fig:screenflow_example} presents a little example of such a screen-flow based application, which uses response from a camera and user input for navigating through the screen-flow.

%, and they can access devices such as the camera for face recognition or the blood pressure measurement tool.

There is a program interpreter running on the robots which can load and execute program files. Screen dialogs get generated and displayed on the touch screen.
%This makes the program behavior replaceable and maintenance can be done easily.
The interpreter processes programs written in Robot Behavior Description Language (RBDL), a particular DSL based on XML, which has been developed specially for the healthcare robotic applications. It allows defining state machine based program behavior and creating UI elements for screen visualizations. An example implementation of one state in RBDL is shown in listing~\ref{lst:RBDL}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{screenflow_example} 
  \caption{Robot screen-flow dialog based interaction example~\cite{robostudio}.}
  \label{fig:screenflow_example}
\end{figure}

\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  showstringspaces=false,
  commentstyle=\color{gray}\upshape
}

\lstdefinelanguage{XML}
{
  morestring=[b]",
  morestring=[s]{>}{<},
  morecomment=[s]{<?}{?>},
  stringstyle=\color{black},
  identifierstyle=\color{darkblue},
  keywordstyle=\color{cyan},
	morekeywords={%no, type_, senderid, receiverid, name_, vartype, label, width, height, x, y, textsize, preconditions
	}% list your attributes here
}

%\lstset{language=XML}
\begin{lstlisting}[language=XML, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a state defined by RBDL~\cite{robostudio}., label=lst:RBDL]
<state no="84">
	<backgroundactions>
		<rocosmessage type="EVENT MESSAGE" senderid="RCP" receiverid="NA" name="FDSessionStart">
			<parameter vartype="text" type="unsigned int" name="id">1</parameter>
			<parameter type="string" name="" />
		</rocosmessage>
	</backgroundactions>
	<screen>
		<components>
			<button label="START" width="250" height="100" x="387" y="600" textsize="40">
				<event name="clicked">
					<action preconditions="no" name="transition">
						<parameter>
							<type>state</type>
							<name>n</name>
							<value>86</value>
						</parameter>
					</action>
				</event>
			</button>
		</components>
	</screen>
	<expectedevents>
		<event name="TimeOut">
			<action preconditions="no" name="transition">
				<parameter>
					<type>state</type>
					<name>n</name>
					<value>VARFirstScreen</value>
				</parameter>
			</action>
		</event>
	</expectedevents>
</state>
\end{lstlisting}

For each state of the program, backgroundactions can be defined. They are are transparent to the user and run in background. Text messages can be sent or external devices can be acessed, for example. The latter is realized with web services to obtain a modular design.
Some states can also have screen definitions with buttons, messages, videos, etc. on it. These are displayed to the user when the particular state is active. Once displayed, buttons can be clicked and trigger events.
The behavior of events can be mapped in the expectedevents tag. It lists all events which are processable by the current state and defines the correspondend behavior such as triggering transitions to other states.

For this work the most important thing about the RBLD to remember is, that it can define state machine behavior with states, transitions and events. This fact was basement for later decisions regarding used safety mechanisms.

%Zugriff auf devices via web services
%Windows is used as operating system on the robots and the programs are executed in Adobe flash and Actionscript. External devices are accessed over web

\section{Robostudio}

In the beginning all programming of the robot behavior was done by directly editing the XML code mentioned in the previous section by using a text editor. But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise. For this Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behavior on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behavior on a visual layer and generates the corresponding program file containing the RBDL code fully automatically. This file can be downloaded to the robot and executed by the interpreter. Figure~\ref{fig:robostudio_proceeding} illustrates what the steps of robot behavior development are.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{robostudio_proceeding} 
  \caption{How Robostudio is used for robot behavior development.}
  \label{fig:robostudio_proceeding}
\end{figure}

%require refactoring and static analysis tools.

Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application.
In several windows and views there are visual tools for editing and understanding robot behavior as well as screen dialog design. As depicted in figure~\ref{fig:robostudio_vpe} the editor has got a State Navigator Window (1) in the top left corner where all existing states are listed. Underneath an overview about all incoming and outgoing transitions is given. New states can be created and existing states can be deleted. A click on one particular state id in the State Navigator Window will cause all other components to show all possible information regarding to the selected state.
The Expected Events Editor Window (2) in the bottom left provides specifying the expected events and their effects.
Heart and center of the editor is the UI Component Layout Editor Window (3). It renders the screen dialog of the selected state and gives a preview of the screen presented on the robot during runtime. Components such as buttons, text boxes or images can be added, changed or removed via drag and drop.
They are provided by the UI Components Palette Window (4) on the right side where the user can choose from all supported UI components.
The Background Actions Editor Window (5) lets the user define background actions, which are executed trasparently to the user such as accessing external devices or sending messages.
Aiming for supporting the developer with helpful graphical tools the State Transition Visualization Window (6) gives an overview about all connections towards and from the selected state. It helps to quickly get an understanding of the local program behavior.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{robostudio_vpe} 
  \caption{Window composition in Robostudio.}
  \label{fig:robostudio_vpe}
\end{figure}

TODO: Vlt herausstellen, dass eine Schwierigkeit darin bestand, Verhaltensmodellierung und graphische Gestaltung unter einen Hut zu bringen (dazu Bild aus datta paper?).

\section{Behavior checking}
TODO: Begriffserklärung: Funktionale Sicherheit

\subsection{State machines}

\subsection{Model checking \& LTL}
\label{sec:modelcheckingandltl}

%CTL and LTL are important concepts for model checking and verification of statemachines. But their powerfulness is reflected in the complexity of their mathematical expressions, what makes it difficult to use for programmers who haven't got deeper expertise in this subject.

NuSMV erwaehnen, erklaeren warum nicht CTL? ltl passt fuer healthcare besser

\section{Related work}

%Was gibt es schon alles?

% 1993 A Visual-Programming Environment for a Temporal Logic Language
% 1995 Visual Specification of Branching Time Temporal Logic
% 2007 HomeTL: A visual formalism, based on temporal logic, for the design of home based care
% 2007 A Visual Editor to Support the Use of Temporal Logic for ADL Monitoring

A first step towards a graphical tool for designing safety constraints is taken by Sisiruca and Ionescu~\cite{332301}. They developed an object-oriented graphical environment for visually creating temporal logic sentences and rules. The tool presents temporal operators as logical modules with boolean inputs and outputs which can be connected to one big expression graph. Del Bimbo et al. worked on a visual tool for temporal logic, but with a special focus on hierarchical representation of formulas~\cite{520786}. Here each node of a tree is described by an abstract operator whose subformulas are branches to their replacement operators. Thus the root node represents the entire formula. They also explored the formula trees' 3D representation within a virtual space. Both approaches are different fundamental ideas of representing temporal formulas. Whereas having a graph  seems not to be applicable for our approach, the idea of hirarchical operator representation of Del Bimbo et al. will be a good basis for our concept of constraint creation and understanding.

%aim for simplifying the development of constraints, but they still demand knowledge of the underlying mathematical formalism of temporal logic.

The step towards a more abstract level of development is taken by HomeTL~\cite{4341725}, a visual formalism for the design of home based care. It allows healthcare professionals to specify rules and sequences of user actions in a quite nontechnical manner. Based on these conditions a monitoring system in a patient's home environment can detect and report abnormal situations within the daily routine. Even though this approach touches the subject of healthcare it barely matches our problem. HomeTL focuses more on monitoring temporal boundaries of a patient's behavior than on ensuring functional safety of an implemented program. It does depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.

%There are already works about constraint generation such as one approach of NASA [4], for example. They used test oracles in order to achieve consistency between program code and specification during development. 
% Das wollen wir auch erzielen mit unseren automatisch generierten constraints...

%But these kind of oracles are for automated test purpose and are rather complex than intuitive. However, our goal is to find nonmathematical and just intuitive ones. An intelligent heuristic must determine key states and possible conjunctions of them in order to build reasonable constraints. Afterwards they get filtered by an algorithm which computes importances for each constraint based on state weights. Finally the developer obtains a set of visual constraints which he can then check for plausibility.

\subsection{Sisiruca and Ionescu}
\subsection{Del Bimbo (3D)}
\subsection{HomeTL}

### 2012-10-11 thesis/tex/5_design.tex
Initial commit of thesis

\chapter{Design}
\label{chap:design}

%On our way towards an useful and simple concept for the defining of safety criteria

%wichtig:
%1. VL
%2. Constraint generierung
%3. integrierung in Robostudio

%different ideas came up

%In this chapter we will intruduce the final design of the visual language and compare it to a first idea of a visual formalism.

In this chapter we introduce a first approach of a visual formalism and later state the final desing of the visual language. Furthermore the idea of automated constraint generation is presented in section~\ref{sec:automatedconstraintgeneration}. The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}. However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots. Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.

%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions. The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually. A common LTL model checker will be added to it as well which fulfils the verification of the translated constraints on the designed statemachine. In case of verification failures the concerning constraints and a visual counter example can be presented.

\section{Visual formalisms}
\label{sec:visualformalisms}

One main difficulty in finding a proper visual formalism was to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points. A first approach, so called template constraints, and its disadvantages are described in section~\ref{sec:templateconstraints}. Subsequently an improved design is presented, which tends to a more abstract direction.

\subsection{Template constraints}
\label{sec:templateconstraints}

What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Leaded by this question, we developed a visual language based on templates. These templates make different propositions about states and their relations. The user can choose from several graphical constraint types and customize them by parameterizing state variables. So far there are two different kinds of template constraints:

\begin{enumerate}
	\item Whenever state (x) is active, state (y) must be visited before state (z) can be reached.

	The idea behind this constraint is, that propositions about temporal dependencies can be made. Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled. Example of a constraint: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z). Going to bed without brushing teeth after meal should never happen.
	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.

	This constraint type enforces at least one particular event to happen after a spezific precondition.
	Example of a constraint: Every time a person enters a supermarket (x) he will eventually decide to buy 
	%a product or not to buy it.
	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can't avoid both possibilities.
	Nitpicking readers might note the person could also steel something without buying it, and thus not buy anything but leave the shop with full hands. Of course this is true; exactly in this case our constraint would signal an undesirable "program" behavior with its invalidity.
\end{enumerate}

Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like. Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot. Furthermore the state labels should sometimes be replaced by the screen displays of the regarding states. All parameters of the template constraints can be customized by just clicking on one robot suface. In the subsequently appearing popup window (figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.7]{templateconstraints}
  \caption{The two template constraint types, parameterized with particular states.}
  \label{fig:templateconstraints}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.7]{templateconstraint_dropdown}
  \caption{Dropdown menu for spezifying template parameter.}
  \label{fig:templateconstraint_dropdown}
\end{figure}

This visual language was presented to potential users for evaluation and feedbak. Soon it turned out that there are some problems and disagreements. The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
A try to solve this problem was to equipp both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
But unfortunately even the textual eplanations don't seem to be a satisfying solution. For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.

However, there is a second lack with the template constraints.
Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can't be described. Not even the check of a state being visited never or at least once is possible. This lack of expressiveness appears to be too big and for this reason the approach of template constraints got dismissed.

%idee: einfache verwendung
%sehr spezialisiert auf healthcare anwendung

\subsection{Operator constraints}
\label{sec:operatorconstraints}

With our second approach we try to gain more expressiveness and uniqueness for the visual language. For this we decided to decrease the abstraction level and design a visual formalism for linear temporal logic (LTL, see~section~\ref{sec:modelcheckingandltl}). Since the visual constraints can be directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness. If it is considered as easy to use, two birds get killed with one stone: Non-experts can use it as well as experts.

% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv. der always operator könnte umgedreht sein.

%The presented visual language is designed to express constraints for state machines.
To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below. In addition the visual language should be capable of expressing constraints about future steps, both (h) any future state and (g) the next state, that (f) events should always happen, and that (i) a property must be true until some future event. Like in Del Bimbo et al.'s formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy. However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
Each block represents either an operator or a proposition and has its specified semantics. Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.

The visual language has support for eight operator types and one proposition as shown in figure~\ref{fig:operators}.

\begin{description}
	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided. Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state. Other types such as numeric or string equations may be added but are less important for our healthcare subject.
	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
	%\Box \Diamond \medcircle
	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true. In addition eventually there has to be a future state with $\psi$ being true.
\end{description}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{table}
  \caption{The visual language consists of eight unary or binary operator types and one state proposition, each of them having its own color flavor. Logical operators are arranged in a vertical row wereas time relevant statements are aligned horizontally.}
  \label{fig:operators}
\end{figure}

To meet the requirements the visual language should be very easy to use, especially for non-experts. Furthermore it should provide an intuitive application which is ideally learnable within little time. So there are some design decisions this visual language must fulfill~\cite{moody-physics-of-notations}:

\begin{itemize}
	\item Each operator has a different significant visual appearance; in this case we have used color. This kind of classification enables a faster subconscious identification.
	\item The layout of constraints (nested operators) has two dimensions with a certain meaning. Figure~\ref{fig:directions} demonstrates the ``logical'' vertical read direction as well as the time releveant horizontal line.
	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear unnecessarily complex. These operators can be merged (see~figure~\ref{fig:nested_and}).
\end{itemize}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{and_simplify} %width=0.7\linewidth
  \caption{In order to simplify the appearance of constraints certain operators might be merged. Merged nested \emph{AND} operators (center) are less complex as default ones (left). However nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
  \label{fig:nested_and}
\end{figure}

Also an editor for this visual language can contribute to usability and clarity. Some requirements are as follows:

\begin{itemize}	
	\item All editing is performed only by simple drag and drop. Operators and propositions can be moved and dropped onto other operators. A tool bar provides all operator and proposition types to be instantiated and there is a trash can for deleting existing operators.
	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer. All higher operators become bleached out so that the hovered operator and its children appear highlighted. It's intended to be similar to line coloring when reading a digital text.
	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
	\item There are different statuses which inform about validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed. Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign (see~figure~\ref{fig:results}).
	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances. Having no additional buttons reduces complexity and keeps it simple.
\end{itemize}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{directions}
  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
  \label{fig:directions}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{results}
  \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
  \label{fig:results}
\end{figure}

\section{Automated constraint generation}
\label{sec:automatedconstraintgeneration}

As already mentioned before, it might be difficult to find all reasonable safety constraints needed for full safety covery. Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
To help prevent this to happen, we introduce the concept of automated constraint generation for state machines. Once a program reaches a complete condition, constraint suggestions which are valid on the current program can be computed fully automatically. On the one hand they may help the developer identifying reasonable constraints he didn't think about as well as contradictions between program and specification, eg. if a computed constraint doesn't make sense at all. On the other hand all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs, automated generation and suggestion of constraints might be helpful. But it's helpful only if the suggested constraints are relevant and reasonable. This however depends on the domain and use case. It has to be mentioned that our approach presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
Regarding the important constraints postulated in chapter~\ref{chap:goals}, all the relevant states fall into one of the following categories:

\begin{enumerate}
	\item States having two or more outgoing transitions ($\alpha$-states).
	\item States in which all paths starting from one $\alpha$-state come together again ($\beta$-states).
	\item All States immediately before $\beta$ states ($\gamma$-states).
\end{enumerate}

In order to compute the required constraints automatically, and possibly additional ones, we deal with subgraphs defined as follows:
%Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

\begin{itemize}
	\item There is a start state ($\alpha$-state) with at least two outgoing transitions.
	\item There is one end state ($\beta$-state) in which all possible paths starting from start state come together again.
	\item All states between the start and end states have no other incoming transitions than the ones originally coming from the start state. That means there is no path to visit these states except through the start state.
	\item The start state may have incoming transitions from states not contained in the subgraph.
	\item The end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
\end{itemize}

\begin{figure}[htbp]
  \centering

  \tikzstyle{state} = [circle,fill=white,draw]
  \tikzstyle{statein} = [circle,fill=blue!20,draw]
  \tikzstyle{every path} = [font=\sffamily\small]

	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=15mm]

	  \node[statein] (1) {$\alpha$};
	  \node[state] (2) [above left of=1] {};
	  \node[state] (3) [below left of=1] {};
	  \node[statein] (4) [above right of=1] {};
	  \node[statein] (5) [below right of=1] {};
	  \node[statein] (8) [right of=5] {$\gamma$};
	  \node[statein] (6) [right of=4] {$\gamma$};
	  \node[statein] (7) [below right of=4] {$\gamma$};

	  \node[statein] (9) [right of=7] {$\beta$};
	  \node[state] (10) [above of=9] {};
	  \node[state] (11) [above right of=9] {};
	  \node[state] (12) [below right of=9] {};

	  \path[dashed]
	    (2) edge node [left] {} (1)
	    (3) edge node [left] {} (1)
	    (10) edge node [left] {} (9)
	    (9) edge node [left] {} (11)
	    (9) edge node [left] {} (12);
		\path[]
	  	(1) edge node [left] {} (4)
	        edge node [left] {} (5)
	    (4) edge node [left] {} (6)
	        edge node [left] {} (7)
	    (5) edge node [left] {} (8)
	    (6) edge node [left] {} (9)
	    (7) edge node [left] {} (9)
	    (8) edge node [left] {} (9)
	    ;

%	  \node[statein] (1) {$\alpha$};
%	  \node[state] (2) [above left of=1] {};
%	  \node[state] (3) [above right of=1] {};
%	  \node[statein] (4) [below left of=1] {};
%	  \node[statein] (6) [below of=4] {$\gamma$};
%	  \node[statein] (7) [below right of=4] {$\gamma$};
%	  \node[statein] (5) [below right of=1] {};
%	  \node[statein] (8) [below of=5] {$\gamma$};
%	  \node[statein] (9) [below of=7] {$\beta$};
%	  \node[state] (10) [right of=9] {};
%	  \node[state] (11) [below left of=9] {};
%	  \node[state] (12) [below right of=9] {};
%	  
%	  \path[dashed]
%	    (2) edge node [left] {} (1)
%	    (3) edge node [left] {} (1)
%	    (10) edge node [left] {} (9)
%	    (9) edge node [left] {} (11)
%	    (9) edge node [left] {} (12);
%		\path[]
%	  	(1) edge node [left] {} (4)
%	        edge node [left] {} (5)
%	    (4) edge node [left] {} (6)
%	        edge node [left] {} (7)
%	    (5) edge node [left] {} (8)
%	    (6) edge node [left] {} (9)
%	    (7) edge node [left] {} (9)
%	    (8) edge node [left] {} (9)
%	    ;

	\end{tikzpicture}
  \caption{Example for a subgraph.}
  \label{fig:subgraph}
\end{figure}

A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine. If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):

\begin{equation} \label{eq:first}
  \models \Box (\alpha \Rightarrow \Diamond \beta)
\end{equation}

\begin{equation} \label{eq:second}
  \models \Box (\alpha \Rightarrow \Diamond (\bigvee_{\gamma} \gamma))
\end{equation}

\begin{equation} \label{eq:third}
  \models \Box (\alpha \Rightarrow [\neg \beta \mathcal{U} \bigvee_{\gamma} \gamma])
\end{equation}

Additionally, for all states not contained in the subgraph (let's call them $\delta$-states, and the proposition $\delta$ stands for ``$\delta$ is current state'') the following formula is true:

\begin{equation} \label{eq:fourth}
  \models \Box (\alpha \Rightarrow [(\bigwedge_{\delta} \neg \delta) \mathcal{U} \beta])
\end{equation}

In case of $\alpha = \beta$ - i.e. when all paths starting from $\alpha$-state eventually return again (cycle) - even another constraint can be assumed to be true:

\begin{equation} \label{eq:fifth}
  \models \Box \Diamond \alpha
\end{equation}

These five constraints apply to each subgraph matching the conditions mentioned above. Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state. Formula~\ref{eq:second} ensures that there will be always at least one $\gamma$-state be visited after each visit of an $\alpha$-state. The quite similar but stricter formula~\ref{eq:third} demands additionally a visit of at least one $\gamma$-state before the $\beta$-state can be reached. Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states. The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

All formulas can be translated to the visual formalism presented in section~\ref{sec:operatorconstraints} and shown to the user. The appropriate LTL formulas for the constraints postulated in chapter~\ref{chap:goals} relate to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:

\begin{equation}
  \models \Box \Diamond \textnormal{'Polling'}
\end{equation}

\begin{equation}
	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
\end{equation}
%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.

\section{Coalescence}
\label{sec:coalescence}

Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in section~\ref{sec:automatedconstraintgeneration} shall be assembled in one tool, the LTLCreator. It shall provide a dashboart with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results. Besides a button is needed for triggering constraint generation.
The LTLCreator should be available for other state machine based programming environments for integration.
The implementation as a java swing component and a clear API enables the tool for a wide range of use.

Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be. The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program. It will automatically revalidate all existing constraints and provide the up-to-date model for constraint generation.
For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\linewidth]{coalescence}
  \caption{Interaction between LTLCreator and programming environment.}
  \label{fig:coalescence}
\end{figure}

TODO: NuSMV hier auch wichtig?

\section{Integration into Robostudio}
\label{sec:integrationintorobostudio}

Robostudio uses NetBeans as a platform, and all tools and features are organized in windows. Thus also the LTLCreator has to be put into a window and added to the Robostudio platform. Since the defining of safety constraints is an important part while developing service robot behavior, the LTLCreator window shall appear in the center of the platform.

### 2012-10-11 thesis/tex/6_implementation.tex
Initial commit of thesis

\chapter{The LTLCreator prototype}
\section{Operator constraints}
\section{Automated constraint generation}
\section{Model checker}

NuSMV
warum den? Davor verwendete erwaehnen?

\section{Interaction with Robostudio}

allein benutzbar, oder ueber robostudio, automatische synchronisierung mit model

%Since Robostudio uses NetBeans as a platform, the LTLCreator has to be integrated as a NetBeans component. Die Datenanbindung zum Austausch des modells (Zustandsmaschine) und der Constraints erfolgt hierbei durch öffentliche

\subsection{Problems}

(auto-revalidation, model-errors, ...?)
nicht EINE zentrale stelle für modeländerungen => zer\-pflück\-te programmierung, wenn man alle änderungen haben will

\subsection{Solutions}

(pre-checker)

### 2012-10-11 thesis/tex/7_prototypeandevaluation.tex
Initial commit of thesis

\chapter{Test scenario and evaluation}
\label{chap:testscenarioandevaluation}

case study, eventuell grosses Beispiel erwaehnen? wenn constraints generiert werden, verlaesst man sich darauf? Hoert man dann auf, selbst mitzudenken? Ist die menge der constraints noch handelbar?

As stated in Sect.~\ref{sec:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language. We therefore designed and implemented a prototype of such an editor. The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated. The implementation is a java swing component, making it easy to integrate the visual language into other java applications. An API enables an underlying state machine which also influences the state propositions available for constraint editing. Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.

Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed in the dashboard in the center of the editor by drag\&drop. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{editor} 
  \caption{Snapshot of the visual editor.}
  \label{fig:editor}
\end{figure}

In order to evaluate the proposed safety functionality the tool was integrated into Robostudio~\cite{robostudio}. This Visual Programming Environment allows the specification of state machine based programs whose behaviour can then be examined by visual constraints. As an example we consider the simplified medication reminder application whose underlying screen-flow is explained in Fig.~\ref{fig:medicationreminder}. Starting from ``Menu'' the user can choose from several services such as blood pressure measurement or entertainment. In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part. In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered. The patient can state if he or she has already taken the medication, or decide whether to take it or not. In the latter he or she may give a reason for it and a caregiver is notified about this incident. Otherwise it will result in ``Well done!'' after completing intake.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{stm.pdf} %width=0.7\linewidth
  \caption{Simplified screen-flow of the medication reminder application.}
  \label{fig:medicationreminder}
\end{figure}

\section{Operator constraints}

Before we can start creating constraints it might be useful to think about a reasonable term we want to express. Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given in the requirements in section~\ref{sec:requirements}.

In order to find a corresponding graphical constraint the sentence has to be analyzed just as you read it. Since ``whenever'' is a semantic equivalent to ``always if'' first of all a \emph{ALWAYS} operator gets dragged to the dashboard directly followed by an \emph{IF} as shown in figure~\ref{fig:sampleconstraint}. The condition for this \emph{IF} is that there is a pending reminder, so a state ``Not taken yet'' has to be added to the upper bucket of the \emph{IF} operator.
Whenever the just mentioned condition becomes true, there also has to be true in future: Medication is taken properly or caregivers get notified about refuse. Accordingly a \emph{FUTURE} operator containing an \emph{OR} forms the second part of the \emph{IF} operator. Finally two states 'Well done!' and 'Staff notified' get added to the disjunction. The resulting visual constraint can now automatically be translated to the corresponding LTL formula by the editor:

\begin{equation} \label{eq:sampleconstraint}
  \models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'}))
\end{equation}

%This expression can now be translated to the visual form as we just have to break down the formula from the outside to the inside. A surrounding \emph{GLOBALLY} encapsulates an \emph{IF} block since it is the substitution for \emph{IMPLIES}. Its first operator is a proposition \emph{'Not taken yet'} whereas the second parameter has to be a \emph{FOLLOWS} block containing a disjunction of two propositions \emph{'Well done!'} and \emph{'Staff notified'}. The resulting visual constraint is shown in figure~\ref{fig:sampleconstraint}.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{sampleconstraint}
  \caption{Visual constraint creation for LTL formula ~\ref{eq:sampleconstraint}.}
  \label{fig:sampleconstraint}
\end{figure}

After each change in the dashboard, the constraint is automatically recompiled and revalidated by using the underlying model checker. For syntactically invalid constraints, an ``incomplete'' sign is displayed in the respective tab. Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign.

Due to the optimized performance of NuSMV, even the validation of constraints on huge and complex programs is fast and allows rapid feedback. In addition, the validation itself runs in the background without locking the dashboard. Therefore we avoid waiting times and disruptions during constraint development, which would be likely with conventional model checking where constraint development and constraint validation are alternating processes. Thus we consider our tool an improvement for the user experience.

\subsection{Expressiveness}

Ausdruckssärke gleich zu LTL, das heißt, dass lediglich Möglichkeiten nicht ausgedrückt werden können. Zudem gilt eine weitere Einschränkung, dass als bedingungen bisher nur Aussagen der Form ``ist in Zustand x'' verwendet werden können. Da die Validierung der healtbot anwendungen derzeit ausschließlich diesen typ braucht, sind andere Bedingungen wie Stringvergleich oder numerische Gleichheit bisher mit der visuellen Sprache nicht möglich. In Figure~\ref{fig:example_propositions} ist aufgezeigt, wie eine Erweiterung des LTLCreators um weitere Bedingungstypen aussehen könnte.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{example_propositions}
  \caption{Possible proposition extensions: numeric and string equation.}
  \label{fig:example_propositions}
\end{figure}

\section{Automated Constraint generation}

The automated constraint generation can be triggered by a click on the magic wand button in the tab area. After activation a dashboard is opened in a new tab for each found constraint, and validation is initiated immediately. For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in Fig.~\ref{fig:generatedconstraints} which match the postulations in the requirements in Sect.~\ref{sec:requirements}.
The constraint used for demonstration in the previous section is also generated, however \emph{b)} forms an intensified restriction of it.

Constraint \emph{a)} ensures the ``Polling'' state is always eventually visited again. If a pending medication is not already taken, constraint \emph{b)} guarantees intake or staff notification before the next reminders can be read from the database.

We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{c)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment or blood pressure measurement.

%** COMMENT: Possible comment here about the scalability and practicality about constraint generation; do you think it will explode in larger cases? Either way a comment might be useful. **

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{generatedconstraints} %width=0.7\linewidth
  \caption{Automatically generated constraints which match the postulations in the requirements.}
  \label{fig:generatedconstraints}
\end{figure}

%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde kann bei groï¿½en programmen (ab ca. 1000 Zustï¿½nden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern. Um den warteprozess ertrï¿½glicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht

\subsection{Performance}

 (on huge state machines) (brute force algorithm)

\subsection{Scalability}

\subsection{Reasonability of generated constraints}

\section{Reliability}

### 2012-10-11 thesis/tex/8_conclusionandfuturework.tex
Initial commit of thesis

\chapter{Conclusion \& future work}
\label{chap:conclusionandfuturework}

first step towards safety certification

In this paper we presented a visual formalism for the defining of safety constraints and first steps towards a concept of automatic constraint generation. Both features are implemented and assembled in an editor for standalone or integrated use. The visual language is reasonably simple which makes it accessible to a wider range of software developers. Nevertheless it is reasonably expressive and allows serious model checking.

The concept of applying hueristics for proposing possible constraints to users of the visual language produced very good results in our use case, the medication reminder application. All postulated constraints were found and we even discovered a new reasonable constraint. Thus we are convinced that this approach, as well as the visual language for defining constraints itself, has the potential to really support users in development of safety critical robot applications in the healthcare domain and also other domains.

Perhaps the most important next step we identified is the application of the language and the constraint generation heuristic to other domains and applications. This will reveal if the identified subgraph-based heuristic for constraint generation provides reasonable results also in different scenarios. Most likely, it will be necessary to identify further heuristics and integrate them in the visual language editor to improve user support. A study among a group of users from the healthcare robotics domain could indicate the potential of usability improvements.

Some improvements on the feature side of the language or the editor were already identified. For example, the state proposition (currently the only supported proposition type) used in constraint formulation might be not enough for other domains. Supporting propositions based on strings or mumerical equations might be necessary. Finally, the only result of constraint checking presented to the user currently is whether the constraint is valid or invalid. In contrast to conventional model checking no additional explanation for the outcome such as a counter example is given. It can be considered to extend the visual language with support for bug identification by a visualization of counter examples, for example.
%\item Substitution of parts of constraints would be nice in order to make it easier to understand. (A simple operator which abstracts from a more complex operator). It's comparable to composite states of hirachical state machines.

%\item A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.

%	\item constraints bewerten (nach gewichtung?) und nur wichtige vorschlagen? Oder zumindest sortieren

%\begin{itemize}
%	\item Yet, there are only states and no variable expressions etc.
%	\item Substitution of parts of constraints would be nice in order to make it easier to understand. (A simple operator which abstracts from a more complex operator). It's comparable to composite states of hirachical state machines.
%	\item A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.
%	\item Problem mit der Zustandsbenennung! Die Zustände müssen passende Namen haben, was aber nicht immer möglich ist. Daher müsste man eigentlich auch Transitionsbedingungen in den Constraints erlauben
%	\item constraints bewerten (nach gewichtung?) und nur wichtige vorschlagen? Oder zumindest sortieren
%	\item übertragbarkeit auf andere domänen untersuchen?
%	\item A bigger generation algorithm. As already metioned our approach only targets the healthcare domain. Especially the constraint generation can be improved to match also other domains.
%\end{itemize}

%1. Save and load functionality of constraints
%2. Yet, there are only states and no variable expressions etc.
%3. Substitution of parts of constraints would be nice in order to make it easier to understand. (A simple operator which abstracts from a more complex operator). It's comparable to composite states of hirachical state machines.
%4. A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.
%5. A bigger generation algorithm
%6. constraints bewerten und nur wichtige vorschlagen? Oder zumindest sortieren
%7. uebertragbarkeit auf andere domaenen untersuchen?

### 2012-10-15 thesis/tex/3_background.tex
initial commit v2 ;)

@@
 But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise.
-For this Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behavior on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behavior on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
+For this reason Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behavior on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behavior on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
 This file can be downloaded to the robot and executed by the interpreter.

### 2012-10-15 thesis/tex/5_design.tex
initial commit v2 ;)

@@
 
-One main difficulty in finding a proper visual formalism was to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points.
+One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points.
 A first approach, so called template constraints, and its disadvantages are described in section~\ref{sec:templateconstraints}.
@@
 Besides a button is needed for triggering constraint generation.
-The LTLCreator should be available for other state machine based programming environments for integration.
+LTLCreator should be available for other state machine based programming environments for integration.
 The implementation as a java swing component and a clear API enables the tool for a wide range of use.
@@
 Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
+LTLCreator consists of a dashboard where constraints can be edited visually and an graphical output of validation results.
+The constraint generator as well as the validator are non UI parts of LTLCreator.
 The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program.
-It will automatically revalidate all existing constraints and provide the up-to-date model for constraint generation.
+This will automatically revalidate all existing constraints and always provide the up-to-date model for constraint generation.
 For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.
@@
 
-TODO: NuSMV hier auch wichtig?
-
 \section{Integration into Robostudio}
@@
 Robostudio uses NetBeans as a platform, and all tools and features are organized in windows.
-Thus also the LTLCreator has to be put into a window and added to the Robostudio platform.
-Since the defining of safety constraints is an important part while developing service robot behavior, the LTLCreator window shall appear in the center of the platform.
+Thus also LTLCreator has to be put into a window and added to the Robostudio platform.
+Since the defining of safety constraints is an important part not only after but also during the process of developing service robot behavior, the LTLCreator window as a vital tool shall be placed in the center of the platform where currently also the UI Component Layout Editor Window is.
+As already proposed in the previous section, all changes to the state machine program made within robostudio must of course cause the LTLCreator window to revalidate all existing constraints.

### 2012-10-15 thesis/tex/6_implementation.tex
initial commit v2 ;)

@@
 \chapter{The LTLCreator prototype}
+
+As stated in chapter~\ref{chap:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language.
+We therefore designed and implemented a prototype of such an editor.
+The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated.
+The implementation is a java swing component, making it easy to integrate the visual language into other java applications.
+An API enables an underlying state machine which also influences the state propositions available for constraint editing.
+Furthermore a particular model checker can be specified for constraint validation.
+As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
+
+Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right.
+It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
+Constraints can be composed in the dashboard in the center of the editor by drag\&drop.
+The tab functionality on the left allows multiple constraints to be managed.
+Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{editor} 
+  \caption{Snapshot of the visual editor.}
+  \label{fig:editor}
+\end{figure}
+
+TODO: Kleine Inhaltsangabe
+
 \section{Operator constraints}
 \section{Automated constraint generation}
+
+% (on huge state machines) (brute force algorithm)
+
+%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde, der saemtliche sub graphs durch Abwandern sämtlicher Pfade der zugrundeliegenden state machine herauszufinden versucht, kann bei grossen programmen (ab ca.
+1000 Zustaenden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern.
+Um den warteprozess ertraelicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
+
 \section{Model checker}
@@
 (pre-checker)
+
+statische analyse

### 2012-10-15 thesis/tex/7_prototypeandevaluation.tex
initial commit v2 ;)

@@
 
-case study, eventuell grosses Beispiel erwaehnen? wenn constraints generiert werden, verlaesst man sich darauf? Hoert man dann auf, selbst mitzudenken? Ist die menge der constraints noch handelbar?
-
-As stated in Sect.~\ref{sec:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language.
-We therefore designed and implemented a prototype of such an editor.
-The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated.
-The implementation is a java swing component, making it easy to integrate the visual language into other java applications.
-An API enables an underlying state machine which also influences the state propositions available for constraint editing.
-Furthermore a particular model checker can be specified for constraint validation.
-As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
-
-Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right.
-It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
-Constraints can be composed in the dashboard in the center of the editor by drag\&drop.
-The tab functionality on the left allows multiple constraints to be managed.
-Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=\linewidth]{editor} 
-  \caption{Snapshot of the visual editor.}
-  \label{fig:editor}
-\end{figure}
-
-In order to evaluate the proposed safety functionality the tool was integrated into Robostudio~\cite{robostudio}.
-This Visual Programming Environment allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
-As an example we consider the simplified medication reminder application whose underlying screen-flow is explained in Fig.~\ref{fig:medicationreminder}.
+In order to evaluate the proposed safety functionality we use the Robostudio version where we already integrated LTLCreator in.
+This visual programming environment allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
+As a test scenario we consider the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}.
 Starting from ``Menu'' the user can choose from several services such as blood pressure measurement or entertainment.
@@
   \centering
-  \includegraphics[width=\linewidth]{stm.pdf} %width=0.7\linewidth
+  \includegraphics[width=\linewidth]{stm.pdf}
   \caption{Simplified screen-flow of the medication reminder application.}
@@
 
+This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections we will evaluate the visual language as well as the capability of constraint generation.
+In section~\ref{sec:reliability} a short statement about reliability of the features is given.
+
 \section{Operator constraints}
@@
 Before we can start creating constraints it might be useful to think about a reasonable term we want to express.
-Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given in the requirements in section~\ref{sec:requirements}.
+Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given as a requirement in chapter~\ref{chap:goals}.
 
@@
 
+We observed the different color flavours of the visual operators to be a good support for fast reading and understanding of constraints.
+Also the two dimensional composition and the round shapes of the operators turned out to
+give the language a schematic but not too rectangular look. Besides it is considered eye-candy what is the best motivation for using this visual language.
+
 \subsection{Expressiveness}
@@
   \includegraphics[scale=0.65]{example_propositions}
-  \caption{Possible proposition extensions: numeric and string equation.}
+  \caption{Possible proposition extensions: numeric and string equations.}
   \label{fig:example_propositions}
@@
 
-\section{Automated Constraint generation}
+\section{Automated constraint generation}
 
@@
 After activation a dashboard is opened in a new tab for each found constraint, and validation is initiated immediately.
-For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in Fig.~\ref{fig:generatedconstraints} which match the postulations in the requirements in Sect.~\ref{sec:requirements}.
+For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in figure~\ref{fig:generatedconstraints} which match the postulations in the requirements in chapter~\ref{chap:goals}.
 The constraint used for demonstration in the previous section is also generated, however \emph{b)} forms an intensified restriction of it. 
-
 Constraint \emph{a)} ensures the ``Polling'' state is always eventually visited again.
@@
 
+The search and generation process finishes within less than a second and thus doesn't make the developer wait a long time.
+
 We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{c)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment or blood pressure measurement.
-
-%** COMMENT: Possible comment here about the scalability and practicality about constraint generation; do you think it will explode in larger cases? Either way a comment might be useful. **
 
@@
   \centering
-  \includegraphics[scale=0.65]{generatedconstraints} %width=0.7\linewidth
-  \caption{Automatically generated constraints which match the postulations in the requirements.}
+  \includegraphics[scale=0.65]{generatedconstraints}
+  \caption{Automatically generated constraints which match the required postulations.}
   \label{fig:generatedconstraints}
@@
 
-%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde kann bei groï¿½en programmen (ab ca.
-1000 Zustï¿½nden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern.
-Um den warteprozess ertrï¿½glicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
-% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht
-
 \subsection{Performance}
 
- (on huge state machines) (brute force algorithm)
+In order to examine the performance of sub graph finding and constraint generation, we need to have a second look at the implementation details again.
+The sub graph finding algorithm is realized as a depth-first search.
+Indeed all healthcare programs are infinite and contain cycles and a depth-first walk-through would never end, but the algorithm doesn't go deeper at already visited states.
+
+The performance of this algorithm is of course highly dependend on the grade of branching.
+Nevertheless, such a depth-first search has an estimated wost case execution time of $O(|V|+|E|)$ in which $|V|$ is the number of states and $|E|$ number of edges respective transitions in our case.
+Since we cope with state machines and thus have directed edges rather than undirected ones, each state can have $|E|=|V|$ transitions; $|V-1|$ to other ones and one to itself.
+As worst case ececution time for the depth-first search we obtain the following result:
+
+\begin{equation}
+O(\textnormal{``depth-first search''})=O(|V|)+O(|V|-1)=O(|V|)
+\end{equation}
+
+In order to find a sub graph with our presented algorithm the depth-first search has to be startet on a start state of a sub graph.
+Since we don't know in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions.
+In the worst case all $|V|$ states match this condition, so the depth-first search has to be executed $|V|$ times.
+This implies for our estimated worst case execution time:
+
+\begin{equation}
+O(\textnormal{``sub graph finding''})=O(|V|)*O(|V|)=O(|V|^2)
+\end{equation}
 
@@
 
+For all the healthcare applications the presented algorithm works brilliant and speedy. But their program sizes are still quite manageable.
+Of course an exponential execution time is bad if the number of states arises.
+A clear slowdown can be felt when the algorithm runs on larger state machines with plenty of states and high grade of branching.
+
+However, sholud this algorithm regardless its tailoring to the healthcare domain be used for other huge programs, it keeps a lot of potential to be improved in efficiency.
+With a sophisticated approach it might be enough to execute just one depth-first search for finding all sub graphs at once what would lead to a linear execution time.
+% wenn noch schreiben will, dann: linear ist cool, weil das tool eh die gefundenen constraints gleich anzeigt und nicht erst am ende alles geballt bringt.
+
 \subsection{Reasonability of generated constraints}
 
+% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht
+%practicality of constraints
+
 \section{Reliability}
+\label{sec:reliability}

### 2012-10-19 src/util/dnd/DndHandler.java
visibility of method changed

@@ -53,7 +53,7 @@ public class DndHandler implements AWTEventListener {
 		return true;		
 	}
 	
-	public void startDragging(Draggable draggable, Point mousePositionRelativeToDraggable) {
+	private void startDragging(Draggable draggable, Point mousePositionRelativeToDraggable) {
 		if (draggable==null) throw new RuntimeException("Parameter can not be null.");
 		draggedOperator = draggable;
 		this.mousePositionRelativeToDraggable = mousePositionRelativeToDraggable;

### 2012-10-19 thesis/MA-Thesis.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

@@
-\documentclass[10pt,a4paper,headsepline,footsepline,DIV13,BCOR12mm]{scrbook}
+\documentclass[12pt,a4paper,headsepline,footsepline,DIV13,BCOR12mm]{scrbook}
 \usepackage[english,ngerman]{babel}
@@
 \usepackage{graphicx}
+
+\usepackage{setspace}
 
@@
 \input{titlepage}
+
+\onehalfspacing
 \newpage
+\thispagestyle{empty}
+
+\centerline{\bfseries ERKLÄRUNG}
+
+\vspace{5cm}
+Hiermit versichere ich, dass ich diese Masterarbeit selbständig verfasst habe.
+Ich habe dazu keine anderen als die angegebenen Quellen und Hilfsmittel
+verwendet.
+
+\vspace{1cm}
+\begin{flushleft}
+%select german for formatting the date
+\selectlanguage{ngerman}
+Augsburg, den \today \hfill Ludwig Nägele
+\end{flushleft}
+
+\newpage
+\thispagestyle{empty}
+\mbox{}
+
+\newpage
+
 %select english as language!

### 2012-10-19 thesis/tex/2_introduction.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

@@
 Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University of Auckland, New Zealand \cite{jayawardena12:_desig}.
-All interaction is guided by showing textual instructions on a
+All interaction is guided by showing textual instructions
 \begin{wrapfigure}{r}{0.58\textwidth}
@@
 \end{wrapfigure}
-display, and by speech generation.
+on a display, and by speech generation.
 An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
@@
 Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
-In section~\ref{sec:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine descriptions that describe robots' behaviors.
+In chapter~\ref{chap:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine definitions that describe robots' behaviors.
 
@@
 This work starts with background information about the healthcare domain, short explanations of common safety mechanisms and insights into related work in chapter~\ref{chap:background}.
-After the goals are mentioned in chapter~\ref{chap:goals}, design decisions are introduced in chapter~\ref{chap:design}.
-A test scenario is described in chapter~\ref{chap:testscenarioandevaluation}, where the presented tools get evaluated.
+After the goals are mentioned in chapter~\ref{chap:goals}, the visual formalisms and the automated constraint generation are introduced in chapters~\ref{chap:visualformalisms} and~\ref{chap:automatedconstraintgeneration}.
+After a prototype implementation is demonstrated in chapter~\ref{chap:theltlcreatorprototype}, a test scenario is described in chapter~\ref{chap:testscenarioandevaluation} where the presented tools get evaluated.
 Chapter~\ref{chap:conclusionandfuturework} gives a short conclusion and an outlook how the work presented in this paper could be continued.

### 2012-10-19 thesis/tex/5_design.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

--- 2012-10-11 thesis/tex/5_design.tex
@@
-\chapter{Design}
-\label{chap:design}
-
-%On our way towards an useful and simple concept for the defining of safety criteria
-
-%wichtig:
-%1. VL
-%2. Constraint generierung
-%3. integrierung in Robostudio
-
-%different ideas came up
-
-%In this chapter we will intruduce the final design of the visual language and compare it to a first idea of a visual formalism.
-
-In this chapter we introduce a first approach of a visual formalism and later state the final desing of the visual language.
+%\chapter{Design}
+%\label{chap:design}
+%
+%
+%In this chapter we introduce a first approach of a visual formalism and later state the final desing of the visual language.
 Furthermore the idea of automated constraint generation is presented in section~\ref{sec:automatedconstraintgeneration}.
@@
 Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.
-
-%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions.
+%
+%%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions.
 The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually.
@@
 
-\section{Visual formalisms}
-\label{sec:visualformalisms}
+\chapter{Visual formalisms}
+\label{chap:visualformalisms}
 
-One main difficulty in finding a proper visual formalism was to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points.
-A first approach, so called template constraints, and its disadvantages are described in section~\ref{sec:templateconstraints}.
+In this chapter we introduce two different approaches of visual formalisms and later state the final desing of the visual language.
+One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points.
+A first approach, so called template constraints and its pros and cons are described in section~\ref{sec:templateconstraints}.
 Subsequently an improved design is presented, which tends to a more abstract direction.
 
-\subsection{Template constraints}
+\section{Template constraints}
 \label{sec:templateconstraints}
 
+%It is not a trivial issue to find a
 What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
@@
 This visual language was presented to potential users for evaluation and feedbak.
-Soon it turned out that there are some problems and disagreements.
+Though most of them named it a good concept, soon it turned out that there are some problems and disagreements.
 The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
-A try to solve this problem was to equipp both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
+A try to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
 These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
@@
 
-\subsection{Operator constraints}
+\section{Operator constraints}
 \label{sec:operatorconstraints}
@@
 
-\section{Automated constraint generation}
-\label{sec:automatedconstraintgeneration}
+%\section{Automated constraint generation}
+%\label{sec:automatedconstraintgeneration}
+\chapter{Automated constraint generation}
+\label{chap:automatedconstraintgeneration}
 
-As already mentioned before, it might be difficult to find all reasonable safety constraints needed for full safety covery.
+As already mentioned before, it might be difficult for developers to find all reasonable safety constraints by hand which are needed for full safety covery.
 Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
@@
 But it's helpful only if the suggested constraints are relevant and reasonable.
-This however depends on the domain and use case.
-It has to be mentioned that our approach presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
-Regarding the important constraints postulated in chapter~\ref{chap:goals}, all the relevant states fall into one of the following categories:
+
+\section{Sub graph definition}
+
+Whether suggested constraints are relevant and reasonable
+%This however
+depends on the domain and use case.
+It has to be mentioned that our approach of finding an algorithm for constraint finding presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
+Regarding the important constraints postulated in chapter~\ref{chap:goals}, we found out that all all the relevant states fall into one of the following categories:
 
@@
 
+%	  \node[statein] (1) {$\alpha$};
+%	  \node[state] (2) [above left of=1] {};
+%	  \node[state] (3) [below left of=1] {};
+%	  \node[statein] (4) [above right of=1] {};
+%	  \node[statein] (5) [below right of=1] {};
+%	  \node[statein] (8) [right of=5] {$\gamma$};
+%	  \node[statein] (6) [right of=4] {$\gamma$};
+%	  \node[statein] (7) [below right of=4] {$\gamma$};
+%	  
+%	  \node[statein] (9) [right of=7] {$\beta$};
+%	  \node[state] (10) [above of=9] {};
+%	  \node[state] (11) [above right of=9] {};
+%	  \node[state] (12) [below right of=9] {};
+%	  
+%	  \path[dashed]
+%	    (2) edge node [left] {} (1)
+%	    (3) edge node [left] {} (1)
+%	    (10) edge node [left] {} (9)
+%	    (9) edge node [left] {} (11)
+%	    (9) edge node [left] {} (12);
+%		\path[]
+%	  	(1) edge node [left] {} (4)
+%	        edge node [left] {} (5)
+%	    (4) edge node [left] {} (6)
+%	        edge node [left] {} (7)
+%	    (5) edge node [left] {} (8)
+%	    (6) edge node [left] {} (9)
+%	    (7) edge node [left] {} (9)
+%	    (8) edge node [left] {} (9)
+%	    ;
+
 	  \node[statein] (1) {$\alpha$};
 	  \node[state] (2) [above left of=1] {};
-	  \node[state] (3) [below left of=1] {};
-	  \node[statein] (4) [above right of=1] {};
+	  \node[state] (3) [above right of=1] {};
+	  \node[statein] (4) [below left of=1] {};
+	  \node[statein] (6) [below of=4] {$\gamma$};
+	  \node[statein] (7) [below right of=4] {$\gamma$};
 	  \node[statein] (5) [below right of=1] {};
-	  \node[statein] (8) [right of=5] {$\gamma$};
-	  \node[statein] (6) [right of=4] {$\gamma$};
-	  \node[statein] (7) [below right of=4] {$\gamma$};
-
-	  \node[statein] (9) [right of=7] {$\beta$};
-	  \node[state] (10) [above of=9] {};
-	  \node[state] (11) [above right of=9] {};
+	  \node[statein] (8) [below of=5] {$\gamma$};
+	  \node[statein] (9) [below of=7] {$\beta$};
+	  \node[state] (10) [right of=9] {};
+	  \node[state] (11) [below left of=9] {};
 	  \node[state] (12) [below right of=9] {};
@@
 
-%	  \node[statein] (1) {$\alpha$};
-%	  \node[state] (2) [above left of=1] {};
-%	  \node[state] (3) [above right of=1] {};
-%	  \node[statein] (4) [below left of=1] {};
-%	  \node[statein] (6) [below of=4] {$\gamma$};
-%	  \node[statein] (7) [below right of=4] {$\gamma$};
-%	  \node[statein] (5) [below right of=1] {};
-%	  \node[statein] (8) [below of=5] {$\gamma$};
-%	  \node[statein] (9) [below of=7] {$\beta$};
-%	  \node[state] (10) [right of=9] {};
-%	  \node[state] (11) [below left of=9] {};
-%	  \node[state] (12) [below right of=9] {};
-%	  
-%	  \path[dashed]
-%	    (2) edge node [left] {} (1)
-%	    (3) edge node [left] {} (1)
-%	    (10) edge node [left] {} (9)
-%	    (9) edge node [left] {} (11)
-%	    (9) edge node [left] {} (12);
-%		\path[]
-%	  	(1) edge node [left] {} (4)
-%	        edge node [left] {} (5)
-%	    (4) edge node [left] {} (6)
-%	        edge node [left] {} (7)
-%	    (5) edge node [left] {} (8)
-%	    (6) edge node [left] {} (9)
-%	    (7) edge node [left] {} (9)
-%	    (8) edge node [left] {} (9)
-%	    ;
-
 	\end{tikzpicture}
@@
 
-A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}.
-Using this specification all possible such subgraphs can be detected within a state machine.
+A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine.
+
+\section{Constraint determination}
+
 If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):
@@
 %	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
-
-\section{Coalescence}
-\label{sec:coalescence}
-
-Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in section~\ref{sec:automatedconstraintgeneration} shall be assembled in one tool, the LTLCreator.
-It shall provide a dashboart with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results.
-Besides a button is needed for triggering constraint generation.
-The LTLCreator should be available for other state machine based programming environments for integration.
-The implementation as a java swing component and a clear API enables the tool for a wide range of use.
-
-Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
-The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program.
-It will automatically revalidate all existing constraints and provide the up-to-date model for constraint generation.
-For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.7\linewidth]{coalescence}
-  \caption{Interaction between LTLCreator and programming environment.}
-  \label{fig:coalescence}
-\end{figure}
-
-TODO: NuSMV hier auch wichtig?
-
-\section{Integration into Robostudio}
-\label{sec:integrationintorobostudio}
-
-Robostudio uses NetBeans as a platform, and all tools and features are organized in windows.
-Thus also the LTLCreator has to be put into a window and added to the Robostudio platform.
-Since the defining of safety constraints is an important part while developing service robot behavior, the LTLCreator window shall appear in the center of the platform.

### 2012-10-19 thesis/tex/6_implementation.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

@@
 \chapter{The LTLCreator prototype}
+\label{chap:theltlcreatorprototype}
 
 As stated in chapter~\ref{chap:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language.
-We therefore designed and implemented a prototype of such an editor.
-The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated.
-The implementation is a java swing component, making it easy to integrate the visual language into other java applications.
-An API enables an underlying state machine which also influences the state propositions available for constraint editing.
-Furthermore a particular model checker can be specified for constraint validation.
-As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
+We therefore designed and implemented a prototype of such an editor which contains the visual language as well as the proposed constraint generation heuristics.
+The implementation is a java swing component and has a simple API definition, making it easy to integrate the LTLCreator tool into other java applications.
 
-Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right.
+First of all some details about the implementation of the visual language are given in section~\ref{sec:prototype:operatorconstraints}, directly followed by an explanation about the realization of the automated constraint generation functionality in section~\ref{sec:prototype:automatedconstraintgeneration}.
+The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}.
+However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots and to be a prototype for evaluation.
+Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.
+
+\section{Operator constraints}
+\label{sec:prototype:operatorconstraints}
+
+Each of the different operator types, i.e.
+IfThenOperator, AndOperator, OrOperator, NotOperator, NextOperator, FutureOperator, AlwaysOperator, UntilOperator and StateOperator, has got its own representing class.
+All of them inherit from one abstract class AbstractOperator which provides the basic functionality beeing used by all operators.
+The AbstractOperator class is shown in figure~\ref{fig:abstractoperator} and its methods are described as follows:
+
+\begin{description}
+	\item[getLTL():] This function returns the LTL formula of an operator and all its children recursively.
+Applied to the most outer operator it returns the complete formula for the constraint.
+This function is abstract and implemented by each operator type separately.
+	\item[getColor():] Each operator type has got is own significant color which can be retrieved over the abstract function getColor().
+	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object dependet on the implementation.
+	\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
+	\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function paintOperator().
+It manages all color and shape painting for the respective operator.
+Additional parameters such as recursive painting including children can be spezified.
+	\item[contains(int, int):] Given two coordinates x and y, it can be determined wheter the herewith defined point lays within an operator.
+	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality.
+That means same operators with same children.
+This functionality is important for avoiding duplicate constraints during constraint generation.
+	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it.
+The hierarchical architecture of AbstractOperators propagates all change events to the root operator.
+The function addChangeListener() allows registering for such change notifications.
+	\item[removeChangeListener(OperatorChangeListener):] According to addChangeListener this function allows deregistration again.
+\end{description}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.6\linewidth]{AbstractOperator} 
+  \caption{AbstractOperator class definition.}
+  \label{fig:abstractoperator}
+\end{figure}
+
+Whereas StateOperators have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with respective one or two possible children.
+The graphical representation of both unary and binary operators thus needs to have place holders and docking stations for sub operators.
+These are realized by buckets which display a small ``drop here'' label and allow operator adding or removing by mouse drops.
+For this a class Bucket is defined with get and set methods for defining a sub operator.
+Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between AbstractOperator, Bucket and operator implementations.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{klassendiagramm_operators} 
+  \caption{Class diagram of operator structure.}
+  \label{fig:klassendiagramm_operators}
+\end{figure}
+
+Once operators are composed in a way that no empty buckets are left over, we retrieve a complete constraint such as the one shown in figure~\ref{fig:operator_tree}.
+This diagram illustrates the correlation of nested operators and buckets.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{operator_tree} 
+  \caption{Instantiation of a constraint with several operators.}
+  \label{fig:operator_tree}
+\end{figure}
+
+All graphical components such as buckets and operators are implemented as java swing JComponent.
+Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.
+
+%PAINT
+
+For lightweight components swing allows the implementation of a paint() method which is responsible for rendering the component.
+This is used for giving every operator its special look.
+We hereby made use of swing's Graphics2D functionality what gives a great support in painting rounded shapes and color gradients.
+An operator is displayed as a rounded rectangle with dents around sub operators and has got a flowing backgound, borders and labels in an operator type specific color.
+For the significant shape multiple rounded rectangles with and without border are arranged in a way that the result appears to be one peace.
+Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the if operator first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be.
+In the third step the same rectangle as from step one just this time without border is painted on top.
+This makes all crossing borders disappear.
+Only labels and buckets have to be added yet and a complete IfThenOperator emerges.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[scale=0.65]{paintprocess} 
+  \caption{Steps of painting an operator.}
+  \label{fig:paintprocess}
+\end{figure}
+
+%Um die paint-Logik muss sich nicht gekümmert werden, lediglich die paint mehtode muss implementiert werden, die dem jeweiligen Operator sein spezielles Aussehen verleiht.
+Dabei wurde auf die Funktionen des Graphics2D zurückgegriffen, was die Erstellung von runden Formen (shapes) und Farbverläufen (stroke?) ermöglicht.
+Jeder Operatortyp hat eine spezifische Farbe, die für den Farbverlauf im Hintergrund, für den Border und als Textfarbe verwendet wird.
+Die shapes der Operatoren wurden so umgesetzt, dass sie sich um die enthaltenen Operatoren schlingen (schmiegen).
+Hierfür wurden auschließlich leere und gefüllte abgerundete Rechtecke gezeichnet, die in der richtigen Reihenfolge übereinandergelegt die gewünschte Endform für den Operator ergibt.
+grafisches beispiel?
+
+%LAYOUT
+
+Swing also provides a mechanism for layouting components, that means depending on child components the aspired sizes and positions of components can be retrieved.
+As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraints} constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for varation in time.
+This fact forms a special requirement for the layouting of operators.
+Whereas the operator's positioning and resizing along the logical direction is only dependend on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult.
+Here sub operators can't just be aligned centered as it is the case vertically.
+To face the determination of horizontal positioning, every operator gets time reference lines as depicted in figure~\ref{fig:referencelines}.
+Logical operators, i.e.
+IfThenOperator, AndOperator, OrOperator, NotOperator and StateOperator, have one reference line, time relevant operators, i.e.
+NextOperator, FutureOperator, AlwaysOperator and UntilOperator, have two such reference lines.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[scale=0.65]{referencelines} 
+  \caption{Time reference lines and their adjustment.}
+  \label{fig:referencelines}
+\end{figure}
+
+%Über den eingebauten Layoutmanager-Mechanismus kann die Anordnung der Operatoren zueinander festgelegt werden.
+Wie aus section xy bekannt, gibt es eine logische wie auch eine zeitliche Achse bei constraints.
+Demnach muss die verticale wie auch horizontale positionierung der operatoren bestimmt werden.
+jeder operator besitzt eine logische Mittellinie, die wird zur mittellinie des übergeordneten Buckets ausgerichtet.
+
+%DRAG&DROP
+
+The inbuilt mouse support of swing was used for identifying operators laying under the current mouse position.
+On the one hand this is important for mouse sensitive highlighting of particular constraint parts, on the other hand this way it can be determined where operators have to be added after drag and drop.
+So far the predefined swing drag and drop has not been used but a simple drag and drop handler has been implemented instead.
+Once a JComponent is registered to the class DndHandler, it automatically manages any drag and drop actions with JComponents which implement either the Draggable or DropTarget interface.
+The former is implemented by all operators, the latter by buckets and the dashboard.
+%Figure~\ref{fig:dndclassstructure} gives a short class overview.
+Furthermore there is a class DraggableProvider which displays a kind of button in the tool bar and lets the user create new Draggables; in our case new operators get instantiated.
+
+%das vorgefertigte Mouse-Handling wurde verwendet, um die jeweilige operator oder bucket componente an der aktuellen mausposition zu bestimmen.
+Die ist zum einen wichtig bei darüberfahrender maus um Teile des constraints hervorzuheben, zum anderen wird damit beim draggen auch bestimmt, welcher operator verschoben werden soll oder an welche stelle einer abgelegt werden soll.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{dndclassstructure} 
+  \caption{Class definition of drag and drop functionality.}
+  \label{fig:dndclassstructure}
+\end{figure}
+
+%operator werden im dashboard, einem weiteren container, zusammengeklickt und angezeigt.
+Das Dashboard hat dabei eine ähnliche Funktionsweise wie ein Bucket.
+die operatoren können dabei vom operatorprovider bezogen werden: Implementierungsdetails?
+trash...
+
+\section{Automated constraint generation}
+\label{sec:prototype:automatedconstraintgeneration}
+
+% (on huge state machines) (brute force algorithm)
+
+%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde, der saemtliche sub graphs durch Abwandern sämtlicher Pfade der zugrundeliegenden state machine herauszufinden versucht, kann bei grossen programmen (ab ca.
+1000 Zustaenden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern.
+Um den warteprozess ertraelicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
+
+\section{Coalescence}
+\label{sec:coalescence}
+
+Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in chapter~\ref{chap:automatedconstraintgeneration} have to be assembled in one tool, the LTLCreator.
+It shall provide a dashboard with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results.
+Besides a button is needed for triggering constraint generation.
+
+TODO: beschreiben dass vl im tab eingebunden wurde, die constraintgenerierung unter den zauberstab.
+
+Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right.
 It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
-Constraints can be composed in the dashboard in the center of the editor by drag\&drop.
+Constraints can be composed in the dashboard in the center of the editor by drag\&drop which holds and displays the operator constraints.
 The tab functionality on the left allows multiple constraints to be managed.
-Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.
+Each tab shows a small thumbnail of the constraint and a symbol indicating its validity; a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.''.
+The magic wand button within the tab pane brings the constraint generation functionality into the editor and triggers the constraint finding process.
 
@@
 
-TODO: Kleine Inhaltsangabe
-
-\section{Operator constraints}
-\section{Automated constraint generation}
-
-% (on huge state machines) (brute force algorithm)
-
-%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde, der saemtliche sub graphs durch Abwandern sämtlicher Pfade der zugrundeliegenden state machine herauszufinden versucht, kann bei grossen programmen (ab ca.
-1000 Zustaenden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern.
-Um den warteprozess ertraelicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
-
-\section{Model checker}
+\subsection{Model checker}
 
@@
 
-\section{Interaction with Robostudio}
+a CTL model checker has been written in java, but...
+
+Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
+
+wie genau wird das getriggert (rekursiver listener an operatoren)
+
+ConstraintValidator
+ValidatorThread
+
+(im hintergrund, mit warteschlange. dabei sollen aber wenn ein contsraint frequently geändert wird, alte validationsaufträge des gleichen constraints aus der warteschlange entfernt werden.
+Callback mit result.
+
+%\subsection{Integratability}
+\subsection{Ease of integration}
+
+LTLCreator should be available for other state machine based programming environments for integration what is ensured by its implementation as a java swing component and a clear API enabling the tool for a wide range of use.
+
+Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
+LTLCreator consists of a dashboard where constraints can be edited visually and an graphical output of validation results.
+The constraint generator as well as the validator are non UI parts of LTLCreator.
+The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program.
+This will automatically revalidate all existing constraints and always provide the up-to-date model for constraint generation.
+For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.7\linewidth]{coalescence}
+  \caption{Interaction between LTLCreator and programming environment.}
+  \label{fig:coalescence}
+\end{figure}
+
+\section{Integration into Robostudio}
+\label{sec:integrationintorobostudio}
+
+kurz erläutern, warum das gemaacht werden soll
+
+Robostudio uses NetBeans as a platform, and all tools and features are organized in windows.
+Thus also LTLCreator has to be put into a window and added to the Robostudio platform.
+Since the defining of safety constraints is an important part not only after but also during the process of developing service robot behavior, the LTLCreator window as a vital tool shall be placed in the center of the platform where currently also the UI Component Layout Editor Window is.
+As already proposed in the previous section, all changes to the state machine program made within robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
 

### 2012-10-19 thesis/tex/7_prototypeandevaluation.tex
* increased font size, from 10 to 12pt * increased linespacing, now: 1,5 * Chapter "Design" is now splittet into "Visual formalisms" and   "Automated constraint generation" * Section "6.1 Operator constraints" now filled (with lots of pictures) * smaller changes due to new font size and linespacing

@@
 This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections we will evaluate the visual language as well as the capability of constraint generation.
-In section~\ref{sec:reliability} a short statement about reliability of the features is given.
+In section~\ref{sec:practicalexperiences} it is described how a non-expert deals with the visual language and what the final judgement is.
 
@@
 
-The search and generation process finishes within less than a second and thus doesn't make the developer wait a long time.
+The search and generation process finishes within less than a second and thus doesn't make the developer wait for a long time.
 
@@
 Nevertheless, such a depth-first search has an estimated wost case execution time of $O(|V|+|E|)$ in which $|V|$ is the number of states and $|E|$ number of edges respective transitions in our case.
-Since we cope with state machines and thus have directed edges rather than undirected ones, each state can have $|E|=|V|$ transitions; $|V-1|$ to other ones and one to itself.
+Since we cope with state machines and thus have directed edges rather than undirected ones, each state can have $|V|$ transitions; $|V-1|$ to other ones and one to itself.
+For the contained tree without cycles for depth-first search we retreive $|E|=|V|$ edges.
 As worst case ececution time for the depth-first search we obtain the following result:
@@
 
-For all the healthcare applications the presented algorithm works brilliant and speedy. But their program sizes are still quite manageable.
+For all the healthcare applications the presented algorithm works efficient and fast.
+But their program sizes are still quite manageable.
+Our test program has eleven states, an average branch grade and a constraint finding execution time of a split second on a standard pc.
 Of course an exponential execution time is bad if the number of states arises.
 A clear slowdown can be felt when the algorithm runs on larger state machines with plenty of states and high grade of branching.
+A test on a state machine with over two hundred and fifty states and partially high branch grade resulted in approximately five minutes execution time.
 
@@
 
-% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht
-%practicality of constraints
+The constraint generator algorithm found six constraints for the medication reminder application.
+Two of them were actually the ones we proposed in the requirements.
+Three of them are either intensified versions of the previous ones or just describe the behaviour of the program and thus make sense, but in matters of safety they aren't fundamental relevant.
+Nevertheless the generator tool actually helped us in finding a new important constraint we didn't think about before.
 
-\section{Reliability}
-\label{sec:reliability}
+To put it in a nutshell, all automatically generated constraints are practical, and approximately half of them are significant.
+We showed that the constraint generator concept works well for the healthcare application and can be a very helpful feature.
+
+\section{Practical experiences}
+\label{sec:practicalexperiences}
+
+eher schreiben: wurde einigen Leuten vorgestellt. Generell blabla... Darunter auch alina kracker...
+
+Alina Kracker, 21 Jahre alt aus Friedberg und Studentin der Erziehungswissenschaften, arbeitet neben dem Studium als Pflegehilfskraft in einer Kurzzteitplfege.
+Dort betreut sie ältere Damen und Herren, die dort kurz- bis mittelfristig untergebracht sind.
+In dieser Einrichtung wohnen Personen mit unterschiedlichsten Krankheiten und Bedürfnissen, von zu beobachtenden aber gesunden Menschen bis hin zu Alzeimerpatienten.
+
+Es wurde getestet, ob Kracker ohne entsprechende Ausbildung Constraints zu dem Beispiel des medication reminder formulieren kann.
+Zum einen möchte die Universität of Auckland die Programmierung solcher Anwendungen auch nicht-professionellen ermöglichen, und da wäre die Erfahrung einer echten Pflegekraft mit dieser Aufgabe wohl sehr interessant.
+Zum anderen kann so aber auch festgestellt werden, wie nicht-Softwareentwickler mit der visuellen Sprache und dem Konzept zur Constrainterstellung zurechtkommen.
+
+Kracker wurde in das Beispiel der medication reminder applikation eingeführt und das Programmverhalten anhand der Zustandsmaschine aus figure 123 erklärt.
+Anschließend wurde sie mit der Oberfläche und Funktionsweise des LTLCreator tools vertraut gemacht.
+
+Ihr wurde nun die Aufgabe gegeben, mit den im Editor zur Verfügung stehenden Mitteln die zwei Bedingungen zu formulieren, die uns in diesem Kapitel bereits begleitet haben.
+Zum einen sollen Medikamenterinnerungen nie vergessen werden, also Zustand ``Polling'' muss regelmäßig besucht werden.
+Zum anderen muss sichergestellt sein, dass bei noch nicht vorgenommener Medikamenteinnahme das Medikament letztendlich eingenommen wird oder das Personal benachrichtigt wird.
+Also bei Besuch von ``Not taken yet'' muss entweder ``Well done!'' oder ``Staff notified'' besucht werden.
+
+Da Kracker mit Zustandsmaschinendarstellungen nicht vertraut ist, wurden ihr die Constraintbedeutungen wie eben beschrieben anhand der Zustandsnamen erläutert.
+Bei der Erstellung der Constraints war sie allerdings anfänglich auf sich alleine gestellt.
+Die erste Bedingung erstellte sie zügig und richtig, und es schien auch keine besondere Herausforderung gewesen zu sein.
+Auch der zweite Constraint wurde
+Bei der Frage nach der Anwendung von FUTURE hakte es dann allerdings etwas, was auf die zeitlichen (...) zurückzuführen war.
+Die Aufforderung, die Bedingung als deutschen Satz zu formulieren, half ihr wieder auf die Sprünge.
+
+Für eine Person, die von modelchecking oder LTL noch nie etwas gehört hat, kam sie mit kleinen Hilfen gut mit der visuellen Sprache zurecht und könnte sie so mit etwas Übung auch verwenden.
+
+Die Nähe zwischen LTL und deutschem Satz ist gut, und dadurch kann man schnell konstraints zusammen klicken (besser als mit fremden mathematischen symbolen).
+
+Auf die Frage, ob sie solche Roboteranwendungen wie der Medicationreminder in der Kurzzeitpflege für sinnvoll erachten würde, antwortet Kracker: ``Für geistig fitte aber dennoch vergessliche und tütelige Leute kann ich mir diese Roboterapplikation super vorstellen.
+Ein Großteil unserer Patienten ist jedoch nicht in der Lage Entscheidungen, die der Roboter abfrägt, wahrheitsgemäß zu treffen.
+Aufgrund der regelmäßig wechselnden Patientenbelegung in der Kurzzeitpflege ist es vielmehr für das Betreuungspersonal eine große Herausforderung, die uhrzeitgebundenen Medikamenteinnahmen der Patienten nicht zu vergessen.
+Daher wäre eine solche Erinnerungsfunktion für das Personal eine große Hilfe.''.

### 2012-10-20 thesis/tex/6_implementation.tex
Section 6.2 (Automated constraint generation) and section 6.3.1 (Model checker) are now finished.

@@
 Each of the different operator types, i.e.
-IfThenOperator, AndOperator, OrOperator, NotOperator, NextOperator, FutureOperator, AlwaysOperator, UntilOperator and StateOperator, has got its own representing class.
-All of them inherit from one abstract class AbstractOperator which provides the basic functionality beeing used by all operators.
-The AbstractOperator class is shown in figure~\ref{fig:abstractoperator} and its methods are described as follows:
+\emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator}, \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator}, \emph{UntilOperator} and \emph{StateOperator}, has got its own representing class.
+All of them inherit from one abstract class \emph{AbstractOperator} which provides the basic functionality beeing used by all operators.
+The \emph{AbstractOperator} class is shown in figure~\ref{fig:abstractoperator} and its methods are described as follows:
 
@@
 This function is abstract and implemented by each operator type separately.
-	\item[getColor():] Each operator type has got is own significant color which can be retrieved over the abstract function getColor().
+	\item[getColor():] Each operator type has got is own significant color which can be retrieved over the abstract function \emph{getColor()}.
 	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object dependet on the implementation.
 	\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
-	\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function paintOperator().
+	\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function \emph{paintOperator()}.
 It manages all color and shape painting for the respective operator.
-Additional parameters such as recursive painting including children can be spezified.
+Additional parameters such as recursive painting including children can be specified.
 	\item[contains(int, int):] Given two coordinates x and y, it can be determined wheter the herewith defined point lays within an operator.
 	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality.
-That means same operators with same children.
+That means same operators with similiar children.
 This functionality is important for avoiding duplicate constraints during constraint generation.
 	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it.
-The hierarchical architecture of AbstractOperators propagates all change events to the root operator.
-The function addChangeListener() allows registering for such change notifications.
-	\item[removeChangeListener(OperatorChangeListener):] According to addChangeListener this function allows deregistration again.
+The hierarchical architecture of \emph{AbstractOperator}s propagates all change events to the root operator.
+The function \emph{addChangeListener()} allows registering for such change notifications.
+	\item[removeChangeListener(OperatorChangeListener):] According to \emph{addChangeListener} this function allows deregistration again.
 \end{description}
@@
 
-Whereas StateOperators have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with respective one or two possible children.
+Whereas \emph{StateOperator}s have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with respective one or two possible children.
 The graphical representation of both unary and binary operators thus needs to have place holders and docking stations for sub operators.
@@
 For this a class Bucket is defined with get and set methods for defining a sub operator.
-Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between AbstractOperator, Bucket and operator implementations.
-
+Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between \emph{AbstractOperator}, \emph{Bucket} and operator implementations.
 \begin{figure}[htbp]
@@
 \end{figure}
-
 Once operators are composed in a way that no empty buckets are left over, we retrieve a complete constraint such as the one shown in figure~\ref{fig:operator_tree}.
@@
 
-All graphical components such as buckets and operators are implemented as java swing JComponent.
+All graphical components such as buckets and operators are implemented as java swing \emph{JComponent}s.
 Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.
@@
 
-For lightweight components swing allows the implementation of a paint() method which is responsible for rendering the component.
+For lightweight components swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component.
 This is used for giving every operator its special look.
-We hereby made use of swing's Graphics2D functionality what gives a great support in painting rounded shapes and color gradients.
+We hereby made use of swing's \emph{Graphics2D} functionality what gives a great support in painting rounded shapes and color gradients.
 An operator is displayed as a rounded rectangle with dents around sub operators and has got a flowing backgound, borders and labels in an operator type specific color.
@@
 This makes all crossing borders disappear.
-Only labels and buckets have to be added yet and a complete IfThenOperator emerges.
+Only labels and buckets have to be added yet and a complete \emph{IfThenOperator} emerges.
 
@@
   \includegraphics[scale=0.65]{paintprocess} 
-  \caption{Steps of painting an operator.}
+  \caption{Four steps of painting an operator.}
   \label{fig:paintprocess}
 \end{figure}
-
-%Um die paint-Logik muss sich nicht gekümmert werden, lediglich die paint mehtode muss implementiert werden, die dem jeweiligen Operator sein spezielles Aussehen verleiht.
-Dabei wurde auf die Funktionen des Graphics2D zurückgegriffen, was die Erstellung von runden Formen (shapes) und Farbverläufen (stroke?) ermöglicht.
-Jeder Operatortyp hat eine spezifische Farbe, die für den Farbverlauf im Hintergrund, für den Border und als Textfarbe verwendet wird.
-Die shapes der Operatoren wurden so umgesetzt, dass sie sich um die enthaltenen Operatoren schlingen (schmiegen).
-Hierfür wurden auschließlich leere und gefüllte abgerundete Rechtecke gezeichnet, die in der richtigen Reihenfolge übereinandergelegt die gewünschte Endform für den Operator ergibt.
-grafisches beispiel?
 
@@
 Logical operators, i.e.
-IfThenOperator, AndOperator, OrOperator, NotOperator and StateOperator, have one reference line, time relevant operators, i.e.
-NextOperator, FutureOperator, AlwaysOperator and UntilOperator, have two such reference lines.
+\emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator} and \emph{StateOperator}, have one reference line, time relevant operators, i.e.
+\emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator} and \emph{UntilOperator}, have two such reference lines.
 
@@
   \includegraphics[scale=0.65]{referencelines} 
-  \caption{Time reference lines and their adjustment.}
+  \caption{Time reference lines and their adjustments.}
   \label{fig:referencelines}
 \end{figure}
-
-%Über den eingebauten Layoutmanager-Mechanismus kann die Anordnung der Operatoren zueinander festgelegt werden.
-Wie aus section xy bekannt, gibt es eine logische wie auch eine zeitliche Achse bei constraints.
-Demnach muss die verticale wie auch horizontale positionierung der operatoren bestimmt werden.
-jeder operator besitzt eine logische Mittellinie, die wird zur mittellinie des übergeordneten Buckets ausgerichtet.
 
@@
 On the one hand this is important for mouse sensitive highlighting of particular constraint parts, on the other hand this way it can be determined where operators have to be added after drag and drop.
-So far the predefined swing drag and drop has not been used but a simple drag and drop handler has been implemented instead.
-Once a JComponent is registered to the class DndHandler, it automatically manages any drag and drop actions with JComponents which implement either the Draggable or DropTarget interface.
+So far
+% the predefined swing drag and drop has not been used but
+a simple drag and drop handler has been implemented.
+% instead. 
+Once a \emph{JComponent} is registered to the \emph{DndHandler}, it automatically manages any drag and drop actions with \emph{JComponent}s which implement either the \emph{Draggable} or \emph{DropTarget} interface.
 The former is implemented by all operators, the latter by buckets and the dashboard.
 %Figure~\ref{fig:dndclassstructure} gives a short class overview.
-Furthermore there is a class DraggableProvider which displays a kind of button in the tool bar and lets the user create new Draggables; in our case new operators get instantiated.
-
-%das vorgefertigte Mouse-Handling wurde verwendet, um die jeweilige operator oder bucket componente an der aktuellen mausposition zu bestimmen.
-Die ist zum einen wichtig bei darüberfahrender maus um Teile des constraints hervorzuheben, zum anderen wird damit beim draggen auch bestimmt, welcher operator verschoben werden soll oder an welche stelle einer abgelegt werden soll.
+Furthermore there is a class \emph{DraggableProvider} which displays a kind of button in the tool bar and lets the user create new \emph{Draggable}s; in our case new operators get instantiated.
 
@@
   \includegraphics[width=\linewidth]{dndclassstructure} 
-  \caption{Class definition of drag and drop functionality.}
+  \caption{Class definition for drag and drop functionality.}
   \label{fig:dndclassstructure}
@@
 
-% (on huge state machines) (brute force algorithm)
+In section~\ref{chap:automatedconstraintgeneration} we stated that automated constraint generation is based on subgraphs.
+For the finding of such subgraphs an algorithm has been developed and the main idea how it works is described as follows.
 
-%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde, der saemtliche sub graphs durch Abwandern sämtlicher Pfade der zugrundeliegenden state machine herauszufinden versucht, kann bei grossen programmen (ab ca.
-1000 Zustaenden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern.
-Um den warteprozess ertraelicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
+We define one start state and generate all possible paths starting from it leading through the state machine with just one restriction that a path ends before it visits states twice.
+Thus we retrieve a finite amount of finite paths since loops are eliminated.
+Now every state gets a number assigned which represents the total number of occurernces of the respective state in all collected paths.
+In other words, the label of a state expresses the number of pahts which start in the start state and lead through the respective state.
+The start state itself has got the number of all paths, of course.
+If during walking through any of the paths a second state can be found with the same label, it means that all paths starting at start state come all together in this state again.
+It is uncovered as potential corresponding end state.
+A last check has to be applied whether there are incoming transitions from outside into the subgraph which don't come over start state. If no such transitions are detected, a subgraph is found.
+If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with start state insofar as the state machine doesn't contain further loops or terminating end states.
+Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrence of each state.
+This procedure is suitable for bringing to light whether a state is the beginning of a subgraph and for discovering the corresponding end state.
+In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state.
+
+\begin{figure}[htbp]
+  \centering
+
+  \tikzstyle{state} = [circle,fill=white,draw]
+  \tikzstyle{statein} = [circle,fill=blue!20,draw]
+  \tikzstyle{none} = [fill=white]
+  \tikzstyle{highlighted} = [fill=yellow]
+  \tikzstyle{every path} = [font=\sffamily\small]
+
+	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=15mm]
+
+	  \node[state] (1) at (0,0) {A};
+	  \node[state] (4) [below left of=1] {B};
+	  \node[state] (6) [below of=4] {D};
+	  \node[state] (7) [below right of=4] {E};
+	  \node[state] (5) [below right of=1] {C};
+	  \node[state] (8) [below of=5] {F};
+	  \node[state] (9) [below of=7] {G};
+	  \node[state] (11) [below left of=9] {H};
+	  \node[state] (12) [right of=9] {I};
+
+	  \node[state] (101) at (3,0) {A};
+	  \node[state] (102) [below of=101] {B};
+	  \node[state] (103) [below of=102] {D};
+	  \node[state] (104) [below of=103] {G};
+	  \node[state] (105) [below of=104] {H};
+
+	  \node[state] (201) at (4.5,0) {A};
+	  \node[state] (202) [below of=201] {B};
+	  \node[state] (203) [below of=202] {D};
+	  \node[state] (204) [below of=203] {G};
+	  \node[state] (205) [below of=204] {I};
+
+	  \node[state] (301) at (6,0) {A};
+	  \node[state] (302) [below of=301] {B};
+	  \node[state] (303) [below of=302] {E};
+	  \node[state] (304) [below of=303] {G};
+	  \node[state] (305) [below of=304] {H};
+
+	  \node[state] (401) at (7.5,0) {A};
+	  \node[state] (402) [below of=401] {B};
+	  \node[state] (403) [below of=402] {E};
+	  \node[state] (404) [below of=403] {G};
+	  \node[state] (405) [below of=404] {I};
+
+	  \node[state] (501) at (9,0) {A};
+	  \node[state] (502) [below of=501] {C};
+	  \node[state] (503) [below of=502] {F};
+	  \node[state] (504) [below of=503] {G};
+	  \node[state] (505) [below of=504] {H};
+
+	  \node[state] (601) at (10.5,0) {A};
+	  \node[state] (602) [below of=601] {C};
+	  \node[state] (603) [below of=602] {F};
+	  \node[state] (604) [below of=603] {G};
+	  \node[state] (605) [below of=604] {I};
+
+	  \node[highlighted] (701) at (12,0) {A=6};
+	  \node[none] (702) at (12,-0.6) {B=4};
+	  \node[none] (703) at (12,-1.2) {C=2};
+	  \node[none] (703) at (12,-1.8) {D=2};
+	  \node[none] (703) at (12,-2.4) {E=2};
+	  \node[none] (703) at (12,-3.0) {F=2};
+	  \node[highlighted] (703) at (12,-3.6) {G=6};
+	  \node[none] (703) at (12,-4.2) {H=3};
+	  \node[none] (703) at (12,-4.8) {I=3};
+
+		%\draw[->] (0, 0) .. controls(1,1) .. (3, 0);
+		\draw[->] (12) to [out=45, in=0] (1);
+
+	  \path[]
+	  	(1) edge node [left] {} (4)
+	        edge node [left] {} (5)
+	    (4) edge node [left] {} (6)
+	        edge node [left] {} (7)
+	    (5) edge node [left] {} (8)
+	    (6) edge node [left] {} (9)
+	    (7) edge node [left] {} (9)
+	    (8) edge node [left] {} (9)
+	    (11) edge [bend left] node [left] {} (9)
+	    (9) edge [bend left] node [left] {} (11)
+	    (9) edge node [left] {} (12)
+
+		  (101) edge node [left] {} (102)
+		  (102) edge node [left] {} (103)
+		  (103) edge node [left] {} (104)
+		  (104) edge node [left] {} (105)
+
+		  (201) edge node [left] {} (202)
+		  (202) edge node [left] {} (203)
+		  (203) edge node [left] {} (204)
+		  (204) edge node [left] {} (205)
+
+		  (301) edge node [left] {} (302)
+		  (302) edge node [left] {} (303)
+		  (303) edge node [left] {} (304)
+		  (304) edge node [left] {} (305)
+
+		  (401) edge node [left] {} (402)
+		  (402) edge node [left] {} (403)
+		  (403) edge node [left] {} (404)
+		  (404) edge node [left] {} (405)
+
+		  (501) edge node [left] {} (502)
+		  (502) edge node [left] {} (503)
+		  (503) edge node [left] {} (504)
+		  (504) edge node [left] {} (505)
+
+		  (601) edge node [left] {} (602)
+		  (602) edge node [left] {} (603)
+		  (603) edge node [left] {} (604)
+		  (604) edge node [left] {} (605)
+		  ;
+
+	\end{tikzpicture}
+  \caption{Example for subgraph finding. The occurrence of each state in the paths lets you determine G as an end state corresponding to A.}
+  \label{fig:subgraph_finding}
+\end{figure}
+
+Following the above description an algorithm has been implemented in java.
+Its realisation differs from the presented concept in one point that paths are not listed but states are directly labelled since saving all possible paths would be storage inefficient.
+The state machine is traversed similarly to a depht-first search and each visited state's label is incremented by one.
+If the currently visited state holds more than one outgoing transition and thus constitutes new paths,  all states laying on the current stack, i.e.
+the previously visited states on the current path, get their labels incremented by $(\textnormal{'number of outgoing transitions'} - 1)$.
+The algorithm turns back whenever it comes across a state which has already been visited within the same path stack of the depth-first search.
+
+The overall implemented way of subgraph finding is illustrated in pseudo code with algorithm~\ref{alg:subgraphfinding}.
+In line 4 the labelling process described above is executed by using one particular start state.
+If a state has got the same label as start state has, it is taken as potential end state.
+Otherwise, the entire state machine is assumed to be a subgraph and start state is taken as end state as well.
+Subsequent checks finally determine whether it can be considered as proper subgraph.
+There are no cycles allowed between start and end state since they're forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
+Furthermore also caving transitions are prohibited.
+Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
+If all checks are successful, start and end states get added to the result list $subgraphs$.
+This proceeding is applied to any state of the state machine what is achieved by a surrounding foreach loop.
+In the end the list $subgraphs$ contains all pairs of start and end states which form subgraphs.
+
+\begin{algorithm}
+\caption{Finding of all subgraphs.}
+\label{alg:subgraphfinding}
+\begin{algorithmic}[1]
+\STATE $subgraphs$: List
+\FORALL{states as $S$}
+  \STATE $start \leftarrow S$
+	\STATE doLabelling($start$);
+	\IF{state $E$ exists with label($E$)$=$label($S$)}
+		\STATE $end \leftarrow E$
+	\ELSE
+		\STATE $end \leftarrow start$
+	\ENDIF
+	\IF{noCycles($start$, $end$) $and$ noExtTransitions($start$, $end$)}
+		\STATE $subgraphs$.add($start$, $end$);
+	\ENDIF
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
+Depending on size and grade of branch of the state machine, the algorithm for finding subgraphs may take more or less time.
+To improve user experience during constraint generation a dialog has been implemented which informs the developer about the current progress.
+This includes the number of found constraints so far and a green bar which gives a percentual estimation of execution time.
+Furthermore a cancel button permits premature stopping of the execution.
+A snapshot of the Dialog is shown in figure~\ref{fig:progressdialog}.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.40\linewidth]{progressdialog} 
+  \caption{Snapshot of the progress dialog displayed during constraint generation.}
+  \label{fig:progressdialog}
+\end{figure}
+
+The algorithm implementation has been infiltrated by code snippets which frequently interacts with so called \emph{ValidationResultListener}s (figure~\ref{fig:validationresultlistener_class}).
+This interface is also implemented by the progress dialog.
+A \emph{ValidationResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about new found subgraphs with invoking \emph{void newSubGraphFound(SubGraph)}.
+The execution will become terminated whenever \emph{boolean continueGeneration()} returns true.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.55\linewidth]{validationresultlistener_class} 
+  \caption{Interface definition of \emph{ValidationResultListener}.}
+  \label{fig:validationresultlistener_class}
+\end{figure}
 
@@
 Besides a button is needed for triggering constraint generation.
-
-TODO: beschreiben dass vl im tab eingebunden wurde, die constraintgenerierung unter den zauberstab.
 
@@
 
-NuSMV
-warum den? Davor verwendete erwaehnen?
+One main requirement of the LTLCreator tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs.
+For this a \emph{ConstraintValidator} class has been defined which provides a method \emph{validate(AbstractOperator)} for proving constraints.
+For the constraint validation itself a particular model checker can be specified.
+As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
+It is an open source project and has been designed to be an open architecture for model checking.
+NuSMV doesn't have a java API but it can be externally accessed over command line.
+For this both the state machine program and the LTL formulas have to be put into file which has to match a NuSMV specific format.
+An example of this input language is given in listing~\ref{lst:nusmv}.
+\singlespacing
+\lstset{
+  basicstyle=\ttfamily,
+  columns=fullflexible,
+  showstringspaces=false,
+  commentstyle=\color{gray}\upshape
+}
+\begin{lstlisting}[float = htbp, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a NuSMV diagram file., label=lst:nusmv]
+MODULE main
+  VAR
+    state: 1..11;
+  ASSIGN
+    init(state) := 2;
+    next(state) := case state = 2 : {3, 1};
+                          state = 3 : 1;
+                          state = 1 : {2, 4};
+                          state = 4 : {1, 5};
+                          state = 5 : {6, 9, 10};
+                          state = 6 : {7, 8};
+                          state = 7 : 8;
+                          state = 8 : 1;
+                          state = 9 : 11;
+                          state = 10 : {9, 11};
+                          state = 11 : 1;
+                          TRUE : state;
+                     esac;
 
-a CTL model checker has been written in java, but...
+LTLSPEC G (state = 5 -> (F (state = 8 | state = 11)))
+\end{lstlisting}
+A state machine is encoded in one variable \emph{state}, and its value represents the active state.
+Thus each state is identified by a number.
+The state machine described in the example contains eleven states what is determined by the number range for \emph{state} from one to eleven.
+All variable assignments are specified under \emph{ASSIGN}, what is the semantic equivalent to transition definitions.
+The initial state can be determined and a case distinction regulates the next possible values, i.e.
+transitions to reachable states.
+If there is more than one outgoing transition, all target state identifiers have to be surrounded by brackets.
+The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
+The last line contains the constraint formula we want to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.
 
-Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.
+This kind of output is generated by the \emph{ConstraintValidator} whenever a constraint needs validation.
+All states are then mapped to numbers and transitions get transormed into case distinctions.
+The result is stored to file and then loaded by NuSMV.
+The corresponding valdidation result can be subsequently read from the command line and displayed in the editor next to the visual constraint.
 
-wie genau wird das getriggert (rekursiver listener an operatoren)
+It can happen that there is more than one validation task at a time.
+This is the case when the constraint generator creates multiple constraints or a change of the state machine causes all constraints to be revalidated at once.
+Since model checking is a quite ressource inefficient activity, it's recommended to execute all validation tasts consecutively rather than in parallel.
+A second advantage is that the developer doesn't have to wait for all of them to finish before he can see any result.
+Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
+All this is managed by the \emph{ValidatorThread}, a queue based mechanism for scheduling validations. Its class definition is presented in figure~\ref{fig:validatorthread_class}.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{validatorthread_class} 
+  \caption{Class definition of \emph{ValidatorThread}.}
+  \label{fig:validatorthread_class}
+\end{figure}
+Every validation task of a constraint gets enqueued in the \emph{ValidatorThread} after calling the method \emph{addToQueue(...)}.
+An identifier is used for each constraint what enables elimination of duplicate tasks for the same constraint in order not to strain the execution time.
+Tasks can also manually be removed from the queue by calling \emph{cancel(...)}.
+After a validation task is finished the result is published via callback to the \emph{ValidationResultListener}.
 
-ConstraintValidator
-ValidatorThread
-
-(im hintergrund, mit warteschlange. dabei sollen aber wenn ein contsraint frequently geändert wird, alte validationsaufträge des gleichen constraints aus der warteschlange entfernt werden.
-Callback mit result.
-
-%\subsection{Integratability}
 \subsection{Ease of integration}
@@
 Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.7\linewidth]{coalescence}
+  \caption{Interaction between LTLCreator and programming environment.}
+  \label{fig:coalescence}
+\end{figure}
 LTLCreator consists of a dashboard where constraints can be edited visually and an graphical output of validation results.
@@
 For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.
+All this functionality can be accessed through methods provided by the component \emph{LTLCreator}.
+Its class definition is shown in figure~\ref{fig:ltlcreator_class}.
+Whenever there is a change to the state machine, \emph{setModel(Fsm)} has to be called with the current state machine as parameter.
+This will cause the editor to revalidate where necessary.
+Constraint exchange can be done by using the methods \emph{addNewDashboard(AbstractOperator, ...)} and \emph{getOperators()}.
 
@@
   \centering
-  \includegraphics[width=0.7\linewidth]{coalescence}
-  \caption{Interaction between LTLCreator and programming environment.}
-  \label{fig:coalescence}
+  \includegraphics[width=0.65\linewidth]{ltlcreator_class}
+  \caption{Class definition of \emph{LTLCreator}.}
+  \label{fig:ltlcreator_class}
 \end{figure}

### 2012-10-20 src/editor/LTLCreator.java

@@ -31,7 +31,7 @@ import editor.validator.impl.NuSMVModelCheckerWrapper;
  * @author ludwig
  * 
  */
-public class Editor extends JComponent {
+public class LTLCreator extends JComponent {
 	private static final long serialVersionUID = 1L;
 	
 	private ConstraintValidator validator = new NuSMVModelCheckerWrapper();
@@ -41,7 +41,7 @@ public class Editor extends JComponent {
 	/**
 	 * Creates a new LTLCreator component. It can be added to any swing container.
 	 */
-	public Editor() {
+	public LTLCreator() {
 		this.setLayout(new BorderLayout());
 		
 		DndHandler.getInstance().registerTopLevelComponent(this);
@@ -107,10 +107,10 @@ public class Editor extends JComponent {
 			final OperatorTab operatorTab = (OperatorTab)graphTab;
 			
 			// Every dashboard has its own listener for changes on either the constraint or the model.
-			OperatorAndModelChangeListener l = new OperatorAndModelChangeListener(operatorTab, Editor.this.validator);
+			OperatorAndModelChangeListener l = new OperatorAndModelChangeListener(operatorTab, LTLCreator.this.validator);
 			operatorTab.getDashboard().addChangeListener(l);
-			Editor.this.validator.addModelChangeListener(l);
-			Editor.this.graphTabs.put(graphTab, l);
+			LTLCreator.this.validator.addModelChangeListener(l);
+			LTLCreator.this.graphTabs.put(graphTab, l);
 			
 			// initial validation
 			SwingUtilities.invokeLater(new Runnable() {
@@ -124,8 +124,8 @@ public class Editor extends JComponent {
 		@Override
 		public void graphTabRemoved(GraphTab graphTab) {
 			OperatorTab operatorTab = (OperatorTab)graphTab;
-			OperatorAndModelChangeListener l = Editor.this.graphTabs.remove(graphTab);
-			Editor.this.validator.removeModelChangeListener(l);
+			OperatorAndModelChangeListener l = LTLCreator.this.graphTabs.remove(graphTab);
+			LTLCreator.this.validator.removeModelChangeListener(l);
 			operatorTab.getDashboard().removeChangeListener(l);
 		}
 	}

### 2012-10-20 src/editor/Test.java

@@ -47,9 +47,9 @@ public class Test {
 	 * 
 	 * @return a new <b>Editor</b>.
 	 */
-	private Editor createEditorComponent() {
-		Editor editor = new Editor();
-		editor.addNewEmptyDashboard(false);
+	private LTLCreator createEditorComponent() {
+		LTLCreator lTLCreator = new LTLCreator();
+		lTLCreator.addNewEmptyDashboard(false);
 		
 		/* medication reminder example */
 		Fsm model = new Fsm();
@@ -86,9 +86,9 @@ public class Test {
 		nohelp.addTransition(welldone);
 		
 		model.setInitialState(menu);
-		editor.setModel(model);
+		lTLCreator.setModel(model);
 		
-		return editor;
+		return lTLCreator;
 	}
 	
 }

### 2012-10-21 thesis/tex/0_abstract.tex
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.

--- 2012-08-08 paper/Paper.tex
@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
+\chapter*{Abstract}
 
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[conference]{IEEEtran}
-% Add the compsoc option for Computer Society conferences.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  \graphicspath{{./images/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.jpg}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  \graphicspath{{./images/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows}
-
-\usepackage{amsmath}
-
-%\usepackage{natbib}
-
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Graphical formalization and automated computing of safety constraints in robotics}
-
-% author names and affiliations
-% use a multiple column layout for up to three different
-% affiliations
-\author{\IEEEauthorblockN{Ludwig N\"{a}gele}
-\IEEEauthorblockA{Department of Software Engineering and\\Programming Languages\\ 
-University of Augsburg\\
-86135 Augsburg, Germany\\
-Email: mail@ludwig-naegele.de}
-\and
-\IEEEauthorblockN{Bruce MacDonald}
-\IEEEauthorblockA{Department of Electrical Engineering\\ 
-The University of Auckland\\
-Auckland 1142, New Zealand\\
-Email: b.macdonald@auckland.ac.nz}
-%\and
-%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
-%\IEEEauthorblockA{Starfleet Academy\\
-%San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212\\
-%Fax: (888) 555--1212}
-}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-%\boldmath
 Nowadays robotic applications are no longer confined to only industrial use but increasingly assigned to jobs within human workspaces.
-Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behavior in order to match the user's individual tasks and requirements.
+Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behavior in order to match individual requirements.
 However both the human robot collaboration as well as the behavior definition by non experts may be a risk to humans within the robots' environment.
 Thus special thinking is needed about how to cope with safety requirements for robot behavior specification by consumers.
-%about how safety could be achieved in this scope.
-
-%to make human robot interaction safe and espacially how to make user defined behavior safe.
-%Special highlevel tools are developed in order to allow behavior definition by non experts.
-%Besides the fact that human robot collaboration is becoming highly important, it has also to be thought about convenient safety mechanisms.
 
@@
 Our approach addresses mainly non-safety-experts and our abstraction from a mathematical temporal logic expression to a more intuitive visual representation is intended to enable a wider range of software developers to do model checking.
-Furthermore we propose a new concept of development support %and qualitätssicherung
-by providing automated constraint generation functionality. It aims to help developers in defining reasonable constraints and finally to increase the probability of finding bugs.
+Furthermore we propose a new concept of development support by providing automated constraint generation functionality.
+It aims to help developers in defining reasonable constraints and finally to increase the probability of finding bugs.
 
 A graphical editor has been developed which demonstrates both visual constraint creation functionality as well as automated constraint generation.
-
-%Our purpose is to target non experts and to make safety checking available for a bigger community.
-Hence we had a special focus on a simple syntax and easy understandability as well as active support  in our approach.
-
-%Among others robots are alredy used in elderly care centers for healthcare purpose.
-
-\end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the conference you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals/conferences frown on
-% math in the abstract anyway.
-
-% no keywords
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
-%----------------------------------------------------------------------
-% SECTION I: Introduction
-%----------------------------------------------------------------------
-\section{Introduction}
-
-%\IEEEPARstart{S}{afety}
-Safety critical applications require extensive testing or formal verification in order to achieve adequate safe and predictable behavior.
-Even if a more highlevel program has no physical movements and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions executed by interacting actors.
-%Even if a more highlevel program has no physical actors and thus cannot cause any damage, it might still have dangerous effects on the environment.
-Imagine a medication reminder application which guides a person through the process of medication intake.
-All interaction is done by showing textual instructions on a display.
-An error in the program logic could cause an already taken medication to be reminded again and may injure an Alzheimer's diseased person by medication overdosage.
-This example is taken from a healthcare robotics project at the University of Auckland, New Zealand.
-Scientists are working on applications such as the medication reminder~\cite{medicationreminder} which run on mobile robots serving in nursing homes for the elderly.
-Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{5649910}\cite{healtbot_new}.
-The applications are meant to be developed by healthcare professionals using Robostudio~\cite{robostudio}, a visual programming environment for rapid authoring and customization of complex services on a personal service robot.
-%state machine based processes and a graphical tool for the designing of human robot interaction.
-Generally these people don't have profound knowledge of the complex mathematical syntax used for formal verification methods such as computation tree logic (CTL) or linear temporal logic (LTL).
-Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthbots.
-In case of the medication reminder mentioned above it mitght be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
-
-So we need a mechanism for the specification of such safety constraints which doesn't require special or expert knowledge and thus is easy to use for all kind of programmers.
-A visual language intended to fill this gap is presented in section~\ref{sec:visualformalism}.
-
-Unfortunately the supply of an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer.
-Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
-To support the user in finding significant constraints it might be useful to automatically compute constraint suggestions which make true testimonies about the current designed program.
-First of all the generated constraints provide an opportunity for the developer to identify reasonable constraints as well as contradictions between constraints and specification.
-The latter would mean there are errors in the program which lead to undesired constraints.
-Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
-
-In section~\ref{sec:automatedconstraintgeneration} we introduce an heuristic for finding reasonable safety constraints based on a state machine in the healthcare robotics domain.
-
-%Once such constraints are defined, they determine the grade of safety coverage.
-Unfortunately, the program may still contain errors if there is at least one significant constraint missed out.
-To support the user in finding such ones the tool could furthermore automatically compute suggestions for possible constraints based on the designed statemachine.
-These are displayed to the programmer and may help him identifying reasonable constraints as well as contradictions between program and specification.
-
-%Our approach presented in this paper has a special focus on the healtcare domain.
-
-%Dieses beispiel ist eine Anwendung eines umfangreichen Healthcare-Projekts, an dem im robotics lab der Universität von Auckland gearbeitet wird.
-Die zustansbasierten Anwendungen laufen auf Cafero, einem mit blutdruckmessgerät und anderen tools austestatteten mobilen Roboter mit Touchdisplay.
-%Die einzelnen Anwendungen werden von Healthcare-Professionals mit dem Tool Robostudio [cite] entwickelt, das eine graphische oberfläsche zur modellierung von zustandsmaschinen externen services und  bietet.
-%Allerdings haben diese not to have profound knowledge of the complex mathematical syntax used for formal verification methods such as computation tree logic (CTL) or linear temporal logic (LTL)).
-%Dennoch möchte man gegenüber den Patienten, die von solchen Robotern gepflegt werden, ein gewisses Level an Sicherheit gewährleisten können.
-Beispielsweise soll in obigem Beispiel sichergestellt werden können, dass ein Medikament nach der Einnahme nicht erneut in einer Erinnerung auftaucht, oder dass bei Nichteinnahme ein Mitarbeiter benachrichtigt wird.
-%Dazu bedarf es allerdings Mechanismen zur spezifizierung solcher Sicherheitseigenschaften, die für den Programmentwickler zugänglich und einfach (am besten ohne spezialwissen) zu benutzen sind.
-
-%Constraints can be easily defined graphically and provide the user with an instant result about validity.
-%, which can make propositions about the behavior of state machines.
-%This language is an abstraction layer above temporal logic and allows a rather intuitive than textual model checking.
-%The abstraction from textual and mathematical temporal logic expression allows a rather intuitive way of model checking and thus makes this language accessible to non-experts.
-
-%CTL and LTL are important concepts for model checking and verification of statemachines.
-But their powerfulness is reflected in the complexity of their mathematical expressions, what makes it difficult to use for programmers who haven't got deeper expertise in this subject.
-
-%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions.
-The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually.
-A common LTL model checker will be added to it as well which fulfils the verification of the translated constraints on the designed statemachine.
-In case of verification failures the concerning constraints and a visual counter example can be presented.
-
-%Once such constraints are defined, they determine the grade of safety coverage.
-Unfortunately, the program may still contain errors if there is at least one significant constraint missed out.
-To support the user in finding such ones the tool could furthermore automatically compute suggestions for possible constraints based on the designed statemachine.
-These are displayed to the programmer and may help him identifying reasonable constraints as well as contradictions between program and specification.
-
-%----------------------------------------------------------------------
-% SECTION I: Introduction
-%----------------------------------------------------------------------
-\section{Related work}
-
-%Was gibt es schon alles?
-
-% 1993 A Visual-Programming Environment for a Temporal Logic Language
-% 1995 Visual Specification of Branching Time Temporal Logic
-% 2007 HomeTL: A visual formalism, based on temporal logic, for the design of home based care
-% 2007 A Visual Editor to Support the Use of Temporal Logic for ADL Monitoring
-
-A first step towards a graphical tool for the designing of safety constraints is taken by Sisiruca and Ionescu~\cite{332301}.
-They developed an object-oriented graphical environment for visually creating temporal logic sentences and rules.
-The tool presents temporal operators as logical modules with boolean inputs and outputs which can be connected to one big expression graph.
-
-Also Del Bimbo et al.
-did work on a visual tool for temporal logic, but with a special focus on hierarchical representations of formulas~\cite{520786}.
-Here each node of a tree is described by an abstract operator whose subformulas are branches to their replacement operators.
-Thus the root node represents the entire formula.
-They also explored the formula trees' 3D representation within a virtual space.
-
-Both approaches are different fundamental ideas of representing temporal formulas.
-Whereas having a graph and inputs and outputs seems not to be applicable for our approach, the idea of hirarchical operator representation of Del Bimbo et al.
-will be a good base for our try to find an easy to use concept for constraint creation and understanding.
-
-%aim for simplifying the development of constraints, but they still demand knowledge of the underlying mathematical formalism of temporal logic.
-
-The step towards a more abstract level of development is taken by HomeTL~\cite{4341725}, a visual formalism for the design of home based care.
-It allows healthcare professionals to specify rules and sequences of user actions in a quite nontechnical manner.
-Based on these conditions a monitoring system in a patient's home environment can detect and report abnormal situations within the daily routine.
-
-Even though this approach touches the subject of healthcare it barely matches our problem.
-HomeTL focuses rather on monitoring temporal boundaries of a patient's behavior than on ensuring functional safety of an implemented program.
-But it depicts how nontechnical constraint design can be realised and its graphical design gives a good example how to handle time dependencies.
-
-%There are already works about constraint generation such as one approach of NASA [4], for example.
-They used test oracles in order to achieve consistency between program code and specification during development.
-% Das wollen wir auch erzielen mit unseren automatisch generierten constraints...
-
-%But these kind of oracles are for automated test purpose and are rather complex than intuitive.
-However, our goal is to find nonmathematical and just intuitive ones.
-An intelligent heuristic must determine key states and possible conjunctions of them in order to build reasonable constraints.
-Afterwards they get filtered by an algorithm which computes importances for each constraint based on state weights.
-Finally the developer obtains a set of visual constraints which he can then check for sanity.
-
-%----------------------------------------------------------------------
-% SECTION II: Requirements
-%----------------------------------------------------------------------
-\section{Requirements}
-\label{sec:requirements}
-
-% soll hier noch gesagt werden, dass wir davon ausgehen, dass es keine endzustände gibt?
-
-We want to introduce a visual language for the defining of safety constraints for state machine definition of robot behaviour which is suitable for non-experts in the formal methods domain.
-Ideally its appliance is easy to use and intuitive to read in order to facilitate maintainability.
-Hence, the graphical formalism should abstract from the usually complex and mathematical concept of conventional formal methods but still stay unreduced in expressiveness compared to LTL as far as possible.
-
-Furthermore a heuristic should be developed of how to identify reasonable constraints on an underlying state machine program. 
-But since the particular domain distinguishes the significance of constraints respectively it might be difficult to show a universal heuristic working for all use cases.
-Hence, in this paper the concept of development support by constraint generation is examined within the domain of healthcare robotics.
-As a special requirement to the heuristic at least the following constraints are to be found for the medication reminder application:
-
-\begin{itemize}
-	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
-	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
-% The next item is left out so that we can say, our heuristic found one we didn't think about yet :)
-%	\item During the medication intake process the program can not switch to other services such as entertainment mode ore blood pressure measurement, for example.
-\end{itemize}
-
-Some of the statements don't make sense of course if patients lie or pretend to take their medication.
-As our approach tries to validate state machine behaviour rather than human attitudes we assume correct collaboration or the use of tools which can recognize fairly medication intake.
-Furthermore also the correctness of external services used by the medication reminder such as the database module holding all reminders is not taken into account.
-
-All the functionality mentioned above shall be combined in one editor and for evaluation purpose later be integrated into a programming environment for state machine editing.
-
-%----------------------------------------------------------------------
-% SECTION II: Design
-%----------------------------------------------------------------------
-\section{Design}
-
-\subsection{Visual formalism}
-\label{sec:visualformalism}
-
-% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv.
-
-The presented visual language is made for expressing constraints for state machines.
-Similar to Del Bimbo's formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchical manner.
-However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
-Each block represents either an operator or a proposition and has its specified semantics.
-Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.
-
-The visual language has support for eight operator types and one proposition as shown in figure~\ref{fig:operators}.
-
-\begin{description}
-	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
-	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
-	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
-Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
-	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
-	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state.
-Other types such as numeric or string equations may be added but are less important for our healthcare subject.
-	\item[(f)] \emph{FUTURE} operator: $F \varphi$.\\At least one time -- now or in a later state -- $\varphi$ must be true.
-	\item[(g)] \emph{ALWAYS} operator: $G \varphi$.\\$\varphi$ must be true now and in all following states.
-	\item[(h)] \emph{NEXT} operator´: $X \varphi$.\\$\varphi$ has to be true in the next state.
-	\item[(i)] \emph{UNTIL} operator: $[\varphi U \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true.
-In addition eventually there has to be a future state with $\psi$ being true.
-\end{description}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.5]{table}
-  \caption{The visual language consists of eight unary or binary operator types and one state proposition, each of them having its own color flavor.
-Logical operators are arranged in a vertical row wereas time relevant statements are aligned horizontally.}
-  \label{fig:operators}
-\end{figure}
-
-Regarding to the requirements the visual language should be very easy to use, especially for non-experts.
-Furthermore it should provide an intuitive appliance which is ideally learnable within little time.
-So there are some design decisions an editor for this visual language has to fulfill:
-
-\begin{itemize}
-	\item Each operator has a different significant color flavor. This kind of classification enables a faster subconscious identification by the user.
-	\item All editing is simply done by only drag and drop.
-Operators and propositions can be moved and dropped onto other operators.
-A tool bar provides all kinds of operator and proposition types to be instantiated and ideally a trash can for deleting existing operators.
-	\item The layout of constraints (nested operators) has two dimensions with a certain meaning.
-Figure~\ref{fig:directions} demonstrates the ``logical'' vertical read direction as well as the time releveant horizontal line.
-	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer.
-All higher operators become bleached out so that the hovered operator and its children appear highlighted.
-It's intended to be similar to line coloring when reading a digital text.
-	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appeared to make the look unnecessarily complex.
-These operators can be merged as shown in figure~\ref{fig:nested_and}.
-	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
-	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances of them.
-Having no additional buttons reduces complexity and keeps it simple.
-\end{itemize}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.5]{directions}
-  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
-  \label{fig:directions}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.5]{and_simplify} %width=0.7\linewidth
-  \caption{In order to simplify the appearance of constraints certain operators might be merged.
-Merged nested \emph{AND} operators (center) are less complex as default ones (left).
-However nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
-  \label{fig:nested_and}
-\end{figure}
-
-\subsection{Automated constraint generation}
-\label{sec:automatedconstraintgeneration}
-
-In order to support the developer in specifying safety constraints, automated generation and suggestion of constraints might be helpful.
-But it's helpful only then, if the suggested constraints are relevant and reasonable.
-This however depends on the domain and use case.
-Regarding to the postulated constraints in the requirements in section~\ref{sec:requirements} it turned out that all therefor relevant states have got one of the following characteristics:
-%following types of states are of a special interest:
-
-\begin{enumerate}
-	\item States having two or more outgoing transitions ($\alpha$-states).
-	\item States in which all paths starting from one $\alpha$-state come together again ($\beta$-states).
-	\item All States immediately before $\beta$ states ($\gamma$-states).
-\end{enumerate}
-
-In order to compute the required constraints and possibly even more of them automatically we manage with so called subgraphs defined as follows:
-%Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
-%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}.
-In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows
-
-\begin{itemize}
-	\item It has a start state ($\alpha$-state) with at least two outgoing transitions.
-	\item There is an end state ($\beta$-state) in which all possible paths starting from start state come together again.
-	\item All states between start and end state have no other incoming transitions than the ones originally coming from start state.
-That means there is no path to visit these states exept through start state.
-	\item start state may have incoming transitions from states not contained in the subgraph.
-	\item end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
-\end{itemize}
-
-\begin{figure}[htbp]
-  \centering
-
-  \tikzstyle{state} = [circle,fill=white,draw]
-  \tikzstyle{statein} = [circle,fill=blue!20,draw]
-  \tikzstyle{every path} = [font=\sffamily\small]
-
-	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=15mm]
-
-	  \node[statein] (1) {$\alpha$};
-	  \node[state] (2) [above left of=1] {};
-	  \node[state] (3) [above right of=1] {};
-	  \node[statein] (4) [below left of=1] {};
-	  \node[statein] (6) [below of=4] {$\gamma$};
-	  \node[statein] (7) [below right of=4] {$\gamma$};
-	  \node[statein] (5) [below right of=1] {};
-	  \node[statein] (8) [below of=5] {$\gamma$};
-	  \node[statein] (9) [below of=7] {$\beta$};
-	  \node[state] (10) [right of=9] {};
-	  \node[state] (11) [below left of=9] {};
-	  \node[state] (12) [below right of=9] {};
-
-	  \path[dashed]
-	    (2) edge node [left] {} (1)
-	    (3) edge node [left] {} (1)
-	    (10) edge node [left] {} (9)
-	    (9) edge node [left] {} (11)
-	    (9) edge node [left] {} (12);
-		\path[]
-	  	(1) edge node [left] {} (4)
-	        edge node [left] {} (5)
-	    (4) edge node [left] {} (6)
-	        edge node [left] {} (7)
-	    (5) edge node [left] {} (8)
-	    (6) edge node [left] {} (9)
-	    (7) edge node [left] {} (9)
-	    (8) edge node [left] {} (9)
-	    ;
-
-	\end{tikzpicture}
-  \caption{Example for a subgraph.}
-  \label{fig:subgraph}
-\end{figure}
-
-Using this specification all possible subgraphs can be detected within a state machine.
-If a subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarily for $\beta$ and all $\gamma$'s):
-
-\begin{equation} \label{eq:first}
-  \models G (\alpha \Rightarrow F \beta)
-\end{equation}
-
-\begin{equation} \label{eq:second}
-  \models G (\alpha \Rightarrow F (\bigvee_{\gamma} \gamma))
-\end{equation}
-
-\begin{equation} \label{eq:third}
-  \models G (\alpha \Rightarrow [\neg \beta U \bigvee_{\gamma} \gamma])
-\end{equation}
-
-Additionaly for all states not contained in the subgraph (let's call them $\delta$-states, and the proposition $\delta$ stands for ``$\delta$ is current state'') the following formula is true:
-
-\begin{equation} \label{eq:fourth}
-  \models G (\alpha \Rightarrow [(\bigwedge_{\delta} \neg \delta) U \beta])
-\end{equation}
-
-In case of $\alpha = \beta$ -- i.e. when all paths starting from $\alpha$-state eventually return again (cycle) -- even one more constraint can be assumed to be true:
-
-\begin{equation} \label{eq:fifth}
-  \models G F \alpha
-\end{equation}
-
-These five constraints apply for each subgraph matching the conditions mentioned above.
-Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state.
-Formula~\ref{eq:second} ensures that there will be always at least one $\gamma$-state be visited after each visit of an $\alpha$-state.
-The quite similar but stricter formula~\ref{eq:third} demands additionally a visit of at least one $\gamma$-state before the $\beta$-state can be reached.
-Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states.
-That the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.
-
-All formulas can be translated to the visual formalism presented in section~\ref{sec:visualformalism} and shown to the user.
-The accordant LTL formulas for the postulated constraints in section~\ref{sec:requirements} belong to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:
-
-\begin{equation}
-  \models G F \textnormal{'Polling'}
-\end{equation}
-
-\begin{equation}
-	\models G (\textnormal{'Not taken yet'} \Rightarrow F (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
-\end{equation}
-%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
-%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
-
-%----------------------------------------------------------------------
-% SECTION III: Prototype and evaluation
-%----------------------------------------------------------------------
-\section{Prototype and evaluation}
-
-To experience and evaluate the visual formalism as well as the usefulness of constraint generation we developed an editor which provides functionalities all proposed functionalities while taking the requirements into account.
-It's implemented as a java swing component, making it easy to integrate the visual language into other java applications.
-An API allows to set the underlying state machine what also influences the state propositions available for constraint editing.
-Furthermore a particular model checker can be specified to be used for constraint validation.
-As default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046}~\cite{NuSMV2} is used for constraint validation.
-
-Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right which contains drag providers for the creating of all operator and proposition types as well as a trash for deleting.
-Constraints can be composed in the dashboard in the middle of the editor by drag and drop.
-The tab functionality on the left allows to manage several constraints.
-Each tab schows a small thumbnail of the constraint and a symbol indicating its validity what is a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress''.
-The magic wand button within the tab pane triggers the constraint generation.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=\linewidth]{editor} 
-  \caption{Snapshot of the visual editor.}
-  \label{fig:editor}
-\end{figure}
-
-In order to evaluate the proposed safety functionalities on various programs the tool was integrated into Robostudio~\cite{robostudio}.
-It allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
-As an example of how to use them we will have a closer look at the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}.
-Starting from ``Menu'' the user can choose from several services such as blood pressure measurement ore entertainment mode.
-In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part.
-In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered.
-The patient can state if he's already taken his medication, or decide whether he wants to take it or not.
-In the latter he may give a reason for it and a caregiver gets notified about this incident.
-Otherwise it will result in ``Well done!'' after completing intake.
-
-%we can now vergleichen die oben genannten bedingungen.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=\linewidth]{stm} %width=0.7\linewidth
-  \caption{Simplified screen-flow of the medication reminder application.}
-  \label{fig:medicationreminder}
-\end{figure}
-
-%Vorstellung des medication reminders
-
-%in particular
-
-% TODO: hier vlt ein komponentendiagramm?
-
-%Der Editor wurde in robostudio eingebunden, ein werkzeug zur erstellung von healthcare applicationen (elderly) basierend auf zustandsmaschinen.
-(bereits in der einleitung erwähnt).
-Über die darin definierbaren zustansmaschinen können wir nun aussagen mit der vl erstellen.
-
-%das tool wurde den healthcare professionals vorgelegt und die ergebnisse werden gespalten in den folgenden zwei unterpunkten beschrieben (vl und constraint generierung).
-
-%Robostudio was used as a platform to integrate the visual language described in this paper .
-%basierend auf diesen healthcare applicationen können nun sicherheits constraints über unser tool definiert werden.
-
-%rs was used as a platform to integrate the safety critical visual programming language described in this paper.
-fs's modular design allows the usage of the state machine model and add plugins to be integrated into the VPE, making it usabble as a plat
-
-%personal service robot
-
-%visual programming environment
-%We embedded the editor into Robostudio~\cite{robostudio}, a tool for the creating of healthcare applications based on state machines.
-
-%an editor for state machine based processes and a graphical tool for the designing of human robot interaction
-
-\subsection{Visual constraints}
-
-Before we can start creating constraints it might be useful to think about a reasonable term we want to express.
-Let's take the example ``Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake'' given in the requirements in section~\ref{sec:requirements}.
-The LTL formula for this constraint is:
-
-\begin{equation} \label{eq:sampleconstraint}
-  \models G (\textnormal{'Not taken yet'} \Rightarrow F (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'}))
-\end{equation}
-
-This expression can now be translated to the visual form as we just have to break down the formula from the outside to the inside.
-A surrounding \emph{GLOBALLY} encapsulates an \emph{IF} block since it is the substitution for \emph{IMPLIES}.
-Its first operator is a proposition \emph{'Not taken yet'} whereas the second parameter has to be a \emph{FOLLOWS} block containing a disjunction of two propositions \emph{'Well done!'} and \emph{'Staff notified'}.
-The resulting visual constraint is shown in figure~\ref{fig:sampleconstraint}.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.5]{sampleconstraint}
-  \caption{Visual constraint creation for LTL formula ~\ref{eq:sampleconstraint}.}
-  \label{fig:sampleconstraint}
-\end{figure}
-
-After each change within the dashboard the constraint is automatically revalidated by using the underlying model checker.
-As long as the constraint is not completed an ``incomplete'' sign is displayed in the regarding tab.
-Otherwise an animated ring indicates validation in progress what finally results in either a ``valid'' or ``invalid'' sign.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.5]{results}
-  \caption{Status indicators displayed during or after validation: incomplete, validation in progress, invalid and valid.}
-  \label{fig:results}
-\end{figure}
-
-Owing to the elaborated performance of NuSMV even the validation of constraints on huge and complex programs is fast and allows quick feedback.
-Besides, the validation itself runs in background without locking the dashboard.
-Therefore we got rid of waiting times and disruptions during constraint development what would be very well the case with conventional model checking where constraint development and constraint validation are alternating occupations.
-Thus it's considered as an improvement of user experience.
-
-\subsection{Constraint generation}
-\label{sec:prototypeandevaluation_constraintgeneration}
-
-The automated constraint generation can be triggered by a click on the magic wand button in the tab area.
-After activation a dashboard gets opened in a new tab for each found constraint, and validation is initiated immediately.
-In case of the medication reminder application six constraints are found, whereof \emph{A)} and \emph{B)} shown in figure~\ref{fig:generatedconstraints} are actually matching the postulations in the requirements in section~\ref{sec:requirements}.
-The constraint used for demonstration in the previous chapter gets also generated, however \emph{B)} forms an intensified restriction of it.
-
-Constraint \emph{A)} ensures the ``Polling'' state to be always eventually visited again.
-If a pending medication is not already taken, constraint \emph{B)} guarantees intake or staff notification before next reminders can be read from database.
-
-We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{C)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment mode ore blood pressure measurement.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=\linewidth]{generatedconstraints} %width=0.7\linewidth
-  \caption{Automatically generated constraints which match the postulations in the requirements.}
-  \label{fig:generatedconstraints}
-\end{figure}
-
-%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde kann bei großen programmen (ab ca.
-1000 Zuständen und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern.
-Um den warteprozess erträglicher zu machen wird während der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
-% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht
-
-%----------------------------------------------------------------------
-% SECTION III: Prototype and evaluation
-%----------------------------------------------------------------------
-\section{Future work}
-
-The editor presented in this paper proofes validity of all constraints and presents the result \emph{valid} or \emph{invalid} to the user.
-But in contrast to conventional model checking no additional explanation for the outcome such as a counter example is given.
-It can be considered to extend the visual language with support for bug identification by a visualization of counter examples, for example.
-%\item Substitution of parts of constraints would be nice in order to make it easier to understand.
-(A simple operator which abstracts from a more complex operator).
-It's comparable to composite states of hirachical state machines.
-
-%\item A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.
-
-%	\item constraints bewerten (nach gewichtung?) und nur wichtige vorschlagen? Oder zumindest sortieren
-
-Since we focus on the healthcare domain in this paper the only proposition type for constraint creation is the state proposition.
-But to suit other domains additional proposition types such as number or string equations might be added.
-Also the portability of the constraint generation heuristic to other domains should be examined and different methods than subgraph based constraint finding might be elaborated.
-
-%\begin{itemize}
-%	\item As already mentioned
-%	\item Yet, there are only states and no variable expressions etc.
-%	\item Substitution of parts of constraints would be nice in order to make it easier to understand.
-(A simple operator which abstracts from a more complex operator).
-It's comparable to composite states of hirachical state machines.
-%	\item A feature which can display an counter example in case a constraint is not valid. It would make the user comprehend the reason in what situation a constraint fails.
-%	\item Problem mit der Zustandsbenennung!
-Die Zustände müssen passende Namen haben, was aber nicht immer möglich ist.
-Daher müsste man eigentlich auch Transitionsbedingungen in den Constraints erlauben
-%	\item constraints bewerten (nach gewichtung?) und nur wichtige vorschlagen? Oder zumindest sortieren
-% \item es wird nur gutes gefunden (also was passieren muss). es werden keine constraints gefunden, die schlechte abläufe darstellenund verbieten
-%	\item übertragbarkeit auf andere domänen untersuchen?
-%	\item A bigger generation algorithm.
-As already metioned our approach only targets the healthcare domain.
-Especially the constraint generation can be improved to match also other domains.
-vielleicht geht beim triggern der constraint generierung ein fenster auf, wo man die heuristic angeben kann (eventuell benannt und gegliedert nach domains).
-%\end{itemize}
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
-
-\section{Conclusion}
-
-In this paper we represented a visual formalism for the defining of safety constraints and a concept of automated constraint generation.
-Both features are implemented and assembled in an editor for standalone or integration use.
-The visual language is pretty simple what makes it accessible to a wider range of software developers.
-Nevertheless its expressiveness is fairly powerful and allows serious model checking.
-
-The concept of constraint generation worked very well for our use case the medication reminder application and all postulated constraints were found.
-Since we even discovered a new reasonable constraint we are convinced this concept as well as the visual language for constraint defining can be a beneficial supporting thing for plenty of developers.
-
-% conference papers do not normally have an appendix
-
-% use section* for acknowledgement
-\section*{Acknowledgment}
-
-The authors would like to thank...
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,Paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-%\bibitem{healthcare}
-%B.~MacDonald, \emph{Healthcare}.
-%\bibitem{robostudio}
-%C.~Datta and B.~MacDonald, \emph{Robostudio}.
-
-%\bibitem{visualtemporallogic1993}
-%\emph{1993 A Visual-Programming Environment for a Temporal Logic Language}.
-%\bibitem{visualtemporallogic1995}
-%\emph{1995 Visual Specification of Branching Time Temporal Logic}.
-%\bibitem{hometl}
-%\emph{2007 HomeTL: A visual formalism, based on temporal logic, for the design of home based care}.
-%\bibitem{adl}
-%\emph{2007 A Visual Editor to Support the Use of Temporal Logic for ADL Monitoring}.
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
-
-% that's all folks
-\end{document}

### 2012-10-21 thesis/tex/2_background.tex
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.

--- 2012-10-15 thesis/tex/3_background.tex
@@
 
-%\begin{wrapfigure}{r}{0.58\textwidth}
+The two newest robots IrobiQ and Cafero are shown in figure~\ref{fig:irobiq_cafero}.
+%\begin{figure}[htbp]
 %  \centering
-%  \includegraphics[width=0.56\textwidth]{irobiq_cafero}
+%  \includegraphics[width=0.6\linewidth]{irobiq_cafero} 
 %  \caption{IrobiQ and Cafero, two robots of the healtcare project.}
 %  \label{fig:irobiq_cafero}
-%\end{wrapfigure}
-The two newest robots IrobiQ and Cafero are shown in figure~\ref{fig:robots}.
-\begin{figure}[htbp]
+%\end{figure}
+These mobile 
+\begin{wrapfigure}{r}{0.60\textwidth}
   \centering
-  \includegraphics[width=0.6\linewidth]{irobiq_cafero} 
+  \includegraphics[width=0.56\textwidth]{irobiq_cafero}
   \caption{IrobiQ and Cafero, two robots of the healtcare project.}
-  \label{fig:robots}
-\end{figure}
-These mobile robots have got a touch display for presenting screens and receiving user input.
-Furthermore they are equipped with additional external tools such as the blood pressure measurement device.
+  \label{fig:irobiq_cafero}
+\end{wrapfigure}
+robots have got a touch display for presenting screens and receiving user input. Furthermore they are equipped with additional external tools such as the blood pressure measurement device.
 Applications running on the robot use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedbak.
@@
 %This makes the program behavior replaceable and maintenance can be done easily.
-The interpreter processes programs written in Robot Behavior Description Language (RBDL), a particular DSL based on XML, which has been developed specially for the healthcare robotic applications.
+The interpreter processes programs written in Robot Behavior Description Language (RBDL), a particular DSL based on XML, which has been developed particularly for the healthcare robotic applications.
 It allows defining state machine based program behavior and creating UI elements for screen visualizations.
@@
 
+\singlespacing
 \lstset{
@@
 %\lstset{language=XML}
-\begin{lstlisting}[language=XML, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a state defined by RBDL~\cite{robostudio}., label=lst:RBDL]
+\begin{lstlisting}[float = htbp, language=XML, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a state defined by RBDL~\cite{robostudio}., label=lst:RBDL]
 <state no="84">
@@
 
+\onehalfspacing
+
 For each state of the program, backgroundactions can be defined.
-They are are transparent to the user and run in background.
+They are transparent to the user and run in background.
 Text messages can be sent or external devices can be acessed, for example.
@@
 
-For this work the most important thing about the RBLD to remember is, that it can define state machine behavior with states, transitions and events.
+For this work the most important thing about the RBDL to remember is, that it can define state machine behavior with states, transitions and events.
 This fact was basement for later decisions regarding used safety mechanisms.

### 2012-10-21 thesis/tex/3_goals.tex
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.

\chapter{Goals}
\label{chap:goals}

% soll hier noch gesagt werden, dass wir davon ausgehen, dass es keine endzustï¿½nde gibt?

Ausgehend vom Ziel, die healthcare applikationen des healthcare projects der Universität Auckland gewissen Sicherheitsstandards näher zu bringen und somit einen Schritt in Richtung Sicherheitszertifizierung zu machen, sollen dem VPE Robostudio hierfür notwendige Werkzeuge hinzugefügt werden. Dieses Tool(VPE) wird zur Entwicklung von Roboterapplikationen in Form von Zustandsmaschinen verwendet und soll zukünftig eine neue Schnittstelle bieten, über die der Programmierer das Roboterverhalten auf funktinale Sicherheit hin überprüfen kann. Diese Erweiterung soll zum einen nahtlos in Robostudio integriert sein, zum anderen soll es so umgesetzt sein, dass es über eine übersichtliche API auch in andere Programme einfach eingebunden werden kann.

We want to introduce a visual language for defining safety constraints in state machine definitions of robot behaviour, that is suitable for people who are not experts in formal methods. The visual language should be easy to use and intuitive to read, in order to facilitate maintainability.
Hence, the graphical formalism should abstract from the usually complex and mathematical concept of conventional temporal logic but still be significantly expressive.

Furthermore we want to integrate some mechanism to support users in finding suitable constraints for a certain application. We assume that applications are modeled by state transition graphs. This modeling paradigm is used in many healthcare robots that employ dialogue systems for communicating with users~\cite{Bickmore2006}. Other possible fields might be shop assistant robots or entertainment robots. For such systems, our approach is suitable. It is based on heuristically analyzing state transition graphs and deriving constraints related to the graph structure. In this work, we want to investigate a particular heuristic as a first step towards user support.

Finally, we will present some results of applying this heuristics to the medication reminder example described before. The minimum goal that should be achieved was the identification of the following constraints:

\begin{itemize}
	\item The application will always eventually check the database for new reminder jobs. Thus we ensure that pending reminders are processed.
%	\item Either a reminded medication is taken properly, or caregivers are notified that the patient refused medication intake.
	\item Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.
% The next item is left out so that we can say, our heuristic found one we didn't think about yet :)
%	\item During the medication intake process the program can not switch to other services such as entertainment mode ore blood pressure measurement, for example.
\end{itemize}

Some of the statements do not make sense if patients lie or pretend to take their medication but don't take it. As our approach validates state machine behaviour rather than human attitudes we assume there is either cooperation from the patient or a separate tool that can verify medication intake. Furthermore we assume the correctness of external services used by the medication reminder such as the database module holding all reminders.

% Bruce: I'm not sure we need this sentence. I think you have integrated it now anyway.
% All the functionality mentioned above shall be combined in one editor and for evaluation purpose later be integrated into a programming environment for state machine editing.

### 2012-10-21 thesis/tex/4_visualformalisms.tex
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.

--- 2012-10-19 thesis/tex/5_design.tex
@@
 \end{figure}
-
-%\section{Automated constraint generation}
-%\label{sec:automatedconstraintgeneration}
-\chapter{Automated constraint generation}
-\label{chap:automatedconstraintgeneration}
-
-As already mentioned before, it might be difficult for developers to find all reasonable safety constraints by hand which are needed for full safety covery.
-Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
-To help prevent this to happen, we introduce the concept of automated constraint generation for state machines.
-Once a program reaches a complete condition, constraint suggestions which are valid on the current program can be computed fully automatically.
-On the one hand they may help the developer identifying reasonable constraints he didn't think about as well as contradictions between program and specification, eg.
-if a computed constraint doesn't make sense at all.
-On the other hand all constraints can be revalidated after each change to the software and thus ensure consistency.
-
-In order to support the developer in developing safe programs, automated generation and suggestion of constraints might be helpful.
-But it's helpful only if the suggested constraints are relevant and reasonable.
-
-\section{Sub graph definition}
-
-Whether suggested constraints are relevant and reasonable
-%This however
-depends on the domain and use case.
-It has to be mentioned that our approach of finding an algorithm for constraint finding presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
-Regarding the important constraints postulated in chapter~\ref{chap:goals}, we found out that all all the relevant states fall into one of the following categories:
-
-\begin{enumerate}
-	\item States having two or more outgoing transitions ($\alpha$-states).
-	\item States in which all paths starting from one $\alpha$-state come together again ($\beta$-states).
-	\item All States immediately before $\beta$ states ($\gamma$-states).
-\end{enumerate}
-
-In order to compute the required constraints automatically, and possibly additional ones, we deal with subgraphs defined as follows:
-%Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
-%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}.
-In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows
-
-\begin{itemize}
-	\item There is a start state ($\alpha$-state) with at least two outgoing transitions.
-	\item There is one end state ($\beta$-state) in which all possible paths starting from start state come together again.
-	\item All states between the start and end states have no other incoming transitions than the ones originally coming from the start state.
-That means there is no path to visit these states except through the start state.
-	\item The start state may have incoming transitions from states not contained in the subgraph.
-	\item The end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
-\end{itemize}
-
-\begin{figure}[htbp]
-  \centering
-
-  \tikzstyle{state} = [circle,fill=white,draw]
-  \tikzstyle{statein} = [circle,fill=blue!20,draw]
-  \tikzstyle{every path} = [font=\sffamily\small]
-
-	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=15mm]
-
-%	  \node[statein] (1) {$\alpha$};
-%	  \node[state] (2) [above left of=1] {};
-%	  \node[state] (3) [below left of=1] {};
-%	  \node[statein] (4) [above right of=1] {};
-%	  \node[statein] (5) [below right of=1] {};
-%	  \node[statein] (8) [right of=5] {$\gamma$};
-%	  \node[statein] (6) [right of=4] {$\gamma$};
-%	  \node[statein] (7) [below right of=4] {$\gamma$};
-%	  
-%	  \node[statein] (9) [right of=7] {$\beta$};
-%	  \node[state] (10) [above of=9] {};
-%	  \node[state] (11) [above right of=9] {};
-%	  \node[state] (12) [below right of=9] {};
-%	  
-%	  \path[dashed]
-%	    (2) edge node [left] {} (1)
-%	    (3) edge node [left] {} (1)
-%	    (10) edge node [left] {} (9)
-%	    (9) edge node [left] {} (11)
-%	    (9) edge node [left] {} (12);
-%		\path[]
-%	  	(1) edge node [left] {} (4)
-%	        edge node [left] {} (5)
-%	    (4) edge node [left] {} (6)
-%	        edge node [left] {} (7)
-%	    (5) edge node [left] {} (8)
-%	    (6) edge node [left] {} (9)
-%	    (7) edge node [left] {} (9)
-%	    (8) edge node [left] {} (9)
-%	    ;
-
-	  \node[statein] (1) {$\alpha$};
-	  \node[state] (2) [above left of=1] {};
-	  \node[state] (3) [above right of=1] {};
-	  \node[statein] (4) [below left of=1] {};
-	  \node[statein] (6) [below of=4] {$\gamma$};
-	  \node[statein] (7) [below right of=4] {$\gamma$};
-	  \node[statein] (5) [below right of=1] {};
-	  \node[statein] (8) [below of=5] {$\gamma$};
-	  \node[statein] (9) [below of=7] {$\beta$};
-	  \node[state] (10) [right of=9] {};
-	  \node[state] (11) [below left of=9] {};
-	  \node[state] (12) [below right of=9] {};
-
-	  \path[dashed]
-	    (2) edge node [left] {} (1)
-	    (3) edge node [left] {} (1)
-	    (10) edge node [left] {} (9)
-	    (9) edge node [left] {} (11)
-	    (9) edge node [left] {} (12);
-		\path[]
-	  	(1) edge node [left] {} (4)
-	        edge node [left] {} (5)
-	    (4) edge node [left] {} (6)
-	        edge node [left] {} (7)
-	    (5) edge node [left] {} (8)
-	    (6) edge node [left] {} (9)
-	    (7) edge node [left] {} (9)
-	    (8) edge node [left] {} (9)
-	    ;
-
-	\end{tikzpicture}
-  \caption{Example for a subgraph.}
-  \label{fig:subgraph}
-\end{figure}
-
-A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine.
-
-\section{Constraint determination}
-
-If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):
-
-\begin{equation} \label{eq:first}
-  \models \Box (\alpha \Rightarrow \Diamond \beta)
-\end{equation}
-
-\begin{equation} \label{eq:second}
-  \models \Box (\alpha \Rightarrow \Diamond (\bigvee_{\gamma} \gamma))
-\end{equation}
-
-\begin{equation} \label{eq:third}
-  \models \Box (\alpha \Rightarrow [\neg \beta \mathcal{U} \bigvee_{\gamma} \gamma])
-\end{equation}
-
-Additionally, for all states not contained in the subgraph (let's call them $\delta$-states, and the proposition $\delta$ stands for ``$\delta$ is current state'') the following formula is true:
-
-\begin{equation} \label{eq:fourth}
-  \models \Box (\alpha \Rightarrow [(\bigwedge_{\delta} \neg \delta) \mathcal{U} \beta])
-\end{equation}
-
-In case of $\alpha = \beta$ - i.e. when all paths starting from $\alpha$-state eventually return again (cycle) - even another constraint can be assumed to be true:
-
-\begin{equation} \label{eq:fifth}
-  \models \Box \Diamond \alpha
-\end{equation}
-
-These five constraints apply to each subgraph matching the conditions mentioned above.
-Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state.
-Formula~\ref{eq:second} ensures that there will be always at least one $\gamma$-state be visited after each visit of an $\alpha$-state.
-The quite similar but stricter formula~\ref{eq:third} demands additionally a visit of at least one $\gamma$-state before the $\beta$-state can be reached.
-Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states.
-The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.
-
-All formulas can be translated to the visual formalism presented in section~\ref{sec:operatorconstraints} and shown to the user.
-The appropriate LTL formulas for the constraints postulated in chapter~\ref{chap:goals} relate to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:
-
-\begin{equation}
-  \models \Box \Diamond \textnormal{'Polling'}
-\end{equation}
-
-\begin{equation}
-	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
-\end{equation}
-%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
-%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.

### 2012-10-21 thesis/tex/5_automatedconstraintgeneration.tex
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.

--- 2012-10-19 thesis/tex/5_design.tex
@@
-%\chapter{Design}
-%\label{chap:design}
-%
-%
-%In this chapter we introduce a first approach of a visual formalism and later state the final desing of the visual language.
-Furthermore the idea of automated constraint generation is presented in section~\ref{sec:automatedconstraintgeneration}.
-The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}.
-However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots.
-Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.
-%
-%%The challenge now is to find a similar intuitive and nonmathematical way of representing reasonable safety constraints for the healthbot domain and to figure out how to translate it to corresponding textual expressions.
-The new concept will be integrated into Robostudio, a design tool for statemachines, in order to enable the user to create safety constraints visually.
-A common LTL model checker will be added to it as well which fulfils the verification of the translated constraints on the designed statemachine.
-In case of verification failures the concerning constraints and a visual counter example can be presented.
-
-\chapter{Visual formalisms}
-\label{chap:visualformalisms}
-
-In this chapter we introduce two different approaches of visual formalisms and later state the final desing of the visual language.
-One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points.
-A first approach, so called template constraints and its pros and cons are described in section~\ref{sec:templateconstraints}.
-Subsequently an improved design is presented, which tends to a more abstract direction.
-
-\section{Template constraints}
-\label{sec:templateconstraints}
-
-%It is not a trivial issue to find a
-What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
-% TODO: wir haben gedacht, dass es einfach zu benutzen ist
-Leaded by this question, we developed a visual language based on templates.
-These templates make different propositions about states and their relations.
-The user can choose from several graphical constraint types and customize them by parameterizing state variables.
-So far there are two different kinds of template constraints:
-
-\begin{enumerate}
-	\item Whenever state (x) is active, state (y) must be visited before state (z) can be reached.
-
-	The idea behind this constraint is, that propositions about temporal dependencies can be made.
-Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled.
-Example of a constraint: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z).
-Going to bed without brushing teeth after meal should never happen.
-	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.
-
-	This constraint type enforces at least one particular event to happen after a spezific precondition.
-	Example of a constraint: Every time a person enters a supermarket (x) he will eventually decide to buy 
-	%a product or not to buy it.
-	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can't avoid both possibilities.
-	Nitpicking readers might note the person could also steel something without buying it, and thus not buy anything but leave the shop with full hands.
-Of course this is true; exactly in this case our constraint would signal an undesirable "program" behavior with its invalidity.
-\end{enumerate}
-
-Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like.
-Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot.
-Furthermore the state labels should sometimes be replaced by the screen displays of the regarding states.
-All parameters of the template constraints can be customized by just clicking on one robot suface.
-In the subsequently appearing popup window (figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.7]{templateconstraints}
-  \caption{The two template constraint types, parameterized with particular states.}
-  \label{fig:templateconstraints}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.7]{templateconstraint_dropdown}
-  \caption{Dropdown menu for spezifying template parameter.}
-  \label{fig:templateconstraint_dropdown}
-\end{figure}
-
-This visual language was presented to potential users for evaluation and feedbak.
-Though most of them named it a good concept, soon it turned out that there are some problems and disagreements.
-The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
-A try to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
-These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
-But unfortunately even the textual eplanations don't seem to be a satisfying solution.
-For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
-
-However, there is a second lack with the template constraints.
-Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can't be described.
-Not even the check of a state being visited never or at least once is possible.
-This lack of expressiveness appears to be too big and for this reason the approach of template constraints got dismissed.
-
-%idee: einfache verwendung
-%sehr spezialisiert auf healthcare anwendung
-
-\section{Operator constraints}
-\label{sec:operatorconstraints}
-
-With our second approach we try to gain more expressiveness and uniqueness for the visual language.
-For this we decided to decrease the abstraction level and design a visual formalism for linear temporal logic (LTL, see~section~\ref{sec:modelcheckingandltl}).
-Since the visual constraints can be directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness.
-If it is considered as easy to use, two birds get killed with one stone: Non-experts can use it as well as experts.
-
-% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv. der always operator könnte umgedreht sein.
-
-%The presented visual language is designed to express constraints for state machines.
-To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below.
-In addition the visual language should be capable of expressing constraints about future steps, both (h) any future state and (g) the next state, that (f) events should always happen, and that (i) a property must be true until some future event.
-Like in Del Bimbo et al.'s formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy.
-However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
-Each block represents either an operator or a proposition and has its specified semantics.
-Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.
-
-The visual language has support for eight operator types and one proposition as shown in figure~\ref{fig:operators}.
-
-\begin{description}
-	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
-	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
-	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
-Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
-	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
-	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state.
-Other types such as numeric or string equations may be added but are less important for our healthcare subject.
-	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
-	%\Box \Diamond \medcircle
-	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
-	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
-	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true.
-In addition eventually there has to be a future state with $\psi$ being true.
-\end{description}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.65]{table}
-  \caption{The visual language consists of eight unary or binary operator types and one state proposition, each of them having its own color flavor.
-Logical operators are arranged in a vertical row wereas time relevant statements are aligned horizontally.}
-  \label{fig:operators}
-\end{figure}
-
-To meet the requirements the visual language should be very easy to use, especially for non-experts.
-Furthermore it should provide an intuitive application which is ideally learnable within little time.
-So there are some design decisions this visual language must fulfill~\cite{moody-physics-of-notations}:
-
-\begin{itemize}
-	\item Each operator has a different significant visual appearance; in this case we have used color. This kind of classification enables a faster subconscious identification.
-	\item The layout of constraints (nested operators) has two dimensions with a certain meaning.
-Figure~\ref{fig:directions} demonstrates the ``logical'' vertical read direction as well as the time releveant horizontal line.
-	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear unnecessarily complex.
-These operators can be merged (see~figure~\ref{fig:nested_and}).
-\end{itemize}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.65]{and_simplify} %width=0.7\linewidth
-  \caption{In order to simplify the appearance of constraints certain operators might be merged.
-Merged nested \emph{AND} operators (center) are less complex as default ones (left).
-However nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
-  \label{fig:nested_and}
-\end{figure}
-
-Also an editor for this visual language can contribute to usability and clarity. Some requirements are as follows:
-
-\begin{itemize}	
-	\item All editing is performed only by simple drag and drop.
-Operators and propositions can be moved and dropped onto other operators.
-A tool bar provides all operator and proposition types to be instantiated and there is a trash can for deleting existing operators.
-	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer.
-All higher operators become bleached out so that the hovered operator and its children appear highlighted.
-It's intended to be similar to line coloring when reading a digital text.
-	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
-	\item There are different statuses which inform about validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed.
-Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign (see~figure~\ref{fig:results}).
-	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances.
-Having no additional buttons reduces complexity and keeps it simple.
-\end{itemize}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.65]{directions}
-  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
-  \label{fig:directions}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.65]{results}
-  \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
-  \label{fig:results}
-\end{figure}
-
-%\section{Automated constraint generation}
-%\label{sec:automatedconstraintgeneration}
 \chapter{Automated constraint generation}
@@
 
-\section{Sub graph definition}
+\section{Subgraph definition}
+\label{sec:subgraphdefinition}
 

### 2012-10-21 thesis/tex/7_testscenarioandevaluation.tex
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.

--- 2012-10-19 thesis/tex/7_prototypeandevaluation.tex
@@
 
-In order to examine the performance of sub graph finding and constraint generation, we need to have a second look at the implementation details again.
-The sub graph finding algorithm is realized as a depth-first search.
+In order to examine the performance of subgraph finding and constraint generation, we need to have a second look at the implementation details again.
+The subgraph finding algorithm is realized as a depth-first search.
 Indeed all healthcare programs are infinite and contain cycles and a depth-first walk-through would never end, but the algorithm doesn't go deeper at already visited states.
@@
 
-In order to find a sub graph with our presented algorithm the depth-first search has to be startet on a start state of a sub graph.
+In order to find a subgraph with our presented algorithm the depth-first search has to be startet on a start state of a subgraph.
 Since we don't know in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions.
@@
 \begin{equation}
-O(\textnormal{``sub graph finding''})=O(|V|)*O(|V|)=O(|V|^2)
+O(\textnormal{``subgraph finding''})=O(|V|)*O(|V|)=O(|V|^2)
 \end{equation}
@@
 However, sholud this algorithm regardless its tailoring to the healthcare domain be used for other huge programs, it keeps a lot of potential to be improved in efficiency.
-With a sophisticated approach it might be enough to execute just one depth-first search for finding all sub graphs at once what would lead to a linear execution time.
+With a sophisticated approach it might be enough to execute just one depth-first search for finding all subgraphs at once what would lead to a linear execution time.
 % wenn noch schreiben will, dann: linear ist cool, weil das tool eh die gefundenen constraints gleich anzeigt und nicht erst am ende alles geballt bringt.

### 2012-10-21 thesis/tex/0_abstract.tex
Sections 6.4, 6.4.1 and 6.4.2 (Integration into Robostudio, Problems, Solutions) are now finished.

@@
 Nowadays robotic applications are no longer confined to only industrial use but increasingly assigned to jobs within human workspaces.
-Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behavior in order to match individual requirements.
-However both the human robot collaboration as well as the behavior definition by non experts may be a risk to humans within the robots' environment.
-Thus special thinking is needed about how to cope with safety requirements for robot behavior specification by consumers.
+Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behaviour in order to match individual requirements.
+However both the human robot collaboration as well as the behaviour definition by non experts may be a risk to humans within the robots' environment.
+Thus special thinking is needed about how to cope with safety requirements for robot behaviour specification by consumers.
 

### 2012-10-21 thesis/tex/1_introduction.tex
Sections 6.4, 6.4.1 and 6.4.2 (Integration into Robostudio, Problems, Solutions) are now finished.

--- 2012-10-19 thesis/tex/2_introduction.tex
@@
 
-Safety critical robot applications require extensive testing or formal verification in order to achieve adequate safe and predictable behavior.
+Safety critical robot applications require extensive testing or formal verification in order to achieve adequate safe and predictable behaviour.
 Even if a program does not control actuators itself and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions involving several actors.
-Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University of Auckland, New Zealand \cite{jayawardena12:_desig}.
-All interaction is guided by showing textual instructions
+Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University
 \begin{wrapfigure}{r}{0.58\textwidth}
@@
 \end{wrapfigure}
-on a display, and by speech generation.
+of Auckland, New Zealand \cite{jayawardena12:_desig}.
+All interaction is guided by showing textual instructions on a display, and by speech generation.
 An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
@@
 Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
-In chapter~\ref{chap:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine definitions that describe robots' behaviors.
+In chapter~\ref{chap:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine definitions that describe robots' behaviours.
 

### 2012-10-21 thesis/tex/2_background.tex
Sections 6.4, 6.4.1 and 6.4.2 (Integration into Robostudio, Problems, Solutions) are now finished.

@@
 There is a program interpreter running on the robots which can load and execute program files. Screen dialogs get generated and displayed on the touch screen.
-%This makes the program behavior replaceable and maintenance can be done easily.
-The interpreter processes programs written in Robot Behavior Description Language (RBDL), a particular DSL based on XML, which has been developed particularly for the healthcare robotic applications.
-It allows defining state machine based program behavior and creating UI elements for screen visualizations.
+%This makes the program behaviour replaceable and maintenance can be done easily.
+The interpreter processes programs written in Robot Behaviour Description Language (RBDL), a particular DSL based on XML, which has been developed particularly for the healthcare robotic applications.
+It allows defining state machine based program behaviour and creating UI elements for screen visualizations.
 An example implementation of one state in RBDL is shown in listing~\ref{lst:RBDL}.
@@
 
-\singlespacing
 \lstset{
@@
 
-\onehalfspacing
-
 For each state of the program, backgroundactions can be defined.
@@
 Once displayed, buttons can be clicked and trigger events.
-The behavior of events can be mapped in the expectedevents tag.
-It lists all events which are processable by the current state and defines the correspondend behavior such as triggering transitions to other states.
+The behaviour of events can be mapped in the expectedevents tag.
+It lists all events which are processable by the current state and defines the correspondend behaviour such as triggering transitions to other states.
 
-For this work the most important thing about the RBDL to remember is, that it can define state machine behavior with states, transitions and events.
+For this work the most important thing about the RBDL to remember is, that it can define state machine behaviour with states, transitions and events.
 This fact was basement for later decisions regarding used safety mechanisms.
@@
 
-In the beginning all programming of the robot behavior was done by directly editing the XML code mentioned in the previous section by using a text editor.
+In the beginning all programming of the robot behaviour was done by directly editing the XML code mentioned in the previous section by using a text editor.
 But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise.
-For this reason Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behavior on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behavior on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
+For this reason Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
 This file can be downloaded to the robot and executed by the interpreter.
-Figure~\ref{fig:robostudio_proceeding} illustrates what the steps of robot behavior development are.
+Figure~\ref{fig:robostudio_proceeding} illustrates what the steps of robot behaviour development are.
 
@@
   \includegraphics[width=\linewidth]{robostudio_proceeding} 
-  \caption{How Robostudio is used for robot behavior development.}
+  \caption{How Robostudio is used for robot behaviour development.}
   \label{fig:robostudio_proceeding}
@@
 Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application.
-In several windows and views there are visual tools for editing and understanding robot behavior as well as screen dialog design.
+In several windows and views there are visual tools for editing and understanding robot behaviour as well as screen dialog design.
 As depicted in figure~\ref{fig:robostudio_vpe} the editor has got a State Navigator Window (1) in the top left corner where all existing states are listed.
@@
 Aiming for supporting the developer with helpful graphical tools the State Transition Visualization Window (6) gives an overview about all connections towards and from the selected state.
-It helps to quickly get an understanding of the local program behavior.
+It helps to quickly get an understanding of the local program behaviour.
 
@@
 
-\section{Behavior checking}
+\section{Behaviour checking}
 TODO: Begriffserklärung: Funktionale Sicherheit
@@
 Even though this approach touches the subject of healthcare it barely matches our problem.
-HomeTL focuses more on monitoring temporal boundaries of a patient's behavior than on ensuring functional safety of an implemented program.
+HomeTL focuses more on monitoring temporal boundaries of a patient's behaviour than on ensuring functional safety of an implemented program.
 It does depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.

### 2012-10-21 thesis/tex/6_theltlcreatorprototype.tex
Sections 6.4, 6.4.1 and 6.4.2 (Integration into Robostudio, Problems, Solutions) are now finished.

--- 2012-10-20 thesis/tex/6_implementation.tex
@@
 An example of this input language is given in listing~\ref{lst:nusmv}.
-\singlespacing
+
 \lstset{
@@
 \subsection{Ease of integration}
+\label{sec:easeofintegration}
 
@@
 
-kurz erläutern, warum das gemaacht werden soll
+As statet in the previous section the LTLCreator provides an API for integration into other java based applications.
+Robostudio is such an application which shall be used for evaluating the visual language as well as the automated constraint generation feature and is also supposed to provide all this functionality for development of service robot behaviour in future.
 
-Robostudio uses NetBeans as a platform, and all tools and features are organized in windows.
+Robostudio uses NetBeans~\cite{netbeans} as a platform and organizes all tools and features in windows.
 Thus also LTLCreator has to be put into a window and added to the Robostudio platform.
-Since the defining of safety constraints is an important part not only after but also during the process of developing service robot behavior, the LTLCreator window as a vital tool shall be placed in the center of the platform where currently also the UI Component Layout Editor Window is.
-As already proposed in the previous section, all changes to the state machine program made within robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
+This is done by placing it into a NetBeans module which is added to Robostudio's modules.
+This module imports the as jar archived LTLCreator tool and further consists of only two classes: \emph{LTLCreatorTopComponent} and \emph{XmlDataToLTLEditorDataWrapper}.
+The former implements the \emph{TopComponent} provided by NetBeans which can be used for displaying UI elements.
+In order to visualize the LTLCreator component it is directly placed on the \emph{LTLCreatorTopComponent}.
+It retreives changes of the model and passes them to the \emph{LTLCreator} component.
+Since the defining of safety constraints is an important issue not only after but also during the process of developing service robot behaviour, the LTLCreator window as a vital tool has been positioned in the center of the platform where also the \emph{UI Component Layout Editor Window} is.
+The \emph{XmlDataToLTLEditorDataWrapper} forms the bridge between Robostudio's own model and the state machine model used by \emph{LTLCreator}.
+All complex data coming from Robostudio has to be detached from unimportant graphical information and transformed to an elementary state machine representation before it can be processed by \emph{LTLCreator}.
 
-allein benutzbar, oder ueber robostudio, automatische synchronisierung mit model
-
-%Since Robostudio uses NetBeans as a platform, the LTLCreator has to be integrated as a NetBeans component.
-Die Datenanbindung zum Austausch des modells (Zustandsmaschine) und der Constraints erfolgt hierbei durch öffentliche
+%TODO: load and save
 
@@
 
-(auto-revalidation, model-errors, ...?)
-nicht EINE zentrale stelle für modeländerungen => zer\-pflück\-te programmierung, wenn man alle änderungen haben will
+As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
+This demands a concept of change notification management which is responsible for receiving all chages and making this information available as a service.
+Thus we tried to extend all parts of the code where model changes are made with short code snippets.
+However this is not as easy to realise in Robostudio since some of the relevant code was generated or imported as external jars what forbids code manipulation.
+
+A second challenge is the current data handling policy in Robostudio.
+Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than syntax checks.
+Thus syntax errors are not detected or even avoided by static analysis ore input restrictions.
+In fact, such features haven't been requirements for Robostudio.
+During integration of \emph{LTLCreator} into Robostudio we encountered problems like transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
+Well, this characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistend data retrieved from Robostudio.
 
@@
 
-(pre-checker)
+To guarantee any all model change causes the constraints to revalidate, all relevant code parts have to report on this.
+But modification of the code hasn't been possible in all cases.
+Whereas all state changes are announced, transition changes stay unnoticed.
+This decision is straight for the prototype, but for later the developers of Robostudio announced an adapted architecture to solve this problem in the next version.
 
-statische analyse
+The problem with inconsistend data could be solved by plausibility and syntax checks during the transformation process of \emph{XmlDataToLTLEditorDataWrapper}.
+After successful transformation the state machine is passed to the \emph{LTLCreator} component.
+Otherwise no validation result is displayed for constraints and an appearing dialog lists all identified syntax errors.
+In matter of safety and reliability a restrictive approach which gives no result at all seemed more applicable than an attempt to interprete the inconsistent data.

### 2012-10-21 thesis/tex/2_background.tex
Translations

@@
 
-Das Thema dieser Arbeit hat sich aus der Situation heraus entwickelt, dass die healthbots der UoA mit sicherheitsmechanismen ausgestattet werden sollen.
-Die Motivation hierfür, die Roboter mitsamt ihrer Anwendungen und deren technischer Hintergrund werden in Section 1 beschrieben.
-Auf das VPE, dass die Programmierung dieser Roboter ermöglicht, wird in section 2 eingegangen.
-Zum Verständnis der sicherstellungskonzepte für sicherheit, auf die diese Arbeit aufbaut, gibt Section 3 eine kurze Einführung in die Welt der funktionalen Sicherheit.
-In Section 4 werden andere Arbeiten und Programme untersucht, die sich ebenfalls mit Sicherheitüberprüfung beschäftigen.
+The subject of this work evolved from the situation, that the healthcare robotic applications of the University of Auckland should be equipped with mechanisms for ensuring safety.
+The motivation for it is described in section~\ref{sec:healthbotapplication} as well as the robots, the kind of applications running on them, and their technical background.
+
+The visual programming environment used for the programming of these robots is presented in section~\ref{sec:robostudio}.
+For the understanding of safety concepts which this work is based on, a short introduction into the world of functional safety is given in section~\ref{sec:behaviourchecking}.
+In section~\ref{sec:relatedwork} other works and programs related to this work's topic are examined which also deal with safety aspects.
 
@@
 \section{Robostudio}
+\label{sec:robostudio}
 
@@
 \section{Behaviour checking}
+\label{sec:behaviourchecking}
 TODO: Begriffserklärung: Funktionale Sicherheit
@@
 \section{Related work}
+\label{sec:relatedwork}
 

### 2012-10-21 thesis/tex/7_testscenarioandevaluation.tex
Translations

@@
 
-%This expression can now be translated to the visual form as we just have to break down the formula from the outside to the inside.
-A surrounding \emph{GLOBALLY} encapsulates an \emph{IF} block since it is the substitution for \emph{IMPLIES}.
-Its first operator is a proposition \emph{'Not taken yet'} whereas the second parameter has to be a \emph{FOLLOWS} block containing a disjunction of two propositions \emph{'Well done!'} and \emph{'Staff notified'}.
-The resulting visual constraint is shown in figure~\ref{fig:sampleconstraint}.
-
 \begin{figure}[htbp]
@@
 
-Ausdruckssärke gleich zu LTL, das heißt, dass lediglich Möglichkeiten nicht ausgedrückt werden können.
-Zudem gilt eine weitere Einschränkung, dass als bedingungen bisher nur Aussagen der Form ``ist in Zustand x'' verwendet werden können.
-Da die Validierung der healtbot anwendungen derzeit ausschließlich diesen typ braucht, sind andere Bedingungen wie Stringvergleich oder numerische Gleichheit bisher mit der visuellen Sprache nicht möglich.
-In Figure~\ref{fig:example_propositions} ist aufgezeigt, wie eine Erweiterung des LTLCreators um weitere Bedingungstypen aussehen könnte.
+Due to its basing on LTL, the visual language's expressivenes is equivalent to LTL's one of course. More precisely this means any logic formula can be expresssed except potentiality.
+Furthermore there is one more limitation in the current visual language with the variety of available proposition operators.
+The visual language was originally designed for validating healthcare applications where just one kind of proposition is needed: ``currently state x is active''.
+Thus other propositions such as equations, greater than etc.
+are not implemented yet, but can be easily extended.
+Figure~\ref{fig:example_propositions} depicts how an extention of \emph{LTLCreator} with additional proposition operators could be realised.
 

### 2012-10-24 thesis/tex/2_background.tex
* Section 7.3 is now translated to english * Minor changes in chapter 2 * Changes to chapter 8

@@
 
-In einigen Ländern, darunter Neuseeland, Japan und Deutschland, ist das Durchschnittsalter der Bevölkerung stets ansteigend.
-Dank hervorragenden Errungenschaften in der Medizin sowie höheren Lebensstandards lebt der Mensch in Durchschnitt immer länger.
-Das hat zur Folge, dass auch die Nachfrage an Plätzen in Altersheimen oder nach Dienstleistungen sozialer Einrichtungen wächst.
-Doch ist in diesen Ländern ein enormer Mangel an professionellen Hilfskräften zu verzeichnen.
-Das mag mitunter daran liegen, dass in der Regel für soziale Arbeit in diesem Umfeld ein niedrigerer Lohn bezahlt wird als vergleichsweise für technische Berufe.
-Doch eine Anhebung der Gehälter ist nicht ohne weiteres möglich, denn auch zu den derzeitigen Preisen können sich viele ältere Leute keine persönliche Pflegekraft leisten.
+%\begin{wrapfigure}{r}{0.38\textwidth}
+%  \centering
+%  \includegraphics[width=0.34\textwidth]{populationpyramide}
+%  \caption{Age structure.}%Population pyramide of germany.}
+%  \label{fig:populationpyramide}
+%\end{wrapfigure}
+In some countries, among New Zealand and Germany, the average age of population is constantly increasing.
+Owing to excellent achievements in medicine as well as higher living standards, humans can enjoy longer lifes.
+Also decreasing birth rates contribute to an upward drift of the dominating age range.
+As a result, the industrial sectors of professional health care and social institutions are required more than ever.
+However, a strong deficit of professionals is denoted in the healthcare branch and availability is far behind demand.
+One reason for this might be below average low salaries for employees in the healthcare sector.
+But, just raising the salary isn't leading to the desired results since the majority of the older people can't even afford a nurse to today's prices.
 
@@
 
-TODO: Vlt herausstellen, dass eine Schwierigkeit darin bestand, Verhaltensmodellierung und graphische Gestaltung unter einen Hut zu bringen (dazu Bild aus datta paper?).
+%One challenge 
+%TODO: Vlt herausstellen, dass eine Schwierigkeit darin bestand, Verhaltensmodellierung und graphische Gestaltung unter einen Hut zu bringen (dazu Bild aus datta paper?).
 

### 2012-10-24 thesis/tex/7_testscenarioandevaluation.tex
* Section 7.3 is now translated to english * Minor changes in chapter 2 * Changes to chapter 8

@@
 
-eher schreiben: wurde einigen Leuten vorgestellt. Generell blabla... Darunter auch alina kracker...
+Some parts of the healthcare robot's behaviour are appropriate to be specified by non-programmers, and that's the purpose of the presented visual language as well.
+Different tools and environments, among Robostudio and LTLCreator, try to provide a particular interface suitable for such people.
+Now, it might be interesting to find out about user's experiences and how they get along when using these tools.
 
-Alina Kracker, 21 Jahre alt aus Friedberg und Studentin der Erziehungswissenschaften, arbeitet neben dem Studium als Pflegehilfskraft in einer Kurzzteitplfege.
-Dort betreut sie ältere Damen und Herren, die dort kurz- bis mittelfristig untergebracht sind.
-In dieser Einrichtung wohnen Personen mit unterschiedlichsten Krankheiten und Bedürfnissen, von zu beobachtenden aber gesunden Menschen bis hin zu Alzeimerpatienten.
+The LTLCreator has been presented to several people, technical engineers as well as ordinary persons.
+The overall feedback was good and the idea and realisation of the visual language was well-received.
+One of the laypersons who have been introduced to LTLCreator was Alina Kracker.
+The 21 yeras old german student of educational science is also working for the elderly in a care facility since more than five years.
+In this social institution Ms Kracker takes care of persons having different disease patterns and necessities.
+This spectrum varies from physical disabilities over mental problems such as Alzheimer's desease to being bedridden.
+Her professional care background makes her feedback about usability and comprehensibility of the visual concept especially interesting since its application and testing was done in the healthcare domain.
 
-Es wurde getestet, ob Kracker ohne entsprechende Ausbildung Constraints zu dem Beispiel des medication reminder formulieren kann.
-Zum einen möchte die Universität of Auckland die Programmierung solcher Anwendungen auch nicht-professionellen ermöglichen, und da wäre die Erfahrung einer echten Pflegekraft mit dieser Aufgabe wohl sehr interessant.
-Zum anderen kann so aber auch festgestellt werden, wie nicht-Softwareentwickler mit der visuellen Sprache und dem Konzept zur Constrainterstellung zurechtkommen.
+Ms Kracker was introduced to the example of medication remider, and the program behaviour was described with the help of figure~\ref{fig:medicationreminder}.
+Afterwards she got familiarized with surface and functionality of LTLCreator.
+Now, she was asked to formulate the two constraints which we postulated in the requirements, just by using the available means of the editor placed at her disposal.
+The first constraint, which ensures the state ``Polling'' always eventually to be visited again, was constructed quickly and accurately.
+The second constraint specifies, that every visit of state ``Not taken yet'' will eventually result in a visit of either state ``Well done!'' or ``Staff notified''.
+It turned out that this constraint held some difficulties for her, the \emph{FUTURE} operator was not considered initially.
+But it was also finisehd after a few consideration breaks.
 
-Kracker wurde in das Beispiel der medication reminder applikation eingeführt und das Programmverhalten anhand der Zustandsmaschine aus figure 123 erklärt.
-Anschließend wurde sie mit der Oberfläche und Funktionsweise des LTLCreator tools vertraut gemacht.
+%indeed
+%surprisingly
+%hm was given as a reason.
+%Probably due to 
+%Due to textnähe und einfachen lesbarkeit der vl
+%da der text nah ist und auch für ... vferständlich ist
+%not familiar with
+%As a first step,
 
-Ihr wurde nun die Aufgabe gegeben, mit den im Editor zur Verfügung stehenden Mitteln die zwei Bedingungen zu formulieren, die uns in diesem Kapitel bereits begleitet haben.
-Zum einen sollen Medikamenterinnerungen nie vergessen werden, also Zustand ``Polling'' muss regelmäßig besucht werden.
-Zum anderen muss sichergestellt sein, dass bei noch nicht vorgenommener Medikamenteinnahme das Medikament letztendlich eingenommen wird oder das Personal benachrichtigt wird.
-Also bei Besuch von ``Not taken yet'' muss entweder ``Well done!'' oder ``Staff notified'' besucht werden.
+Ms Kracker didn't know in beforehand what state machines are and never came in contact with LTL or model checking in general.
+But she dealt with this example and the visual language surprisingly well, and she's convinced that she could use it after some exercise.
+As reason for the success she pointed out the easy readability of the visual language and its close relationship to spoken language.
+Furthermore the intuitive drag and drop allowed faster editing.
+Her examination on the visual language gives us evidence that we succeeded with our approach of creating an easy to use concept for defining LTL formulas.
+The visual language we developed enables even non-professionals to use the usually quite complex concept of model checking.
 
-Da Kracker mit Zustandsmaschinendarstellungen nicht vertraut ist, wurden ihr die Constraintbedeutungen wie eben beschrieben anhand der Zustandsnamen erläutert.
-Bei der Erstellung der Constraints war sie allerdings anfänglich auf sich alleine gestellt.
-Die erste Bedingung erstellte sie zügig und richtig, und es schien auch keine besondere Herausforderung gewesen zu sein.
-Auch der zweite Constraint wurde
-Bei der Frage nach der Anwendung von FUTURE hakte es dann allerdings etwas, was auf die zeitlichen (...) zurückzuführen war.
-Die Aufforderung, die Bedingung als deutschen Satz zu formulieren, half ihr wieder auf die Sprünge.
+%Für eine Person, die von modelchecking oder LTL noch nie etwas gehört hat, kam sie mit kleinen Hilfen gut mit der visuellen Sprache zurecht und könnte sie so mit etwas Übung auch verwenden.
 
-Für eine Person, die von modelchecking oder LTL noch nie etwas gehört hat, kam sie mit kleinen Hilfen gut mit der visuellen Sprache zurecht und könnte sie so mit etwas Übung auch verwenden.
+%Die Nähe zwischen LTL und deutschem Satz ist gut, und dadurch kann man schnell konstraints zusammen klicken (besser als mit fremden mathematischen symbolen).
 
-Die Nähe zwischen LTL und deutschem Satz ist gut, und dadurch kann man schnell konstraints zusammen klicken (besser als mit fremden mathematischen symbolen).
-
-Auf die Frage, ob sie solche Roboteranwendungen wie der Medicationreminder in der Kurzzeitpflege für sinnvoll erachten würde, antwortet Kracker: ``Für geistig fitte aber dennoch vergessliche und tütelige Leute kann ich mir diese Roboterapplikation super vorstellen.
+To the question, whether she would recommend such robotic services like the medication reminder application for use in the care facility she's working in, Ms Kracker replied: ``I can envision such robot applications beeing an excellent assistance for the elderly who are mentally fit but maybe a little doolally.
+However, in our facilities most of the patients aren't capable enough anymore of making decisions based on what the robots ask.
+In fact, due to frequent changes of the patients in our facilities it is a big issue for the care givers not to forget time-dependend medication intakes of their patients.
+Thus such a reminding robot application would constitute an excellent support for us staff!''.
+%Auf die Frage, ob sie solche Roboteranwendungen wie der Medicationreminder in der Kurzzeitpflege für sinnvoll erachten würde, antwortet Kracker:
+%``Für geistig fitte aber dennoch vergessliche und tütelige Leute kann ich mir diese Roboterapplikation super vorstellen.
 Ein Großteil unserer Patienten ist jedoch nicht in der Lage Entscheidungen, die der Roboter abfrägt, wahrheitsgemäß zu treffen.

### 2012-10-24 thesis/tex/2_background.tex
have written section 2.3

@@
 The two newest robots IrobiQ and Cafero are shown in figure~\ref{fig:irobiq_cafero}.
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=0.6\linewidth]{irobiq_cafero} 
-%  \caption{IrobiQ and Cafero, two robots of the healtcare project.}
-%  \label{fig:irobiq_cafero}
-%\end{figure}
 These mobile 
@@
 	morekeywords={%no, type_, senderid, receiverid, name_, vartype, label, width, height, x, y, textsize, preconditions
-	}% list your attributes here
+	}
 }
 
-%\lstset{language=XML}
 \begin{lstlisting}[float = htbp, language=XML, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a state defined by RBDL~\cite{robostudio}., label=lst:RBDL]
@@
 
-%require refactoring and static analysis tools.
-
 Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application.
@@
 
-%One challenge 
-%TODO: Vlt herausstellen, dass eine Schwierigkeit darin bestand, Verhaltensmodellierung und graphische Gestaltung unter einen Hut zu bringen (dazu Bild aus datta paper?).
+%TODO: Vlt herausstellen, dass eine Schwierigkeit darin bestand, Verhaltensmodellierung und graphische Gestaltung unter einen Hut zu bringen (dazu Bild aus datta paper?). One challenge...
 
@@
 \label{sec:behaviourchecking}
-TODO: Begriffserklärung: Funktionale Sicherheit
+
+The use of software is versatile from text editing programs running on personal computers, over control software for high-precision applications in aeronautics, right up to trigger mechanisms for deathly weapons.
+For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also human lifes.
+In order to eliminate software malfunction there are a lot of mathematical concepts for verifying software behaviour.
+One of them is model checking wich allows checking for functional safety on state machine programs.
+Since we deal with state machines in the healthbot applications mentioned in section~\ref{sec:healthbotapplication}, this concept of behaviour checking is applicable and important for our kind of programs.
+At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an intorduction into model checking in section~\ref{sec:modelcheckingandltl}.
 
 \subsection{State machines}
+\label{sec:statemachines}
 
@@
 
-NuSMV erwaehnen, erklaeren warum nicht CTL? ltl passt fuer healthcare besser
+%NuSMV erwaehnen, erklaeren warum nicht CTL? ltl passt fuer healthcare besser
 

### 2012-10-27 thesis/tex/0_abstract.tex
correction, v.1.

@@
 Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behaviour in order to match individual requirements.
-However both the human robot collaboration as well as the behaviour definition by non experts may be a risk to humans within the robots' environment.
-Thus special thinking is needed about how to cope with safety requirements for robot behaviour specification by consumers.
+However, both human robot collaboration as well as the robot's behaviour definition by non experts may be a risk to humans within the robots' environment.
+Thus, special thinking is needed about how to cope with safety requirements for robot behaviour specification by consumers.
 
-In this paper we present a visual language for the defining of safety constraints for state machine definition of robot behaviour.
-Our approach addresses mainly non-safety-experts and our abstraction from a mathematical temporal logic expression to a more intuitive visual representation is intended to enable a wider range of software developers to do model checking.
-Furthermore we propose a new concept of development support by providing automated constraint generation functionality.
+In this work a visual language for the defining of safety constraints for state machine definition of robot behaviour is presented.
+The approach addresses mainly non-safety-experts and introduces abstraction from a mathematical temporal logic expression to a more intuitive visual representation to enable a wider range of software developers to do model checking.
+Furthermore, a new concept of development support by providing automated constraint generation functionality is proposed.
 It aims to help developers in defining reasonable constraints and finally to increase the probability of finding bugs.

### 2012-10-27 thesis/tex/1_introduction.tex
correction, v.1.

@@
 Even if a program does not control actuators itself and thus cannot directly cause any damage, its output might still be a trigger for dangerous actions involving several actors.
-Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, such as the healthcare robotics project at the University
+Imagine a medication reminder application running on a service robot which guides a person through the process of medication intake, as investigated in the healthcare robotics project at the
 \begin{wrapfigure}{r}{0.58\textwidth}
@@
 \end{wrapfigure}
-of Auckland, New Zealand \cite{jayawardena12:_desig}.
+University of Auckland, New Zealand \cite{jayawardena12:_desig}.
 All interaction is guided by showing textual instructions on a display, and by speech generation.
@@
 With service robots coexisting with people and acting within their work\-spa\-ces, coping with reliability and safety issues is essential for trust and acceptance of these robots.
-In some cases safety certifications based on international standards for functional safety, such as ISO 13482 and IEC 61508, might be even required before robots get released for use or obtain concession for sale.
+In some cases safety certifications based on international standards for functional safety might be even required before robots get released for use or obtain concession for sale.
+One such international standard is IEC 61508~\cite{iec61508}, a basic functional safety standard targeting all kinds of industry.
+ISO 13482~\cite{iso13482, iso13482-2} is still under development and is intended to be a harmonised European standard for safety requirements for robots in personal care applications.
 However, considering functional safety while developing service robots is both complex and difficult; maybe one reason for the absense of safety checks and warranties in the current healthbot versions at the University of Auckland.
@@
 %state machine based processes and a graphical tool for the designing of human robot interaction.
-Generally these people don't have profound knowledge of the complex mathematical syntax used for formal temporal logic such as computation tree logic (CTL) or linear temporal logic (LTL).
-Nevertheless certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthcare robots.
-In case of the medication reminder mentioned above it might be important to ensure notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medications which have already been taken.
-So a mechanism is needed for the specification of such safety constraints; one that does not require special or expert knowledge and thus is easy to use for all kind of programmers.
-A visual language intended to fill this gap is presented in section~\ref{sec:operatorconstraints}.
+Generally these people don't have profound knowledge of the complex mathematical syntax used for formal methods which allow software behaviour verification.
+%temporal logic such as computation tree logic (CTL) or linear temporal logic (LTL).
+Nevertheless, certain safety guarantees must somehow be delivered in order to gain users' trust and to prevent any harm or injuries by the healthcare robots.
+In case of the medication reminder mentioned above it might be important to ensure that notifications are sent to staff members whenever medication is not taken by the patient, or to avoid duplicate reminders for medication which has already been taken.
+A mechanism like this is needed for the specification of such safety constraints; one that does not require special or expert knowledge and is thus easy to use for all kind of programmers.
+A visual language intended to fill this gap is presented in chapter~\ref{chap:visualformalismsforsafetyconstraints}.
 
-Unfortunately the supply of an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer.
+Unfortunately, supplying an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer.
 Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
@@
 The latter would mean there are errors in the program which lead to undesired constraints.
-Furthermore once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
-In chapter~\ref{chap:automatedconstraintgeneration} we introduce a heuristic for finding reasonable safety constraints based on state machine definitions that describe robots' behaviours.
+Furthermore, once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
+Chapter~\ref{chap:automatedgenerationofsafetyconstraints} introduces a heuristics for finding reasonable safety constraints based on state machine definitions that describe robots' behaviours.
 
-Both ideas are to be integrated into Robostudio in order to simplify respectively facilitate the use of safety mechanisms for application development for service robots.
-Thereby a desired safety certification for the healthcare service robots can move closer one step.
+Both ideas have been integrated into Robostudio in order to simplify respectively facilitate the use of safety mechanisms for application development for service robots.
+Thereby the desired safety certification for the healthcare service robots can move closer one step.
 
 This work starts with background information about the healthcare domain, short explanations of common safety mechanisms and insights into related work in chapter~\ref{chap:background}.
-After the goals are mentioned in chapter~\ref{chap:goals}, the visual formalisms and the automated constraint generation are introduced in chapters~\ref{chap:visualformalisms} and~\ref{chap:automatedconstraintgeneration}.
-After a prototype implementation is demonstrated in chapter~\ref{chap:theltlcreatorprototype}, a test scenario is described in chapter~\ref{chap:testscenarioandevaluation} where the presented tools get evaluated.
-Chapter~\ref{chap:conclusionandfuturework} gives a short conclusion and an outlook how the work presented in this paper could be continued.
+After the goals have been mentioned in chapter~\ref{chap:goals}, the visual formalisms and the automated constraint generation are introduced in chapters~\ref{chap:visualformalismsforsafetyconstraints} and~\ref{chap:automatedgenerationofsafetyconstraints}.
+A prototype implementation is demonstrated in chapter~\ref{chap:theltlcreatorprototype}, and a test scenario is described in chapter~\ref{chap:testscenarioandevaluation} where the presented tools get evaluated.
+Chapter~\ref{chap:conclusionandfuturework} gives a short conclusion and an outlook how the presented work could be continued.
 

### 2012-10-27 thesis/tex/2_background.tex
correction, v.1.

@@
 
-The subject of this work evolved from the situation, that the healthcare robotic applications of the University of Auckland should be equipped with mechanisms for ensuring safety.
-The motivation for it is described in section~\ref{sec:healthbotapplication} as well as the robots, the kind of applications running on them, and their technical background.
+The topic of this work evolved from the situation that the healthcare robotic applications of the University of Auckland should be equipped with mechanisms for ensuring safety.
+The motivation for it is described in section~\ref{sec:healthbotapplication} as well as the robots used, the kind of applications running on them, and their technical background.
 
 The visual programming environment used for the programming of these robots is presented in section~\ref{sec:robostudio}.
-For the understanding of safety concepts which this work is based on, a short introduction into the world of functional safety is given in section~\ref{sec:behaviourchecking}.
-In section~\ref{sec:relatedwork} other works and programs related to this work's topic are examined which also deal with safety aspects.
+For the understanding of safety concepts which this work is based on, a short introduction about functional safety is given in section~\ref{sec:behaviourchecking}.
+In section~\ref{sec:relatedwork} other work and programs related to this work are examined.
 
-\section{Healhbot application}
+\section{Healthbot application}
 \label{sec:healthbotapplication}
@@
 
-Aiming for fillig the constantly increasing gap in professional personnel for healtcare of the elderly and bypassing die financial burden, service robots are supposed to more and more undertake the task of care.
+Aiming for fillig the constantly increasing gap in professional personnel for healthcare of the elderly and bypassing the financial burden, service robots are supposed to more and more undertake the task of care.
 There are lots of possible tasks a robot could do such as taking the blood pressure, guiding medication intake, detecting falls or just entertaining with music, video or internet.
-This challenge as a goal, the healtcare robotics team at the University of Auckland researches this topic and builds service robots for supporting the elderlies in their daily life.
+This challenge as a goal, the healthcare robotics team at the University of Auckland researches this topic and builds service robots for supporting the elderly in their daily life.
 
@@
 \end{wrapfigure}
-robots have got a touch display for presenting screens and receiving user input. Furthermore they are equipped with additional external tools such as the blood pressure measurement device.
-Applications running on the robot use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedbak.
+robots are equipped with a touch display for presenting screens and receiving user input. Furthermore, they are equipped with additional external tools such as the blood pressure measurement device.
+Applications running on the robot use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedback.
 User inputs or response from devices cause dialogs to change.
-Figure~\ref{fig:screenflow_example} presents a little example of such a screen-flow based application, which uses response from a camera and user input for navigating through the screen-flow.
+Figure~\ref{fig:screenflow_example} presents a small example of such a screen-flow based application, which uses response from a camera and user input for navigating through the screen-flow.
 
@@
 %This makes the program behaviour replaceable and maintenance can be done easily.
-The interpreter processes programs written in Robot Behaviour Description Language (RBDL), a particular DSL based on XML, which has been developed particularly for the healthcare robotic applications.
+The interpreter processes programs written in Robot Behaviour Description Language (RBDL), a domain specific language (DSL) based on XML, which has been developed particularly for the healthcare robotics applications.
 It allows defining state machine based program behaviour and creating UI elements for screen visualizations.
@@
 
-For each state of the program, backgroundactions can be defined.
+For each state of the program, background actions can be defined.
 They are transparent to the user and run in background.
 Text messages can be sent or external devices can be acessed, for example.
-The latter is realized with web services to obtain a modular design.
-Some states can also have screen definitions with buttons, messages, videos, etc.
-on it.
-These are displayed to the user when the particular state is active.
+The latter is realized with web services~\cite{webservices} to obtain a modular design.
+Some states can also have screen definitions with buttons, messages, videos, etc., on it.
+These are displayed to the user when the respective state is active.
 Once displayed, buttons can be clicked and trigger events.
 The behaviour of events can be mapped in the expectedevents tag.
-It lists all events which are processable by the current state and defines the correspondend behaviour such as triggering transitions to other states.
+It lists all events which are processable by the current state and defines the corresponding behaviour such as triggering transitions to other states.
 
-For this work the most important thing about the RBDL to remember is, that it can define state machine behaviour with states, transitions and events.
-This fact was basement for later decisions regarding used safety mechanisms.
+For this work the most important thing about the RBDL to remember is that it can define state machine behaviour with states, transitions and events.
+This fact was the basis for later decisions regarding used safety mechanisms.
 
@@
 But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise.
-For this reason Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
+For this reason, Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
 This file can be downloaded to the robot and executed by the interpreter.
-Figure~\ref{fig:robostudio_proceeding} illustrates what the steps of robot behaviour development are.
+Figure~\ref{fig:robostudio_proceeding} illustrates the steps of robot behaviour development.
 
@@
 Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application.
-In several windows and views there are visual tools for editing and understanding robot behaviour as well as screen dialog design.
-As depicted in figure~\ref{fig:robostudio_vpe} the editor has got a State Navigator Window (1) in the top left corner where all existing states are listed.
-Underneath an overview about all incoming and outgoing transitions is given.
+Several windows and views provide visual tools for editing and understanding robot behaviour as well as screen dialog design.
+As depicted in figure~\ref{fig:robostudio_vpe}, the editor integrates a \emph{State Navigator Window} (1) in the top left corner where all existing states are listed.
+Below an overview about all incoming and outgoing transitions is given.
 New states can be created and existing states can be deleted.
-A click on one particular state id in the State Navigator Window will cause all other components to show all possible information regarding to the selected state.
-The Expected Events Editor Window (2) in the bottom left provides specifying the expected events and their effects.
-Heart and center of the editor is the UI Component Layout Editor Window (3).
+A click on one particular state id in the \emph{State Navigator Window} will cause all other components to show all possible information regarding the selected state.
+The \emph{Expected Events Editor Window} (2) in the bottom left can be used for specifying the expected events and their effects.
+Heart and center of the editor is the \emph{UI Component Layout Editor Window} (3).
 It renders the screen dialog of the selected state and gives a preview of the screen presented on the robot during runtime.
 Components such as buttons, text boxes or images can be added, changed or removed via drag and drop.
-They are provided by the UI Components Palette Window (4) on the right side where the user can choose from all supported UI components.
-The Background Actions Editor Window (5) lets the user define background actions, which are executed trasparently to the user such as accessing external devices or sending messages.
-Aiming for supporting the developer with helpful graphical tools the State Transition Visualization Window (6) gives an overview about all connections towards and from the selected state.
+They are provided by the \emph{UI Components Palette Window} (4) on the right side where the user can choose from all supported UI components.
+The \emph{Background Actions Editor Window} (5) lets the user define background actions, which are executed transparently to the user such as accessing external devices or sending messages.
+Aiming to support the developer with helpful graphical tools, the \emph{State Transition Visualization Window} (6) gives an overview about all connections towards and from the selected state.
 It helps to quickly get an understanding of the local program behaviour.
@@
 Here each node of a tree is described by an abstract operator whose subformulas are branches to their replacement operators.
-Thus the root node represents the entire formula.
+Thus, the root node represents the entire formula.
 They also explored the formula trees' 3D representation within a virtual space.
 Both approaches are different fundamental ideas of representing temporal formulas.
-Whereas having a graph  seems not to be applicable for our approach, the idea of hirarchical operator representation of Del Bimbo et al.
+Whereas having a graph  seems not to be applicable for our approach, the idea of hierarchical operator representation of Del Bimbo et al.
 will be a good basis for our concept of constraint creation and understanding.
@@
 HomeTL focuses more on monitoring temporal boundaries of a patient's behaviour than on ensuring functional safety of an implemented program.
-It does depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.
+It does however depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.
 

### 2012-10-27 thesis/tex/4_visualformalismsforsafetyconstraints.tex
correction, v.1.

--- 2012-10-21 thesis/tex/4_visualformalisms.tex
@@
 
-\chapter{Visual formalisms}
-\label{chap:visualformalisms}
+\chapter{Visual formalisms for safety constraints}
+\label{chap:visualformalismsforsafetyconstraints}
 
-In this chapter we introduce two different approaches of visual formalisms and later state the final desing of the visual language.
+This chapter introduces two different approaches of visual formalisms and later states the final desing of the visual language.
 One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points.
-A first approach, so called template constraints and its pros and cons are described in section~\ref{sec:templateconstraints}.
-Subsequently an improved design is presented, which tends to a more abstract direction.
+A first approach, so called template constraints, and its pros and cons are described in section~\ref{sec:templateconstraintformalism}.
+Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
 
-\section{Template constraints}
-\label{sec:templateconstraints}
+\section{Template constraint formalism}
+\label{sec:templateconstraintformalism}
 
@@
 % TODO: wir haben gedacht, dass es einfach zu benutzen ist
-Leaded by this question, we developed a visual language based on templates.
+Led by this question, a visual language based on templates has been developed.
 These templates make different propositions about states and their relations.
@@
 
-	The idea behind this constraint is, that propositions about temporal dependencies can be made.
+	The idea behind this constraint is that propositions about temporal dependencies can be made.
 Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled.
@@
 
-	This constraint type enforces at least one particular event to happen after a spezific precondition.
-	Example of a constraint: Every time a person enters a supermarket (x) he will eventually decide to buy 
+	This constraint type enforces at least one particular event to happen after a specific precondition.
+	An example of such a constraint would be: Every time a person enters a supermarket (x) he will eventually decide to buy 
 	%a product or not to buy it.
 	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can't avoid both possibilities.
-	Nitpicking readers might note the person could also steel something without buying it, and thus not buy anything but leave the shop with full hands.
-Of course this is true; exactly in this case our constraint would signal an undesirable "program" behavior with its invalidity.
+	%Nitpicking readers might note 
+	Of course the person could also steal something without buying it, and thus not buy anything but leave the shop with full hands.
+Of course this is true; In fact, exactly in this case the constraint would signal an undesirable ``program'' behaviour with its invalidity.
 \end{enumerate}
@@
 Furthermore the state labels should sometimes be replaced by the screen displays of the regarding states.
-All parameters of the template constraints can be customized by just clicking on one robot suface.
-In the subsequently appearing popup window (figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.
+All parameters of the template constraints can be directly customized by the robot surface symbols.
+In the subsequently appearing popup window (see figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.
 
@@
   \includegraphics[scale=0.7]{templateconstraint_dropdown}
-  \caption{Dropdown menu for spezifying template parameter.}
+  \caption{Dropdown menu for specifying template parameter.}
   \label{fig:templateconstraint_dropdown}
@@
 
-This visual language was presented to potential users for evaluation and feedbak.
+This visual language was presented to potential users for evaluation and feedback.
 Though most of them named it a good concept, soon it turned out that there are some problems and disagreements.
-The presented constraints were interpreted differently, the semantic seemd to be unclear and not unique.
-A try to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
+The presented constraints were interpreted differently, the semantic seemed to be unclear and ambiguous.
+An attempt to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
 These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
-But unfortunately even the textual eplanations don't seem to be a satisfying solution.
+But unfortunately even the textual explanations do not seem to be a satisfying solution.
 For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
 
-However, there is a second lack with the template constraints.
+However, there is a second problem with the template constraints.
 Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can't be described.
@@
 
-\section{Operator constraints}
-\label{sec:operatorconstraints}
+weitere constraint templates...
+
+\section{Operator constraint formalism}
+\label{sec:operatorconstraintformalism}
 
@@
 Since the visual constraints can be directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness.
-If it is considered as easy to use, two birds get killed with one stone: Non-experts can use it as well as experts.
+If it is considered as easy to use,
+%two birds get killed with one stone: N
+non-experts can use it as well as experts.
 
@@
 To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below.
-In addition the visual language should be capable of expressing constraints about future steps, both (h) any future state and (g) the next state, that (f) events should always happen, and that (i) a property must be true until some future event.
-Like in Del Bimbo et al.'s formula representation~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy.
-However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
-Each block represents either an operator or a proposition and has its specified semantics.
-Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.
-
-The visual language has support for eight operator types and one proposition as shown in figure~\ref{fig:operators}.
 
@@
 Other types such as numeric or string equations may be added but are less important for our healthcare subject.
+
+In addition the visual language should be capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i).
+
 	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
@@
 \end{description}
+
+Like the formula representation introduced by Del Bimbo et al.~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy.
+However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
+Each block represents either an operator or a proposition and has its specified semantics.
+Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.
+
+The visual language has support for eight operator types and one proposition. A suggestion for their graphical representation is shown in figure~\ref{fig:operators}.
 

### 2012-10-27 thesis/tex/5_automatedgenerationofsafetyconstraints.tex
correction, v.1.

--- 2012-10-21 thesis/tex/5_automatedconstraintgeneration.tex
@@
-\chapter{Automated constraint generation}
-\label{chap:automatedconstraintgeneration}
+\chapter{Automated generation of safety constraints}
+\label{chap:automatedgenerationofsafetyconstraints}
 
-As already mentioned before, it might be difficult for developers to find all reasonable safety constraints by hand which are needed for full safety covery.
-Hence it also can happen, that a prior software version was correct and over time mistakes crept in because of minor adjustments once in a while.
-To help prevent this to happen, we introduce the concept of automated constraint generation for state machines.
-Once a program reaches a complete condition, constraint suggestions which are valid on the current program can be computed fully automatically.
-On the one hand they may help the developer identifying reasonable constraints he didn't think about as well as contradictions between program and specification, eg.
-if a computed constraint doesn't make sense at all.
-On the other hand all constraints can be revalidated after each change to the software and thus ensure consistency.
+Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications.
+In order to discover whether a program meets its safety requirements, its behaviour is checked by the constraints.
+This happens mostly as a last check before software gets released.
+But regarding updating and maintenace work, often further changes are made to the software even after release.
+%But even after possible changes to the software in the context of updating and maintenance work,
+However, safety constraints can be applied again and assure the adherence of all properties.
+Safety constraints thus are also suitable for sustaining consistency during the different phases of software life.
+In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that.
+The more constraints are defined the the better can be the reliability.
+But as already mentioned before, it might be difficult for developers to find all reasonable %safety
+constraints by hand. % which are needed for full safety covery.
+In this case it can happen that a prior correct software version becomes faulty over time because of minor adjustments once in a while which are not covered by constraints.
+%To help prevent this to happen
 
-In order to support the developer in developing safe programs, automated generation and suggestion of constraints might be helpful.
+As a solution to this problem, the concept of automated constraint generation for state machines is proposed.
+Once the initial implementation of a program is finished, its behaviour can be analysed.
+Sgnificant parts of it can be determined automatically and put into constraint representation which are valid on the current program of course.
+%constraint suggestions which are valid on the current program can be computed fully automatically. 
+On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg.
+if a computed constraint does not make sense at all.
+On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.
+
+In order to support the developer in developing safe programs and to simplify maintainability, automated generation and suggestion of constraints might be helpful.
 But it's helpful only if the suggested constraints are relevant and reasonable.
@@
 depends on the domain and use case.
-It has to be mentioned that our approach of finding an algorithm for constraint finding presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
-Regarding the important constraints postulated in chapter~\ref{chap:goals}, we found out that all all the relevant states fall into one of the following categories:
+It has to be mentioned that the approach of finding an algorithm for constraint generation presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
+Regarding the important safety constraints postulated in chapter~\ref{chap:goals}, all the relevant states seem to fall into one of the following categories:
 
@@
 
-In order to compute the required constraints automatically, and possibly additional ones, we deal with subgraphs defined as follows:
-%Um die in requirements definierten constraints und vielleicht noch mehr zu finden, behelfen wir uns 
+%All associated states which match these properties can be combined to a 
+In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined.
+Such a group of states is named subgrap from now on.
+It is defined as follows:
 %Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}.
@@
 \end{itemize}
+
+A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}.
+Using this specification all possible subgraphs of this kind can be detected within a state machine.
+An algorithm how subgraphs can be computed programmatically is discussed later in section~\ref{sec:prototype:automatedconstraintgeneration}.
 
@@
 
-A graphical example of such a subgraph is shown in Fig.~\ref{fig:subgraph}. Using this specification all possible such subgraphs can be detected within a state machine.
-
 \section{Constraint determination}
 
-If the subgraph contains neither inner (infinite) loops nor final states (states having no outgoing transition) we can derive the following valid LTL formulas for each subgraph (with $\alpha$ as proposition for ``$\alpha$ is current state'' and similarly for $\beta$ and $\gamma$):
+For the healthcare applications, all important states which concern safety constraints belong to subgraphs. Thus, several safety properties can be formulated based on the structure of subgraphs.
+For each subgraph the following valid LTL formulas can be derived if it contains neither inner (infinite) loops nor final states (states having no outgoing transition).
+In the following $\alpha$ is proposition for ``$\alpha$ is current state'' and similarly $\beta$ and $\gamma$:
 
@@
 Formula~\ref{eq:first} says that an $\alpha$-state is always eventually followed by a $\beta$-state.
-Formula~\ref{eq:second} ensures that there will be always at least one $\gamma$-state be visited after each visit of an $\alpha$-state.
-The quite similar but stricter formula~\ref{eq:third} demands additionally a visit of at least one $\gamma$-state before the $\beta$-state can be reached.
+Formula~\ref{eq:second} ensures that there will always at least one $\gamma$-state be eventually visited after each visit of an $\alpha$-state.
+The quite similar but stricter formula~\ref{eq:third} additionally demands a visit of at least one $\gamma$-state before the $\beta$-state can be reached.
 Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states.
@@
 
-All formulas can be translated to the visual formalism presented in section~\ref{sec:operatorconstraints} and shown to the user.
+All formulas can be translated to the visual formalism presented in section~\ref{sec:operatorconstraintformalism} and shown to the user.
 The appropriate LTL formulas for the constraints postulated in chapter~\ref{chap:goals} relate to number \ref{eq:second} and \ref{eq:fifth} of the subgraph based formula templates and are as follows:
 
-\begin{equation}
+\begin{equation} \label{eq:sixth}
   \models \Box \Diamond \textnormal{'Polling'}
@@
 
-\begin{equation}
+\begin{equation} \label{eq:seventh}
 	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
@@
 %	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
+
+Formula~\ref{eq:sixth} ensures that state ``Polling'' will always eventually be visited again, and formula~\ref{eq:seventh} specifies that either state ``Well done!'' or state ``Staff notified'' will be eventuelly visited after each visit of ``Not taken yet''.

### 2012-10-27 thesis/tex/6_theltlcreatorprototype.tex
correction, v.1.

@@
 As stated in chapter~\ref{chap:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language.
-We therefore designed and implemented a prototype of such an editor which contains the visual language as well as the proposed constraint generation heuristics.
-The implementation is a java swing component and has a simple API definition, making it easy to integrate the LTLCreator tool into other java applications.
+Therefore a prototype of such an editor has been designed and implemented which contains the visual language as well as the proposed constraint generation heuristics.
+The implementation is a Java Swing component and has a simple API definition, making it easy to integrate the LTLCreator tool into other java applications.
 
@@
 However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots and to be a prototype for evaluation.
-Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio ought to be.
+Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio is possible.
 
@@
 
-Each of the different operator types, i.e.
-\emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator}, \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator}, \emph{UntilOperator} and \emph{StateOperator}, has got its own representing class.
+All different operator types, i.e.
+\emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator}, \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator}, \emph{UntilOperator} and \emph{StateOperator}, are realized by dedicated classes.
 All of them inherit from one abstract class \emph{AbstractOperator} which provides the basic functionality beeing used by all operators.
@@
 This function is abstract and implemented by each operator type separately.
-	\item[getColor():] Each operator type has got is own significant color which can be retrieved over the abstract function \emph{getColor()}.
-	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object dependet on the implementation.
+	\item[getColor():] Each operator type has got is own significant color which can be retrieved by the abstract function \emph{getColor()}.
+	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object, depending on the implementation.
 	\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
@@
 Additional parameters such as recursive painting including children can be specified.
-	\item[contains(int, int):] Given two coordinates x and y, it can be determined wheter the herewith defined point lays within an operator.
+	\item[contains(int, int):] Given two coordinates x and y, it can be determined whether the herewith defined point lays within an operator.
 	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality.
-That means same operators with similiar children.
+Equality is given when the compared operators are of the same type and similarly their children.
 This functionality is important for avoiding duplicate constraints during constraint generation.
@@
 
-Whereas \emph{StateOperator}s have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with respective one or two possible children.
+Whereas \emph{StateOperator}s have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two possible children respectively.
 The graphical representation of both unary and binary operators thus needs to have place holders and docking stations for sub operators.
@@
 
-All graphical components such as buckets and operators are implemented as java swing \emph{JComponent}s.
+All graphical components such as buckets and operators are implemented as Java Swing \emph{JComponent}s.
 Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.
@@
 
-For lightweight components swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component.
+For lightweight components Swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component.
 This is used for giving every operator its special look.
-We hereby made use of swing's \emph{Graphics2D} functionality what gives a great support in painting rounded shapes and color gradients.
-An operator is displayed as a rounded rectangle with dents around sub operators and has got a flowing backgound, borders and labels in an operator type specific color.
-For the significant shape multiple rounded rectangles with and without border are arranged in a way that the result appears to be one peace.
-Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the if operator first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be.
-In the third step the same rectangle as from step one just this time without border is painted on top.
+We hereby made use of Swing's \emph{Graphics2D} functionality what gives a great support in painting rounded shapes and color gradients.
+An operator is displayed as a rounded rectangle with dents around sub operators and has a flowing backgound, borders and labels in an operator type specific color.
+For the significant shape multiple rounded rectangles with and without border are arranged in a way that the result appears to be one piece.
+Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the \emph{IfThenOperator}, first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be.
+In the third step the same rectangle as from step one is painted on top, this time without border.
 This makes all crossing borders disappear.
-Only labels and buckets have to be added yet and a complete \emph{IfThenOperator} emerges.
+Only labels and buckets still have to be added to create a complete \emph{IfThenOperator}.
 
@@
 
-Swing also provides a mechanism for layouting components, that means depending on child components the aspired sizes and positions of components can be retrieved.
-As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraints} constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for varation in time.
+Swing also provides a mechanism for layouting components, that means determining the aspired sizes and positions of components depending on child components.
+As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for variation in time.
 This fact forms a special requirement for the layouting of operators.
-Whereas the operator's positioning and resizing along the logical direction is only dependend on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult.
+Whereas the operator's positioning and resizing along the logical direction only depends on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult.
 Here sub operators can't just be aligned centered as it is the case vertically.
@@
 
-The inbuilt mouse support of swing was used for identifying operators laying under the current mouse position.
-On the one hand this is important for mouse sensitive highlighting of particular constraint parts, on the other hand this way it can be determined where operators have to be added after drag and drop.
+The built-in mouse support of Swing was used for identifying operators positioned under the mouse cursor.
+On the one hand this is important for context sensitive highlighting of particular constraint parts, on the other hand this way the position where operators have to be added after drag and drop can be determined.
 So far
-% the predefined swing drag and drop has not been used but
-a simple drag and drop handler has been implemented.
-% instead. 
-Once a \emph{JComponent} is registered to the \emph{DndHandler}, it automatically manages any drag and drop actions with \emph{JComponent}s which implement either the \emph{Draggable} or \emph{DropTarget} interface.
-The former is implemented by all operators, the latter by buckets and the dashboard.
-%Figure~\ref{fig:dndclassstructure} gives a short class overview.
-Furthermore there is a class \emph{DraggableProvider} which displays a kind of button in the tool bar and lets the user create new \emph{Draggable}s; in our case new operators get instantiated.
-
+% the predefined Swing drag and drop has not been used but
+a simple drag and drop mechanism has been implemented whose class structure is shown in figure~\ref{fig:dndclassstructure}.
 \begin{figure}[htbp]
@@
   \label{fig:dndclassstructure}
-\end{figure}
+\end{figure}% instead. 
+Once a \emph{JComponent} is registered to the \emph{DndHandler}, it automatically manages any drag and drop actions with \emph{JComponent}s which implement either the \emph{Draggable} or \emph{DropTarget} interface.
+The former is implemented by all operators, the latter by buckets and the dashboard.
+%Figure~\ref{fig:dndclassstructure} gives a short class overview.
+Furthermore there is a class \emph{DraggableProvider} which displays a kind of button in the tool bar and lets the user create new \emph{Draggable}s; in our case new operators get instantiated.
 
@@
 
-In section~\ref{chap:automatedconstraintgeneration} we stated that automated constraint generation is based on subgraphs.
-For the finding of such subgraphs an algorithm has been developed and the main idea how it works is described as follows.
+Section~\ref{sec:subgraphdefinition} states that automated constraint generation is based on subgraphs. For the finding of such subgraphs an algorithm has been developed. It is described as follows:
 
-We define one start state and generate all possible paths starting from it leading through the state machine with just one restriction that a path ends before it visits states twice.
-Thus we retrieve a finite amount of finite paths since loops are eliminated.
-Now every state gets a number assigned which represents the total number of occurernces of the respective state in all collected paths.
-In other words, the label of a state expresses the number of pahts which start in the start state and lead through the respective state.
-The start state itself has got the number of all paths, of course.
+First of all one start state ($\alpha$) is defined and all possible paths are generated which start from it and lead through the state machine.
+Excluded are all paths which contain states more than once.
+The result is a finite amount of finite paths since loops are eliminated.
+Now every state gets assigned a number which represents the total number of occurrences of the respective state in all collected paths.
+In other words, the label of a state expresses the number of paths which start in the start state and lead through the respective state.
+The start state itself is labelsed the number of all paths.
 If during walking through any of the paths a second state can be found with the same label, it means that all paths starting at start state come all together in this state again.
-It is uncovered as potential corresponding end state.
-A last check has to be applied whether there are incoming transitions from outside into the subgraph which don't come over start state. If no such transitions are detected, a subgraph is found.
+It is determined to be a potential corresponding end state.
+A last check has to be applied whether there are incoming transitions from outside into the subgraph which don't originally come from start state.
+If no such transitions are detected, a subgraph is found.
 If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with start state insofar as the state machine doesn't contain further loops or terminating end states.
 Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrence of each state.
-This procedure is suitable for bringing to light whether a state is the beginning of a subgraph and for discovering the corresponding end state.
+This procedure is suitable for checking whether a state is the beginning of a subgraph and for discovering the corresponding end state.
 In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state.
@@
 
-Following the above description an algorithm has been implemented in java.
-Its realisation differs from the presented concept in one point that paths are not listed but states are directly labelled since saving all possible paths would be storage inefficient.
-The state machine is traversed similarly to a depht-first search and each visited state's label is incremented by one.
-If the currently visited state holds more than one outgoing transition and thus constitutes new paths,  all states laying on the current stack, i.e.
+Following the above description an algorithm has been implemented in Java.
+Its realisation differs from the presented concept in one point: Paths are not listed but states are directly labelled since saving all possible paths would be memory inefficient.
+The state machine is traversed similarly to a depth-first search and each visited state's label is incremented by one.
+If the currently visited state holds more than one outgoing transition and thus constitutes new paths,  all states on the current stack, i.e.
 the previously visited states on the current path, get their labels incremented by $(\textnormal{'number of outgoing transitions'} - 1)$.
@@
 In line 4 the labelling process described above is executed by using one particular start state.
-If a state has got the same label as start state has, it is taken as potential end state.
-Otherwise, the entire state machine is assumed to be a subgraph and start state is taken as end state as well.
+If a state has the same label as the start state, it is assumed as potential end state.
+Otherwise, the entire state machine is assumed to be a subgraph and start state is defined as end state as well.
 Subsequent checks finally determine whether it can be considered as proper subgraph.
 There are no cycles allowed between start and end state since they're forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
-Furthermore also caving transitions are prohibited.
+Furthermore also foreign transitions are prohibited.
 Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
 If all checks are successful, start and end states get added to the result list $subgraphs$.
-This proceeding is applied to any state of the state machine what is achieved by a surrounding foreach loop.
-In the end the list $subgraphs$ contains all pairs of start and end states which form subgraphs.
+This procedure is applied to all states of the state machine by a surrounding foreach loop.
+In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.
 
@@
 
-Depending on size and grade of branch of the state machine, the algorithm for finding subgraphs may take more or less time.
+The runtime of the algorithm depends on size and grade of branch of the state machineh.
 To improve user experience during constraint generation a dialog has been implemented which informs the developer about the current progress.
-This includes the number of found constraints so far and a green bar which gives a percentual estimation of execution time.
+This includes the number of constraints found so far and a progress bar which gives a percentual estimation of execution time.
 Furthermore a cancel button permits premature stopping of the execution.
@@
 
-The algorithm implementation has been infiltrated by code snippets which frequently interacts with so called \emph{ValidationResultListener}s (figure~\ref{fig:validationresultlistener_class}).
+The algorithm implementation has been extended by code snippets which frequently interact with so called \emph{ValidationResultListener}s (figure~\ref{fig:validationresultlistener_class}).
 This interface is also implemented by the progress dialog.
-A \emph{ValidationResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about new found subgraphs with invoking \emph{void newSubGraphFound(SubGraph)}.
-The execution will become terminated whenever \emph{boolean continueGeneration()} returns true.
+A \emph{ValidationResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about newly found subgraphs by invoking \emph{void newSubGraphFound(SubGraph)}.
+The execution will terminate whenever \emph{boolean continueGeneration()} returns false.
 
@@
 
-Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in chapter~\ref{chap:automatedconstraintgeneration} have to be assembled in one tool, the LTLCreator.
-It shall provide a dashboard with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results.
-Besides a button is needed for triggering constraint generation.
+Both features, the visual language presented in section~\ref{sec:operatorconstraintformalism} and the automated constraint generation in chapter~\ref{chap:automatedgenerationofsafetyconstraints}, were assembled in one tool, the LTLCreator.
+It provides a dashboard with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results.
+Besides a button is provided for triggering constraint generation.
 
@@
 It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
-Constraints can be composed in the dashboard in the center of the editor by drag\&drop which holds and displays the operator constraints.
+Constraints can be composed by drag\&drop in the dashboard in the center of the editor which holds and displays the operator constraints.
 The tab functionality on the left allows multiple constraints to be managed.
 Each tab shows a small thumbnail of the constraint and a symbol indicating its validity; a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.''.
-The magic wand button within the tab pane brings the constraint generation functionality into the editor and triggers the constraint finding process.
+The magic wand button within the tab pane makes the constraint generation functionality accessible within the editor and triggers the constraint finding process.
 
@@
 
-One main requirement of the LTLCreator tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs.
+One main requirement to the LTLCreator tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs.
 For this a \emph{ConstraintValidator} class has been defined which provides a method \emph{validate(AbstractOperator)} for proving constraints.
@@
 NuSMV doesn't have a java API but it can be externally accessed over command line.
-For this both the state machine program and the LTL formulas have to be put into file which has to match a NuSMV specific format.
+For this both the state machine program and the LTL formulas have to be translated to an input file which has to match a NuSMV specific format.
 An example of this input language is given in listing~\ref{lst:nusmv}.
@@
 The state machine described in the example contains eleven states what is determined by the number range for \emph{state} from one to eleven.
-All variable assignments are specified under \emph{ASSIGN}, what is the semantic equivalent to transition definitions.
+All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions.
 The initial state can be determined and a case distinction regulates the next possible values, i.e.
@@
 This kind of output is generated by the \emph{ConstraintValidator} whenever a constraint needs validation.
-All states are then mapped to numbers and transitions get transormed into case distinctions.
-The result is stored to file and then loaded by NuSMV.
+All states are then mapped to numbers and transitions get transformed into case distinctions.
+The result is stored to a file and then loaded by NuSMV.
 The corresponding valdidation result can be subsequently read from the command line and displayed in the editor next to the visual constraint.
@@
 This is the case when the constraint generator creates multiple constraints or a change of the state machine causes all constraints to be revalidated at once.
-Since model checking is a quite ressource inefficient activity, it's recommended to execute all validation tasts consecutively rather than in parallel.
-A second advantage is that the developer doesn't have to wait for all of them to finish before he can see any result.
+Model checking is a quite ressource inefficient activity and parallel execution of several checks might overload the main memory what causes slow data swapping to start.
+Since each constraint is checked in a new instance of NuSMV, a parallel execution would furthermore strain the CPU scheduler and slow down the overall execution time.
+Thus it's recommended to execute all validation tasks consecutively rather than in parallel.
+An extra advantage is that the developer doesn't have to wait for all tasks to finish before he can see any result.
 Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
@@
 Every validation task of a constraint gets enqueued in the \emph{ValidatorThread} after calling the method \emph{addToQueue(...)}.
-An identifier is used for each constraint what enables elimination of duplicate tasks for the same constraint in order not to strain the execution time.
+An identifier is used for each constraint which enables elimination of duplicate tasks for the same constraint in order not to strain the execution time.
 Tasks can also manually be removed from the queue by calling \emph{cancel(...)}.
@@
 
-LTLCreator should be available for other state machine based programming environments for integration what is ensured by its implementation as a java swing component and a clear API enabling the tool for a wide range of use.
+LTLCreator should be integratable into other state machine based programming environments, what is ensured by its implementation as a Java Swing component and a clear API enabling a wide range of use of the tool.
 
-Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
+Figure~\ref{fig:coalescence} points out interactions between LTLCreator and programming environments it is integrated in.
 \begin{figure}[htbp]
@@
 The constraint generator as well as the validator are non UI parts of LTLCreator.
-The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program.
+The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the model.
 This will automatically revalidate all existing constraints and always provide the up-to-date model for constraint generation.
@@
 
-As statet in the previous section the LTLCreator provides an API for integration into other java based applications.
-Robostudio is such an application which shall be used for evaluating the visual language as well as the automated constraint generation feature and is also supposed to provide all this functionality for development of service robot behaviour in future.
+As stated in the previous section, the LTLCreator provides an API for integration into other Java based applications.
+Robostudio is such an application which was used for evaluating the visual language as well as the automated constraint generation feature and is also supposed to provide all this functionality for development of service robot behaviour in future.
 
 Robostudio uses NetBeans~\cite{netbeans} as a platform and organizes all tools and features in windows.
-Thus also LTLCreator has to be put into a window and added to the Robostudio platform.
-This is done by placing it into a NetBeans module which is added to Robostudio's modules.
-This module imports the as jar archived LTLCreator tool and further consists of only two classes: \emph{LTLCreatorTopComponent} and \emph{XmlDataToLTLEditorDataWrapper}.
+Thus also LTLCreator was put into a window and added to the Robostudio platform.
+This was done by placing it into a NetBeans module which was added to Robostudio's modules.
+This module imports the as jar archived LTLCreator tool, archived as a jar file, and further consists of only two classes: \emph{LTLCreatorTopComponent} and \emph{XmlDataToLTLEditorDataWrapper}.
 The former implements the \emph{TopComponent} provided by NetBeans which can be used for displaying UI elements.
 In order to visualize the LTLCreator component it is directly placed on the \emph{LTLCreatorTopComponent}.
-It retreives changes of the model and passes them to the \emph{LTLCreator} component.
-Since the defining of safety constraints is an important issue not only after but also during the process of developing service robot behaviour, the LTLCreator window as a vital tool has been positioned in the center of the platform where also the \emph{UI Component Layout Editor Window} is.
+It retrieves changes of the model and passes them to the \emph{LTLCreator} component.
+Since the defining of safety constraints is an important issue not only after but also during the process of developing service robot behaviour, the LTLCreator window as a vital tool has been positioned in the center of the platform where also the \emph{UI Component Layout Editor Window} is located.
 The \emph{XmlDataToLTLEditorDataWrapper} forms the bridge between Robostudio's own model and the state machine model used by \emph{LTLCreator}.
-All complex data coming from Robostudio has to be detached from unimportant graphical information and transformed to an elementary state machine representation before it can be processed by \emph{LTLCreator}.
+All complex data coming from Robostudio has to be cleaned from unimportant layout information and transformed to an elementary state machine representation before it can be processed by \emph{LTLCreator}.
 
@@
 As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
-This demands a concept of change notification management which is responsible for receiving all chages and making this information available as a service.
+This demands a concept of change notification management which is responsible for receiving all changes and making this information available as a service.
 Thus we tried to extend all parts of the code where model changes are made with short code snippets.
@@
 Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than syntax checks.
-Thus syntax errors are not detected or even avoided by static analysis ore input restrictions.
+Thus syntax errors are not detected or even ignored by static analysis or input restrictions.
 In fact, such features haven't been requirements for Robostudio.
 During integration of \emph{LTLCreator} into Robostudio we encountered problems like transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
-Well, this characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistend data retrieved from Robostudio.
+This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistend data retrieved from Robostudio.
 
@@
 
-To guarantee any all model change causes the constraints to revalidate, all relevant code parts have to report on this.
+To guarantee that all model changes cause the constraints to revalidate, all relevant code parts have to report on this.
 But modification of the code hasn't been possible in all cases.
 Whereas all state changes are announced, transition changes stay unnoticed.
-This decision is straight for the prototype, but for later the developers of Robostudio announced an adapted architecture to solve this problem in the next version.
+This decision is straight for the prototype, but the developers of Robostudio announced an adapted architecture to solve this problem in the next version.
 
@@
 After successful transformation the state machine is passed to the \emph{LTLCreator} component.
-Otherwise no validation result is displayed for constraints and an appearing dialog lists all identified syntax errors.
-In matter of safety and reliability a restrictive approach which gives no result at all seemed more applicable than an attempt to interprete the inconsistent data.
+Otherwise no validation result is displayed for constraints and a dialog lists all identified syntax errors.
+In the context of safety and reliability a restrictive approach which gives no result at all seemed more applicable than an attempt to interpret the inconsistent data.

### 2012-10-27 thesis/tex/7_testscenarioandevaluation.tex
have written section 2.3
correction, v.1.

@@
 
-In order to evaluate the proposed safety functionality we use the Robostudio version where we already integrated LTLCreator in.
-This visual programming environment allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
-As a test scenario we consider the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}.
+Both, the visual language and the automated constraint generation functionality, have been integrated into Robostudio in order to evaluate the proposed safety functionality.
+Robostudio is a visual programming environment which allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
+As a test scenario the simplified medication reminder application is considered whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}.
 Starting from ``Menu'' the user can choose from several services such as blood pressure measurement or entertainment.
-In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part.
-In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered.
+In the given example these services are simplified to just one state ``additional services'' since the focus concentrates on the medication reminder part.
+In state ``Polling'' the database is checked for a pending reminder and, if there is one, the medication intake guidance is triggered.
 The patient can state if he or she has already taken the medication, or decide whether to take it or not.
-In the latter he or she may give a reason for it and a caregiver is notified about this incident.
-Otherwise it will result in ``Well done!'' after completing intake.
+In the latter case, he or she may give a reason for it and a caregiver is notified about this incident.
+Otherwise the progress will result in the state ``Well done!'' after completing intake.
 
@@
 
-This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections we will evaluate the visual language as well as the capability of constraint generation.
-In section~\ref{sec:practicalexperiences} it is described how a non-expert deals with the visual language and what the final judgement is.
+This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections we will evaluate the visual language as well as the potential of constraint generation.
+Section~\ref{sec:practicalexperiences} describes how a non-expert deals with the visual language and what the final judgement is.
 
@@
 
-Before we can start creating constraints it might be useful to think about a reasonable term we want to express.
+Before constraints can be created it might be useful to think about a reasonable term to express.
 Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given as a requirement in chapter~\ref{chap:goals}.
 
-In order to find a corresponding graphical constraint the sentence has to be analyzed just as you read it.
+In order to find a corresponding graphical constraint the sentence has to be analyzed just as developers read it.
 Since ``whenever'' is a semantic equivalent to ``always if'' first of all a \emph{ALWAYS} operator gets dragged to the dashboard directly followed by an \emph{IF} as shown in figure~\ref{fig:sampleconstraint}.
@@
 
+\subsection{Usability}
+
 Due to the optimized performance of NuSMV, even the validation of constraints on huge and complex programs is fast and allows rapid feedback.
 In addition, the validation itself runs in the background without locking the dashboard.
-Therefore we avoid waiting times and disruptions during constraint development, which would be likely with conventional model checking where constraint development and constraint validation are alternating processes.
-Thus we consider our tool an improvement for the user experience.
+Therefore waiting times and disruptions during constraint development can be avoided, which would be likely with conventional model checking where constraint development and constraint validation are alternating processes.
+Thus our tool is considered an improvement for the user experience.
 
-We observed the different color flavours of the visual operators to be a good support for fast reading and understanding of constraints.
-Also the two dimensional composition and the round shapes of the operators turned out to
-give the language a schematic but not too rectangular look. Besides it is considered eye-candy what is the best motivation for using this visual language.
+It could be observed that the different color flavours of the visual operators are a good support for fast reading and understanding of constraints.
+Also the two dimensional composition and the round shapes of the operators turned out to give the language a schematic but not too rectangular look.
+Besides it is considered eye-candy what is the best motivation for using this visual language.
 
@@
 
-Due to its basing on LTL, the visual language's expressivenes is equivalent to LTL's one of course. More precisely this means any logic formula can be expresssed except potentiality.
-Furthermore there is one more limitation in the current visual language with the variety of available proposition operators.
+Due to being based on LTL, the visual language's expressivenes is equivalent to LTL's one of course. More precisely this means any temporal logic formula can be expresssed except potentiality.
+Furthermore there is one more limitation in the current visual language concerning the variety of available proposition operators.
 The visual language was originally designed for validating healthcare applications where just one kind of proposition is needed: ``currently state x is active''.
 Thus other propositions such as equations, greater than etc.
-are not implemented yet, but can be easily extended.
-Figure~\ref{fig:example_propositions} depicts how an extention of \emph{LTLCreator} with additional proposition operators could be realised.
+are not implemented yet, but can be easily integrated.
+Figure~\ref{fig:example_propositions} depicts how an extension of \emph{LTLCreator} with additional proposition operators could be realised.
 
@@
 The automated constraint generation can be triggered by a click on the magic wand button in the tab area.
-After activation a dashboard is opened in a new tab for each found constraint, and validation is initiated immediately.
+After activation a dashboard is opened in a new tab for each constraint found, and validation is initiated immediately.
 For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in figure~\ref{fig:generatedconstraints} which match the postulations in the requirements in chapter~\ref{chap:goals}.
@@
 
-The search and generation process finishes within less than a second and thus doesn't make the developer wait for a long time.
+The search and generation process finishes within less than a second and thus doesn't let developers wait for a long time.
 
-We showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{c)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment or blood pressure measurement.
+It was showed that the subgraph approach is working for the medication reminder application, and there was even one more reasonable constraint found by the heuristic: Constraint \emph{c)} ensures that during the medication intake process the program can not switch back to the menu or other services such as entertainment or blood pressure measurement.
 
@@
 In order to examine the performance of subgraph finding and constraint generation, we need to have a second look at the implementation details again.
-The subgraph finding algorithm is realized as a depth-first search.
-Indeed all healthcare programs are infinite and contain cycles and a depth-first walk-through would never end, but the algorithm doesn't go deeper at already visited states.
+The subgraph finding algorithm is realized similar to a depth-first search.
+Indeed all healthcare programs are infinite and contain cycles and a depth-first walk-through would never end, but as stated in section~\ref{sec:prototype:automatedconstraintgeneration} the algorithm does not revisit already visited states.
 
-The performance of this algorithm is of course highly dependend on the grade of branching.
-Nevertheless, such a depth-first search has an estimated wost case execution time of $O(|V|+|E|)$ in which $|V|$ is the number of states and $|E|$ number of edges respective transitions in our case.
-Since we cope with state machines and thus have directed edges rather than undirected ones, each state can have $|V|$ transitions; $|V-1|$ to other ones and one to itself.
+The performance of this algorithm of course highly depends on the grade of branching.
+In contrast to a usual depth-first search in trees, the modified depth-first search is applicable for even cyclic state machines and will walk through all possible paths starting from one particular start state without visiting states twice.
+Assuming the worst case in which every state has transitions to all other states, each of the remaining $|V|-1$ states can be possibly visited next, similarly $|V|-2$ for the third step and so on.
+As a worst case execution time the following result is obtained:
+
+%Nevertheless, such a depth-first search has a worst case execution time of $O(|V|+|E|)$ in which $|V|$ is the number of states and $|E|$ number of edges respective transitions in our case.
+%Since we cope with state machines and thus have directed edges rather than undirected ones, each state can have $|V|$ transitions; $|V|-1$ to other ones and one to itself.
 For the contained tree without cycles for depth-first search we retreive $|E|=|V|$ edges.
-As worst case ececution time for the depth-first search we obtain the following result:
+%As worst case ececution time for the depth-first search we obtain the following result:
 
 \begin{equation}
-O(\textnormal{``depth-first search''})=O(|V|)+O(|V|-1)=O(|V|)
+O(\textnormal{``depth-first search''})=O(1*(|V|-1)*(|V|-2)*...*2*1)=O((|V|-1)!)
 \end{equation}
 
-In order to find a subgraph with our presented algorithm the depth-first search has to be startet on a start state of a subgraph.
+In order to find a subgraph with the presented algorithm the depth-first search has to be startet on a start state of a subgraph.
 Since we don't know in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions.
-In the worst case all $|V|$ states match this condition, so the depth-first search has to be executed $|V|$ times.
-This implies for our estimated worst case execution time:
+In the worst case all $|V|$ states fulfil this condition, so the depth-first search has to be executed $|V|$ times.
+This implies for the estimated worst case execution time:
 
 \begin{equation}
-O(\textnormal{``subgraph finding''})=O(|V|)*O(|V|)=O(|V|^2)
+O(\textnormal{``subgraph finding''})=O(|V|)*O((|V|-1)!)=O(|V|!)
 \end{equation}
@@
 
-For all the healthcare applications the presented algorithm works efficient and fast.
+For all the healthcare applications that we examined the presented algorithm works efficient and fast.
 But their program sizes are still quite manageable.
-Our test program has eleven states, an average branch grade and a constraint finding execution time of a split second on a standard pc.
-Of course an exponential execution time is bad if the number of states arises.
-A clear slowdown can be felt when the algorithm runs on larger state machines with plenty of states and high grade of branching.
-A test on a state machine with over two hundred and fifty states and partially high branch grade resulted in approximately five minutes execution time.
+The test program has eleven states, an average branch grade of $1.\overline{63}$ outgoing transitions per state and a constraint finding execution time of a split second on a standard computer.
+Of course an exponential or factorial execution time is bad if the number of states arises.
+A clear slowdown can be felt when the algorithm runs on larger state machines with plenty of states and a high grade of branching.
+A test on a state machine with over 250 states and partially high branch grade resulted in approximately five minutes execution time.
 
-However, sholud this algorithm regardless its tailoring to the healthcare domain be used for other huge programs, it keeps a lot of potential to be improved in efficiency.
+However, in case that this algorithm is used for other huge programs regardless its tailoring to the healthcare domain, it keeps a lot of potential to be improved in efficiency.
 With a sophisticated approach it might be enough to execute just one depth-first search for finding all subgraphs at once what would lead to a linear execution time.
@@
 The constraint generator algorithm found six constraints for the medication reminder application.
-Two of them were actually the ones we proposed in the requirements.
+Two of them were actually the ones proposed in the requirements.
 Three of them are either intensified versions of the previous ones or just describe the behaviour of the program and thus make sense, but in matters of safety they aren't fundamental relevant.
-Nevertheless the generator tool actually helped us in finding a new important constraint we didn't think about before.
+Nevertheless the generator tool actually found a new important constraint which had not been considered before.
+It is constraint \emph{c)} in figure~\ref{fig:generatedconstraints}, section~\ref{sec:prototype:automatedconstraintgeneration}.
 
-To put it in a nutshell, all automatically generated constraints are practical, and approximately half of them are significant.
-We showed that the constraint generator concept works well for the healthcare application and can be a very helpful feature.
+To put it in a nutshell, in context of this example all automatically generated constraints are practical, and approximately half of them are significant.
+This shows that the constraint generator concept works well for the healthcare application and can be a very helpful feature.
 
@@
 
-%indeed
-%surprisingly
-%hm was given as a reason.
-%Probably due to 
-%Due to textnähe und einfachen lesbarkeit der vl
-%da der text nah ist und auch für ... vferständlich ist
-%not familiar with
-%As a first step,
-
 Ms Kracker didn't know in beforehand what state machines are and never came in contact with LTL or model checking in general.
@@
 
-%Für eine Person, die von modelchecking oder LTL noch nie etwas gehört hat, kam sie mit kleinen Hilfen gut mit der visuellen Sprache zurecht und könnte sie so mit etwas Übung auch verwenden.
-
-%Die Nähe zwischen LTL und deutschem Satz ist gut, und dadurch kann man schnell konstraints zusammen klicken (besser als mit fremden mathematischen symbolen).
-
 To the question, whether she would recommend such robotic services like the medication reminder application for use in the care facility she's working in, Ms Kracker replied: ``I can envision such robot applications beeing an excellent assistance for the elderly who are mentally fit but maybe a little doolally.
@@
 Thus such a reminding robot application would constitute an excellent support for us staff!''.
-%Auf die Frage, ob sie solche Roboteranwendungen wie der Medicationreminder in der Kurzzeitpflege für sinnvoll erachten würde, antwortet Kracker:
-%``Für geistig fitte aber dennoch vergessliche und tütelige Leute kann ich mir diese Roboterapplikation super vorstellen.
-Ein Großteil unserer Patienten ist jedoch nicht in der Lage Entscheidungen, die der Roboter abfrägt, wahrheitsgemäß zu treffen.
-Aufgrund der regelmäßig wechselnden Patientenbelegung in der Kurzzeitpflege ist es vielmehr für das Betreuungspersonal eine große Herausforderung, die uhrzeitgebundenen Medikamenteinnahmen der Patienten nicht zu vergessen.
-Daher wäre eine solche Erinnerungsfunktion für das Personal eine große Hilfe.''.

### 2012-10-27 thesis/tex/8_conclusionandfuturework.tex
* Section 7.3 is now translated to english * Minor changes in chapter 2 * Changes to chapter 8
correction, v.1.

@@
 
-first step towards safety certification
+%first step towards safety certification
 
-In this paper we presented a visual formalism for the defining of safety constraints and first steps towards a concept of automatic constraint generation.
+This work presented a visual formalism for the defining of safety constraints and first steps towards a concept of automatic constraint generation.
 Both features are implemented and assembled in an editor for standalone or integrated use.
@@
 
-The concept of applying hueristics for proposing possible constraints to users of the visual language produced very good results in our use case, the medication reminder application.
-All postulated constraints were found and we even discovered a new reasonable constraint.
+The concept of applying hueristics for proposing possible constraints to users of the visual language produced very good results in the healthcare use case, the medication reminder application.
+All postulated constraints were found and even a new reasonable constraint was discovered.
 Thus we are convinced that this approach, as well as the visual language for defining constraints itself, has the potential to really support users in development of safety critical robot applications in the healthcare domain and also other domains.
@@
 For example, the state proposition (currently the only supported proposition type) used in constraint formulation might be not enough for other domains.
-Supporting propositions based on strings or mumerical equations might be necessary.
+Supporting propositions based on strings or numerical equations might be necessary.
+For reasons of enhancing the visual language, a substitution operator could be provided.
+Already existing constraints could be reused as parts of more complex constraints and thus simplify the overall look.
 Finally, the only result of constraint checking presented to the user currently is whether the constraint is valid or invalid.

### 2012-10-28 thesis/tex/2_background.tex
2.3.1 finished

@@
 For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also human lifes.
-In order to eliminate software malfunction there are a lot of mathematical concepts for verifying software behaviour.
-One of them is model checking wich allows checking for functional safety on state machine programs.
-Since we deal with state machines in the healthbot applications mentioned in section~\ref{sec:healthbotapplication}, this concept of behaviour checking is applicable and important for our kind of programs.
+In order to eliminate software malfunction there are different mathematical concepts for verifying software behaviour.
+One of them is model checking wich allows checking state machine programs for functional safety.
+Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realised wth state machines, this concept of behaviour checking is applicable and important for our kind of programs.
 At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an intorduction into model checking in section~\ref{sec:modelcheckingandltl}.
@@
 \label{sec:statemachines}
+
+A state machine is a model which can describe the behaviour of a system. It contains an aggregation of all different statuses of a system and possible changeovers from one status to another.
+
+Each system status is represented by one \emph{state} which stands for a particular and unique constellation of system variables and properties.
+One state can be declared as \emph{start state} which is the very first state when the system starts working.
+The state representing the current system status is called \emph{current state}.
+The current state changes whenever there is a modification to the system.
+Such transfers of the system from the current state to another one are defined by \emph{transitions}.
+Thus, they define the possible system behaviour.
+All influences of the environment which cause a change to the system are called \emph{events}.
+They trigger transitions which transfer the system into a new state (current state).
+The fact that a state machine reacts to events is the reason why state machines are also called \emph{reactive systems}.
+
+State machines can be visualized, as shown in figure~\ref{fig:statemachine} which gives a small example of a state machine.
+\begin{figure}[htbp]
+  \centering
+  \tikzstyle{state} = [circle,fill=white,draw]
+  \tikzstyle{stateblack} = [circle,fill=black!100,draw]
+  \tikzstyle{every path} = [font=\sffamily\small]
+
+	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=25mm]
+	  \node[state] (1) {state 1};
+	  \node[state] (2) [below right of=1] {state 2};
+	  \node[state] (3) [above right of=2] {state 3};
+	  \node[stateblack] (4) [above of=1,node distance=17mm] {};
+
+	  \path[]
+	  	(1) edge [bend right] node [left] {event b} (2)
+	        edge [bend left] node [above] {event a} (3)
+	    (2) edge [bend right] node [right] {event d} (3)
+	    (3) edge [bend left] node [above] {event c} (1)
+	        edge [loop right] node {event e} (3)
+	    (4) edge [right] node [right] {} (1)
+	    ;
+	\end{tikzpicture}
+  \caption{General example for a state machine.}
+  \label{fig:statemachine}
+\end{figure}
+All states are drawn as circles which contain their names. Transitions are painted as arrows between source and destination states, and the regarding trigger event names are written next to it.
+
+The following example clarifies the concept of state machines.
+Assuming a door which can be opened, closed and locked by a key-operated bolt lock, the behaviour can be reduced to a state machine as shown in figure~\ref{fig:doorstatemachine}.
+\begin{figure}[htbp]
+  \centering
+  \tikzstyle{state} = [circle,fill=white,draw]
+  \tikzstyle{stateblack} = [circle,fill=black!100,draw]
+  \tikzstyle{every path} = [font=\sffamily\small]
+
+	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=35mm]
+	  \node[state] (1) {closed};
+	  \node[state] (2) [below of=1] {locked};
+	  \node[state] (3) [right of=1] {opened};
+	  \node[state] (4) [below of=3] {blocked};
+	  \node[stateblack] (5) [left of=1,node distance=16mm] {};
+
+	  \path[]
+	  	(1) edge [bend left] node [above] {open door} (3)
+	        edge [bend left] node [above right] {turn key} (2)
+	    (2) edge [bend left] node [left] {turn key} (1)
+	    (3) edge [bend left] node [above] {close door} (1)
+	        edge [bend left] node [right] {turn key} (4)
+	    (4) edge [bend left] node [below left] {turn key} (3)
+	    (5) edge [right] node [right] {} (1)    
+	    ;
+	\end{tikzpicture}
+  \caption{Door behaviour as a state machine.}
+  \label{fig:doorstatemachine}
+\end{figure}
+Four states represent the ``system statuses'' of a lockable door: closed but unlocked (closed), opened and unlocked (opened), closed and locked (locked), opened but locked (blocked).
+The latter case is when the extended bolt prevents the door from being closed.
+
+The black dot defines the point of entry for this state machine, the door is closed but unlocked in the beginning.
+Through an event of the environment - in this case a human acting at the door - the door can be opened ,what makes ``opened'' the current state, and later be closed again.
+A turn of the key locks respectively blocks the door and no further opening or closing is possible.
+A second turn of the key un(b)locks the door again.
+
+The short description given in this section does not contain a complete explanation about the concept of statemachines at all, but it gives a basic overview which is needed as background information for this work.
+Also the formal definition of state machine is not relevant and can be read in other literature~\cite{statemachines, statemachinesandformalmethods}.
+
+%There is also a mathematical way of expressing state machines
+
+%mathematical model of computation
+%program behaviour
+%finite (endliche anzahl zustände)
+
+%States
+%Events
+%Transitions
+%Mathematical expression
+%Graphic
+%Incomplete description
 

### 2012-10-28 thesis/tex/2_background.tex
2.3.2 finished

@@
 The short description given in this section does not contain a complete explanation about the concept of statemachines at all, but it gives a basic overview which is needed as background information for this work.
-Also the formal definition of state machine is not relevant and can be read in other literature~\cite{statemachines, statemachinesandformalmethods}.
-
-%There is also a mathematical way of expressing state machines
+Also the formal definition of state machines is not relevant and can be read in other literature~\cite{statemachines, statemachinesandformalmethods}.
 
 %mathematical model of computation
-%program behaviour
 %finite (endliche anzahl zustände)
-
-%States
-%Events
-%Transitions
-%Mathematical expression
-%Graphic
-%Incomplete description
 
@@
 
-%CTL and LTL are important concepts for model checking and verification of statemachines.
-But their powerfulness is reflected in the complexity of their mathematical expressions, what makes it difficult to use for programmers who haven't got deeper expertise in this subject.
+Whenever a software or hardware system has a special requirement in robustness and reliability, different formal methods are available to be used in order to verify the design.
+Model checking is such a method which allows automatically verifying a system description against a specification by means of an exhaustive search of all states which could possibly be visited by the system during its execution.
+As a result, either the system description fulfills all propositions or a counter example is provided which gives evidence about a broken property.
+Figure~\ref{fig:modelchecking} gives a schematic view about the concept of model checking.
 
-%NuSMV erwaehnen, erklaeren warum nicht CTL? ltl passt fuer healthcare besser
+\begin{figure}[htbp]
+  \centering
+  \tikzstyle{state} = [circle,fill=white,draw]
+  \tikzstyle{stateblack} = [circle,fill=black!100,draw]
+  \tikzstyle{every path} = [font=\sffamily\small]
+  \definecolor{egg}{RGB}{252,253,206}
+
+	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=30mm]
+
+		\node[cloud, cloud puffs=15.7, cloud ignores aspect, align=center, draw, fill=egg] (model) at (0,5) {\parbox{2cm}{\centering system\\model}};
+
+		\node[ellipse, draw, fill=egg] (properties) at (5,5) {\parbox{2cm}{\centering system\\properties}};
+
+		\node[rectangle, draw, minimum width=2cm, minimum height=2cm, fill=egg] (checker) at (2.5,3) {\parbox{2cm}{\centering model\\checker}};
+
+		\node[state, minimum width=2cm, minimum height=2cm,fill=red!40] (no) at (0,0) {\parbox{1.3cm}{\centering counter\\example}};
+
+		\node[state, minimum width=2cm, minimum height=2cm,fill=green!40] (yes) at (5,0) {};
+
+		\path[]
+	  	(model)      edge [bend right] node {} (checker)
+	    (properties) edge [bend left]  node {} (checker)
+	    (checker)    edge [bend left]  node [above left] {not satisfied} (no)
+	    						 edge [bend right] node [above right] {satisfied} (yes)
+	    ;
+
+	\end{tikzpicture}
+  \caption{Concept of model checking.}
+  \label{fig:modelchecking}
+\end{figure}
+
+For representing the system description - which can be a computer program, for example - a formal language is used.
+Also state machines are applicable to serve as a formal system description.
+In contrast, the specification used for model checking contains certain properties that express the desired and safe system behaviour in form of mathematical formulas.
+Linear temporal logic (LTL) is an example for a logic that is used to define such formulas on state machines.
+It is a temporal logic which can express formulas concerning the future of paths, for example, that something will eventually happen.
+
+LTL is based on propositional logic, thus, it can express propositional variables ($p_1$, $p_2$, ...) and logical connectives ($\wedge$, $\vee$, $\neg$, $\rightarrow$ and $\leftrightarrow$).
+Additionally further temporal modal operators are provided which enable propositions concerning the future of paths, among three unary operators:
+
+\begin{description}
+	\item[$\Box \varphi$] This formula expresses that $\varphi$ is true now and in all following states. The meaning of the rectangle is ``GLOBALLY''.
+	\item[$\medcircle \varphi$] With a circle representing ``NEXT'', $\varphi$ will be true in the next state.
+	\item[$\Diamond \varphi$] At least one time - now or eventually in a later state - $\varphi$ is or will be true. The diamond stands for ``FUTURE''.
+\end{description}
+
+Furthermore, LTL provides binary operators as follows:
+
+\begin{description}
+	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
+	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true through the first occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
+	\item[$\varphi \mathcal{W} \psi$] ``WEAK UNTIL'' represents the same as ``UNTIL'', except that an occurrence of $\psi$	is not necessary. In this case, $\varphi$ is true infinetely.
+\end{description}
+
+%Verknüpfung von operatoren?
 

### 2012-10-30 thesis/tex/2_background.tex
Section 2.4 Related work now completely finished

@@
 
-A first step towards a graphical tool for designing safety constraints is taken by Sisiruca and Ionescu~\cite{332301}.
+%aim for simplifying the development of constraints, but they still demand knowledge of the underlying mathematical formalism of temporal logic.
+
+The idea of a visual environment for the defining of system properties is not new and already pursued by some researchers.
+Although their focus does not completely match the subject of safety for healthcare robots, they present concepts of visualisation which might be relevant
+% can be quite interesting 
+for this work.
+In section~\ref{sec:sisiruca}, the work of Sisirucá and Ionescu is presented.
+Section~\ref{sec:delbimbo} dexcribes a three dimensional concept by Del Bimbo et al., followed by an introduction of HomeTL in section~\ref{sec:hometl}.
+
+\subsection{Sisirucá and Ionescu}
+\label{sec:sisiruca}
+
+A first step towards a graphical tool for designing safety constraints is taken by Sisirucá and Ionescu~\cite{332301}.
 They developed an object-oriented graphical environment for visually creating temporal logic sentences and rules.
-The tool presents temporal operators as logical modules with boolean inputs and outputs which can be connected to one big expression graph.
-Del Bimbo et al.
-worked on a visual tool for temporal logic, but with a special focus on hierarchical representation of formulas~\cite{520786}.
-Here each node of a tree is described by an abstract operator whose subformulas are branches to their replacement operators.
-Thus, the root node represents the entire formula.
-They also explored the formula trees' 3D representation within a virtual space.
-Both approaches are different fundamental ideas of representing temporal formulas.
+As shown in figure~\ref{fig:sisirucaandionescu}, the tool presents all operators as blocks which have inputs and outputs. 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.6\linewidth]{sisirucaandionescu} 
+  \caption{The visual programming environment. An example showing the ruel construction process~\cite{332301}.}
+  \label{fig:sisirucaandionescu}
+\end{figure}
+Whereas logical operators such as \emph{AND} and \emph{OR} have their boolean output and inputs represented by arrows pointing to the right, generic blocks for temporal expressions additionally have an input symbolized by an arrow pointing downward where the temporal operator itself is to be attached.
+By adding and connecting the operators a big expression graph can be built.
+In case that a diagram becomes too complex, an implode feature allows to compress the graph to a single box.
+It is also applicable for the reuse of diagrams.
+
+The presented environment aims to be an effective tool for developing temporal logic sentences and rules.
+In fact, it gives a schematic overview of even complex formulas and allows substitution and reuse.
+The visual formalism is an alienation of LTL which might be suitable for researchers, but probably not for novices.
+
+%The tool presents temporal operators as logical modules with boolean inputs and outputs which can be connected to one big expression graph.
+
+\subsection{Del Bimbo et al. (3D)}
+\label{sec:delbimbo}
+
+Del Bimbo and his colleagues worked on a visual tool for temporal logic, too, but with a special focus on hierarchical representation of formulas~\cite{520786}.
+It is based on computation tree logic (CTL) which is a branching-time logic with support for quantifiers.
+In contrast to LTL where properties are fulfilled by all possible paths, CTL expresses properties which depend on the future branching, for example, that at least one path is existend which has a particular property.
+As indicated by figure~\ref{fig:delbimbo}, the visual tool provides all operators known from LTL 
+%The visual tool provides visual operators as shown in figure~\ref{fig:delbimbo}.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.8\linewidth]{delbimbo} 
+  \caption{Some of the visual operators of Del Bimbo et al.~\cite{520786}.}
+  \label{fig:delbimbo}
+\end{figure}
+except that temporal operators are extended by an existential or universal quantifier (``at least on one path...'', ``on all paths...'').
+
+Except propositions all operators are abstract and have placeholders representing their subformulas which can be connected to other operators.
+%Here each node of a tree is described by an abstract operator whose subformulas are represented by placeholders which can be connected to other operators.
+The positioning of these operators is organized in layers where the parent operator lays one level above its children. It can be compared to a tree upside down with the topmost operator as root.
+This formula structure can be visualized as a 3D branching tree within a virual space. Figure~\ref{fig:delbimbo3d} shows the result of considerations done by Del Bimbo and his colleagues.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1\linewidth]{delbimbo3d} 
+  \caption{Formula trees' 3D representation within a virual space~\cite{520786}.}
+  \label{fig:delbimbo3d}
+\end{figure}
+
+The presented visual formalism directly maps the syntax of the temporal logic onto graphic symbols and composition patterns.
+The researchers found well expressive graphical representations for all operator especially the temporal ones, what makes the tool fairly intuitive.
+The tree dimensional visualisation gives a good overview of the structure, although it shuld not be used for editing.
+Computer monitors can naturally display only two dimensions and input devices such as mouse and keyboard are customised to it.
+Editing of the third dimension needs special gesture definitions and 3D modelling techniques the most users might not be familiar with.
+Thus, a 3D representation for editing should only be used if necessary.
+This perception has been tried to be considered in this work.
+
+%Die visualisierungen der Operatoren sind aussagekräftig, da sie Zukunft von Pfaden andeuten.
+%A visual formalism has been presented which embeds the formal nucleus of this tool within a visual shell providing a one-to-one mapping of the syntax of the model onto graphic symbols and composition patterns.
+
+%By exploiting 3D graphics, this visual representation breaks the native sequentiality of textual expressions and provides an explicit and intuitive representation for the three natural dimensions of branching time Temporal Formulae, i.e.
+time progression, parallelism among execution paths, and condition nesting.
+
+%Both approaches are different fundamental ideas of representing temporal formulas.
 Whereas having a graph  seems not to be applicable for our approach, the idea of hierarchical operator representation of Del Bimbo et al.
@@
 
-%aim for simplifying the development of constraints, but they still demand knowledge of the underlying mathematical formalism of temporal logic.
+\subsection{HomeTL}
+\label{sec:hometl}
 
@@
 Based on these conditions a monitoring system in a patient's home environment can detect and report abnormal situations within the daily routine.
-Even though this approach touches the subject of healthcare it barely matches our problem.
+The visual notation provides propositional variables and logical connectives, the latter is shown in figure~\ref{fig:hometloperators}.
+In contrast to LTL, the temporal operators of HomeTL use absolute time limits (seconds, for example) which are defined by $\tau$.
+Thus, not only the relative order of states is defined but also a time by which a particular action has to happen.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\linewidth]{hometloperators} 
+  \caption{Graphical representation of visual notation for temporal
+operators within HomeTL user interface~\cite{4341725}.}
+  \label{fig:hometloperators}
+\end{figure}
+
+This notation uses special meanings for the two dimensions: The horizontal dimension shows the temporal evolution of states, while the vertical dimension expresses the logical relationships at the respective current state.
+A programming environment allows editing and creating of constraints by using the atomar operators.
+Figure~\ref{fig:hometleditor} presents a snapshot of this editor.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\linewidth]{hometleditor} 
+  \caption{HomeTL Visual Editor~\cite{4341725}.}
+  \label{fig:hometleditor}
+\end{figure}
+
+Even though this approach touches the subject of healthcare it barely matches the problem stated in this work.
 HomeTL focuses more on monitoring temporal boundaries of a patient's behaviour than on ensuring functional safety of an implemented program.
@@
 % Das wollen wir auch erzielen mit unseren automatisch generierten constraints...
-
 %But these kind of oracles are for automated test purpose and are rather complex than intuitive.
@@
 Finally the developer obtains a set of visual constraints which he can then check for plausibility.
-
-\subsection{Sisiruca and Ionescu}
-\subsection{Del Bimbo (3D)}
-\subsection{HomeTL}

### 2012-10-30 thesis/tex/4_visualformalismsforsafetyconstraints.tex
Section 2.4 Related work now completely finished

@@
 
-This visual language was presented to potential users for evaluation and feedback.
-Though most of them named it a good concept, soon it turned out that there are some problems and disagreements.
-The presented constraints were interpreted differently, the semantic seemed to be unclear and ambiguous.
-An attempt to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
-These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
-But unfortunately even the textual explanations do not seem to be a satisfying solution.
-For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
-
-However, there is a second problem with the template constraints.
-Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can't be described.
-Not even the check of a state being visited never or at least once is possible.
-This lack of expressiveness appears to be too big and for this reason the approach of template constraints got dismissed.
-
-%idee: einfache verwendung
-%sehr spezialisiert auf healthcare anwendung
-
-weitere constraint templates...
-
 \section{Operator constraint formalism}
@@
 	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state.
-Other types such as numeric or string equations may be added but are less important for our healthcare subject.
+Other types such as numeric or string equations may be added but are less important for the healthcare subject.
+\end{description}
 
@@
 
+\begin{description}
 	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
@@
   \centering
+  \includegraphics[scale=0.65]{directions}
+  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
+  \label{fig:directions}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
   \includegraphics[scale=0.65]{and_simplify} %width=0.7\linewidth
@@
 Merged nested \emph{AND} operators (center) are less complex as default ones (left).
-However nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
+However, nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
   \label{fig:nested_and}
@@
   \centering
-  \includegraphics[scale=0.65]{directions}
-  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
-  \label{fig:directions}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
   \includegraphics[scale=0.65]{results}
@@
 \end{figure}
+
+\section{Comparison}
+
+The visual language for template constraints was presented to potential users for evaluation and feedback.
+Though most of them named it a good concept, soon it turned out that there are some problems and disagreements.
+The presented constraints were interpreted differently, the semantic seemed to be unclear and ambiguous.
+An attempt to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
+These additional expressions are also shown in figure~\ref{fig:templateconstraints} of section~\ref{sec:templateconstraintformalism}.
+But unfortunately even the textual explanations do not seem to be a satisfying solution.
+For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
+
+However, there is a second problem with the template constraints.
+Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can not be described.
+Not even the check of a state being visited never or at least once is possible.
+This lack of expressiveness appears to be too big and for this reason the approach of template constraints was discarded.
+
+%In contrast, the operator constraints were successful.

### 2012-11-03 thesis/tex/4_visualformalismsforsafetyconstraints.tex
correction update, chapter 4 has a new structure

@@
 One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points.
-A first approach, so called template constraints, and its pros and cons are described in section~\ref{sec:templateconstraintformalism}.
-Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
+A first approach, so called template constraints,
+% and its pros and cons 
+are described in section~\ref{sec:templateconstraintformalism}. Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
+Section~\ref{sec:comparison} gives a comparison of the two formalisms presented before and states which one is used for the prototype.
 
@@
 These templates make different propositions about states and their relations.
-The user can choose from several graphical constraint types and customize them by parameterizing state variables.
+The user can choose from several graphical constraint types and customize them by parameterising state variables.
 So far there are two different kinds of template constraints:
@@
 Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled.
-Example of a constraint: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z).
-Going to bed without brushing teeth after meal should never happen.
+An example constraint for this template is: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z).
+Thus, going to bed without brushing teeth after meal should never happen.
 	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.
@@
 	%a product or not to buy it.
-	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can't avoid both possibilities.
+	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can not avoid both possibilities.
 	%Nitpicking readers might note 
 	Of course the person could also steal something without buying it, and thus not buy anything but leave the shop with full hands.
-Of course this is true; In fact, exactly in this case the constraint would signal an undesirable ``program'' behaviour with its invalidity.
+In fact, exactly in this case the constraint would signal an undesirable ``program'' behaviour with its invalidity.
 \end{enumerate}
@@
 Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot.
-Furthermore the state labels should sometimes be replaced by the screen displays of the regarding states.
-All parameters of the template constraints can be directly customized by the robot surface symbols.
-In the subsequently appearing popup window (see figure~\ref{fig:templateconstraint_dropdown}) all states contained in the underlying program are listed and can be chosen.
+Furthermore the state labels could be replaced by the real screen displays of the regarding states.
+All parameters of the template constraints shall be directly customizable by the robot surface symbols.
+Figure~\ref{fig:templateconstraint_dropdown} demonstrates how states could be choosen as parameters.
 
@@
   \includegraphics[scale=0.7]{templateconstraint_dropdown}
-  \caption{Dropdown menu for specifying template parameter.}
+  \caption{Possible way of specifying template parameter.}
   \label{fig:templateconstraint_dropdown}
 \end{figure}
+
+All template constraints are additionally equipped with textual expressions in order to explain their semantics.
+It is a template sentence with placeholders for the states available for parameterisation.
+Whenever there is a change to the constraint, te text is being updated automatically regarding to the parameter states.
+These additional expressions are also shown in figure~\ref{fig:templateconstraints}.
+
+Each template constraint describes a certain behaviour which might be later used for checks.
+Thus, specific formulas have to be provided which represent the semantics of the constraints.
+They have to be in a format so that they can be processed by a model checker.
+For the first kind of template operator, a linear temporal logic fomula (LTL, see~section~\ref{sec:modelcheckingandltl}) expresses its semantics (with $x$ as ``x is current state'' and similarly for $y$ and $z$):
+
+\begin{equation} \label{eq:template_one}
+  \models \Box (x \rightarrow ([(\neg z) \mathcal{U} y] \vee \Box \neg z))
+\end{equation}
+
+``Globally is true: Whenever x is the current state, either z is not visited until y or z is never visited''.
+Compared to the example above it means: Whenever a person eats something (x), he can not go to bed (z) until he brushes his teeth (y).
+Otherwise, if he never goes to bed, he does not necessarily need to brush his teeth.
+
+The semantics of the second operator template is more complex since it has a secondary condition that all y, z, ...
+states are actually reachable after a visit of x.
+Possibilities can not be expressed by LTL, thus, it needs to use computation tree logic (CTL) for describing this auxiliary condition in a second formula.
+A short explanation about CTL is already given in section~\ref{sec:delbimbo}, but the two new constructs used for this formula are described below.
+The main formula for the second template is still using LTL:
+\begin{equation} \label{eq:template_two_one}
+  \models \Box (x \rightarrow \Diamond (y \vee z (\vee ...)))
+\end{equation}
+``Globally is true: Whenever x is the current state, in a future step either y or z (or ...) will be visited''.
+In context of the example mentioned above it means: Every time a person enters a supermarket (x) he will eventually buy something (y) or buy nothing (z).
+The fact that both states are really potentially reachable is expressed by the following CTL formula.
+\emph{AG} replaces the $\Box$ sign and stands for ``on all paths is globally true...'', \emph{EF} is related to $\Diamond$ and expresses ``a path exists where is true in future...''.
+\begin{equation} \label{eq:template_two_two}
+  \models AG (x \rightarrow ((EF y) \wedge (EF z) (\wedge ...)))
+\end{equation}
+``It is globally true on all pahts: Whenever x is the current state, there is at least one path leading to y and at least one path leading to z''.
 
@@
 
-With our second approach we try to gain more expressiveness and uniqueness for the visual language.
-For this we decided to decrease the abstraction level and design a visual formalism for linear temporal logic (LTL, see~section~\ref{sec:modelcheckingandltl}).
-Since the visual constraints can be directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness.
-If it is considered as easy to use,
-%two birds get killed with one stone: N
-non-experts can use it as well as experts.
+The operator constraint formalism is a new approach with a special focus on expressiveness and uniqueness of the visual language.
+%For this the abstraction level is adapted in order to 
+It is designed as a visual formalism for linear temporal logic having a visual operator for each LTL operator.
+Since the visual constraints are directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness.
+If it is considered as easy to use, non-experts can use it as well as experts who already know LTL.
 
-% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv. der always operator könnte umgedreht sein.
+\subsection{Requirements}
 
-%The presented visual language is designed to express constraints for state machines.
-To provide the required functionality, the fundamental logical operators are required, as shown in (a) though (e) below.
+To provide the required functionality, the fundamental logical operators are required, as shown in (a) through (e) below.
 
@@
 	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
-	\item[(c)] \emph{IF} operator: $\varphi \Rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
+	\item[(c)] \emph{IF} operator: $\varphi \rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
 Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
@@
 
-In addition the visual language should be capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i).
+In addition to these five logical operators the visual language should be furthermore capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i).
 
@@
 	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
-	%\Box \Diamond \medcircle
 	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
 	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
-	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true until there is a state with $\psi$ being true.
+	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true at least until there is a state with $\psi$ being true.
 In addition eventually there has to be a future state with $\psi$ being true.
@@
 
+All the operators mentioned above shall be suitable for constraint editing in a way that the visual language is very easy to use, especially for non-experts.
+This implies simple editing as well as easy readability of constraints.
+Furthermore it should provide an intuitive application which is ideally learnable within little time.
+
+\subsection{Design decisions}
+
 Like the formula representation introduced by Del Bimbo et al.~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy.
-However we surrender the three dimensional idea and focus on two dimensional blocks and their compositions instead, what is also used by HomeTL~\cite{4341725}.
-Each block represents either an operator or a proposition and has its specified semantics.
-Thus a visual formula can be recursively translated to a linear temporal logic (LTL) sentence and be checked using any ordinary model checker.
+However, the new visual formalism surrenders the three dimensional idea and focuses on two dimensional blocks and their compositions instead, which is used by HomeTL~\cite{4341725} as well.
+Each operator of the visual language is represented by such a block.
 
-The visual language has support for eight operator types and one proposition. A suggestion for their graphical representation is shown in figure~\ref{fig:operators}.
+The visual language has support for eight unary or binary operator types and one proposition which have their specific semantics.
+In order to support the reading of constraints, different significant visual appearances of the operator types shall enable easy classification and a faster subconscious identification~\cite{moody-physics-of-notations}.
+In this case, different colors are used for each operator type.
+% aiming for faster perceptibility and understandability. 
+A suggestion for the operators' graphical representation is shown in figure~\ref{fig:operators}.
 
@@
   \includegraphics[scale=0.65]{table}
-  \caption{The visual language consists of eight unary or binary operator types and one state proposition, each of them having its own color flavor.
-Logical operators are arranged in a vertical row wereas time relevant statements are aligned horizontally.}
+  \caption{Visual representation of all supported operators.}
   \label{fig:operators}
@@
 
-To meet the requirements the visual language should be very easy to use, especially for non-experts.
-Furthermore it should provide an intuitive application which is ideally learnable within little time.
-So there are some design decisions this visual language must fulfill~\cite{moody-physics-of-notations}:
+A second effort for making the perceptibility of the visual language as intuitive as possible is done with the orientation of constraints. 
+Operators which consist of nested operators have particular meanings for the two dimensions: The ``logical'' vertical read direction and the horizontal direction for variation in time.
+Thus, logical operators are arranged in a vertical row wereas time relevant operators are aligned horizontally.
+Figure~\ref{fig:directions} demonstrates an example constraint composed along the logical and time axes.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[scale=0.65]{directions}
+  \caption{Easy understanding of visual constraints due to intuitive read directions.}
+  \label{fig:directions}
+\end{figure}
+It is read from left to right and top to bottom:
+``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''
 
-\begin{itemize}
-	\item Each operator has a different significant visual appearance; in this case we have used color. This kind of classification enables a faster subconscious identification.
-	\item The layout of constraints (nested operators) has two dimensions with a certain meaning.
-Figure~\ref{fig:directions} demonstrates the ``logical'' vertical read direction as well as the time releveant horizontal line.
-	\item Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear unnecessarily complex.
-These operators can be merged (see~figure~\ref{fig:nested_and}).
-\end{itemize}
+The two dimensions form a special requirement also for the layouting of nested operators.
+Each operator adapts its size so that it surrounds all recursively nested operators.
+But also the positioning of nested operators within teir parents have to be taken into account.
+Whereas the operator's centered positioning and resizing along the logical direction only depends on the sizes of all sub operators, the arrangement within the time axis turns out to be more complex.
+Here, sub operators can not just be aligned centered as it is the case vertically.
+To face the determination of horizontal positioning, every operator gets time reference lines which specify its positioning within its parent and the positioning of its own children within itself.
+As depicted in figure~\ref{fig:referencelines}, logical operators - i.e.
+\emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator} and \emph{StateOperator} - have one reference line.
+Time relevant operators - i.e.
+\emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator} and \emph{UntilOperator} - have two such reference lines, one for their own positioning within their parents and one for the positioning of their children.
 
@@
   \centering
-  \includegraphics[scale=0.65]{directions}
-  \caption{Easy understanding of visual constraints due to intuitive read directions: ``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''}
-  \label{fig:directions}
+  \includegraphics[scale=0.65]{referencelines} 
+  \caption{Time reference lines and their adjustments.}
+  \label{fig:referencelines}
 \end{figure}
+
+As postulated in the requirements for the visual language, constraints shall be easy to read.
+In fact, there is a possibility to improve the presentation of certain visual constraints.
+Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear graphically unnecessarily complex.
+These operators can be visually merged in the manner of unifying their border lines, for example.
+As a result, all associated operators appear as just one operator with a couple of sub operators (see~figure~\ref{fig:nested_and}).
 
@@
   \centering
-  \includegraphics[scale=0.65]{and_simplify} %width=0.7\linewidth
-  \caption{In order to simplify the appearance of constraints certain operators might be merged.
-Merged nested \emph{AND} operators (center) are less complex as default ones (left).
-However, nested \emph{AND} operators with mouse over (right) are displayed separated in order to retain full editability.}
+  \includegraphics[scale=0.65]{and_simplify}
+  \caption{Default nested \emph{AND} operator and visually merged nested \emph{AND} operator.}
   \label{fig:nested_and}
@@
 
-Also an editor for this visual language can contribute to usability and clarity. Some requirements are as follows:
+\subsection{Guidelines for an editor}
+
+Also an editor for this visual language which holds the functionality for composing operators can contribute to usability and clarity. Some requirements are as follows:
 
@@
 Operators and propositions can be moved and dropped onto other operators.
-A tool bar provides all operator and proposition types to be instantiated and there is a trash can for deleting existing operators.
-	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer.
-All higher operators become bleached out so that the hovered operator and its children appear highlighted.
-It's intended to be similar to line coloring when reading a digital text.
+A tool bar provides all operator and proposition types to be instantiated and a possibility for deleting existing operators.
+	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer. It is intended to be similar to line coloring when reading a digital text.
 	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
-	\item There are different statuses which inform about validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed.
-Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign (see~figure~\ref{fig:results}).
 	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances.
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.65]{results}
-  \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
-  \label{fig:results}
-\end{figure}
+\section{Comparison}
+\label{sec:comparison}
 
-\section{Comparison}
+Due to its arrow based design the visual language for template constraints is a good visualisation of time dependencies.
+Furthermore, it allows expressing possibilities by the use of CTL formulas which is not possible with the operator costraint formalism.
+Nevertheless, the template visualisations have some trouble with clearness of their semantics.
+Even though there are well defined formulas behind each constraint and textual expressions which explain the meanings of constraints, the visual presentation seems to be ambiguous in some settings what leads to the risk of being interpreted differently.
+For example, in the context of the visual language it is unclear whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
+It might also be problematic to interprete whether more than one of the states $y$, $z$, ...
+can be actually visited after a visit of $x$ in the second constraint.
+In contrast, the operator constraint formalism is not ambiguous due to to its direct mapping to LTL wich is consistently defined.
+All visual operator expressions represent exactly the same semantics as their corresponding textual LTL formulas.
 
-The visual language for template constraints was presented to potential users for evaluation and feedback.
-Though most of them named it a good concept, soon it turned out that there are some problems and disagreements.
-The presented constraints were interpreted differently, the semantic seemed to be unclear and ambiguous.
-An attempt to solve this problem was to equip both constraint templates with textual explanations which are updated automatically whenever there is a change in parameterization.
-These additional expressions are also shown in figure~\ref{fig:templateconstraints} of section~\ref{sec:templateconstraintformalism}.
-But unfortunately even the textual explanations do not seem to be a satisfying solution.
-For example, one question came up whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
+%The visual language for template constraints was presented to potential users for evaluation and feedback.
+Though most of them named it a good concept, soon it turned out that they had problems and disagreements concerning the issues mentioned above.
+The presented constraints were interpreted differently, the meaning seemed to be unclear and ambiguous.
 
 However, there is a second problem with the template constraints.
-Whereas lots of safety constraints specific to the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are a lot of other common constraints which have no matching template and thus can not be described.
-Not even the check of a state being visited never or at least once is possible.
-This lack of expressiveness appears to be too big and for this reason the approach of template constraints was discarded.
+Whereas the safety constraints specific to the example within the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are other common constraints which have no matching template and thus can not be described.
+For example, the check of a state being visited never or at least once is not possible.
+For every new constraint type a new template has to be created what means an infinite amount of templates is needed in order to achieve full expressiveness.
+On the contrary, the visual language for operator constraints is not limited in expressiveness compared to LTL, disregarding that there is only one proposition type available.
+But this is an extension issue of the prototype rather than a lack in expressiveness.
+% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv. der always operator könnte umgedreht sein.
 
-%In contrast, the operator constraints were successful.
+Since a visual language has to be developed which suits non-professionals as well as experts, the risk of ambiguity is undesirable.
+Especially for a tool for safety and consistency it would form a really bad basis.
+For this reason, the approach of template constraints is discarded and the operator constraint formallism is used for the editor prototype instead.
+Owing to the recursive nesting of operator constraints, a visual formula can be recursively translated to a LTL sentence which can be processed by the prototype using any ordinary model checker.

### 2012-11-03 thesis/tex/6_theltlcreatorprototype.tex
correction update, chapter 4 has a new structure

@@
 
-Swing also provides a mechanism for layouting components, that means determining the aspired sizes and positions of components depending on child components.
+%Swing also provides a mechanism for layouting components, that means determining the aspired sizes and positions of components depending on child components.
 As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for variation in time.
 This fact forms a special requirement for the layouting of operators.
-Whereas the operator's positioning and resizing along the logical direction only depends on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult.
+%Whereas the operator's positioning and resizing along the logical direction only depends on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult.
 Here sub operators can't just be aligned centered as it is the case vertically.
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[scale=0.65]{referencelines} 
-  \caption{Time reference lines and their adjustments.}
-  \label{fig:referencelines}
-\end{figure}
+%\begin{figure}[htbp]
+%  \centering
+%  \includegraphics[scale=0.65]{referencelines} 
+%  \caption{Time reference lines and their adjustments.}
+%  \label{fig:referencelines}
+%\end{figure}
 
@@
 trash...
+
+TODO
+
+trash can
+
+%In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer.
+All higher operators become bleached out so that the hovered operator and its children appear highlighted.
+%It is intended to be similar to line coloring when reading a digital text.
+
+%As postulated in the requirements for the visual language, constraints shall be easy to read.
+In fact, there is a possibility to improve the presentation of certain visual constraints.
+Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear graphically unnecessarily complex.
+These operators can be visually merged in the manner of unifying their border lines, for example.
+As a result, all associated operators appear as just one operator with a couple of sub operators (see~figure~\ref{fig:nested_and}).
+
+In order to retain full editability, hovered operators are still displayed separated (see~figure~\ref{fig:nested_and_mouseover}).
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[scale=0.65]{and_simplify_mouseover}
+  \caption{Default nested \emph{AND} operator, visually merged nested \emph{AND} operator and visually nested \emph{AND} operator with mouse over.
+  }
+  \label{fig:nested_and_mouseover}
+\end{figure}
 
@@
 
+TODO
+
+There are different statuses which inform about validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed.
+Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign (see~figure~\ref{fig:results}).
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[scale=0.65]{results}
+  \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
+  \label{fig:results}
+\end{figure}
+
 \subsection{Model checker}
@@
 As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
-This demands a concept of change notification management which is responsible for receiving all changes and making this information available as a service.
-Thus we tried to extend all parts of the code where model changes are made with short code snippets.
-However this is not as easy to realise in Robostudio since some of the relevant code was generated or imported as external jars what forbids code manipulation.
+This demands a concept of change notification management which is responsible for receiving all changes and making this information available for the \emph{LTLCreator} component.
+As a solution, all parts of the code where model changes are made should be extended with short code snippets which send notifications.
+However, this is not as easy to realise in Robostudio since some of the relevant code was generated or even imported as external jars which makes code manipulation impossible.
 

### 2012-11-03 thesis/tex/7_testscenarioandevaluation.tex
correction update, chapter 4 has a new structure

@@
 Finally two states 'Well done!' and 'Staff notified' get added to the disjunction.
-The resulting visual constraint can now automatically be translated to the corresponding LTL formula by the editor:
+As soon as the visual constraint is complete, it is translated to the following corresponding textual LTL formula by the editor:
 
 \begin{equation} \label{eq:sampleconstraint}
-  \models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'}))
+  \models \Box (\textnormal{'Not taken yet'} \rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'}))
 \end{equation}
@@
 
-However, in case that this algorithm is used for other huge programs regardless its tailoring to the healthcare domain, it keeps a lot of potential to be improved in efficiency.
+However, this algorithm for subgraph finding is not sophisticated regarding efficiency.
+It is rather a prototype to show the concept of automated constraint generation working and to provide this feature in the editor for evaluation.
+In case that this algorithm shall be used for other huge programs regardless its tailoring to the healthcare domain, the author is convinced that a more efficient algorithm for subgraph finding might be found by researchers.
+
+%it keeps a lot of potential to be improved in efficiency.
 With a sophisticated approach it might be enough to execute just one depth-first search for finding all subgraphs at once what would lead to a linear execution time.

### 2012-11-07 thesis/tex/1_introduction.tex
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.

@@
 An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
-Scientists are working on applications such as the medication reminder~\cite{p11:_feasib} which run on mobile robots serving in nursing homes for the elderly.
+Scientists are working on applications such as the medication reminder~\cite{p11:_feasib} which run on mobile robots (so called healthbots) serving in nursing homes for the elderly.
 Equipped with a touch display and several external devices such as a blood pressure measurement tool the robots aim to support and entertain the elderly in their daily activities~\cite{jayawardena12:_desig,5649910}.
@@
 ISO 13482~\cite{iso13482, iso13482-2} is still under development and is intended to be a harmonised European standard for safety requirements for robots in personal care applications.
-However, considering functional safety while developing service robots is both complex and difficult; maybe one reason for the absense of safety checks and warranties in the current healthbot versions at the University of Auckland.
+However, considering functional safety while developing service robots is both complex and difficult; maybe one reason for the absense of safety checks and assurances in the current healthbot versions at the University of Auckland.
 But facing the fact that safety requirements for service robots are expected to increase in future, there is a plan of slowly integrating safety functionality into the healthcare robotics project as a next step starting with the concepts introduced in this paper.
@@
 
-Unfortunately, supplying an easy-to-use editor for constraint creation doesn't guarantee adequate thinking about sufficent constraints by the developer.
+Unfortunately, supplying an easy-to-use editor for constraint creation does not guarantee adequate thinking about sufficent constraints by the developer.
 Unmotivated or just unexperienced developers may miss important constraints needed in order to achieve a certain safety level.
@@
 
-Both ideas have been integrated into Robostudio in order to simplify respectively facilitate the use of safety mechanisms for application development for service robots.
-Thereby the desired safety certification for the healthcare service robots can move closer one step.
+Both ideas, the viual language and the automated constraint generation, have been developed and integrated by the author into Robostudio in order to simplify respectively facilitate the use of safety mechanisms for application development for service robots.
+Thereby the desired safety certification for the healthcare service robots can move closer by one step.
 
 This work starts with background information about the healthcare domain, short explanations of common safety mechanisms and insights into related work in chapter~\ref{chap:background}.
-After the goals have been mentioned in chapter~\ref{chap:goals}, the visual formalisms and the automated constraint generation are introduced in chapters~\ref{chap:visualformalismsforsafetyconstraints} and~\ref{chap:automatedgenerationofsafetyconstraints}.
+After the goals have been described in chapter~\ref{chap:goals}, the visual formalisms and the automated constraint generation are introduced in chapters~\ref{chap:visualformalismsforsafetyconstraints} and~\ref{chap:automatedgenerationofsafetyconstraints}.
 A prototype implementation is demonstrated in chapter~\ref{chap:theltlcreatorprototype}, and a test scenario is described in chapter~\ref{chap:testscenarioandevaluation} where the presented tools get evaluated.

### 2012-11-07 thesis/tex/2_background.tex
correction update, chapter 4 has a new structure
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.

@@
 For the understanding of safety concepts which this work is based on, a short introduction about functional safety is given in section~\ref{sec:behaviourchecking}.
-In section~\ref{sec:relatedwork} other work and programs related to this work are examined.
+In section~\ref{sec:relatedwork} other works and programs related to this work are examined.
 
@@
 One reason for this might be below average low salaries for employees in the healthcare sector.
-But, just raising the salary isn't leading to the desired results since the majority of the older people can't even afford a nurse to today's prices.
+But, just raising the salary will not lead to the desired results since the majority of the older people can not even afford a nurse to today's prices.
 
-Aiming for fillig the constantly increasing gap in professional personnel for healthcare of the elderly and bypassing the financial burden, service robots are supposed to more and more undertake the task of care.
+Aiming for fillig the constantly increasing gap in professional personnel for healthcare of the elderly and bypassing the financial burden, service robots are supposed to undertake more and more the task of care.
 There are lots of possible tasks a robot could do such as taking the blood pressure, guiding medication intake, detecting falls or just entertaining with music, video or internet.
-This challenge as a goal, the healthcare robotics team at the University of Auckland researches this topic and builds service robots for supporting the elderly in their daily life.
+This challenge as a goal, the healthcare robotics team at the University of Auckland researches this topic and develops service robots for supporting the elderly in their daily life.
 
@@
   \includegraphics[width=0.56\textwidth]{irobiq_cafero}
-  \caption{IrobiQ and Cafero, two robots of the healtcare project.}
+  \caption{IrobiQ and Cafero, two robots of the healthcare project.}
   \label{fig:irobiq_cafero}
@@
 robots are equipped with a touch display for presenting screens and receiving user input. Furthermore, they are equipped with additional external tools such as the blood pressure measurement device.
-Applications running on the robot use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedback.
-User inputs or response from devices cause dialogs to change.
+Applications running on the robots use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedback.
+User inputs or responses from devices cause dialogs to change.
 Figure~\ref{fig:screenflow_example} presents a small example of such a screen-flow based application, which uses response from a camera and user input for navigating through the screen-flow.
@@
 The interpreter processes programs written in Robot Behaviour Description Language (RBDL), a domain specific language (DSL) based on XML, which has been developed particularly for the healthcare robotics applications.
-It allows defining state machine based program behaviour and creating UI elements for screen visualizations.
+It allows defining state machine based program behaviour and creating user interface (UI) elements for screen visualizations.
 An example implementation of one state in RBDL is shown in listing~\ref{lst:RBDL}.
@@
   \includegraphics[width=\linewidth]{screenflow_example} 
-  \caption{Robot screen-flow dialog based interaction example~\cite{robostudio}.}
+  \caption{Interaction example based on robot screen-flow dialogs~\cite{robostudio}.}
   \label{fig:screenflow_example}
@@
 
-\begin{lstlisting}[float = htbp, language=XML, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a state defined by RBDL~\cite{robostudio}., label=lst:RBDL]
+\begin{lstlisting}[float = htbp, language=XML, captionpos=b, breaklines=true, showspaces=false, showtabs=false, tabsize=2, caption=Example of a state defined with RBDL~\cite{robostudio}., label=lst:RBDL]
 <state no="84">
@@
 
-For this work the most important thing about the RBDL to remember is that it can define state machine behaviour with states, transitions and events.
-This fact was the basis for later decisions regarding used safety mechanisms.
+Very important for this work is a fundamental feature of RBDL, the ability to define state machine behaviour with states, transitions and events.
+This fact was the basis for later decisions regarding the applied safety mechanisms.
 
@@
 In the beginning all programming of the robot behaviour was done by directly editing the XML code mentioned in the previous section by using a text editor.
-But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise.
-For this reason, Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
+However, huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise.
+For this reason, Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL.
+This environment- it is called Robostudio~\cite{robostudio} - allows to edit easily the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
 This file can be downloaded to the robot and executed by the interpreter.
@@
 Components such as buttons, text boxes or images can be added, changed or removed via drag and drop.
-They are provided by the \emph{UI Components Palette Window} (4) on the right side where the user can choose from all supported UI components.
-The \emph{Background Actions Editor Window} (5) lets the user define background actions, which are executed transparently to the user such as accessing external devices or sending messages.
+They are provided by the \emph{UI Components Palette Window} (4) on the right side where the developer can choose from all supported UI components.
+The \emph{Background Actions Editor Window} (5) lets the developer define background actions, which are executed transparently to the developer such as accessing external devices or sending messages.
 Aiming to support the developer with helpful graphical tools, the \emph{State Transition Visualization Window} (6) gives an overview about all connections towards and from the selected state.
-It helps to quickly get an understanding of the local program behaviour.
+It helps to get quickly an understanding of the local program behaviour.
 
@@
 
-The use of software is versatile from text editing programs running on personal computers, over control software for high-precision applications in aeronautics, right up to trigger mechanisms for deathly weapons.
-For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also human lifes.
+The use and application of software is versatile and varies from text editing programs running on personal computers, over control software for high-precision applications in aeronautics, right up to trigger mechanisms for deathly weapons.
+For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also to human lifes.
 In order to eliminate software malfunction there are different mathematical concepts for verifying software behaviour.
-One of them is model checking wich allows checking state machine programs for functional safety.
+One of them is model checking which allows checking state machine programs for functional safety.
 Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realised wth state machines, this concept of behaviour checking is applicable and important for our kind of programs.
-At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an intorduction into model checking in section~\ref{sec:modelcheckingandltl}.
+At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an introduction into model checking in section~\ref{sec:modelcheckingandltl}.
 
@@
 \begin{description}
-	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
+	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true at least until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
 	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true through the first occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.

### 2012-11-07 thesis/tex/4_visualformalismsforsafetyconstraints.tex
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.

@@
 \begin{description}
-	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
-	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
+	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.\\This conjunction evaluates to true only when both $\varphi$ and $\psi$ are true.
+	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.\\At least one $\varphi$ or $\psi$ being true makes this construct true. The result is false when both parameters are false.
 	\item[(c)] \emph{IF} operator: $\varphi \rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
 Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
-	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
+	\item[(d)] \emph{NOT} operator: $\neg \varphi$. This operator negates the value of $\varphi$.
 	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state.
@@
 \subsection{Design decisions}
+\label{sec:designdecisions}
 

### 2012-11-07 thesis/tex/6_theltlcreatorprototype.tex
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.

--- 2012-10-27 thesis/tex/6_theltlcreatorprototype.tex
@@
 
-First of all some details about the implementation of the visual language are given in section~\ref{sec:prototype:operatorconstraints}, directly followed by an explanation about the realization of the automated constraint generation functionality in section~\ref{sec:prototype:automatedconstraintgeneration}.
+First of all the model implementation for representing state machines is described in section~\ref{sec:statemachinemodel}.
+Then, some details about the implementation of the visual language are given in section~\ref{sec:prototype:operatorconstraints}, directly followed by an explanation about the realization of the automated constraint generation functionality in section~\ref{sec:prototype:automatedconstraintgeneration}.
 The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}.
@@
 Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio is possible.
+
+\section{State machine model}
+\label{sec:statemachinemodel}
+
+For the checking of constraints as well as for the constraint generation a model is needed which represents the state machine program.
+Three classes are responsible for holding the state machine program with all its necessary information: The class \emph{Fsm} represents the finite state machine and lists all contained states.
+It is called finite state machine, because the number of states is finite which applies to the domain of this work.
+\emph{State} represents the states of the machine and holds all transitions from and to the respective state.
+Transitions are represented by the class \emph{Transition} which specifies the source and a target state.
+Figure~\ref{fig:fsm_class} shows all three class definitions for the state machine model.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.6\linewidth]{fsm_class} 
+  \caption{State machine model class definitions.}
+  \label{fig:fsm_class}
+\end{figure}
 
@@
 	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object, depending on the implementation.
-	\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
-	\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function \emph{paintOperator()}.
+	%\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
+	%\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function \emph{paintOperator()}.
 It manages all color and shape painting for the respective operator.
 Additional parameters such as recursive painting including children can be specified.
-	\item[contains(int, int):] Given two coordinates x and y, it can be determined whether the herewith defined point lays within an operator.
+	%\item[contains(int, int):] Given two coordinates x and y, it can be determined whether the herewith defined point lays within an operator.
 	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality.
 Equality is given when the compared operators are of the same type and similarly their children.
-This functionality is important for avoiding duplicate constraints during constraint generation.
+This functionality is important for avoiding duplicate constraints during constraint generation, for example.
 	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it.
@@
 
-Whereas \emph{StateOperator}s have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two possible children respectively.
-The graphical representation of both unary and binary operators thus needs to have place holders and docking stations for sub operators.
-These are realized by buckets which display a small ``drop here'' label and allow operator adding or removing by mouse drops.
-For this a class Bucket is defined with get and set methods for defining a sub operator.
+Whereas \emph{StateOperator}s have no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two possible nested children respectively.
+The graphical representation of both unary and binary operators thus needs to have place holders and docking stations
+\begin{wrapfigure}{r}{0.40\textwidth}
+  \centering
+  \includegraphics[scale=0.65]{bucket_example.jpg}
+  \caption{Buckets for graphical nesting of operators.}
+  \label{fig:bucket_example}
+\end{wrapfigure}
+for sub operators. These are realized by buckets which display a small ``drop'' label as shown in figure~\ref{fig:bucket_example} and allow operator adding or removing by mouse drops.
+For this a class \emph{Bucket} is defined with get and set methods for defining a sub operator.
 Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between \emph{AbstractOperator}, \emph{Bucket} and operator implementations.
+Once operators are composed in a way that no empty buckets are left over, a complete constraint such as the one shown in figure~\ref{fig:operator_tree} is retrieved.
+This diagram illustrates the correlation of nested operators and buckets.
 \begin{figure}[htbp]
@@
 \end{figure}
-Once operators are composed in a way that no empty buckets are left over, we retrieve a complete constraint such as the one shown in figure~\ref{fig:operator_tree}.
-This diagram illustrates the correlation of nested operators and buckets.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{operator_tree} 
+  \includegraphics[width=0.95\linewidth]{operator_tree} 
   \caption{Instantiation of a constraint with several operators.}
@@
 
-All graphical components such as buckets and operators are implemented as Java Swing \emph{JComponent}s.
+For the prototype implementation, all graphical components such as buckets and operators are implemented as Java Swing \emph{JComponent}.
 Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.
@@
 
-For lightweight components Swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component.
+Swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component.
 This is used for giving every operator its special look.
-We hereby made use of Swing's \emph{Graphics2D} functionality what gives a great support in painting rounded shapes and color gradients.
+Swing's \emph{Graphics2D} functionality provides a great support in painting rounded shapes and color gradients.
 An operator is displayed as a rounded rectangle with dents around sub operators and has a flowing backgound, borders and labels in an operator type specific color.
-For the significant shape multiple rounded rectangles with and without border are arranged in a way that the result appears to be one piece.
+For the significant shape multiple rounded rectangles with and without borders are arranged in a way that the result appears to be one piece.
 Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the \emph{IfThenOperator}, first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be.
@@
 
-Swing also provides a mechanism for layouting components, that means determining the aspired sizes and positions of components depending on child components.
-As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for variation in time.
-This fact forms a special requirement for the layouting of operators.
-Whereas the operator's positioning and resizing along the logical direction only depends on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult.
-Here sub operators can't just be aligned centered as it is the case vertically.
-To face the determination of horizontal positioning, every operator gets time reference lines as depicted in figure~\ref{fig:referencelines}.
-Logical operators, i.e.
-\emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator} and \emph{StateOperator}, have one reference line, time relevant operators, i.e.
-\emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator} and \emph{UntilOperator}, have two such reference lines.
+As proposed in the requirements for the visual language in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: A vertical read direction and the horizontal direction for variation in time.
+Swing's mechanism for layouting components is used for the positioning of operators.
+After determining the aspired sizes of child components which represent sub operators, reference lines which have been already demonstrated with figure~\ref{fig:referencelines} in section~\ref{sec:operatorconstraintformalism} are computed mathematically.
+Based on this information operators are sized and positioned correctly.
 
+%MOUSE OVER
+
+Regarding to the requirement for simple reading of constraints based on mouse hovering, the editor prototype supports context sensitive highlighting of particular constraint parts.
+The built-in mouse support of Swing is used for identifying the operator positioned under the mouse cursor.
+All operators except the hovered one and its sub operators are presented bleached out what causes the hovered operator to appear emphasized.
+The most right operator composition in figure~\ref{fig:nested_and_mouseover} gives a demonstration of this idea.
+
+%PROBLEM WITH MOUSE OVER AND MERGING
+
+However, the concept of operator highlighting causes trouble in combination with the graphical merging of visual operators which was introduced in section~\ref{sec:designdecisions}.
+Since a merged operator does not have its own border, it can not be displayed as desired when hovered.
+This fact is also a problem for the editing: The graphical presentation does not support changing of merged operators, because it seems that they can not be picked by the mouse.
+In order to retain full editability, hovered operators are excluded from the merge rule and still have to be displayed separated (see~figure~\ref{fig:nested_and_mouseover}).
 \begin{figure}[htbp]
   \centering
-  \includegraphics[scale=0.65]{referencelines} 
-  \caption{Time reference lines and their adjustments.}
-  \label{fig:referencelines}
+  \includegraphics[scale=0.65]{and_simplify_mouseover}
+  \caption{Default nested \emph{AND} operator, visually merged nested \emph{AND} operator and visually nested \emph{AND} operator with mouse over.
+  }
+  \label{fig:nested_and_mouseover}
 \end{figure}
@@
 
-The built-in mouse support of Swing was used for identifying operators positioned under the mouse cursor.
-On the one hand this is important for context sensitive highlighting of particular constraint parts, on the other hand this way the position where operators have to be added after drag and drop can be determined.
-So far
+Besides its importance for mouse hovering, Swing's mouse support is also used for determining the position where operators have to be added after drag and drop edit actions. So far
 % the predefined Swing drag and drop has not been used but
-a simple drag and drop mechanism has been implemented whose class structure is shown in figure~\ref{fig:dndclassstructure}.
+a simple drag and drop mechanism called \emph{DndHandler} has been implemented whose class structure is shown in figure~\ref{fig:dndclassstructure}.
 \begin{figure}[htbp]
@@
 Once a \emph{JComponent} is registered to the \emph{DndHandler}, it automatically manages any drag and drop actions with \emph{JComponent}s which implement either the \emph{Draggable} or \emph{DropTarget} interface.
-The former is implemented by all operators, the latter by buckets and the dashboard.
+The former is implemented by all operators and specifies that this object can be moved from and to buckets or the dashboard, the latter is implemented by buckets and the dashboard and determines that it can accept and contain \emph{Draggable}s such as operators.
 %Figure~\ref{fig:dndclassstructure} gives a short class overview.
-Furthermore there is a class \emph{DraggableProvider} which displays a kind of button in the tool bar and lets the user create new \emph{Draggable}s; in our case new operators get instantiated.
+Furthermore there is a class \emph{DraggableProvider} which is similar to \emph{Draggable} except that it is not moved itself but creates a new instance of the respective \emph{Draggable} implementation type which is then moved instead.
 
-%operator werden im dashboard, einem weiteren container, zusammengeklickt und angezeigt.
-Das Dashboard hat dabei eine ähnliche Funktionsweise wie ein Bucket.
-die operatoren können dabei vom operatorprovider bezogen werden: Implementierungsdetails?
-trash...
+%TOOLBAR
+
+\emph{LLTCreator} is equipped with a tool bar which provides all functionality needed for constraint editing: For each operator type a \emph{DraggableProvider} is displayed as an icon in the respective operator color.
+Each time the mouse is pressed on it (drag action start), a new instance of the respective operator type is created which can be moved and dropped to the dashboard.
+Additionally, a trash can icon is displayed within the tool bar.
+It implements the \emph{DropTarget} interface and can thus receive operators.
+Its purpose is to provide a remove functionality for already instantiated operators which are not needed anymore.
 
@@
 
-Section~\ref{sec:subgraphdefinition} states that automated constraint generation is based on subgraphs. For the finding of such subgraphs an algorithm has been developed. It is described as follows:
+As stated in section~\ref{sec:subgraphdefinition}, the automated generation of constraints is based on subgraphs.
+For the finding of such subgraphs an algorithm has been developed.
+Section~\ref{sec:subgraphfindingconcept} aims for explaining the basic concept of the algorithm, section~\ref{sec:subgraphfindingimplementation} introduces the slightly modified algorithm which is used for the prototype.
+%It is described as follows:
 
-First of all one start state ($\alpha$) is defined and all possible paths are generated which start from it and lead through the state machine.
+\subsection{Subgraph finding concept}
+\label{sec:subgraphfindingconcept}
+
+First of all one start state ($\alpha$) is defined which has two or more outgoing transitions and all possible paths are generated which start from it and lead through the state machine.
 Excluded are all paths which contain states more than once.
@@
 Now every state gets assigned a number which represents the total number of occurrences of the respective state in all collected paths.
-In other words, the label of a state expresses the number of paths which start in the start state and lead through the respective state.
-The start state itself is labelsed the number of all paths.
-If during walking through any of the paths a second state can be found with the same label, it means that all paths starting at start state come all together in this state again.
-It is determined to be a potential corresponding end state.
-A last check has to be applied whether there are incoming transitions from outside into the subgraph which don't originally come from start state.
+In other words, the label of a state expresses the number of paths which start in the $\alpha$-state and lead through the respective state.
+The $\alpha$-state itself is labeled with the total number of paths.
+If during walking through any of the paths a second state can be found with the same label, it means that all paths starting at $\alpha$-state come all together in this state again.
+It is determined to be a potential corresponding end state ($\beta$).
+A last check has to be applied whether there are incoming transitions from outside into the subgraph which do not originally come from $\alpha$-state.
 If no such transitions are detected, a subgraph is found.
-If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with start state insofar as the state machine doesn't contain further loops or terminating end states.
+If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with $\alpha$-state insofar as the state machine does not contain further loops or terminating end states.
 Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrence of each state.
-This procedure is suitable for checking whether a state is the beginning of a subgraph and for discovering the corresponding end state.
-In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state.
-
 \begin{figure}[htbp]
@@
 	\end{tikzpicture}
-  \caption{Example for subgraph finding. The occurrence of each state in the paths lets you determine G as an end state corresponding to A.}
+  \caption{Example for subgraph finding. The occurrence of each state in the paths lets you determine G as an end state ($\beta$) corresponding to A.}
   \label{fig:subgraph_finding}
 \end{figure}
+This procedure is suitable for checking whether a state is the beginning of a subgraph and for discovering the corresponding $\beta$-state.
+In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state wich has two or more outgoing transitions.
+In order to improve the further finding of subgraphs and walking through the state machine, all already found subgraphs can be assumed as single states since it is already known where their paths come finally together again: In their $\beta$-states.
 
-Following the above description an algorithm has been implemented in Java.
-Its realisation differs from the presented concept in one point: Paths are not listed but states are directly labelled since saving all possible paths would be memory inefficient.
+\subsection{Subgraph finding implementation}
+\label{sec:subgraphfindingimplementation}
+
+Following the previous description an algorithm has been implemented in Java.
+Its realisation differs from the presented concept in one point: Paths are not listed but states are directly labelled since not saving all possible paths is more memory efficient.
 The state machine is traversed similarly to a depth-first search and each visited state's label is incremented by one.
@@
 The overall implemented way of subgraph finding is illustrated in pseudo code with algorithm~\ref{alg:subgraphfinding}.
-In line 4 the labelling process described above is executed by using one particular start state.
-If a state has the same label as the start state, it is assumed as potential end state.
-Otherwise, the entire state machine is assumed to be a subgraph and start state is defined as end state as well.
-Subsequent checks finally determine whether it can be considered as proper subgraph.
-There are no cycles allowed between start and end state since they're forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
-Furthermore also foreign transitions are prohibited.
-Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
-If all checks are successful, start and end states get added to the result list $subgraphs$.
-This procedure is applied to all states of the state machine by a surrounding foreach loop.
-In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.
-
 \begin{algorithm}
@@
 \end{algorithm}
+In line 4 the labelling process described above is executed by using one particular start state.
+If a state has the same label as the start state, it is assumed as potential end state.
+Otherwise, the entire state machine is assumed to be a subgraph and start state is defined as end state as well.
+Subsequent checks finally determine whether it can be considered as proper subgraph.
+There are no cycles allowed between start and end state since they are forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
+Furthermore also foreign transitions are prohibited.
+Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
+If all checks are successful, start and end states get added to the result list $subgraphs$.
+This procedure is applied to all states of the state machine by a surrounding foreach loop.
+In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.
 
-The runtime of the algorithm depends on size and grade of branch of the state machineh.
-To improve user experience during constraint generation a dialog has been implemented which informs the developer about the current progress.
+\subsection{Constraint generation}
+
+After determining a subgraph, constraints are derived from it as defined in section~\ref{sec:constraintdetermination}.
+The entire process of constraint generation is illustrated by figure~\ref{fig:processofconstraintgeneration}.
+Once the user triggers the constraint generation, the subgraph finding algorithm starts working.
+When a subgraph is found, constraints are created and passed to the user.
+For each subgraph multiple constraint types can apply.
+This procedure is done iterative for each subgraph found.
+
+\begin{figure}
+  \centering
+
+  \begin{sequencediagram}
+		\tikzstyle{inststyle}+=[top color=yellow!30,bottom
+color=white]
+		\newthread{u}{User/Editor}{blue}
+		\newthread[2.6]{cc}{Constraint creator}{red}
+		\newinst[1]{alg}{Subgraph finding algorithm}
+
+		\mess{u}{start(stm)}{cc}
+		\begin{call}[3]{cc}{start(stm)}{alg}{finish}
+			\begin{sdblock}[yellow!20]{For each subgraph}{}
+				\setthreadbias{east}
+				\begin{call}[3]{alg}{newSubgraph(subgraph)}{cc}{}
+					\begin{sdblock}[yellow!20]{For each constraint}{}
+						\begin{call}[3]{cc}{newConstraint(constraint)}{u}{}
+						\end{call}
+					\end{sdblock}
+				\end{call}
+				\setthreadbias{center}
+			\end{sdblock}
+		\end{call}
+		\mess{cc}{finish}{u}
+	\end{sequencediagram}
+
+  \caption{Process of constraint generation.}
+  \label{fig:processofconstraintgeneration}
+\end{figure}
+
+The runtime of the algorithm depends on size and grade of branch of the state machine.
+An estimation about the execution time is presented later in section~\ref{sec:performance}.
+In order to improve waiting times and thus user experience during constraint generation the entire result might not be returned after termination but each subgraph should be processed as soon as it is found.
+For this, the algorithm implementation has been extended by code snippets which frequently interact with so called \emph{ResultListener}s (figure~\ref{fig:resultlistener_class}).
+%This interface is also implemented by the progress dialog.
+An implementation of \emph{ResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about newly found subgraphs by invoking \emph{void newSubGraphFound(SubGraph)}.
+The method \emph{boolean continueGeneration()} is called frequently and the execution will terminate as soon as it returns false.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.55\linewidth]{resultlistener_class} 
+  \caption{Interface definition of \emph{ResultListener}.}
+  \label{fig:resultlistener_class}
+\end{figure}
+
+As a second approach aiming for better user experience during constraint generation, a dialog has been implemented which informs the developer continuously about the current progress.
 This includes the number of constraints found so far and a progress bar which gives a percentual estimation of execution time.
@@
 A snapshot of the Dialog is shown in figure~\ref{fig:progressdialog}.
-
 \begin{figure}[htbp]
@@
   \label{fig:progressdialog}
-\end{figure}
-
-The algorithm implementation has been extended by code snippets which frequently interact with so called \emph{ValidationResultListener}s (figure~\ref{fig:validationresultlistener_class}).
-This interface is also implemented by the progress dialog.
-A \emph{ValidationResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about newly found subgraphs by invoking \emph{void newSubGraphFound(SubGraph)}.
-The execution will terminate whenever \emph{boolean continueGeneration()} returns false.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.55\linewidth]{validationresultlistener_class} 
-  \caption{Interface definition of \emph{ValidationResultListener}.}
-  \label{fig:validationresultlistener_class}
 \end{figure}
@@
 
-Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right.
-It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
-Constraints can be composed by drag\&drop in the dashboard in the center of the editor which holds and displays the operator constraints.
-The tab functionality on the left allows multiple constraints to be managed.
-Each tab shows a small thumbnail of the constraint and a symbol indicating its validity; a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.''.
-The magic wand button within the tab pane makes the constraint generation functionality accessible within the editor and triggers the constraint finding process.
-
+Figure~\ref{fig:editor} shows a snapshot of the editor's environment:
 \begin{figure}[htbp]
@@
 \end{figure}
+All functionality needed for constraint editing is provided in the tool bar on the right.
+It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
+Constraints can be composed by drag\&drop in the dashboard in the center of the editor which holds and displays the operator constraints.
+The tab functionality on the left allows multiple constraints to be managed.
+Each tab shows a small thumbnail of the constraint and a symbol indicating its current status of validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed.
+Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign as shown in figure~\ref{fig:results}.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[scale=0.65]{results}
+  \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
+  \label{fig:results}
+\end{figure}	
+The magic wand button within the tab pane makes the constraint generation functionality accessible within the editor and triggers the constraint finding process.
 
@@
 A state machine is encoded in one variable \emph{state}, and its value represents the active state.
-Thus each state is identified by a number.
-The state machine described in the example contains eleven states what is determined by the number range for \emph{state} from one to eleven.
+Thus, each state is identified by a number.
+The state machine described in the example contains eleven states, this is determined by the number range for \emph{state} from one to eleven.
 All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions.
@@
 Since each constraint is checked in a new instance of NuSMV, a parallel execution would furthermore strain the CPU scheduler and slow down the overall execution time.
-Thus it's recommended to execute all validation tasks consecutively rather than in parallel.
-An extra advantage is that the developer doesn't have to wait for all tasks to finish before he can see any result.
+Thus it is recommended to execute all validation tasks consecutively rather than in parallel.
+An extra advantage is that the developer does not have to wait for all tasks to finish before he can see any result.
 Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
@@
 
-LTLCreator should be integratable into other state machine based programming environments, what is ensured by its implementation as a Java Swing component and a clear API enabling a wide range of use of the tool.
+LTLCreator should be integratable into other state machine based programming environments.
+This is ensured by its implementation as a Java Swing component and a clear API which enables a wide range of use for the tool.
 

### 2012-11-13 thesis/MA-Thesis.tex
Section 6.2 (Automated constraint generation) and section 6.3.1 (Model checker) are now finished.
The .tex file structure and names have been changed. Now they match the numbers and titles in the table of contents.
have written section 2.3
correction, v.1.
2.3.2 finished
correction update, chapter 4 has a new structure
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.
Final version?

@@
 
+%\usepackage[backend=biber,style=unsrt]{biblatex}
+
 \usepackage{graphicx}
@@
 \usepackage{tikz}
-\usetikzlibrary{arrows}
+\usetikzlibrary{arrows, shapes}
+\usepackage{pgf-umlsd}
 \usepackage{amsmath}
@@
 
+%for bookmark bar
+\usepackage[colorlinks,pdfpagelabels,pdfstartview=FitH,bookmarksopen=false,bookmarksnumbered=false,linkcolor=black,plainpages=false,hypertexnames=false,citecolor=black]{hyperref}
+
+\usepackage{algorithm}
+\usepackage{algorithmic}
+
 \begin{document}
@@
 
-\onehalfspacing
 \newpage
@@
 \newpage
+\onehalfspacing
+\selectlanguage{ngerman}
+\input{tex/0_zusammenfassung}
+\selectlanguage{english}
+\input{tex/0_abstract}
+
+\chapter*{Acknowledgment}
+%\thispagestyle{empty}
+
+For my master thesis it was my wish to do research abroad in a foreign country.
+Professor Wolfgang Reif supported me in my plan and managed to give me the unique possibility to visit New Zealand for half an year.
+Therefore I am deeply grateful to him.
+
+My thesis project was hostet at the University of Auckland, New Zealand, by Professor Bruce MacDonald.
+I want to thank him for the opportunity to stay and to work at his lab, for all the meetings and discussions we had and for all the support he gave to me.
+
+I also want to thank Felix Kaser, a fellow student of mine, who was with me in New Zealand.
+During several coffee breaks we frequently discussed ideas for this work and he helped me a lot with his good suggestions.
+
+Chandan Datta, PhD student at the University of Auckland and developer of Robostudio, always offered his help regarding the use of Robostudio or general questions about the healthcare robotics project.
+He furthermore supported me in getting familiar with the lab and its research topics and with Auckland as a city to live in.
+Especially in the first time after my arrival in New Zealand I was very grateful for this.
+
+Sincere thanks is also given to Alwin Hoffmann and my supervising tutor Andreas Angerer for assisting me by discussions and hints.
+They did not only inspect and correct the thesis writing regularly, but they also assisted me in finding the thesis topic in the very beginning and contributed to the results of this work decisively by providing ideas and feedback during the research phase.
+
+Last but not least I want to thank my father Rudolf Nägele for all the effort he put in reading and correcting my work.
+Since this work relates to different mathematical concepts and formalisms such as LTL, model checking, RBDL, etc., he had to spend plenty of hours in front of the computer in order to get familiar with all these concepts and to check this work for plausibility and correctness.
+
+%\newpage
+%\thispagestyle{empty}
+%\mbox{}
+\newpage
 
 %select english as language!
-\selectlanguage{english}
+%\selectlanguage{english}
 %\pagenumbering{roman}
-\pagenumbering{arabic}
+%\pagenumbering{arabic}
+
+\pdfbookmark[1]{Inhaltsverzeichnis}{toc}
 \tableofcontents
@@
 \newpage
-\listoftables
+%\listoftables
+\lstlistoflistings
 %\pagenumbering{arabic}
@@
 
-\input{tex/1_abstract}
-\input{tex/2_introduction}
-\input{tex/3_background}
-\input{tex/4_goals}
-\input{tex/5_design}
-\input{tex/6_implementation}
-\input{tex/7_prototypeandevaluation}
+\input{tex/1_introduction}
+\input{tex/2_background}
+\input{tex/3_goals}
+\input{tex/4_visualformalismsforsafetyconstraints}
+\input{tex/5_automatedgenerationofsafetyconstraints}
+\input{tex/6_theltlcreatorprototype}
+\input{tex/7_testscenarioandevaluation}
 \input{tex/8_conclusionandfuturework}
-
-\chapter*{Acknowledgment}
 

### 2012-11-13 thesis/tex/0_abstract.tex
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.
Final version?

@@
 
-Nowadays robotic applications are no longer confined to only industrial use but increasingly assigned to jobs within human workspaces.
+Nowadays robotic applications are no longer confined to only industrial use but increasingly assigned to tasks within human workspaces and living environments.
 Even the programming of robots is shifting from experienced software developers to the consumers themselves who need to customize the robot's behaviour in order to match individual requirements.
@@
 
-A graphical editor has been developed which demonstrates both visual constraint creation functionality as well as automated constraint generation.
+A graphical editor has been developed which integrates and demonstrates both visual constraint creation functionality as well as automated constraint generation and makes them usable in an uniform manner.

### 2012-11-13 thesis/tex/0_zusammenfassung.tex
Final version?

\chapter*{Zusammenfassung}

Robotik-Anwendungen sind heutzutage nicht mehr nur in der Industrie zu finden, sondern zunehmend auch im menschlichen Bereich und Lebensumfeld, wo die Roboter in direkter Nähe zu Menschen arbeiten. Auch die Programmierung solcher Roboter wird nicht mehr zwangsläufig von Softwareentwicklern erledigt. Der Kunde möchte das Verhalten der Roboter je nach deren Einsatzgebiet individuell selbst definieren und anpassen können. Doch solche Roboter im Umfeld von Menschen stellen aufgrund der Verletzungsgefahr ein Sicherheitsrisiko dar, zumal wenn das Roboterverhalten nicht von geschulten Ingenieuren entwickelt wurde. Daher ist nach Lösungen zu suchen, wie man mit Sicherheitsanforderungen für Roboter im Zusammenhang mit kundeneigener Programmierung umgeghen soll.

Mit dieser Arbeit wird eine visuelle Sprache vorgestellt, die der Erstellung von Sicherheitseigenschaften für zustandsbasierte Roboterprogramme dient.
Bei der Entwicklung dieser Sprache wurde besonders darauf Wert gelegt, dass auch Laien, die insbesondere nicht mit Sicherheitsaspekten vertraut sind, damit umgehen und arbeiten können. Dazu wurde zu einer mathematischen temporalen Beschreibungssprache eine intuitive visuelle Darstellung erarbeitet, die einem größeren Personenkreis Zugang zu Methoden der Programmverifikation verschaffen soll.
Außerdem wird in dieser Arbeit ein neues Konzept vorgestellt, das mittels einer automatischen Bestimmung von Sicherheitsbedingungen den Entwickler bei der Programmierung unterstützt. Es soll dem Entwickler dabei helfen wichtige Bedingungen zu finden, und letztendlich die Wahrscheinlichkeit erhöhen, dass Programmfehler gefunden werden.

Ein graphischer Editor wurde entwickelt, der die visuelle Sprache zur Definition und Bearbeitung von Sicherheitsbedingungen wie auch die automatisierte Bedingungsfindung integrierend zusammenführt und gemeinsam nutzbar macht.

### 2012-11-13 thesis/tex/1_introduction.tex
Final version?

@@
 \end{wrapfigure}
-University of Auckland, New Zealand \cite{jayawardena12:_desig}.
-All interaction is guided by showing textual instructions on a display, and by speech generation.
+Uni\-ver\-si\-ty of Auckland, New Zealand \cite{jayawardena12:_desig}.
+All interaction is gui\-ded by showing textual instructions on a display, and by speech generation.
 An error in the program logic could cause an already taken medication to be reminded again and may injure a patient with Alzheimer's disease by causing a medication overdose.
@@
 %state machine based processes and a graphical tool for the designing of human robot interaction.
-Generally these people don't have profound knowledge of the complex mathematical syntax used for formal methods which allow software behaviour verification.
+Generally these people do not have profound knowledge of the complex mathematical syntax used for formal methods which allow software behaviour verification.
 %temporal logic such as computation tree logic (CTL) or linear temporal logic (LTL).
@@
 A mechanism like this is needed for the specification of such safety constraints; one that does not require special or expert knowledge and is thus easy to use for all kind of programmers.
-A visual language intended to fill this gap is presented in chapter~\ref{chap:visualformalismsforsafetyconstraints}.
+Within the scope of this work a visual language has been developed intending to fill this gap.
 
@@
 The latter would mean there are errors in the program which lead to undesired constraints.
-Furthermore, once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the development process.
-Chapter~\ref{chap:automatedgenerationofsafetyconstraints} introduces a heuristics for finding reasonable safety constraints based on state machine definitions that describe robots' behaviours.
+Furthermore, once constraints are created and checked for sanity they can be validated after every program change and thus ensure integrity during the entire development process.
+This work introduces a new heuristics for finding such reasonable safety constraints based on state machine definitions that describe robots' behaviours.
 

### 2012-11-13 thesis/tex/2_background.tex
Final version?

@@
 %\end{wrapfigure}
-In some countries, among New Zealand and Germany, the average age of population is constantly increasing.
+In some countries, among New Zealand and Germany, the average age of population is constantly increasing~\cite{altersstatistik,altersstatistik2}.
 Owing to excellent achievements in medicine as well as higher living standards, humans can enjoy longer lifes.
@@
 \end{wrapfigure}
-robots are equipped with a touch display for presenting screens and receiving user input. Furthermore, they are equipped with additional external tools such as the blood pressure measurement device.
+robots are equip\-ped with a touch display for presenting screens and receiving user input. Furthermore, they are equipped with additional external tools such as the blood pressure measurement device.
 Applications running on the robots use the touch display to present screen dialogs to the user and use speech synthesis in order to give also accoustic feedback.
@@
 They are transparent to the user and run in background.
-Text messages can be sent or external devices can be acessed, for example.
+Text messages can be sent or external devices can be accessed, for example.
 The latter is realized with web services~\cite{webservices} to obtain a modular design.
@@
 
-Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application.
+Robostudio is an editor fully written in Java and implemented as a NetBeans rich client application~\cite{netbeans}.
 Several windows and views provide visual tools for editing and understanding robot behaviour as well as screen dialog design.
@@
 
-\section{Behaviour checking}
+\section{Behaviour description and analysis}
 \label{sec:behaviourchecking}
@@
 For most of them safety properties are irrelevant, but a few are ranked highly dangerous to their environment and also to human lifes.
-In order to eliminate software malfunction there are different mathematical concepts for verifying software behaviour.
+In order to eliminate software malfunction, there are different mathematical concepts for verifying software behaviour.
 One of them is model checking which allows checking state machine programs for functional safety.
-Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realised wth state machines, this concept of behaviour checking is applicable and important for our kind of programs.
+Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realised with state machines, the concept of behaviour checking is applicable and important for this kind of programs.
 At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an introduction into model checking in section~\ref{sec:modelcheckingandltl}.
@@
 The state representing the current system status is called \emph{current state}.
-The current state changes whenever there is a modification to the system.
+The current state changes (maybe into itself) whenever there is a modification to the system.
 Such transfers of the system from the current state to another one are defined by \emph{transitions}.
@@
 All influences of the environment which cause a change to the system are called \emph{events}.
-They trigger transitions which transfer the system into a new state (current state).
+They trigger transitions which transfer the system into a new (current) state.
 The fact that a state machine reacts to events is the reason why state machines are also called \emph{reactive systems}.
@@
 \end{figure}
-All states are drawn as circles which contain their names. Transitions are painted as arrows between source and destination states, and the regarding trigger event names are written next to it.
+All states are represented by circles which contain their names. Transitions are drawm as arrows between source and destination states, and the regarding trigger event names are written next to it.
 
@@
 
-The black dot defines the point of entry for this state machine, the door is closed but unlocked in the beginning.
-Through an event of the environment - in this case a human acting at the door - the door can be opened ,what makes ``opened'' the current state, and later be closed again.
+The black circle defines the point of entry for this state machine, the door is closed but unlocked in the beginning.
+By an event of the environment - in this case a human acting at the door - the door can be opened, which turns ``opened'' into the current state, and closed later on again.
 A turn of the key locks respectively blocks the door and no further opening or closing is possible.
-A second turn of the key un(b)locks the door again.
+A second turn of the key unlocks respective unblocks the door again.
 
-The short description given in this section does not contain a complete explanation about the concept of statemachines at all, but it gives a basic overview which is needed as background information for this work.
-Also the formal definition of state machines is not relevant and can be read in other literature~\cite{statemachines, statemachinesandformalmethods}.
+The short description given in this section does not contain a complete explanation about the concept of state machines at all, but it gives a basic overview which is needed as background information for this work.
+Also the explanation of state machines above is no formal definition; it can be read found in literature~\cite{statemachines, statemachinesandformalmethods}.
+%Also the formal definition of state machines is not relevant and can be found in literature~\cite{statemachines, statemachinesandformalmethods}.
 
@@
 
-Whenever a software or hardware system has a special requirement in robustness and reliability, different formal methods are available to be used in order to verify the design.
-Model checking is such a method which allows automatically verifying a system description against a specification by means of an exhaustive search of all states which could possibly be visited by the system during its execution.
+Whenever a software or hardware system has a special requirement regarding robustness and reliability, different formal methods are available to be used in order to verify the design.
+Model checking~\cite{principles-of-model-checking} is such a method which allows automatically verifying a system description against a specification by means of an exhaustive search for all states which could possibly be visited by the system during its execution.
 As a result, either the system description fulfills all propositions or a counter example is provided which gives evidence about a broken property.
@@
 In contrast, the specification used for model checking contains certain properties that express the desired and safe system behaviour in form of mathematical formulas.
-Linear temporal logic (LTL) is an example for a logic that is used to define such formulas on state machines.
+Linear temporal logic (LTL)~\cite{principles-of-model-checking} is an example for a logic that is used to define such formulas on state machines.
 It is a temporal logic which can express formulas concerning the future of paths, for example, that something will eventually happen.
@@
 \begin{description}
-	\item[$\Box \varphi$] This formula expresses that $\varphi$ is true now and in all following states. The meaning of the rectangle is ``GLOBALLY''.
-	\item[$\medcircle \varphi$] With a circle representing ``NEXT'', $\varphi$ will be true in the next state.
-	\item[$\Diamond \varphi$] At least one time - now or eventually in a later state - $\varphi$ is or will be true. The diamond stands for ``FUTURE''.
+	\item[$\Box \varphi \textnormal{ }$] This formula expresses that $\varphi$ is true now and in all following states. The meaning of the rectangle is ``GLOBALLY''.
+	\item[$\medcircle \varphi \textnormal{ }$] With a circle representing ``NEXT'', $\varphi$ will be true in the next state.
+	\item[$\Diamond \varphi \textnormal{ }$] At least one time - now or eventually in a later state - $\varphi$ is or will be true. The diamond stands for ``FUTURE''.
 \end{description}
@@
 	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true at least until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
-	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true through the first occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
-	\item[$\varphi \mathcal{W} \psi$] ``WEAK UNTIL'' represents the same as ``UNTIL'', except that an occurrence of $\psi$	is not necessary. In this case, $\varphi$ is true infinetely.
+	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true caused by the first occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
+	\item[$\varphi \mathcal{W} \psi$] ``WEAK UNTIL'' represents the same as ``UNTIL'', except that an occurrence of $\psi$	is not necessary. In this case, $\varphi$ remains true infinitely.
 \end{description}
@@
 The idea of a visual environment for the defining of system properties is not new and already pursued by some researchers.
-Although their focus does not completely match the subject of safety for healthcare robots, they present concepts of visualisation which might be relevant
+Although their focus does not completely match the subject of safety for healthcare robots, they present concepts of visualization which might be relevant
 % can be quite interesting 
@@
 A first step towards a graphical tool for designing safety constraints is taken by Sisirucá and Ionescu~\cite{332301}.
-They developed an object-oriented graphical environment for visually creating temporal logic sentences and rules.
+They developed an object-oriented graphical environment for creating temporal logic sentences and rules visually.
 As shown in figure~\ref{fig:sisirucaandionescu}, the tool presents all operators as blocks which have inputs and outputs. 
@@
   \includegraphics[width=0.6\linewidth]{sisirucaandionescu} 
-  \caption{The visual programming environment. An example showing the ruel construction process~\cite{332301}.}
+  \caption{The visual programming environment. An example showing the rule construction process~\cite{332301}.}
   \label{fig:sisirucaandionescu}
 \end{figure}
-Whereas logical operators such as \emph{AND} and \emph{OR} have their boolean output and inputs represented by arrows pointing to the right, generic blocks for temporal expressions additionally have an input symbolized by an arrow pointing downward where the temporal operator itself is to be attached.
+Whereas logical operators such as \emph{AND} and \emph{OR} have each one boolean output and several boolean inputs represented by arrows pointing to the right, generic blocks for temporal expressions additionally have an input symbolized by an arrow pointing downward where the temporal operator itself is to be attached.
 By adding and connecting the operators a big expression graph can be built.
@@
 In fact, it gives a schematic overview of even complex formulas and allows substitution and reuse.
-The visual formalism is an alienation of LTL which might be suitable for researchers, but probably not for novices.
+The visual formalism is an alienation of LTL which might be suitable for researchers, but probably not for non-experts.
 
@@
 It is based on computation tree logic (CTL) which is a branching-time logic with support for quantifiers.
-In contrast to LTL where properties are fulfilled by all possible paths, CTL expresses properties which depend on the future branching, for example, that at least one path is existend which has a particular property.
+In contrast to LTL where properties are fulfilled by all possible paths, CTL expresses properties which depend on the future branching, for example, that at least one path exists which has a particular property.
 As indicated by figure~\ref{fig:delbimbo}, the visual tool provides all operators known from LTL 
@@
   \includegraphics[width=0.8\linewidth]{delbimbo} 
-  \caption{Some of the visual operators of Del Bimbo et al.~\cite{520786}.}
+  \caption{Visual representation of some of the operators of Del Bimbo et al.~\cite{520786}.}
   \label{fig:delbimbo}
 \end{figure}
-except that temporal operators are extended by an existential or universal quantifier (``at least on one path...'', ``on all paths...'').
+except that temporal operators are extended by an existential quantifier (e.g.\ ``at least on one path...'') or an universal quantifier (e.g.\ ``on all paths...'').
 
@@
 The positioning of these operators is organized in layers where the parent operator lays one level above its children. It can be compared to a tree upside down with the topmost operator as root.
-This formula structure can be visualized as a 3D branching tree within a virual space. Figure~\ref{fig:delbimbo3d} shows the result of considerations done by Del Bimbo and his colleagues.
+This formula structure can be visualized as a 3D branching tree within a virtual space. Figure~\ref{fig:delbimbo3d} shows the result of considerations done by Del Bimbo and his colleagues.
 
@@
   \centering
-  \includegraphics[width=1\linewidth]{delbimbo3d} 
-  \caption{Formula trees' 3D representation within a virual space~\cite{520786}.}
+  \includegraphics[width=0.98\linewidth]{delbimbo3d} 
+  \caption{Formula trees' 3D representation within a virtual space~\cite{520786}.}
   \label{fig:delbimbo3d}
@@
 The presented visual formalism directly maps the syntax of the temporal logic onto graphic symbols and composition patterns.
-The researchers found well expressive graphical representations for all operator especially the temporal ones, what makes the tool fairly intuitive.
-The tree dimensional visualisation gives a good overview of the structure, although it shuld not be used for editing.
-Computer monitors can naturally display only two dimensions and input devices such as mouse and keyboard are customised to it.
+The researchers found well expressive graphical representations for all operator especially for the temporal ones, which makes the tool fairly intuitive.
+The three dimensional visualization gives a good overview of the structure, although it shuld not be used for editing.
+Computer monitors can naturally display only two dimensions and also input devices such as mouse and keyboard are customised to it.
 Editing of the third dimension needs special gesture definitions and 3D modelling techniques the most users might not be familiar with.
@@
 The step towards a more abstract level of development is taken by HomeTL~\cite{4341725}, a visual formalism for the design of home based care.
-It allows healthcare professionals to specify rules and sequences of user actions in a quite nontechnical manner.
+It allows healthcare professionals to specify rules and sequences of actions in a quite nontechnical manner.
 Based on these conditions a monitoring system in a patient's home environment can detect and report abnormal situations within the daily routine.
 The visual notation provides propositional variables and logical connectives, the latter is shown in figure~\ref{fig:hometloperators}.
-In contrast to LTL, the temporal operators of HomeTL use absolute time limits (seconds, for example) which are defined by $\tau$.
+In contrast to LTL, the temporal operators of HomeTL use absolute time limits (seconds, for example) which are represented by $\tau$ in figure~\ref{fig:hometloperators}.
 Thus, not only the relative order of states is defined but also a time by which a particular action has to happen.
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{hometloperators} 
-  \caption{Graphical representation of visual notation for temporal
-operators within HomeTL user interface~\cite{4341725}.}
+  \includegraphics[width=0.54\linewidth]{hometloperators} 
+  \caption{Visual notation for temporal operators within HomeTL user interface~\cite{4341725}.}
   \label{fig:hometloperators}
@@
 
-This notation uses special meanings for the two dimensions: The horizontal dimension shows the temporal evolution of states, while the vertical dimension expresses the logical relationships at the respective current state.
-A programming environment allows editing and creating of constraints by using the atomar operators.
+This notation gives special meanings to the two dimensions: The horizontal dimension shows the temporal evolution of states, while the vertical dimension (see~figure~\ref{fig:hometleditor}) expresses the logical relationships at the respective current state.
+A programming environment allows editing and creating of constraints by using the elementary operators shown in figure~\ref{fig:hometloperators}.
 Figure~\ref{fig:hometleditor} presents a snapshot of this editor.

### 2012-11-13 thesis/tex/4_visualformalismsforsafetyconstraints.tex
Final version?

--- 2012-11-03 thesis/tex/4_visualformalismsforsafetyconstraints.tex
@@
 
-This chapter introduces two different approaches of visual formalisms and later states the final desing of the visual language.
-One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points.
-A first approach, so called template constraints,
+This chapter introduces two different approaches of visual formalisms and states later on the final desing of the visual language.
+A main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since these two characteristics appear to be mutually exclusive in some points.
+A first approach, the so called template constraint formalism,
 % and its pros and cons 
-are described in section~\ref{sec:templateconstraintformalism}. Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
-Section~\ref{sec:comparison} gives a comparison of the two formalisms presented before and states which one is used for the prototype.
+is described in section~\ref{sec:templateconstraintformalism}. Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
+Section~\ref{sec:comparison} gives a comparison of the two formalisms presented before and states which one is suitable to be used for the prototype.
 
@@
 %It is not a trivial issue to find a
-What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
+Which visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
 % TODO: wir haben gedacht, dass es einfach zu benutzen ist
 Led by this question, a visual language based on templates has been developed.
-These templates make different propositions about states and their relations.
-The user can choose from several graphical constraint types and customize them by parameterising state variables.
+These templates make different propositions about visits of states and their relations.
+Application developers can choose from several graphical constraint types and customize them by parameterising state variables.
 So far there are two different kinds of template constraints:
@@
 
+	In this constraint template, x, y and z are state variables which can be replaced by real states.
 	The idea behind this constraint is that propositions about temporal dependencies can be made.
-Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled.
+Under certain circumstances the visit of a particular state is allowed only if a specific precondition is fulfilled.
 An example constraint for this template is: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z).
@@
 
+	As before, x, y, z and w are state variables.
 	This constraint type enforces at least one particular event to happen after a specific precondition.
@@
 	%a product or not to buy it.
-	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can not avoid both possibilities.
+	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can not avoid each of both possibilities.
 	%Nitpicking readers might note 
@@
 Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like.
-Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot.
-Furthermore the state labels could be replaced by the real screen displays of the regarding states.
-All parameters of the template constraints shall be directly customizable by the robot surface symbols.
+Each state of a constraint is visualised by the image of the robot appearance to make the developer think he is directly working with the robot.
+Furthermore, the state labels could be replaced by the real screen displays of the regarding states.
+It has to be noticed that the arrows used for visualizing template constraints are not necessarily equal to transitions of the state machine.
+All parameters for the template constraints shall be directly customizable with the robot surface symbols.
 Figure~\ref{fig:templateconstraint_dropdown} demonstrates how states could be choosen as parameters.
@@
 All template constraints are additionally equipped with textual expressions in order to explain their semantics.
-It is a template sentence with placeholders for the states available for parameterisation.
+These textual expressions are template sentences with placeholders for the states available for parameterisation.
 Whenever there is a change to the constraint, te text is being updated automatically regarding to the parameter states.
@@
 Thus, specific formulas have to be provided which represent the semantics of the constraints.
-They have to be in a format so that they can be processed by a model checker.
+They have to be in a specially defined format so that they can be processed by a model checker.
 For the first kind of template operator, a linear temporal logic fomula (LTL, see~section~\ref{sec:modelcheckingandltl}) expresses its semantics (with $x$ as ``x is current state'' and similarly for $y$ and $z$):
@@
 \begin{equation} \label{eq:template_one}
-  \models \Box (x \rightarrow ([(\neg z) \mathcal{U} y] \vee \Box \neg z))
+  \models \Box (x \rightarrow [(\neg z) \mathcal{W} y])
 \end{equation}
 
-``Globally is true: Whenever x is the current state, either z is not visited until y or z is never visited''.
-Compared to the example above it means: Whenever a person eats something (x), he can not go to bed (z) until he brushes his teeth (y).
+``Globally is true: When x is the current state, z can not not be visited before y is visited''.
+Compared to the example above it means: Whenever a person eats something (x), he can not go to bed (z) before he brushes his teeth (y).
 Otherwise, if he never goes to bed, he does not necessarily need to brush his teeth.
@@
 A short explanation about CTL is already given in section~\ref{sec:delbimbo}, but the two new constructs used for this formula are described below.
-The main formula for the second template is still using LTL:
+The main formula for the second template uses LTL:
 \begin{equation} \label{eq:template_two_one}
@@
 \end{equation}
-``Globally is true: Whenever x is the current state, in a future step either y or z (or ...) will be visited''.
-In context of the example mentioned above it means: Every time a person enters a supermarket (x) he will eventually buy something (y) or buy nothing (z).
-The fact that both states are really potentially reachable is expressed by the following CTL formula.
+``Globally is true: When x is the current state, in a future step either y or z (or ...) will be visited''.
+%In context of the example mentioned above it means:
+Applied to the example mentioned above it means: every time a person enters a supermarket (x) he will eventually buy something (y) or buy nothing (z).
+The fact that paths really exist in the program which eventually lead to x and y
+% both states are really potentially reachable
+is expressed by the following CTL formula.
 \emph{AG} replaces the $\Box$ sign and stands for ``on all paths is globally true...'', \emph{EF} is related to $\Diamond$ and expresses ``a path exists where is true in future...''.
@@
 Since the visual constraints are directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness.
-If it is considered as easy to use, non-experts can use it as well as experts who already know LTL.
+If it turns out to be usable easily though it is a mathematical concept, it can also be used by non-experts who do not already know LTL.
 
@@
 \begin{description}
-	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
-	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
-	\item[(c)] \emph{IF} operator: $\varphi \rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
-Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
-	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
-	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state.
+	\item[(a)] \emph{AND} (\emph{AndOperator}): $\varphi \wedge \psi$.\\This conjunction evaluates to true only if both $\varphi$ and $\psi$ are true.
+	\item[(b)] \emph{OR} (\emph{OrOperator}): $\varphi \vee \psi$.\\At least one $\varphi$ or $\psi$ being true makes this construct true. The result is false when both parameters are false.
+	\item[(c)] \emph{IF THEN} (\emph{IfThenOperator}): $\varphi \rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided.
+Its meaning seems to be more intuitive for non-experts; moreover it is a common construct in almost every programming language.
+	\item[(d)] \emph{NOT} (\emph{NotOperator}): $\neg \varphi$. This operator negates the value of $\varphi$.
+	\item[(e)] Proposition (\emph{StateOperator}): $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state.
 Other types such as numeric or string equations may be added but are less important for the healthcare subject.
@@
 
-In addition to these five logical operators the visual language should be furthermore capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i).
+In addition to these five logical operators the visual language should be furthermore capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i):
 
 \begin{description}
-	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
-	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
-	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
-	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true at least until there is a state with $\psi$ being true.
+	\item[(f)] \emph{ALWAYS} (\emph{AlwaysOperator}): $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
+	\item[(g)] \emph{NEXT} (\emph{NextOperator}): $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
+	\item[(h)] \emph{FUTURE} (\emph{FutureOperator}): $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
+	\item[(i)] \emph{UNTIL} (\emph{UntilOperator}): $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true at least until there is a state with $\psi$ being true.
 In addition eventually there has to be a future state with $\psi$ being true.
@@
 \subsection{Design decisions}
+\label{sec:designdecisions}
 
 Like the formula representation introduced by Del Bimbo et al.~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy.
-However, the new visual formalism surrenders the three dimensional idea and focuses on two dimensional blocks and their compositions instead, which is used by HomeTL~\cite{4341725} as well.
+However, the new visual formalism surrenders the three dimensional idea and focuses on two dimensional graphical blocks and their compositions instead, which is used by HomeTL~\cite{4341725} as well.
 Each operator of the visual language is represented by such a block.
@@
 In order to support the reading of constraints, different significant visual appearances of the operator types shall enable easy classification and a faster subconscious identification~\cite{moody-physics-of-notations}.
-In this case, different colors are used for each operator type.
+For this purpose an individual color is used for each operatot type.
 % aiming for faster perceptibility and understandability. 
@@
 Operators which consist of nested operators have particular meanings for the two dimensions: The ``logical'' vertical read direction and the horizontal direction for variation in time.
-Thus, logical operators are arranged in a vertical row wereas time relevant operators are aligned horizontally.
+Thus, logical operators are arranged in a vertical row whereas time relevant operators are aligned horizontally.
 Figure~\ref{fig:directions} demonstrates an example constraint composed along the logical and time axes.
@@
 But also the positioning of nested operators within teir parents have to be taken into account.
-Whereas the operator's centered positioning and resizing along the logical direction only depends on the sizes of all sub operators, the arrangement within the time axis turns out to be more complex.
-Here, sub operators can not just be aligned centered as it is the case vertically.
+Whereas the operator's centered positioning and resizing along the logical direction only depends on the sizes of all suboperators, the arrangement within the time axis turns out to be more complex.
+Here, suboperators can not just be aligned centered as it is the case vertically.
 To face the determination of horizontal positioning, every operator gets time reference lines which specify its positioning within its parent and the positioning of its own children within itself.
@@
 These operators can be visually merged in the manner of unifying their border lines, for example.
-As a result, all associated operators appear as just one operator with a couple of sub operators (see~figure~\ref{fig:nested_and}).
+As a result, all associated operators appear as just one operator with a couple of suboperators (see~figure~\ref{fig:nested_and}).
 
@@
 	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer. It is intended to be similar to line coloring when reading a digital text.
-	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
-	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances.
+	\item Every individual edit step to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint.
+The result is immediately displayed to the user.
+	\item There is no button for constraint creation respectively editing except operator and proposition providers, which let the developer create new instances.
 Having no additional buttons reduces complexity and keeps it simple.
@@
 
-Due to its arrow based design the visual language for template constraints is a good visualisation of time dependencies.
+Due to its arrow based design the visual language for template constraints is a good visualization of time dependencies.
 Furthermore, it allows expressing possibilities by the use of CTL formulas which is not possible with the operator costraint formalism.
-Nevertheless, the template visualisations have some trouble with clearness of their semantics.
-Even though there are well defined formulas behind each constraint and textual expressions which explain the meanings of constraints, the visual presentation seems to be ambiguous in some settings what leads to the risk of being interpreted differently.
+Nevertheless, the template visualizations have some trouble with clearness of their semantics.
+Even though there are well defined formulas behind each constraint and textual expressions which explain the meanings of constraints, the visual presentation seems to be ambiguous in some settings which leads to the risk of being interpreted differently.
 For example, in the context of the visual language it is unclear whether the first constraint template allows state (x) to be visited more than once before state (z) is reached.
-It might also be problematic to interprete whether more than one of the states $y$, $z$, ...
-can be actually visited after a visit of $x$ in the second constraint.
-In contrast, the operator constraint formalism is not ambiguous due to to its direct mapping to LTL wich is consistently defined.
+It might also be problematic to interpret in the second constraint whether more than one of the states $y$, $z$, ...
+can actually be visited after a visit of $x$.
+In contrast, the operator constraint formalism is not ambiguous due to to its direct mapping to LTL which is consistently defined.
 All visual operator expressions represent exactly the same semantics as their corresponding textual LTL formulas.
@@
 
-However, there is a second problem with the template constraints.
+Additionally, there is a second problem with the template constraints.
 Whereas the safety constraints specific to the example within the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are other common constraints which have no matching template and thus can not be described.
-For example, the check of a state being visited never or at least once is not possible.
-For every new constraint type a new template has to be created what means an infinite amount of templates is needed in order to achieve full expressiveness.
+For example, checks of states like ``being visited never'' or ``being visited at least once'' are not possible.
+For every new constraint type a new template has to be created which means an infinite amount of templates is needed in order to achieve full expressiveness.
 On the contrary, the visual language for operator constraints is not limited in expressiveness compared to LTL, disregarding that there is only one proposition type available.
@@
 Especially for a tool for safety and consistency it would form a really bad basis.
-For this reason, the approach of template constraints is discarded and the operator constraint formallism is used for the editor prototype instead.
+For this reason, the approach of template constraints is discarded and the operator constraint formalism is used for the editor prototype instead.
 Owing to the recursive nesting of operator constraints, a visual formula can be recursively translated to a LTL sentence which can be processed by the prototype using any ordinary model checker.

### 2012-11-13 thesis/tex/5_automatedgenerationofsafetyconstraints.tex
correction update, chapter 4 has a new structure
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.
Final version?

@@
 Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications.
-In order to discover whether a program meets its safety requirements, its behaviour is checked by the constraints.
+In order to discover whether a program meets certain safety requirements, its behaviour is checked by the constraints.
 This happens mostly as a last check before software gets released.
@@
 As a solution to this problem, the concept of automated constraint generation for state machines is proposed.
-Once the initial implementation of a program is finished, its behaviour can be analysed.
-Sgnificant parts of it can be determined automatically and put into constraint representation which are valid on the current program of course.
-%constraint suggestions which are valid on the current program can be computed fully automatically. 
-On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg.
-if a computed constraint does not make sense at all.
+Once the initial implementation of a program is finished, the behaviour of this state machine can be analyzed.
+In particular heuristics can be elaborated which determine significant parts of the state machine automatically and put them into constraint representation.
+These constraints are of course valid on the current program.
+On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg.\ if a computed constraint does not make sense at all.
 On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.
 
-In order to support the developer in developing safe programs and to simplify maintainability, automated generation and suggestion of constraints might be helpful.
-But it's helpful only if the suggested constraints are relevant and reasonable.
+In order to support the developer in developing safe programs and to simplify maintainability, automated generation and suggestion of constraints for state machines might be helpful.
+But it is helpful only if the suggested constraints are relevant and reasonable.
 
@@
 depends on the domain and use case.
-It has to be mentioned that the approach of finding an algorithm for constraint generation presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
-Regarding the important safety constraints postulated in chapter~\ref{chap:goals}, all the relevant states seem to fall into one of the following categories:
+It has to be mentioned that the approach of finding an algorithm for constraint generation presented in this work has a special focus on the healtcare use cases and is not necessarily applicable for other domains.
+Among others, each safety constraint postulated in chapter~\ref{chap:goals} was analyzed on a state machine program which represents the medication reminder application of the health care project.
+It turned out that all states which are named in the constraint are always connected in an identical manner and seem to fall into one of the following categories:
 
@@
 In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined.
-Such a group of states is named subgrap from now on.
+Such a group of states of a state machine is named subgraph from now on.
 It is defined as follows:
@@
 	\item There is a start state ($\alpha$-state) with at least two outgoing transitions.
-	\item There is one end state ($\beta$-state) in which all possible paths starting from start state come together again.
-	\item All states between the start and end states have no other incoming transitions than the ones originally coming from the start state.
+	\item There is one end state ($\beta$-state) which is the first state where all possible paths of the state machine starting from start state come together again.
+	\item All states between the start and end state also belong to the subgraph.
+They have no other incoming transitions than the ones originally coming from the start state.
 That means there is no path to visit these states except through the start state.
 	\item The start state may have incoming transitions from states not contained in the subgraph.
-	\item The end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
+	\item The end state may have self transitions or incoming and outgoing transitions from and to states which are not contained in the subgraph.
+Transitions back into the subgraph are not allowed except to the start state.
+	\item Except self transitions of the end state, no cycles are allowed in the subgraph.
 \end{itemize}
@@
 \section{Constraint determination}
+\label{sec:constraintdetermination}
 
@@
 \begin{equation} \label{eq:first}
-  \models \Box (\alpha \Rightarrow \Diamond \beta)
+  \models \Box (\alpha \rightarrow \Diamond \beta)
 \end{equation}
@@
 \begin{equation} \label{eq:second}
-  \models \Box (\alpha \Rightarrow \Diamond (\bigvee_{\gamma} \gamma))
+  \models \Box (\alpha \rightarrow \Diamond (\bigvee_{\gamma} \gamma))
 \end{equation}
@@
 \begin{equation} \label{eq:third}
-  \models \Box (\alpha \Rightarrow [\neg \beta \mathcal{U} \bigvee_{\gamma} \gamma])
+  \models \Box (\alpha \rightarrow [\neg \beta \mathcal{U} \bigvee_{\gamma} \gamma])
 \end{equation}
@@
 \begin{equation} \label{eq:fourth}
-  \models \Box (\alpha \Rightarrow [(\bigwedge_{\delta} \neg \delta) \mathcal{U} \beta])
+  \models \Box (\alpha \rightarrow [(\bigwedge_{\delta} \neg \delta) \mathcal{U} \beta])
 \end{equation}
@@
 \begin{equation} \label{eq:seventh}
-	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
+	\models \Box (\textnormal{'Not taken yet'} \rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
 \end{equation}

### 2012-11-13 thesis/tex/6_theltlcreatorprototype.tex
Final version?

@@
 Therefore a prototype of such an editor has been designed and implemented which contains the visual language as well as the proposed constraint generation heuristics.
-The implementation is a Java Swing component and has a simple API definition, making it easy to integrate the LTLCreator tool into other java applications.
+The implementation is a Java Swing component and has a simple API definition, making it easy to integrate the \emph{LTLCreator} tool into other java applications.
 
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{fsm_class} 
-  \caption{State machine model class definitions.}
+  \includegraphics[width=0.65\linewidth]{fsm_class} 
+  \caption{Class definitons for the state machine model.}
   \label{fig:fsm_class}
@@
 
-All different operator types, i.e.
+All operator types, i.e.
 \emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator}, \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator}, \emph{UntilOperator} and \emph{StateOperator}, are realized by dedicated classes.
@@
 	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality.
-Equality is given when the compared operators are of the same type and similarly their children.
+Equality is given if the compared operators are of the same type and similarly their children.
 This functionality is important for avoiding duplicate constraints during constraint generation, for example.
-	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it.
+	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or to one of its children, an operator change listener gets notified about it.
 The hierarchical architecture of \emph{AbstractOperator}s propagates all change events to the root operator.
@@
   \includegraphics[width=0.6\linewidth]{AbstractOperator} 
-  \caption{AbstractOperator class definition.}
+  \caption{Class definition for \emph{AbstractOperator}.}
   \label{fig:abstractoperator}
@@
 
-Whereas \emph{StateOperator}s have no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two possible nested children respectively.
+Whereas \emph{StateOperator}s have no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two optional nested children respectively.
 The graphical representation of both unary and binary operators thus needs to have place holders and docking stations
@@
 Swing's \emph{Graphics2D} functionality provides a great support in painting rounded shapes and color gradients.
-An operator is displayed as a rounded rectangle with dents around sub operators and has a flowing backgound, borders and labels in an operator type specific color.
+An operator is displayed as a rounded rectangle with dents around sub operators; it is displayed with a flowing backgound and borders and labels in the operator type's specific color.
 For the significant shape multiple rounded rectangles with and without borders are arranged in a way that the result appears to be one piece.
-Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the \emph{IfThenOperator}, first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be.
+Figure~\ref{fig:paintprocess} demonstrates how the process of painting works: For painting the \emph{IfThenOperator}, first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where the two child operators shall be later on.
 In the third step the same rectangle as from step one is painted on top, this time without border.
@@
 The built-in mouse support of Swing is used for identifying the operator positioned under the mouse cursor.
-All operators except the hovered one and its sub operators are presented bleached out what causes the hovered operator to appear emphasized.
+All operators except the hovered one and its sub operators are presented bleached out which causes the hovered operator to appear emphasized.
 The most right operator composition in figure~\ref{fig:nested_and_mouseover} gives a demonstration of this idea.
@@
   \includegraphics[scale=0.65]{and_simplify_mouseover}
-  \caption{Default nested \emph{AND} operator, visually merged nested \emph{AND} operator and visually nested \emph{AND} operator with mouse over.
+  \caption{Default nested \emph{AND} operator, visually merged nested \emph{AND} operator and visually nested \emph{AND} operator with mouse over (from left to right).
   }
@@
 
-\emph{LLTCreator} is equipped with a tool bar which provides all functionality needed for constraint editing: For each operator type a \emph{DraggableProvider} is displayed as an icon in the respective operator color.
+\emph{LTLCreator} is equipped with a tool bar which provides all functionalities needed for constraint editing: For each operator type a \emph{DraggableProvider} is displayed as an icon in the respective operator color.
 Each time the mouse is pressed on it (drag action start), a new instance of the respective operator type is created which can be moved and dropped to the dashboard.
@@
 
-As stated in section~\ref{sec:subgraphdefinition}, the automated generation of constraints is based on subgraphs.
+As stated in section~\ref{sec:subgraphdefinition}, the automated generation of constraints for a state machine is based on subgraphs.
 For the finding of such subgraphs an algorithm has been developed.
-Section~\ref{sec:subgraphfindingconcept} aims for explaining the basic concept of the algorithm, section~\ref{sec:subgraphfindingimplementation} introduces the slightly modified algorithm which is used for the prototype.
+Section~\ref{sec:subgraphfindingconcept} aims to explain the basic concept of the algorithm, section~\ref{sec:subgraphfindingimplementation} introduces the slightly modified algorithm which is used for the prototype.
 %It is described as follows:
@@
 
-First of all one start state ($\alpha$) is defined which has two or more outgoing transitions and all possible paths are generated which start from it and lead through the state machine.
-Excluded are all paths which contain states more than once.
-The result is a finite amount of finite paths since loops are eliminated.
-Now every state gets assigned a number which represents the total number of occurrences of the respective state in all collected paths.
-In other words, the label of a state expresses the number of paths which start in the $\alpha$-state and lead through the respective state.
+First of all a start state ($\alpha$) is defined which has two or more outgoing transitions.
+All possible paths are generated which start from it and lead through the state machine.
+Every path ends before it visits states twice.
+The result is a finite set of finite paths since loops are eliminated.
+Now every state gets as label assigned a number which represents the total number of occurrences of this state in all collected paths.
+In other words, the label of a state expresses the number of paths which start in the $\alpha$-state and contain this state.
 The $\alpha$-state itself is labeled with the total number of paths.
@@
 If no such transitions are detected, a subgraph is found.
-If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with $\alpha$-state insofar as the state machine does not contain further loops or terminating end states.
-Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrence of each state.
+If there is no state with the same label, or a transition exists which comes from outside, the state machine as a whole can be assumed as subgraph beginning with $\alpha$-state insofar as the state machine does not contain further loops or terminating end states.
+Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrences of each state.
 \begin{figure}[htbp]
@@
 	\end{tikzpicture}
-  \caption{Example for subgraph finding. The occurrence of each state in the paths lets you determine G as an end state ($\beta$) corresponding to A.}
+  \caption{Example for the subgraph finding algorithm: The occurrence labels of the states lead to
+   %of each state in the paths lets you determine 
+   G as an end state ($\beta$) corresponding to A.}
   \label{fig:subgraph_finding}
@@
 This procedure is suitable for checking whether a state is the beginning of a subgraph and for discovering the corresponding $\beta$-state.
-In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state wich has two or more outgoing transitions.
-In order to improve the further finding of subgraphs and walking through the state machine, all already found subgraphs can be assumed as single states since it is already known where their paths come finally together again: In their $\beta$-states.
+In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state which has two or more outgoing transitions.
+In order to improve the further finding of subgraphs and walking through the state machine, all already found subgraphs can be assumed as single states since it is already known where their paths come finally together again.%: In their $\beta$-states.
 
@@
 Following the previous description an algorithm has been implemented in Java.
-Its realisation differs from the presented concept in one point: Paths are not listed but states are directly labelled since not saving all possible paths is more memory efficient.
+Its realization differs from the presented concept: Paths are not listed but states are directly labelled since not saving all possible paths is more memory efficient.
 The state machine is traversed similarly to a depth-first search and each visited state's label is incremented by one.
-If the currently visited state holds more than one outgoing transition and thus constitutes new paths,  all states on the current stack, i.e.
-the previously visited states on the current path, get their labels incremented by $(\textnormal{'number of outgoing transitions'} - 1)$.
+If the current state holds more than one outgoing transition and thus constitutes new paths,  all states on the current stack, i.e.
+the previously visited states on the current path, get their labels incremented by ('number of outgoing transitions' - 1).
 The algorithm turns back whenever it comes across a state which has already been visited within the same path stack of the depth-first search.
@@
 \end{algorithm}
+Line 1 defines a variable $subgraphs$ which will contain all results of the algorithm after execution.
 In line 4 the labelling process described above is executed by using one particular start state.
-If a state has the same label as the start state, it is assumed as potential end state.
-Otherwise, the entire state machine is assumed to be a subgraph and start state is defined as end state as well.
-Subsequent checks finally determine whether it can be considered as proper subgraph.
+Line 5 checks whether there is a state having the same label as the start state.
+If so, it is assumed as potential end state (line 6).
+Otherwise, the entire state machine is assumed to be a subgraph and the start state is defined as end state as well (line 8).
+Subsequent checks in line 9 finally determine whether it can be considered as proper subgraph.
 There are no cycles allowed between start and end state since they are forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
-Furthermore also foreign transitions are prohibited.
-Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
-If all checks are successful, start and end states get added to the result list $subgraphs$.
-This procedure is applied to all states of the state machine by a surrounding foreach loop.
+Furthermore also foreign transitions are prohibited: Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
+If all checks are successful, start and end states get added to the result list $subgraphs$ in line 11.
+This procedure is applied to all states of the state machine by the surrounding foreach loop (line 2).
 In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.
@@
 The entire process of constraint generation is illustrated by figure~\ref{fig:processofconstraintgeneration}.
-Once the user triggers the constraint generation, the subgraph finding algorithm starts working.
-When a subgraph is found, constraints are created and passed to the user.
-For each subgraph multiple constraint types can apply.
-This procedure is done iterative for each subgraph found.
-
 \begin{figure}
@@
 \end{figure}
+Once the user triggers the constraint generation for a state machine (stm), the subgraph finding algorithm starts working.
+When a subgraph is found, constraints are created and passed to the user.
+For each subgraph multiple constraint types can apply.
+This procedure is done iterative for each subgraph found.
 
-The runtime of the algorithm depends on size and grade of branch of the state machine.
+The runtime of the algorithm depends on size and grade of branching of the state machine.
 An estimation about the execution time is presented later in section~\ref{sec:performance}.
 In order to improve waiting times and thus user experience during constraint generation the entire result might not be returned after termination but each subgraph should be processed as soon as it is found.
-For this, the algorithm implementation has been extended by code snippets which frequently interact with so called \emph{ResultListener}s (figure~\ref{fig:resultlistener_class}).
+For this, the implementation for the algorithm has been extended by code snippets which frequently interact with so called \emph{ResultListener}s (figure~\ref{fig:resultlistener_class}).
 %This interface is also implemented by the progress dialog.
@@
   \centering
-  \includegraphics[width=0.55\linewidth]{resultlistener_class} 
+  \includegraphics[width=0.50\linewidth]{resultlistener_class} 
   \caption{Interface definition of \emph{ResultListener}.}
@@
   \centering
-  \includegraphics[width=0.40\linewidth]{progressdialog} 
+  \includegraphics[width=0.34\linewidth]{progressdialog} 
   \caption{Snapshot of the progress dialog displayed during constraint generation.}
@@
 
-\section{Coalescence}
+\section{The LTLCreator}
 \label{sec:coalescence}
 
-Both features, the visual language presented in section~\ref{sec:operatorconstraintformalism} and the automated constraint generation in chapter~\ref{chap:automatedgenerationofsafetyconstraints}, were assembled in one tool, the LTLCreator.
-It provides a dashboard with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results.
-Besides a button is provided for triggering constraint generation.
+Both features, the editor for the visual language presented in section~\ref{sec:operatorconstraintformalism} and the automated constraint generation in chapter~\ref{chap:automatedgenerationofsafetyconstraints}, were assembled in one tool, the \emph{LTLCreator}.
+Figure~\ref{fig:LTLCreator} shows the assembled parts of LTLCreator: A dashboard is provided where constraints can be edited visually by the developer and result displays indicate the validation results of constraints graphically.
+For validating the constraints an external model checker is linked to \emph{LTLCreator}.
+This component does not have an UI as well as the constraint generator, which is also part of \emph{LTLCreator}.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.65\linewidth]{LTLCreator}
+  \caption{Integration of LTLCreator.}
+  \label{fig:LTLCreator}
+\end{figure}
+The \emph{LTLCreator} itself does not provide functionalities for editing state machines although a state machine model is needed for the visual language as well as for constraint generation.
+Therefore the standalone version of \emph{LTLCreator} uses a hard coded state machine as model which represents an example program for test purpose.
 
@@
   \includegraphics[width=\linewidth]{editor} 
-  \caption{Snapshot of the visual editor.}
+  \caption{Snapshot of the visual editor LTLCreator.}
   \label{fig:editor}
 \end{figure}
-All functionality needed for constraint editing is provided in the tool bar on the right.
-It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting.
-Constraints can be composed by drag\&drop in the dashboard in the center of the editor which holds and displays the operator constraints.
-The tab functionality on the left allows multiple constraints to be managed.
+All functionalities needed for constraint editing are provided in the tool bar on the right (1).
+It contains draggable elements for creating all operator and proposition types as well as a trash can for deleting.
+Constraints can be composed by drag\&drop in the dashboard in the center of the editor (2) which holds and displays the operator constraints.
+The tab functionality on the left allows multiple constraints to be managed (3).
 Each tab shows a small thumbnail of the constraint and a symbol indicating its current status of validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed.
@@
   \centering
-  \includegraphics[scale=0.65]{results}
+  \includegraphics[scale=0.7]{results}
   \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
@@
 
-One main requirement to the LTLCreator tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs.
+A main requirement to the \emph{LTLCreator} tool is that all constraints can be validated immediately and allow fast feedback about the correctness of programs.
 For this a \emph{ConstraintValidator} class has been defined which provides a method \emph{validate(AbstractOperator)} for proving constraints.
@@
 It is an open source project and has been designed to be an open architecture for model checking.
-NuSMV doesn't have a java API but it can be externally accessed over command line.
+NuSMV does not have a java API but it can be externally accessed over command line.
 For this both the state machine program and the LTL formulas have to be translated to an input file which has to match a NuSMV specific format.
+For LTL formulas NuSMV uses the char ``G'' (for globally) instead of $\Box$ and the char ``F'' (for future) instead of $\Diamond$.
 An example of this input language is given in listing~\ref{lst:nusmv}.
@@
 \end{lstlisting}
-A state machine is encoded in one variable \emph{state}, and its value represents the active state.
+A state machine is encoded in the variable \emph{state}, and its value represents the active state.
 Thus, each state is identified by a number.
-The state machine described in the example contains eleven states, this is determined by the number range for \emph{state} from one to eleven.
+The state machine described in the example of listing~\ref{lst:nusmv} contains eleven states, this is determined by the number range for \emph{state} from one to eleven.
 All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions.
@@
 transitions to reachable states.
-If there is more than one outgoing transition, all target state identifiers have to be surrounded by brackets.
+If there are more than one outgoing transition, all target state identifiers have to be surrounded by brackets.
 The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
@@
 The corresponding valdidation result can be subsequently read from the command line and displayed in the editor next to the visual constraint.
+Due to issues of this command line API, each constraint is checked in a new NuSMV instance.
 
-It can happen that there is more than one validation task at a time.
-This is the case when the constraint generator creates multiple constraints or a change of the state machine causes all constraints to be revalidated at once.
-Model checking is a quite ressource inefficient activity and parallel execution of several checks might overload the main memory what causes slow data swapping to start.
-Since each constraint is checked in a new instance of NuSMV, a parallel execution would furthermore strain the CPU scheduler and slow down the overall execution time.
-Thus it is recommended to execute all validation tasks consecutively rather than in parallel.
-An extra advantage is that the developer does not have to wait for all tasks to finish before he can see any result.
-Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
-All this is managed by the \emph{ValidatorThread}, a queue based mechanism for scheduling validations. Its class definition is presented in figure~\ref{fig:validatorthread_class}.
+Model checking can be a quite ressource inefficient activity and due to high temporary data generation for each check, a parallel execution of multiple checks might overload the main memory which causes slow data swapping to start.
+This phenomen causes the overall execution to significantly slow down even on modern multicore CPUs since every process or thread has to wait for the memory.
+To avoid this it might be recommended to execute all validation tasks consecutively rather than in parallel.
+However, it can happen that there is more than one validation task at a time.
+This is the case when the constraint generator creates multiple constraints in succession or a change of the state machine causes all constraints to be revalidated at once.
+%An extra advantage is that the developer does not have to wait for all tasks to finish before he can see any result.
+%Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
+In order to avoid the concurrent execution of multiple NuSMV instances, the \emph{ValidatorThread}, a queue based mechanism for scheduling validations, manages a sequential processing of all validation tasks.
+Its class definition is presented in figure~\ref{fig:validatorthread_class}.
 \begin{figure}[htbp]
@@
 Every validation task of a constraint gets enqueued in the \emph{ValidatorThread} after calling the method \emph{addToQueue(...)}.
-An identifier is used for each constraint which enables elimination of duplicate tasks for the same constraint in order not to strain the execution time.
+An identifier is used for each constraint which enables elimination of duplicate tasks for the same constraint in order to minimize the execution time.
 Tasks can also manually be removed from the queue by calling \emph{cancel(...)}.
@@
 
-\subsection{Ease of integration}
+\subsection{External interface of LTLCreator}
 \label{sec:easeofintegration}
 
-LTLCreator should be integratable into other state machine based programming environments.
-This is ensured by its implementation as a Java Swing component and a clear API which enables a wide range of use for the tool.
+\emph{LTLCreator} should be integratable into other state machine based programming environments.
+This is ensured by its implementation as a Java Swing component and a clear API which enables a wide range of use for this tool.
 
-Figure~\ref{fig:coalescence} points out interactions between LTLCreator and programming environments it is integrated in.
+Figure~\ref{fig:coalescence} points out interactions between \emph{LTLCreator} and programming environments it is integrated in.
 \begin{figure}[htbp]
@@
 \end{figure}
-LTLCreator consists of a dashboard where constraints can be edited visually and an graphical output of validation results.
-The constraint generator as well as the validator are non UI parts of LTLCreator.
-The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the model.
-This will automatically revalidate all existing constraints and always provide the up-to-date model for constraint generation.
-For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.
+The programming environment provides the state machine model and notifies the \emph{LTLCreator} component about all changes made to the model.
+This will automatically revalidate all existing constraints and always hold the up-to-date model for constraint generation.
+For load and save functionality, all existing constraints can be received and new ones can be added to the \emph{LTLCreator} component programmatically.
 All this functionality can be accessed through methods provided by the component \emph{LTLCreator}.
@@
   \includegraphics[width=0.65\linewidth]{ltlcreator_class}
-  \caption{Class definition of \emph{LTLCreator}.}
+  \caption{Class definition for \emph{LTLCreator}.}
   \label{fig:ltlcreator_class}
@@
 
-As stated in the previous section, the LTLCreator provides an API for integration into other Java based applications.
-Robostudio is such an application which was used for evaluating the visual language as well as the automated constraint generation feature and is also supposed to provide all this functionality for development of service robot behaviour in future.
+As stated in the previous section, the \emph{LTLCreator} provides an API for integration into other Java based applications. Robostudio,
+a visual programming environment for rapid authoring and customization of complex services on a personal service robot, is such an application which was used for evaluating the visual language as well as the automated constraint generation feature and is also supposed to provide all this functionality for development of service robot behaviour in future.
 
 Robostudio uses NetBeans~\cite{netbeans} as a platform and organizes all tools and features in windows.
-Thus also LTLCreator was put into a window and added to the Robostudio platform.
+Thus also \emph{LTLCreator} was put into a window and added to the Robostudio platform.
 This was done by placing it into a NetBeans module which was added to Robostudio's modules.
-This module imports the as jar archived LTLCreator tool, archived as a jar file, and further consists of only two classes: \emph{LTLCreatorTopComponent} and \emph{XmlDataToLTLEditorDataWrapper}.
+This module imports the \emph{LTLCreator} tool, archived as a jar file, and further consists of only two classes: \emph{LTLCreatorTopComponent} and \emph{XmlDataToLTLEditorDataWrapper}.
 The former implements the \emph{TopComponent} provided by NetBeans which can be used for displaying UI elements.
-In order to visualize the LTLCreator component it is directly placed on the \emph{LTLCreatorTopComponent}.
+In order to visualize the \emph{LTLCreator} component it is directly placed on the \emph{LTLCreatorTopComponent}.
 It retrieves changes of the model and passes them to the \emph{LTLCreator} component.
-Since the defining of safety constraints is an important issue not only after but also during the process of developing service robot behaviour, the LTLCreator window as a vital tool has been positioned in the center of the platform where also the \emph{UI Component Layout Editor Window} is located.
+Since the defining of safety constraints is an important issue not only after but also during the process of developing service robot behaviour, the \emph{LTLCreator} window as a vital tool has been positioned in the center of the platform where also the \emph{UI Component Layout Editor Window} is located.
 The \emph{XmlDataToLTLEditorDataWrapper} forms the bridge between Robostudio's own model and the state machine model used by \emph{LTLCreator}.
@@
 
-As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the LTLCreator window to revalidate all existing constraints.
+As already proposed in section~\ref{sec:easeofintegration}, all changes to the state machine program made within Robostudio must of course cause the \emph{LTLCreator} module to revalidate all existing constraints.
 This demands a concept of change notification management which is responsible for receiving all changes and making this information available for the \emph{LTLCreator} component.
@@
 A second challenge is the current data handling policy in Robostudio.
-Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than syntax checks.
+Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than for syntax checks.
 Thus syntax errors are not detected or even ignored by static analysis or input restrictions.
 In fact, such features haven't been requirements for Robostudio.
-During integration of \emph{LTLCreator} into Robostudio we encountered problems like transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
-This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistend data retrieved from Robostudio.
+During integration of \emph{LTLCreator} into Robostudio problems came up like transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
+This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistent data retrieved from Robostudio.
 
@@
 
-To guarantee that all model changes cause the constraints to revalidate, all relevant code parts have to report on this.
-But modification of the code hasn't been possible in all cases.
+To guarantee that all model changes cause the constraints to be revalidated, all relevant code parts have to report on this.
+But modification of the code has not been possible in all cases.
 Whereas all state changes are announced, transition changes stay unnoticed.
-This decision is straight for the prototype, but the developers of Robostudio announced an adapted architecture to solve this problem in the next version.
+This decision is sufficient for the prototype, but the developers of Robostudio announced an adapted architecture to solve this problem in the next version.
 
-The problem with inconsistend data could be solved by plausibility and syntax checks during the transformation process of \emph{XmlDataToLTLEditorDataWrapper}.
+The problem with inconsistent data could be solved by plausibility and syntax checks during the transformation process of \emph{XmlDataToLTLEditorDataWrapper}.
 After successful transformation the state machine is passed to the \emph{LTLCreator} component.

### 2012-11-13 thesis/tex/7_testscenarioandevaluation.tex
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.
Final version?

@@
 
-This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections we will evaluate the visual language as well as the potential of constraint generation.
+This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections~\ref{sec:testscenarioandevaluation_operatorconstraints} and~\ref{sec:testscenarioandevaluation_automatedconstraintgeneration} the visual language as well as the potential of constraint generation shall be evaluated.
 Section~\ref{sec:practicalexperiences} describes how a non-expert deals with the visual language and what the final judgement is.
@@
 \section{Operator constraints}
+\label{sec:testscenarioandevaluation_operatorconstraints}
 
-Before constraints can be created it might be useful to think about a reasonable term to express.
+Before constraints can be created it might be useful to think about a reasonable term to be expressed.
 Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given as a requirement in chapter~\ref{chap:goals}.
@@
 
-After each change in the dashboard, the constraint is automatically recompiled and revalidated by using the underlying model checker.
+After each change on the dashboard, the constraint is automatically recompiled and revalidated by using the underlying model checker.
 For syntactically invalid constraints, an ``incomplete'' sign is displayed in the respective tab.
@@
 In addition, the validation itself runs in the background without locking the dashboard.
-Therefore waiting times and disruptions during constraint development can be avoided, which would be likely with conventional model checking where constraint development and constraint validation are alternating processes.
+Therefore waiting times and disruptions during constraint development can be avoided, which would likely be the case with conventional model checking where constraint development and constraint validation are alternating processes.
 Thus our tool is considered an improvement for the user experience.
@@
 
-Due to being based on LTL, the visual language's expressivenes is equivalent to LTL's one of course. More precisely this means any temporal logic formula can be expresssed except potentiality.
+Due to being based on LTL, the visual language's expressivenes is equivalent to LTL's one of course. More precisely, that means any temporal logic formula can be expresssed except potentiality.
 Furthermore there is one more limitation in the current visual language concerning the variety of available proposition operators.
 The visual language was originally designed for validating healthcare applications where just one kind of proposition is needed: ``currently state x is active''.
-Thus other propositions such as equations, greater than etc.
-are not implemented yet, but can be easily integrated.
+Thus other propositions such as equations, comparisons with greater than etc.\ are not implemented yet, but can be easily integrated.
 Figure~\ref{fig:example_propositions} depicts how an extension of \emph{LTLCreator} with additional proposition operators could be realised.
@@
 \section{Automated constraint generation}
+\label{sec:testscenarioandevaluation_automatedconstraintgeneration}
 
@@
 If a pending medication is not already taken, constraint \emph{b)} guarantees intake or staff notification before the next reminders can be read from the database.
-
-The search and generation process finishes within less than a second and thus doesn't let developers wait for a long time.
+Two more constraints are found which might be quite interesting for other programs; however, they do not contribute to safe behaviour of the medication reminder example and are less relevant.
+The search and generation process finishes within less than a second and thus does not let developers wait for a long time.
 
@@
   \includegraphics[scale=0.65]{generatedconstraints}
-  \caption{Automatically generated constraints which match the required postulations.}
+  \caption{Automatically generated constraints which match the required pos\-tu\-la\-tions.}
   \label{fig:generatedconstraints}
@@
 \subsection{Performance}
+\label{sec:performance}
 
-In order to examine the performance of subgraph finding and constraint generation, we need to have a second look at the implementation details again.
+In order to examine the performance of subgraph finding and constraint generation, the implementation details of the algorithm have to be analyzed.
 The subgraph finding algorithm is realized similar to a depth-first search.
@@
 In contrast to a usual depth-first search in trees, the modified depth-first search is applicable for even cyclic state machines and will walk through all possible paths starting from one particular start state without visiting states twice.
-Assuming the worst case in which every state has transitions to all other states, each of the remaining $|V|-1$ states can be possibly visited next, similarly $|V|-2$ for the third step and so on.
+Being $S$ the set of all states and $|S|$ the number of all states, and assuming the worst case in which every state has transitions to all other states, each of the remaining $|S|-1$ states can be possibly visited next, similarly $|S|-2$ for the third step and so on.
 As a worst case execution time the following result is obtained:
@@
 \begin{equation}
-O(\textnormal{``depth-first search''})=O(1*(|V|-1)*(|V|-2)*...*2*1)=O((|V|-1)!)
+O(\textnormal{``depth-first search''})=O((|S|-1)*(|S|-2)*...*2*1)=O((|S|-1)!)
 \end{equation}
@@
 Since we don't know in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions.
-In the worst case all $|V|$ states fulfil this condition, so the depth-first search has to be executed $|V|$ times.
+In the worst case all $|S|$ states fulfil this condition, so the depth-first search has to be executed $|S|$ times.
 This implies for the estimated worst case execution time:
@@
 \begin{equation}
-O(\textnormal{``subgraph finding''})=O(|V|)*O((|V|-1)!)=O(|V|!)
+O(\textnormal{``subgraph finding''})=O(|S|)*O((|S|-1)!)=O(|S|!)
 \end{equation}
@@
 
-For all the healthcare applications that we examined the presented algorithm works efficient and fast.
+For all the healthcare applications that were examined, the presented algorithm works efficient and fast.
 But their program sizes are still quite manageable.
-The test program has eleven states, an average branch grade of $1.\overline{63}$ outgoing transitions per state and a constraint finding execution time of a split second on a standard computer.
+The test program has eleven states, an average branching grade of $1.\overline{63}$ outgoing transitions per state and a constraint finding execution time of a split second on a standard computer.
 Of course an exponential or factorial execution time is bad if the number of states arises.
-A clear slowdown can be felt when the algorithm runs on larger state machines with plenty of states and a high grade of branching.
-A test on a state machine with over 250 states and partially high branch grade resulted in approximately five minutes execution time.
+A clear slowdown can therefore be felt if the algorithm runs on larger state machines with plenty of states and a high grade of branching.
+A test on a state machine with over 250 states and partially high branching grade resulted in approximately five minutes execution time.
 
@@
 Two of them were actually the ones proposed in the requirements.
-Three of them are either intensified versions of the previous ones or just describe the behaviour of the program and thus make sense, but in matters of safety they aren't fundamental relevant.
+Three of them are either intensified versions of the proposed ones or just describe the behaviour of the program and therefore make sense, but in matters of safety they are not fundamental relevant.
 Nevertheless the generator tool actually found a new important constraint which had not been considered before.
-It is constraint \emph{c)} in figure~\ref{fig:generatedconstraints}, section~\ref{sec:prototype:automatedconstraintgeneration}.
+It is constraint \emph{c)} in figure~\ref{fig:generatedconstraints} of section~\ref{sec:testscenarioandevaluation_automatedconstraintgeneration}.
 
@@
 The LTLCreator has been presented to several people, technical engineers as well as ordinary persons.
-The overall feedback was good and the idea and realisation of the visual language was well-received.
+The overall feedback was good and the idea and realization of the visual language was well-received.
 One of the laypersons who have been introduced to LTLCreator was Alina Kracker.
-The 21 yeras old german student of educational science is also working for the elderly in a care facility since more than five years.
-In this social institution Ms Kracker takes care of persons having different disease patterns and necessities.
+The 21 yeras old German student of educational science is also working for the elderly in a care facility since more than five years.
+In this social institution Ms.
+Kracker takes care of persons having different disease patterns and necessities.
 This spectrum varies from physical disabilities over mental problems such as Alzheimer's desease to being bedridden.
@@
 
-Ms Kracker was introduced to the example of medication remider, and the program behaviour was described with the help of figure~\ref{fig:medicationreminder}.
+Ms.
+Kracker was introduced to the example of medication reminder, and the program behaviour was described with the help of figure~\ref{fig:medicationreminder}.
 Afterwards she got familiarized with surface and functionality of LTLCreator.
 Now, she was asked to formulate the two constraints which we postulated in the requirements, just by using the available means of the editor placed at her disposal.
-The first constraint, which ensures the state ``Polling'' always eventually to be visited again, was constructed quickly and accurately.
+The first constraint, which ensures the state ``Polling'' always eventually to be visited again, was constructed quickly and accurately by her.
 The second constraint specifies, that every visit of state ``Not taken yet'' will eventually result in a visit of either state ``Well done!'' or ``Staff notified''.
-It turned out that this constraint held some difficulties for her, the \emph{FUTURE} operator was not considered initially.
-But it was also finisehd after a few consideration breaks.
+It turned out that this constraint held some difficulties for her: The \emph{FUTURE} operator was not considered initially by her.
+But it was also finished after a few consideration breaks.
 
-Ms Kracker didn't know in beforehand what state machines are and never came in contact with LTL or model checking in general.
+Ms.
+Kracker did not know in beforehand what state machines are and never came in contact with LTL or model checking in general.
 But she dealt with this example and the visual language surprisingly well, and she's convinced that she could use it after some exercise.
@@
 Furthermore the intuitive drag and drop allowed faster editing.
-Her examination on the visual language gives us evidence that we succeeded with our approach of creating an easy to use concept for defining LTL formulas.
-The visual language we developed enables even non-professionals to use the usually quite complex concept of model checking.
+Her examination on the visual language gives the evidence of success with the approach of creating an easy to use concept for defining LTL formulas.
+The visual language developed in this work enables even non-professionals to use the usually quite complex concept of model checking.
 
-To the question, whether she would recommend such robotic services like the medication reminder application for use in the care facility she's working in, Ms Kracker replied: ``I can envision such robot applications beeing an excellent assistance for the elderly who are mentally fit but maybe a little doolally.
+To the question, whether she would recommend such robotic services like the medication reminder application for use in the care facility she's working in, Ms.
+Kracker replied: ``I can envision such robot applications beeing an excellent assistance for the elderly who are mentally fit but maybe a little doolally.
 However, in our facilities most of the patients aren't capable enough anymore of making decisions based on what the robots ask.

### 2012-11-13 thesis/tex/8_conclusionandfuturework.tex
Final version?

@@
 
-Perhaps the most important next step we identified is the application of the language and the constraint generation heuristic to other domains and applications.
-This will reveal if the identified subgraph-based heuristic for constraint generation provides reasonable results also in different scenarios.
-Most likely, it will be necessary to identify further heuristics and integrate them in the visual language editor to improve user support.
-A study among a group of users from the healthcare robotics domain could indicate the potential of usability improvements.
+Both the visual language and the automated constraint generation have been indeed developed in the context of the healthcare domain, particularly of the rather simple medication reminder example.
+But the goals achieved in this work are not limited to it at all.
+The results, in particular the visual formalism and the concept of developer support by automated constraint finding, can rather be transferred to completely different domains and use cases.
+
+Therefore, perhaps the most important next step which can be identified is the application of the language and the constraint generation heuristic to other domains and applications.
+This will reveal if the identified subgraph-based heuristics for constraint generation provides reasonable results also in different scenarios.
+Most likely, it will be necessary to identify further heuristics and integrate them in the visual language editor to improve developer support.
+A study among a group of developers from the healthcare robotics domain could indicate the potential of usability improvements.
 
@@
 In contrast to conventional model checking no additional explanation for the outcome such as a counter example is given.
-It can be considered to extend the visual language with support for bug identification by a visualization of counter examples, for example.
+It can, for example, be considered to extend the visual language with support for bug identification by a visualization of counter examples.
 %\item Substitution of parts of constraints would be nice in order to make it easier to understand.

### 2012-11-15 thesis/MA-Thesis.tex
final!

@@
 \selectlanguage{ngerman}
-Augsburg, den \today \hfill Ludwig Nägele
+Augsburg, den 16. November 2012 \hfill Ludwig Nägele %\today
 \end{flushleft}
@@
 For my master thesis it was my wish to do research abroad in a foreign country.
-Professor Wolfgang Reif supported me in my plan and managed to give me the unique possibility to visit New Zealand for half an year.
+Professor Wolfgang Reif, supervisor of my thesis, supported me in my plan and managed to give me the unique possibility to visit New Zealand for half an year.
 Therefore I am deeply grateful to him.
@@
 My thesis project was hostet at the University of Auckland, New Zealand, by Professor Bruce MacDonald.
-I want to thank him for the opportunity to stay and to work at his lab, for all the meetings and discussions we had and for all the support he gave to me.
+I want to thank him for the opportunity to stay and to work at his robotics lab, for all the meetings and discussions we had and for all the support he gave to me.
 
@@
 
-Sincere thanks is also given to Alwin Hoffmann and my supervising tutor Andreas Angerer for assisting me by discussions and hints.
+Sincere thanks is also given to Alwin Hoffmann and my supervising tutor Andreas Angerer, both doctoral candidates at the University of Augsburg, Germany, for assisting me by discussions and hints.
 They did not only inspect and correct the thesis writing regularly, but they also assisted me in finding the thesis topic in the very beginning and contributed to the results of this work decisively by providing ideas and feedback during the research phase.
@@
 
-\pdfbookmark[1]{Inhaltsverzeichnis}{toc}
+\pdfbookmark[1]{Contents}{toc}
 \tableofcontents

### 2012-11-15 thesis/tex/2_background.tex
final!

@@
 
-%Zugriff auf devices via web services
-%Windows is used as operating system on the robots and the programs are executed in Adobe flash and Actionscript. External devices are accessed over web
-
 \section{Robostudio}
@@
 
-In the beginning all programming of the robot behaviour was done by directly editing the XML code mentioned in the previous section by using a text editor.
+In the beginning all programming of the robot behaviour was done by directly editing the RBDL code mentioned in the previous section by using a text editor.
 However, huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise.
 For this reason, Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behaviour on top of RBDL.
-This environment- it is called Robostudio~\cite{robostudio} - allows to edit easily the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
+This environment - it is called Robostudio~\cite{robostudio} - allows to easily edit the program behaviour on a visual layer and generates the corresponding program file containing the RBDL code fully automatically.
 This file can be downloaded to the robot and executed by the interpreter.
@@
 They are provided by the \emph{UI Components Palette Window} (4) on the right side where the developer can choose from all supported UI components.
-The \emph{Background Actions Editor Window} (5) lets the developer define background actions, which are executed transparently to the developer such as accessing external devices or sending messages.
+The \emph{Background Actions Editor Window} (5) lets the developer define background actions, which are executed transparently such as accessing external devices or sending messages.
 Aiming to support the developer with helpful graphical tools, the \emph{State Transition Visualization Window} (6) gives an overview about all connections towards and from the selected state.
-It helps to get quickly an understanding of the local program behaviour.
+It helps to quickly get an understanding of the local program behaviour.
 
@@
 One of them is model checking which allows checking state machine programs for functional safety.
-Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realised with state machines, the concept of behaviour checking is applicable and important for this kind of programs.
+Since the healthbot applications mentioned in section~\ref{sec:healthbotapplication} are realized with state machines, the concept of behaviour checking is applicable and important for this kind of programs.
 At first the basic idea of state machines is described in section~\ref{sec:statemachines}, followed by an introduction into model checking in section~\ref{sec:modelcheckingandltl}.
@@
 They trigger transitions which transfer the system into a new (current) state.
-The fact that a state machine reacts to events is the reason why state machines are also called \emph{reactive systems}.
+The fact that a state machine reacts to events is the reason why state machines are also known as \emph{reactive systems}.
 
@@
 The short description given in this section does not contain a complete explanation about the concept of state machines at all, but it gives a basic overview which is needed as background information for this work.
-Also the explanation of state machines above is no formal definition; it can be read found in literature~\cite{statemachines, statemachinesandformalmethods}.
-%Also the formal definition of state machines is not relevant and can be found in literature~\cite{statemachines, statemachinesandformalmethods}.
-
-%mathematical model of computation
-%finite (endliche anzahl zustände)
+Also the explanation of state machines above is no formal definition; it can be found in literature~\cite{statemachines, statemachinesandformalmethods}.
 
@@
 	\item[$\varphi \mathcal{U} \psi$] The ``UNTIL'' operator makes $\varphi$ true at least until the first occurrence of a state where $\psi$ is true. Besides, $\psi$ will eventually be true.
-	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true caused by the first occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
+	\item[$\varphi \mathcal{R} \psi$] It expresses that $\psi$ is true at least as long as there is no occurrence of a state where $\varphi$ is true. Its meaning is $\varphi$ ``RELEASES'' $\psi$.
 	\item[$\varphi \mathcal{W} \psi$] ``WEAK UNTIL'' represents the same as ``UNTIL'', except that an occurrence of $\psi$	is not necessary. In this case, $\varphi$ remains true infinitely.
@@
 HomeTL focuses more on monitoring temporal boundaries of a patient's behaviour than on ensuring functional safety of an implemented program.
-It does however depict how nontechnical constraint design can be realised and its graphical design is a good example of handling time dependencies.
+It does however depict how nontechnical constraint design can be realized and its graphical design is a good example of handling time dependencies.
 

### 2012-11-15 thesis/tex/5_automatedgenerationofsafetyconstraints.tex
final!

@@
 In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that.
-The more constraints are defined the the better can be the reliability.
+The more constraints are defined the better can be the reliability.
 But as already mentioned before, it might be difficult for developers to find all reasonable %safety
@@
 Once the initial implementation of a program is finished, the behaviour of this state machine can be analyzed.
-In particular heuristics can be elaborated which determine significant parts of the state machine automatically and put them into constraint representation.
+In particular, heuristics can be elaborated which determine significant parts of the state machine automatically and put them into constraint representation.
 These constraints are of course valid on the current program.
@@
 %All associated states which match these properties can be combined to a 
-In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined.
+In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\gamma$-states and the one $\beta$-state have to be determined.
 Such a group of states of a state machine is named subgraph from now on.
@@
 The quite similar but stricter formula~\ref{eq:third} additionally demands a visit of at least one $\gamma$-state before the $\beta$-state can be reached.
-Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can't be visited between $\alpha$- and $\beta$-states.
+Formula~\ref{eq:fourth} simply says that a state not contained in the subgraph can not be visited between $\alpha$- and $\beta$-states.
 The constraint that the $\alpha$-state will always eventually be visited again is expressed by formula~\ref{eq:fifth}.

### 2012-11-15 thesis/tex/6_theltlcreatorprototype.tex
final!

@@
 It is called finite state machine, because the number of states is finite which applies to the domain of this work.
-\emph{State} represents the states of the machine and holds all transitions from and to the respective state.
+\emph{State} represents the states of the machine and holds a list of all transitions from or to itself.
 Transitions are represented by the class \emph{Transition} which specifies the source and a target state.
@@
 
-As proposed in the requirements for the visual language in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: A vertical read direction and the horizontal direction for variation in time.
+As proposed in the design decisions for the visual language in section~\ref{sec:designdecisions}, constraints have particular meanings for the two dimensions: A vertical read direction and the horizontal direction for variation in time.
 Swing's mechanism for layouting components is used for the positioning of operators.
-After determining the aspired sizes of child components which represent sub operators, reference lines which have been already demonstrated with figure~\ref{fig:referencelines} in section~\ref{sec:operatorconstraintformalism} are computed mathematically.
+After determining the aspired sizes of child components which represent sub operators, reference lines which have been already demonstrated with figure~\ref{fig:referencelines} in section~\ref{sec:designdecisions} are computed mathematically.
 Based on this information operators are sized and positioned correctly.
@@
 There are no cycles allowed between start and end state since they are forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
-Furthermore also foreign transitions are prohibited: Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph.
+Furthermore also foreign transitions are prohibited: Transitions coming from a state outside the subgraph are allowed to lead to the start state or the end state but not to other states of the subgraph.
 If all checks are successful, start and end states get added to the result list $subgraphs$ in line 11.
@@
 Both features, the editor for the visual language presented in section~\ref{sec:operatorconstraintformalism} and the automated constraint generation in chapter~\ref{chap:automatedgenerationofsafetyconstraints}, were assembled in one tool, the \emph{LTLCreator}.
-Figure~\ref{fig:LTLCreator} shows the assembled parts of LTLCreator: A dashboard is provided where constraints can be edited visually by the developer and result displays indicate the validation results of constraints graphically.
+Figure~\ref{fig:LTLCreator} shows the assembled parts of \emph{LTLCreator}: A dashboard is provided where constraints can be edited visually by the developer and result displays indicate the validation results of constraints graphically.
 For validating the constraints an external model checker is linked to \emph{LTLCreator}.
@@
 It is an open source project and has been designed to be an open architecture for model checking.
-NuSMV does not have a java API but it can be externally accessed over command line.
+NuSMV does not have a Java API but it can be externally accessed over command line.
 For this both the state machine program and the LTL formulas have to be translated to an input file which has to match a NuSMV specific format.
@@
 The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
-The last line contains the constraint formula we want to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.
+The last line contains the constraint formula which has to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.
 
@@
 As a solution, all parts of the code where model changes are made should be extended with short code snippets which send notifications.
-However, this is not as easy to realise in Robostudio since some of the relevant code was generated or even imported as external jars which makes code manipulation impossible.
+However, this is not as easy to realize in Robostudio since some of the relevant code was generated or even imported as external jars which makes code manipulation impossible.
 
@@
 Since this platform is mainly made for developers who used to edit RBDL xml code manually, Robostudio aims for visual rapid editing rather than for syntax checks.
-Thus syntax errors are not detected or even ignored by static analysis or input restrictions.
-In fact, such features haven't been requirements for Robostudio.
-During integration of \emph{LTLCreator} into Robostudio problems came up like transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
-This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete state machine as input and doesn't know how to cope with inconsistent data retrieved from Robostudio.
+Thus, syntax errors are not detected or even ignored by methods like static analysis or input restrictions.
+In fact, such features have not been requirements for Robostudio.
+During integration of \emph{LTLCreator} into Robostudio problems came up when loading already existing RBDL programs of the healthbot project: Transitions leading to non existing states, transitions leading to nowhere (null) or transitions whose target is depending on runtime facts.
+This characteristic seems to be a problem for the \emph{LTLCreator} which needs a complete and correct state machine as input and does not know how to cope with inconsistent data retrieved from Robostudio.
 

### 2012-11-15 thesis/tex/7_testscenarioandevaluation.tex
final!

@@
 In order to find a corresponding graphical constraint the sentence has to be analyzed just as developers read it.
-Since ``whenever'' is a semantic equivalent to ``always if'' first of all a \emph{ALWAYS} operator gets dragged to the dashboard directly followed by an \emph{IF} as shown in figure~\ref{fig:sampleconstraint}.
+Since ``whenever'' is a semantic equivalent to ``always if'' first of all an \emph{ALWAYS} operator gets dragged to the dashboard directly followed by an \emph{IF} as shown in figure~\ref{fig:sampleconstraint}.
 The condition for this \emph{IF} is that there is a pending reminder, so a state ``Not taken yet'' has to be added to the upper bucket of the \emph{IF} operator.
@@
 Therefore waiting times and disruptions during constraint development can be avoided, which would likely be the case with conventional model checking where constraint development and constraint validation are alternating processes.
-Thus our tool is considered an improvement for the user experience.
+Thus the \emph{LTLCreator} tool is considered an improvement for the developer experience.
 
@@
 Thus other propositions such as equations, comparisons with greater than etc.\ are not implemented yet, but can be easily integrated.
-Figure~\ref{fig:example_propositions} depicts how an extension of \emph{LTLCreator} with additional proposition operators could be realised.
+Figure~\ref{fig:example_propositions} depicts how an extension of \emph{LTLCreator} with additional proposition operators could be realized.
 
@@
   \includegraphics[scale=0.65]{example_propositions}
-  \caption{Possible proposition extensions: numeric and string equations.}
+  \caption{Possible proposition extensions: Numeric and string equations.}
   \label{fig:example_propositions}
@@
 The performance of this algorithm of course highly depends on the grade of branching.
-In contrast to a usual depth-first search in trees, the modified depth-first search is applicable for even cyclic state machines and will walk through all possible paths starting from one particular start state without visiting states twice.
+In contrast to a usual depth-first search in trees, the modified depth-first search is applicable for even cyclic state machines and will walk through all possible paths starting from one particular start state but without visiting states twice.
 Being $S$ the set of all states and $|S|$ the number of all states, and assuming the worst case in which every state has transitions to all other states, each of the remaining $|S|-1$ states can be possibly visited next, similarly $|S|-2$ for the third step and so on.
@@
 In order to find a subgraph with the presented algorithm the depth-first search has to be startet on a start state of a subgraph.
-Since we don't know in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions.
+Since it is not known in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions.
 In the worst case all $|S|$ states fulfil this condition, so the depth-first search has to be executed $|S|$ times.
@@
 
-Some parts of the healthcare robot's behaviour are appropriate to be specified by non-programmers, and that's the purpose of the presented visual language as well.
-Different tools and environments, among Robostudio and LTLCreator, try to provide a particular interface suitable for such people.
+Some parts of the healthcare robot's behaviour are appropriate to be specified by non-programmers, and that is the purpose of the presented visual language as well.
+Different tools and environments, among Robostudio and \emph{LTLCreator}, try to provide a particular interface suitable for such people.
 Now, it might be interesting to find out about user's experiences and how they get along when using these tools.
 
-The LTLCreator has been presented to several people, technical engineers as well as ordinary persons.
+The \emph{LTLCreator} has been presented to several people, technical engineers as well as ordinary persons.
 The overall feedback was good and the idea and realization of the visual language was well-received.
-One of the laypersons who have been introduced to LTLCreator was Alina Kracker.
+One of the laypersons who have been introduced to \emph{LTLCreator} was Alina Kracker.
 The 21 yeras old German student of educational science is also working for the elderly in a care facility since more than five years.
@@
 Kracker was introduced to the example of medication reminder, and the program behaviour was described with the help of figure~\ref{fig:medicationreminder}.
-Afterwards she got familiarized with surface and functionality of LTLCreator.
-Now, she was asked to formulate the two constraints which we postulated in the requirements, just by using the available means of the editor placed at her disposal.
+Afterwards she got familiarized with surface and functionality of \emph{LTLCreator}.
+Now, she was asked to formulate the two constraints postulated in the requirements, just by using the available means of the editor placed at her disposal.
 The first constraint, which ensures the state ``Polling'' always eventually to be visited again, was constructed quickly and accurately by her.
@@
 Kracker did not know in beforehand what state machines are and never came in contact with LTL or model checking in general.
-But she dealt with this example and the visual language surprisingly well, and she's convinced that she could use it after some exercise.
+But she dealt with this example and the visual language surprisingly well, and she is convinced that she could use it after some exercise.
 As reason for the success she pointed out the easy readability of the visual language and its close relationship to spoken language.
-Furthermore the intuitive drag and drop allowed faster editing.
+Furthermore the intuitive drag and drop allowed fast editing.
 Her examination on the visual language gives the evidence of success with the approach of creating an easy to use concept for defining LTL formulas.
@@
 
-To the question, whether she would recommend such robotic services like the medication reminder application for use in the care facility she's working in, Ms.
+To the question, whether she would recommend such robotic services like the medication reminder application for use in the care facility she is working in, Ms.
 Kracker replied: ``I can envision such robot applications beeing an excellent assistance for the elderly who are mentally fit but maybe a little doolally.
-However, in our facilities most of the patients aren't capable enough anymore of making decisions based on what the robots ask.
-In fact, due to frequent changes of the patients in our facilities it is a big issue for the care givers not to forget time-dependend medication intakes of their patients.
-Thus such a reminding robot application would constitute an excellent support for us staff!''.
+However, in our facilities most of the patients are not capable enough anymore of making decisions based on what the robots ask.
+In fact, due to frequent changes of the patients in our facilities it is a big issue for us care givers not to forget time-dependend medication intakes of our patients.
+Thus, such a reminding robot application would constitute an excellent support for us staff!''.

### 2012-11-15 src/model/impl/UntilOperator.java

@@ -89,11 +89,13 @@ public class UntilOperator extends AbstractOperator {
 		Dimension ps1 = this.a.getPreferredSize();
 		Dimension ps2 = this.b.getPreferredSize();
 		
-		int top1 = (this.getHeight() - ps1.height) / 2;
-		int top2 = (this.getHeight() - ps2.height) / 2;
+		int max_height = Math.max(ps1.height, ps2.height);
 		
-		this.a.setBounds(insets.left, top1, ps1.width, ps1.height);
-		this.b.setBounds(insets.left + ps1.width + space_between_buckets, top2, ps2.width, ps2.height);
+		int top1 = (max_height - ps1.height) / 2;
+		int top2 = (max_height - ps2.height) / 2;
+		
+		this.a.setBounds(insets.left, insets.top + top1, ps1.width, ps1.height);
+		this.b.setBounds(insets.left + ps1.width + space_between_buckets, insets.top + top2, ps2.width, ps2.height);
 	}
 
 	@Override

### 2012-11-15 src/editor/validator/impl/NuSMVModelCheckerWrapper.java
CD which is with the thesis is added

@@ -3,6 +3,8 @@ package editor.validator.impl;
 import java.io.File;
 import java.io.FileWriter;
 import java.io.IOException;
+import java.io.UnsupportedEncodingException;
+import java.net.URLDecoder;
 import java.util.ArrayList;
 import java.util.HashMap;
 import java.util.List;
@@ -24,8 +26,24 @@ import editor.validator.ValidationCanceledException;
 public class NuSMVModelCheckerWrapper extends ConstraintValidator {
 	
 	private boolean askForNuSmv = true;
-	private String nuSmvFile = "C:\\Users\\ludwig\\Documents\\TUM\\12 SS\\Thesis\\NuSMV\\bin\\NuSMV.exe";
+//	private String nuSmvFile = "C:\\Users\\ludwig\\Documents\\TUM\\12 SS\\Thesis\\NuSMV\\bin\\NuSMV.exe";
+	private String nuSmvFile = null;
 	private Map<String, String> replacementNames = new HashMap<String, String>();
+	
+	public NuSMVModelCheckerWrapper() {
+		try {
+			this.nuSmvFile = new File(ClassLoader.getSystemClassLoader().getResource("./NuSMV/NuSMV.exe").toURI()).getPath();
+			System.out.println(this.nuSmvFile);
+			try {
+				this.nuSmvFile = URLDecoder.decode(this.nuSmvFile, "UTF-8");
+			} catch (UnsupportedEncodingException e) {
+				e.printStackTrace();
+			}
+			System.out.println(this.nuSmvFile);
+		} catch (Exception e2) {
+			e2.printStackTrace();
+		}
+	}
 	
 	@Override
 	public String getLTLForVariable(String variable) {
@@ -84,8 +102,8 @@ public class NuSMVModelCheckerWrapper extends ConstraintValidator {
 	 * 
 	 * @return
 	 */
-	private String getNuSmvFile() {
-		if (this.nuSmvFile!=null && new File(this.nuSmvFile).isFile()) {
+	private String getNuSmvFile() {		
+		if (this.nuSmvFile!=null && new File(this.nuSmvFile).isFile() && new File(this.nuSmvFile).exists()) {
 			return this.nuSmvFile;
 		}
 		
